### 2020-08-20 manuscript.Rmd
misc

---
title: "Ribosome accumulation during early phase resistance training"
author: Daniel Hammarström^1,2,£^&ast;,Sjur Øfsteng^2,£^  Bent R. Rønnestad^2^, Stian Ellefsen^1,3^
date: ""
header-includes:
  - \usepackage{wasysym}
csl: resources/experimental-physiology.csl
bibliography: resources/bibliography.bib
output:
  word_document:
    reference_docx: resources/manuscript_word_template.docx
editor_options: 
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

<!-- The formatting of the manuscript is based on faseb journal, find csl-files at https://github.com/citation-style-language/styles -->

<!-- Figure texts are included in the end of the manuscript file -->
<!-- Tables in a separate file -->

```{r, include = FALSE}

source("./R/libs.R")
# Settings for chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```

^1^ Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway.

^2^ Swedish School of Sport and Health Sciences, Stockholm, Sweden.

^3^ Innlandet Hospital Trust, Lillehammer, Norway. </div>

^£^ These authors contributed equally to this work

<div custom-style="correspondence_style">
&ast; Correspondance

Daniel Hammarström, Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway. Email: daniel.hammarstrom@inn.no  </div>

<div custom-style="runningtitle_style">**Running title:** Ribosome biogenesis in early phase resistace training</div>

\pagebreak

# Abstract
<div custom-style="abstract_style">
<!-- Word limit 200 -->
Abstract...

**Keywords:** Resistance training, ribosome biogenesis

</div>

\pagebreak

# Introduction

# Methods

## Study overview

## Muscle strength and thickness

## Muscle biopsy sampling

## RNA and protein extraction

## Quantitative polymerase chain reaction (qPCR)

## Immunoblotting

## Immunohistochemistry

## Statistics and data analysis

# Results

# Discussion

# References
<div id="refs"></div>

### 2020-08-20 R/libs.R
misc

@@ -4,6 +4,8 @@
 library(tidyverse)
 library(readxl)
 library(qpcR)
+
+# devtools::install_github("dhammarstrom/qpcrpal")
 library(qpcrpal)
 library(ggrepel)
 

### 2020-08-20 manuscript.Rmd
updates in methods section

@@
 ## Muscle strength and thickness
+Muscle strength was assessed as maximal knee extension isokinetic (90&deg; sec<sup>-1</sup>) and isometric (60&deg; angle, fully extended leg 0&deg;) torque.
+After a brief warm-up (5-min cycling, RPE 12-14), participants were seated and secured in the individually adjusted dynamometer.
+The test protocol consisted of unilateral knee-extensions.
+Participants were instructed to gradually increase their effort during three warm-up repetitions (50, 60 and, 70% of subjective maximal effort).
+After a 30-sec restitution period participants were instructed to perform three repetitions with maximal effort in the concentric phase.
+Sixty seconds after the isokinetic test the lever automatically moved to a 60&deg; angle and participants were instructed to push against the lever enough to see feedback from the visual feedback system.
+After an additional 15-sec restitution period, participants were instructed to push against the lever with maximal effort.
+During the same assessment session, participants remained seated in the dynamometer for meassurement from both legs.
+The first measurement was alternated between legs and sessions.
+For statistical treatment of the data, all successful attempts were used.
+Successful attempts were decided based on participants (...).
+The last strength assessment at baseline was performed at least **seven** days prior to the first biopsy sampling.
+At least one strength was performed on separate days with two sessions allowed perform on the same day with a short rest between sessions.
+Post training assessments were performed approximately 48 hours after the last session and after seven days of no training.
+
+Muscle thickness was assessed using (...) ultra sound.
 
 ## Muscle biopsy sampling 
+Muscle specimens were sampled bilaterally from *m.
+vastus laterlis* under local anesthesia (lidocaion %, Manufacturer, Oslo, Norway) using a disposable needle (12-14 gauge, Universal plus, Medax, Italy), operated with a spring loaded device (Bard Magnum).
+Two to four passes were made to get sufficient material.
+One to two aliquots of the samples were quickly dissected free from connective and fat tissue, weighed and frozen in isopentan chilled to -80&deg;C and stored at -80&deg;C until further processing.
+At baseline, after twelve weeks of training and, after the de-training period, a further aliquot from each sample was selected for immunohistochemistry analysis and subsequently immersed in 4% formalin for fixation.
 
 ## RNA and protein extraction
+Frozen muscle tissue was homogenized in 1 ml of Trizol (ThermoFisher Scientific, Oslo, Norway) spiked with 0.04 ng of an external, non-mammalian, RNA spike-in (Lambda, Takara Bioscience).
+Mechanical disruption of the samples was achieved using Zirconium Oxide Beads (0.5 mm, Next Advance, Inc., New York, USA) and a bead mill (Bullet blender, Next Advance).
+Chloroform (200 &#956;l) was prior to centrifugation (12000 g, 15 min at 4&deg;C) to achieve phase separation.
+Four houndred fifty &#956;l of the upper aqueous phase was transferred to a fresh tube and 500 &#956;l of isopropanol was added to precipitate the RNA.
+After a 10 min incubation at room temperature, samples were centrifuged (12000 g, 10 min at 4&deg;C), after which a pellet formed.
+The pellet was washed three times in chilled 75% ethanol with centrifugation between each wash (7500 g, 5 min at 4&deg;C).
+After the final wash all ethanol was removed and the pellet was eluted in 0.1X Tris-EDTA buffer.
+RNA concentration and purity was assessed by spectrophotometry.
+All samples had 260/280 ratios > 1.95.
+
+Protein was extracted from Trizol preparations according to manufacturers instructions and [@KOPEC] with modifications.
+The remaining aqueous phase was removed and DNA was precipitated by the addition of XXX &#956;l of absolute ethanol followed by gentle centrifugation (2000 g, 5 min at room temperature).
+An aliquot of the phenol-ethanol phase, corresponding to ~1.75 mg of tissue, was transferred to to a fresh tube.
+After addition of at least two volumes of isopropanol and incubation (10 min at room temperature), samples were centrifuged (7500 g, 10 min 4&deg;C) and a pellet formed.
+The pellet was washed three times in 95% ethanol with each wash separated by centrifugation (5000g, 5 min at room temperature).
+After the last wash all liquid was removed and 45 &#956;l of Kopec buffer was added (4% SDS, 300 mM Tris ..., pH; containing protease and phosphatase inhibitors).
+Pellets were incubated at 50&deg;C for three hours after which the majority of samples were dissolved.
+Any undissolved material was sedimented by centrifugation (10000 g, 10 min at room temperature).
+Protein concentrations were measured (Pierce Detergent Compatible Bradford Assay, ThermoFisher Scientific).
+Sample were normalized in Laemmli buffer (Bio-Rad), boiled (95&deg;C, 5 min) and stored at -20&deg;C before later use.
 

### 2020-08-21 manuscript.Rmd
update manuscript

@@
 
+Our skeletal muscles not only constitutes our locomotive apparatus but also serve as our primary storage site of amino acids.
+This makes the musculoskeletal system a critical target for interventions that promotes health across the lifespan [@RN763].
+The beneficial effects of resistance-training (RT) are in large part mediated through exercise-induced changes in the balance between muscular protein breakdown and synthesis, with successful RT leading to net protein accretion and muscle hypertrophy.
+A traditional view of exercise-induced muscle hypertrophy has been that muscle protein accumulates as a consequence of repeated episodes of elevated post-exercise protein synthesis in the acute phase, resulting in a net-positive protein balance over time.
+Indeed, one bout of resistance exercise acutely up-regulates skeletal muscle protein synthesis up to 48-hr after exercise [@RN786] and subsequently, repeated bouts would lead to accumulation of muscle protein over time [@RN1272; @RN1642].
+In recent years, this view has been supplemented with evidence showing that chronic RT also leads to an increased resting synthetic rate of muscle protein [@RN1521; @RN1866; @RN1897].
+An increase in the resting protein synthetic rate has been postulated to be related to an increased translational capacity, i.e.
+accumulation of ribosomes [@RN1897; @RN1912], a notion supported by observations relating exercise-induced changes in total RNA, a proxy marker for ribosome abundance, to muscle hypertrophy [@RN1644, @RN1755].
+Furthermore, in vitro studies shows that direct inhibition of ribosomal RNA (rRNA) transcription or inhibition of up-stream transcription factors leading to halted biosynthesis of ribosomes results in diminished cellular growth [@RN1755; @RN1754; @RN884].
+Formation of new ribosomes is a complex and energy demanding process involving the whole transcriptional apparatus [@RN1920, @RN1820].
+In skeletal muscle, a single bout of resistance exercise elevates markers of ribosome biogenesis in skeletal muscle evident from both Figueiredo et al.
+and Nader et al.
+who evaluated immature forms of rRNA in response to exercise in the untrained state [@RN1644; @RN1037].
+Studies’ examining repeated bouts of resistance exercise suggests that RT leads to accumulation of ribosomes.
+Bickel et al.
+showed that two subsequent bouts of a hypertrophic stimulus increased total RNA at 72 h after a second bout of RT [@RN1520].
+Following three weeks of RT, Brook et al.
+reported an increase in total RNA with no further increase after three additional weeks of training [@RN1642].
+The findings of Brook et al.
+fits with observations from our laboratory where we saw a ~30% increase in total RNA per tissue weight after two weeks of RT.
+Following an additional 10 weeks of RT the RNA content tended to normalized without reaching baseline values (Hammarström et al.
+2017 unpublished observation).
+These data suggests that ribosome accumulation per tissue weight, reaches a plateau in the early phase of RT.
+Interestingly, in the study by Figueiredo and colleagues, acute exercise in the trained state (after eight weeks of RT), rRNA and pre-rRNA rapidly decreases in response to acute exercise [@RN1644].
+Based on these observations we are hypothesizing that (1) ribosome accumulation occurs in the early phase (3-4 weeks) of RT and (2) reaches a plateau during this phase in response to constant volume RT and finally (3) that fluctuations in ribosome abundance occurs in response to altered training volume and training cessation.
+Subsequently the aim of this study is to characterize fluctuations in ribosomal biogenesis during the initial phase of RT, in response to altered training volumes and in response to a short-term de-training period.
+
 # Methods
@@
 Mechanical disruption of the samples was achieved using Zirconium Oxide Beads (0.5 mm, Next Advance, Inc., New York, USA) and a bead mill (Bullet blender, Next Advance).
-Chloroform (200 &#956;l) was prior to centrifugation (12000 g, 15 min at 4&deg;C) to achieve phase separation.
+Chloroform (200 &#956;l) was added prior to centrifugation (12000 g, 15 min at 4&deg;C) to achieve phase separation.
 Four houndred fifty &#956;l of the upper aqueous phase was transferred to a fresh tube and 500 &#956;l of isopropanol was added to precipitate the RNA.
@@
 Protein concentrations were measured (Pierce Detergent Compatible Bradford Assay, ThermoFisher Scientific).
-Sample were normalized in Laemmli buffer (Bio-Rad), boiled (95&deg;C, 5 min) and stored at -20&deg;C before later use.
+Sample were normalized in 4X Laemmli buffer (Bio-Rad Norway AS, Oslo, Norway), boiled (95&deg;C, 5 min) and stored at -20&deg;C before later use.
 
 ## Quantitative polymerase chain reaction (qPCR)
+Complementary DNA (cDNA) was synthesized in technical duplicates from 500 ng of total RNA using random hexemer and anchored Oligo-dT primers (Thermo Scientific) together with Superscript IV (ThermoFisher Scientific) according to manufacturer's instruction.
+qPCR reactions were performed with diluted cDNA (2 &#956;l, 1:25 dilution), a SYBR-green based commercial master mix (PowerUp™ SYBR™ Green Master Mix, Thermo Fisher) and, target-specific primers (500 nM) in 10 &#956;l reaction volumes using a real-time detection system (QuantStudio 5 Real-Time PCR System, Thermo Fisher).
+Fast cycling was used (1 sec denaturing, 30 sec annealing) after UNG (2 min, 50&deg;C) and polymerase  (2 min, 95&deg;C) activation.
+Melt curves were collected from all reactions to confirm single product amplification.
+Primers were further evaluated by agarose gel electrophoresis an non-template control experiments.
+Primer sequences and their respective average performances are shown in Table 2.
+
+Raw fluorescence data was exported from the QuantStudio software and estimates of quantification cycles (Cq) and amplification efficiencies were derived using the qpcR package[qpcR-package].
+Relative abundance data was linearized using average efficiencies per primer pair and reaction specific Cq-values before any further statistical treatment.
 
 ## Immunoblotting
+Protein samples (20 &#956;g) were separated on 4-20% Tris-Glycin gels (Criterion TGX Precast Gels, Bio-Rad) at 250 V for 45 min using the recommended running buffer (25 mM Tris, 192 mM Glycin, 0.1% SDS).
+All samples from the same participant were run on the same gel and all samples were run in at leat duplicates.
+Separated samples were transferred to &#956m PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 Glycin, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h.
+Membranes were then stained to confirm transfer and enable total protein quantification using a reversible protein stain (Pierce Reversible Protein Stain, ThermoFisher Scientiﬁc).
+Primary antibodies were acquired to detect UBF (F-9, sc-13125, Santa-Cruz Biotechnology) and ribosomal protein S6 (54D2, #2317, Cell signaling technology).
+After blocking (Tris-buffered saline blocking buffer, 20 mM Tris, 150 mM NaCl, 5% fat-free milk, 0.1% Tween-20), membranes were incubated over-night with primary antibodies diluted in blocking buffer (UBF, 1:200; S6, 1:1000) followed by incubation with a secondary antibody conjugated to horseradish peroxidase (Anti-mouse IgG, #7076, Cell Singlaing Technology, 1:10000).
+Membranes were washed 6 &#215; 5 min after incubation with primary antibodies and 8 &#215; 5 min after incubation with the secondary antibody.
+All incubation and washing steps were performed at 4&deg;C using an automatic membrane processor (BlotCycler, Precision Biosystems, Mansﬁeld, MA, USA).
+Chemiluminescent signals from membranes were detected after 5 min incubation in substrate (Super Signal West Femto Maximum Sensitivity Substrate, ThermoFisher Scientiﬁc) using a documentation system ().
+Total protein content was quantified whole membrane images defined as the mean gray value of the whole lane.
+Between-lane gray values were used as background subtracted from protein values.
+Total protein quantification was done using ImageJ [REF].
+Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA).
 

### 2020-08-24 Figures/Figure-rrna-qpcr.R
Small changes

@@ -1,0 +1,39 @@
+
+
+#### Figure X RNA/qPCR ###################
+
+
+
+## Panels
+# A: rRNA primers 
+
+
+
+
+
+### rRNA and primers #########
+
+
+# Sequences can be found : https://www.ncbi.nlm.nih.gov/nuccore/NR_146144.1
+# Add primer sequences to plot and annotate mature segments
+
+
+
+data.frame(x = c(1, 13500), 
+           y = c(1, 2)) %>%
+  ggplot(aes(x, y)) +
+  
+  # Full 45S segment 
+  annotate("segment", x = 1, xend = 13351, y = 1.5, yend = 1.5, size = 0.5) +
+  # 18S segment
+  annotate("segment", x = 3655, xend = 5523, y = 1.5, yend = 1.5, size = 2.5) +
+  # 5.8S segment
+  annotate("segment", x = 6601, xend = 6757, y = 1.5, yend = 1.5, size = 2.5) +
+  # 28S segment
+  annotate("segment", x = 7925, xend = 12990, y = 1.5, yend = 1.5, size = 2.5) 
+
+
+
+
+
+

### 2020-08-24 Figures/Figure1.R

@@ -1,0 +1,206 @@
+############# Figure 1 ##################################
+
+
+## Panels
+# A: Study design
+# B: Performed volume per condition
+# C/D: Strength and body composition changes. 
+
+
+source("./R/figure-source.R")
+
+
+
+
+
+# Colors for plot
+
+control_color <- "white"
+
+const_color <- "#56B4E9"
+var_color <- "#009E73"
+experiment_color <- "#4abdcb"
+
+
+
+line_size <- 0.2
+textsize <- 2.5
+
+
+# design figure
+
+biopsy_glyph <- "\U2193"
+strength_glyph <-  "\U2299"
+us_glyph <- "\U2295"
+
+#### Design graph: EXP ####
+
+design_exp <- expand.grid(cond = c("Variable volume", "Constant volume"),
+                               session = seq(1:12), 
+                               sets = rep(6)) %>%
+  mutate(sets = if_else(cond == "Variable volume" & session %in% c(5, 6, 7, 8), 3, 
+                        if_else(cond == "Variable volume" & session %in% c(9, 10, 11, 12), 9, sets))) %>%
+  ggplot(aes(session, sets, fill = cond)) +
+  
+  annotate("text", x = -2, y = 11.0, label = c("Experimental group"), size = textsize + 0.5) +  
+  
+  geom_bar(stat = "identity", position = position_dodge(width = 0.55),
+           width = 0.45, 
+           color = "black", 
+           size = 0) +
+  scale_x_continuous(breaks = seq(1:12), limits = c(-4,15.5), expand = c(0, 0)) +
+  
+  scale_y_continuous(breaks = c(0, 3, 6, 9, 12), limits = c(0,12), expand = c(0,0), 
+                     labels = c("", "3", "6", "9", "")) +
+  scale_fill_manual(values = c(const_color, var_color)) +
+
+  annotate("text", x = c(-3.8, -3.8, -3.8), y = c(9.1, 8.2, 6), 
+           label = c("Muscle biopsy", "Muscle thickness", "Strength tests"), 
+           vjust = 0, hjust = 0, alpha = 0.8, size = textsize) + 
+  
+  # Biopsies
+  annotate("text", x = c(0.5, 1.5, 4.5, 5.5, 8.5, 9.5, 12.5, 14.5), y = rep(9, 8), 
+           label = rep(biopsy_glyph, 8), vjust = 0, family = "Cambria") +
+  
[…]

### 2020-08-24 R/figure-source.R

@@ -1,0 +1,27 @@
+##### Figure source ###############################
+
+
+
+# Libraries for figures ####
+
+library(tidyverse)
+library(cowplot)
+
+
+
+# Functions for plotting  #####
+
+
+
+
+
+
+# Themes, color, scales etc. #####
+
+
+
+
+
+
+
+

### 2020-08-24 manuscript.Rmd
Small changes

@@
 ## Study overview
+Eighteen volunteers were recruited to the study.
+Eligible participants had to be non-smokers, between 18 and 35 years of age, report a training history of less than one RT session per week during the six months leading up to the study and not consume dietary supplements or medication with known effects on muscle metabolism.
+Potential participants were further screened for injuries causing impaired strength and/or affecting their ability to perform RT, symptoms or history of disease or known adverse reactions to local anaesthetics.
+Included participants were allocated to either an experimental group (EXP, *n* = 11) or a non-training control group (CTRL, *n* = 8; see Table 1 for participant characteristics).
+EXP performed a 12 session RT protocol consisting unilateral knee-extension to allow for within-participant comparison exercise volume.
+In EXP, one leg performed constant volume (CONST, 6 sets per session) and the other leg performed variable volume over the course of the study (VAR, four sessions each with 6, 3 and 9 sets per session; Figure 1A).
+CTRL did not partake in RT and were instructed to continue their everyday activities.
+Muscle biopsies were sampled bilaterally in EXP before and 48 h after the first session and 48 h after the fourth, fifth, eight, ninth and twelfth session as wells as after a seven day de-training period.
+Muscle biopsies were obtained from CTRL at three occasions with 48 h separating the first two and 2-4 weeks separating second and third sampling.
+EXP and CTRL performed strength assessments at least seven days prior to the first biopsy sampling and 72 h after the twelwth session (EXP) and 24 h after the last biopsy (EXP and CTRL).
+Appendicular lean mass (Dual-energy X-ray absorptiometry, DXA) and muscle thickness of *m.
+vastus lateralis* were assessed prior to the first biopsy (EXP and CTRL) as wells as before the second to last (EXP) and last (EXP and CTRL) biopsy (Figure 1A).
 
-## Muscle strength and thickness
-Muscle strength was assessed as maximal knee extension isokinetic (90&deg; sec<sup>-1</sup>) and isometric (60&deg; angle, fully extended leg 0&deg;) torque.
+## RT protocol
+Prior to all RT-sessions, participants performed a standardized warm-up consisting of 5 min ergometer cycling (RPE 12-14), followed by ten repetitions each of push-ups, sit-ups and back-extensions.
+After warm-up participants performed unilateral knee-extension with the prescribed number of sets.
+Each set was prescribed with 10 repetitions maximum (RM).
+When sets was completed with either fewer (8) or more (12) repetitions than 10, the resistance was adjusted accordingly.
+Inter-set rest periods were 90 sec.
+Participants performed the first set alternating between the right and left leg every other session and the set of the contralateral leg was performed in the rest-period of the first leg.
+The second session of each four-session block (session 2, 6 and 10) were performed at a sub-maximal resistance (~90% of the previous session) with the same number of repetitions.
+In each session, after the knee-extension exercise, participants completed two sets each of three upper-body exercises (bench press, lateral pull-down and shoulder press).
+Prior to each session, participants were asked to rate muscle soreness on a scale from 0 to 10 (0 = no soreness, 10 = hurting very bad, e.g.
+not being able to walk properly) [@Vickers2010].
+After each session, participants were asked to rate their perceived exertion (Borg CR10).
+After the completion of each session participants were given a standardized drink to aid recovery (0.15 g × kg-1 protein, 11.2 g × kg-1 carbohydrates and 0.5 g × kg-1 fat).
+
+## Muscle strength, body composition and muscle thickness assessments
+Muscle strength was assessed as maximal voluntary isokinetic (90&deg; sec<sup>-1</sup>) and isometric (60&deg; angle, fully extended leg 0&deg;) knee extension torque.
 After a brief warm-up (5-min cycling, RPE 12-14), participants were seated and secured in the individually adjusted dynamometer.
-The test protocol consisted of unilateral knee-extensions.
 Participants were instructed to gradually increase their effort during three warm-up repetitions (50, 60 and, 70% of subjective maximal effort).
@@
 
-Muscle thickness was assessed using (...) ultra sound.
+Muscle thickness was assessed using (...) ultra sound. DXA.
 

### 2020-08-24 manuscript.Rmd
update manuscript

@@
 
-Our skeletal muscles not only constitutes our locomotive apparatus but also serve as our primary storage site of amino acids.
-This makes the musculoskeletal system a critical target for interventions that promotes health across the lifespan [@RN763].
+Our skeletal muscles not only constitutes our locomotive apparatus but also serve as our primary storage site of amino-acids.
+This makes the skeletal muscle system a critical target for interventions that promotes health across the lifespan [@RN763]<!-- Wolfe 2006 -->.
 The beneficial effects of resistance-training (RT) are in large part mediated through exercise-induced changes in the balance between muscular protein breakdown and synthesis, with successful RT leading to net protein accretion and muscle hypertrophy.
 A traditional view of exercise-induced muscle hypertrophy has been that muscle protein accumulates as a consequence of repeated episodes of elevated post-exercise protein synthesis in the acute phase, resulting in a net-positive protein balance over time.
-Indeed, one bout of resistance exercise acutely up-regulates skeletal muscle protein synthesis up to 48-hr after exercise [@RN786] and subsequently, repeated bouts would lead to accumulation of muscle protein over time [@RN1272; @RN1642].
-In recent years, this view has been supplemented with evidence showing that chronic RT also leads to an increased resting synthetic rate of muscle protein [@RN1521; @RN1866; @RN1897].
-An increase in the resting protein synthetic rate has been postulated to be related to an increased translational capacity, i.e.
-accumulation of ribosomes [@RN1897; @RN1912], a notion supported by observations relating exercise-induced changes in total RNA, a proxy marker for ribosome abundance, to muscle hypertrophy [@RN1644, @RN1755].
-Furthermore, in vitro studies shows that direct inhibition of ribosomal RNA (rRNA) transcription or inhibition of up-stream transcription factors leading to halted biosynthesis of ribosomes results in diminished cellular growth [@RN1755; @RN1754; @RN884].
-Formation of new ribosomes is a complex and energy demanding process involving the whole transcriptional apparatus [@RN1920, @RN1820].
-In skeletal muscle, a single bout of resistance exercise elevates markers of ribosome biogenesis in skeletal muscle evident from both Figueiredo et al.
-and Nader et al.
-who evaluated immature forms of rRNA in response to exercise in the untrained state [@RN1644; @RN1037].
-Studies’ examining repeated bouts of resistance exercise suggests that RT leads to accumulation of ribosomes.
-Bickel et al.
-showed that two subsequent bouts of a hypertrophic stimulus increased total RNA at 72 h after a second bout of RT [@RN1520].
-Following three weeks of RT, Brook et al.
-reported an increase in total RNA with no further increase after three additional weeks of training [@RN1642].
+Indeed, one bout of resistance exercise acutely up-regulates skeletal muscle protein synthesis up to 48-hr after exercise [@RN786]<!-- Phillips 1997 -->
+and subsequently, repeated bouts would lead to accumulation of muscle protein over time
+[@RN1272<!-- Phillips review 2014 -->;
+@RN1642]<!-- Brook 2015 experiment -->. 
+In recent years, this view has been supplemented with evidence showing that chronic RT also leads to an increased resting synthetic rate of muscle protein
+[@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> @RN1866;<!-- Wilkingson 2008 -->
+@RN1897]<!-- Reidy 2017 Increased rested state protein synthesis -->.
+An increase in the resting protein synthetic rate has been postulated to be related to an increased translational capacity, i.e. accumulation of ribosomes
+[@RN1897;<!-- Reidy 2017 -->
+@RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->,
+a notion supported by observations relating exercise-induced changes in total RNA, a proxy marker for ribosome abundance, to muscle hypertrophy
+[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->@RN1755; <!-- Stec Ribosome bio higher in high responders -->
+@Hammarstrm2020]<!-- Volume dependent RNA accum relates to volume depend hyper -->.
+Furthermore, inhibition of ribosomal RNA (rRNA) transcription or inhibition of up-stream transcription factors leads to halted biosynthesis of ribosomes resulting in diminished cellular growth
+[@RN1755<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->;
+@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
+@vonWalden2016]<!-- Inhibition of Pol 1 rDNA tracsription and myotube growth -->.
+
+Formation of new ribosomes is a complex and energy demanding process involving the whole transcriptional apparatus
+[@RN1920<!-- Warner review yeast -->,
+@RN1820;<!-- Moss review -->
+@vonWalden2016]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
+In skeletal muscle, a single bout of resistance exercise elevates markers of ribosome biogenesis in skeletal muscle evident from studies investigating immature forms of rRNA in response to exercise in the untrained state
+[@REFNEEDED<!-- References on acute 45S changes -->;
+@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->].
+Studies examining repeated bouts of resistance exercise suggests that RT leads to accumulation of ribosomes.
+Bickel *et al*.
+showed that two subsequent bouts of a hypertrophic stimulus increased total RNA at 72 h after a second bout of RT
+[@RN1520]<!-- Total RNA increase 72 h after a secound bout -->.
+Following three weeks of RT, Brook et al. reported an increase in total RNA with no further increase after three additional weeks of training
+[@RN1642]<!-- Brook 2015 6 week time-course -->.
 The findings of Brook et al.

### 2020-08-24 manuscript.Rmd
manuscript updates

@@
 @RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->].
-Studies examining repeated bouts of resistance exercise suggests that RT leads to accumulation of ribosomes.
+Studies examining repeated bouts of resistance exercise suggests that RT leads to accumulation of ribosomes, measured as total RNA, with increased number of completed sessions
+[@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
+@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
+@Hammarstrom2020; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
+@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
+@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
+@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
+@RN1897<!-- 11% 36 sessions young -->].
+Time-course data is however sparse as only a limited number of studies have investigated exercise induced accumulation over multiple time-points.
 Bickel *et al*.
-showed that two subsequent bouts of a hypertrophic stimulus increased total RNA at 72 h after a second bout of RT
-[@RN1520]<!-- Total RNA increase 72 h after a secound bout -->.
-Following three weeks of RT, Brook et al. reported an increase in total RNA with no further increase after three additional weeks of training
-[@RN1642]<!-- Brook 2015 6 week time-course -->.
-The findings of Brook et al.
-fits with observations from our laboratory where we saw a ~30% increase in total RNA per tissue weight after two weeks of RT.
-Following an additional 10 weeks of RT the RNA content tended to normalized without reaching baseline values (Hammarström et al.
-2017 unpublished observation).
+showed that two subsequent bouts of a hypertrophic stimulus increased total RNA to the highest levels at 72 h after a second bout of RT [@RN1520]<!-- Total RNA increase 72 h after a second bout -->.
+We have previously reported peak RNA levels 48 h after the fourth training session with a tendency towards normalization after 31 sessions
+[@Hammarstrom2020].
+This is similar to what was reported by Brook *et al.* who reported peak values after nine sessions followed by a slight decrease to after 18 sessions
+[@RN1809].
 These data suggests that ribosome accumulation per tissue weight, reaches a plateau in the early phase of RT.

### 2020-08-25 tables.Rmd
started analysis dxa and us

---
title: "tables"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

source("./R/libs.R")

```

```{r table1-characteristics}

# Read dxa data

dxa_whole <- read_excel("./data/tr010_dxa_standarddefinitions.xlsx") %>%
  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:`HeleVev%fett`) %>%
  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:`HeleVev%fett`) %>%
  mutate(value = gsub("kg", "", value), 
         value = gsub("[[:space:]]", "", value, fixed = FALSE), 
         value = as.numeric(value)) %>%

  separate(Etternavn, into = c("Study", "participant")) %>%
  dplyr::select(participant, Date = Målingsdato, variable, value) %>%
  filter(!(participant %in% c("P08", "P20", "P17"))) %>%
  group_by(participant) %>%
  mutate(time = if_else(Date == min(Date), "S0", 
                        if_else(Date == max(Date), "post1w", 
                                "S12")), 
         leg = if_else(variable == "BeinhøyreMagermasse", "R", "L")) %>%
  print()

# DXA weight stature

read_excel("./data/tr010_dxa.xlsx") %>%
    inner_join(read_excel("./data/leg_randomization.xlsx") %>%
               dplyr::select(participant, sex)) %>%

  group_by(participant, sex, group) %>%
  summarise(mass = mean(bodymass_manual.weight, na.rm = TRUE), 
            stature = mean(height, na.rm = TRUE)) %>%
  group_by(group, sex) %>%
  summarise(mass = mean(stature), 
            stature = mean(stature), 
            n = n()) %>%

  print()

```

### 2020-08-25 R/figure-source.R

@@ -7,12 +7,48 @@
 library(tidyverse)
 library(cowplot)
 
+# Color scale 
 
+multiple.color <- "#636363"
+single.color <- "#bdbdbd"
+
+# line sizes
+line_size <- 0.3
+
+# text sizes
+label.size <- 10
+
+text.size <- 8
+
+
+
+sequential.colors <- c("#1b9e77",
+                       "#d95f02",
+                       "#7570b3",
+                       "#e7298a",
+                       "#66a61e",
+                       "#e6ab02")
 
 # Functions for plotting  #####
 
-
-
+plot_theme <- function() {
+  
+  theme_bw() +
+    theme(panel.grid = element_blank(), 
+          panel.border = element_blank(), 
+          axis.line = element_line(size = line_size), 
+          axis.ticks = element_line(size = line_size), 
+          axis.text = element_text(color = "black", size = 7), 
+          axis.title = element_text(color = "black", size = 7),
+          legend.title = element_blank(), 
+          legend.background = element_rect(fill = "white"),
+          legend.margin = margin(t = 0, r = 1, b = 1, l = 1, unit = "pt"),
+          legend.key = element_rect(fill = "white"),
+          legend.position = c(0.85, 0.9)) 
+  
+  
+  
+}
 
 
 

### 2020-08-25 R/lean-mass-us-analysis.R

@@ -1,0 +1,365 @@
+
+### Leg lean mass and muscle thickness ##############
+
+
+
+
+
+# Load libraries
+source("R/libs.R")
+
+
+
+# Leg randomization data ##############
+
+leg <- read_excel("./data/leg_randomization.xlsx")
+
+
+# DXA data ################
+
+# This file has a modified ROI using only the thigh muscle
+dxa_modified_roi <- read_excel("./data/tr010_dxa.xlsx") %>%
+  dplyr::select(participant, group, time, leg, lean_g) %>%
+  print()
+
+
+
+# Full data set
+dxa <- read_excel("./data/tr010_dxa_standarddefinitions.xlsx") %>%
+  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:`HeleVev%fett`) %>%
+  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:`HeleVev%fett`) %>%
+  mutate(value = gsub("kg", "", value), 
+         value = gsub("[[:space:]]", "", value, fixed = FALSE), 
+         value = as.numeric(value)) %>%
+  filter(variable %in% c("BeinhøyreMagermasse", "BeinvenstreMagermasse")) %>%
+  separate(Etternavn, into = c("Study", "participant")) %>%
+  dplyr::select(participant, Date = Målingsdato, variable, value) %>%
+  filter(!(participant %in% c("P08", "P20", "P17"))) %>%
+  group_by(participant) %>%
+  mutate(time = if_else(Date == min(Date), "S0", 
+                        if_else(Date == max(Date), "post1w", 
+                                "S12")), 
+         leg = if_else(variable == "BeinhøyreMagermasse", "R", "L")) %>%
+  dplyr::select(participant, Date, time, leg, lean.mass = value) %>%
+  inner_join(leg) %>%
+  mutate(cond = if_else(cond == "ctrl_leg", "ctrl", cond), 
+         time = factor(time, levels = c("S0", "S12", "post1w")), 
+         group = if_else(cond == "ctrl", "ctrl", "exp")) %>%
+  dplyr::select(-Date) %>%
+  inner_join(dxa_modified_roi) %>%
+  mutate(lean.mean = (lean.mass + lean_g)/2, 
+         group = factor(group, levels = c("ctrl", "exp")), 
+         time = factor(time, levels = c("S12", "S0", "post1w"))) %>%
+#  filter( group != "ctrl") %>%
[…]

### 2020-08-27 manuscript.Rmd
updates figures

@@
 Primer sequences and their respective average performances are shown in Table 2.
+```{r check-primer-sizes, include = FALSE}
+
+## To check primer sizes ##
+# Be aware that the gel is skewed (overestimating sizes in last wells)
+
+#
+# mw <- read_excel("./data/tr010_primers.xlsx", sheet = "molecular_marker") %>%
+#  mutate(logx = log(X_coord)) %>%
+#   print()
+# mw_primers <- read_excel("./data/tr010_primers.xlsx", sheet = "primers_mw") %>%
+#     mutate(logx = log(X_coord)) %>%
+#   print()
+# 
+# 
+# m1 <- lm(MW ~ poly(logx,3), data = mw)
+#
+# mw %>%
+#  ggplot(aes(logx, MW)) + 
+#  geom_point() + 
+#  geom_smooth(method = "lm", formula = y ~ poly(x, 3))
+# 
+# 
+# mw_primers$mw <-predict(m1, newdata = mw_primers)
+# 
+# 
+# library(ggrepel)
+# read_excel("./data/tr010_primers.xlsx") %>%
+#   mutate(predicted_product_size = rstart - fstart) %>%
+#   dplyr::select(primer = primer_id, predicted_product_size) %>%
+#   inner_join(mw_primers) %>%
+#   ggplot(aes(predicted_product_size, mw, label = primer)) + 
+#   geom_point() + 
+#   ggrepel::geom_text_repel() + 
+#   scale_x_continuous(limits = c(50, 160)) +
+#   scale_y_continuous(limits = c(50, 160))+
+#   geom_abline(slope = 1, intercept = 0)
+#
+
+```
 

### 2020-08-27 Figures/Figure1.R

@@ -15,6 +15,8 @@ source("./R/figure-source.R")
 
 # Colors for plot
 
+
+
 control_color <- "white"
 
 const_color <- "#56B4E9"
@@ -30,8 +32,8 @@ textsize <- 2.5
 # design figure
 
 biopsy_glyph <- "\U2193"
-strength_glyph <-  "\U2299"
-us_glyph <- "\U2295"
+strength_glyph <-  "\U2020"
+us_glyph <- "\U2021"
 
 #### Design graph: EXP ####
 
@@ -45,28 +47,31 @@ design_exp <- expand.grid(cond = c("Variable volume", "Constant volume"),
   annotate("text", x = -2, y = 11.0, label = c("Experimental group"), size = textsize + 0.5) +  
   
   geom_bar(stat = "identity", position = position_dodge(width = 0.55),
+                 
            width = 0.45, 
-           color = "black", 
            size = 0) +
+  
+  
+  
   scale_x_continuous(breaks = seq(1:12), limits = c(-4,15.5), expand = c(0, 0)) +
   
   scale_y_continuous(breaks = c(0, 3, 6, 9, 12), limits = c(0,12), expand = c(0,0), 
                      labels = c("", "3", "6", "9", "")) +
-  scale_fill_manual(values = c(const_color, var_color)) +
+  scale_fill_manual(values = c(color.scale[2], color.scale[3])) +
 
-  annotate("text", x = c(-3.8, -3.8, -3.8), y = c(9.1, 8.2, 6), 
+  annotate("text", x = c(0, 0, -2), y = c(9.1, 8.2, 6), 
            label = c("Muscle biopsy", "Muscle thickness", "Strength tests"), 
-           vjust = 0, hjust = 0, alpha = 0.8, size = textsize) + 
+           vjust = 0, hjust = 1, alpha = 0.8, size = textsize) + 
   
   # Biopsies
-  annotate("text", x = c(0.5, 1.5, 4.5, 5.5, 8.5, 9.5, 12.5, 14.5), y = rep(9, 8), 
-           label = rep(biopsy_glyph, 8), vjust = 0, family = "Cambria") +
+  annotate("text", x = c(0.5, 1.5, 4.5, 5.5, 8.5, 9.5, 12.5, 14.5), y = rep(9.2, 8), 
+           label = rep(biopsy_glyph, 8), vjust = 0) +
   
[…]

### 2020-08-27 R/figure-source.R

@@ -6,8 +6,13 @@
 
 library(tidyverse)
 library(cowplot)
+library(ggpattern)
 
 # Color scale 
+color.scale <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
+                 "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+
 
 multiple.color <- "#636363"
 single.color <- "#bdbdbd"

### 2020-08-27 R/scraps/lit-review-data.R

@@ -1,0 +1,25 @@
+
+
+temp <- read_excel("./review/rna-literature-review.xlsx") %>%
+  filter(subject != "all") %>%
+  print()
+
+
+
+
+
+## Digitize 
+# library(digitize)
+# bickel <- digitize("./review/digitize/Bickel2005.tif")
+
+
+
+
+read_excel("./review/rna-literature-review.xlsx") %>%
+  filter(statistics != "individual") %>%
+  group_by(subject, study_name, n_sessions) %>%
+  summarise(rna_fc = mean(rna_fc, na.rm = TRUE)) %>%
+  arrange(n_sessions, study_name) %>%
+  ggplot(aes(n_sessions, rna_fc, group = paste(study_name, subject))) + 
+  geom_point() + geom_line() 
+

### 2020-08-27 tables.Rmd
updates figures

@@
 ---
-title: "tables"
-output: html_document
+title: "Tables"
+output: rmarkdown::word_document
 editor_options: 
@@
 source("./R/libs.R")
+library(flextable)
 
@@
          leg = if_else(variable == "BeinhøyreMagermasse", "R", "L")) %>%
-  print()
+  filter(variable %in% c("HeleBeinmasse", 
+                         "HeleFettmasse", 
+                         "HeleVevmasse", 
+                         "HeleMagermasse" ,
+                         "Vekt", 
+                         "HeleTotalmasse")) %>%
+  pivot_wider(names_from = variable, values_from = value) %>%
+  mutate(fat_rel = 100*(HeleFettmasse/HeleTotalmasse)) %>%
+  dplyr::select(participant, time, Vekt,HeleTotalmasse, fat_rel) %>%
+  filter(time == "S0")
 
@@
 
-read_excel("./data/tr010_dxa.xlsx") %>%
-    inner_join(read_excel("./data/leg_randomization.xlsx") %>%
-               dplyr::select(participant, sex)) %>%
+read_excel("./data/tr010_age_participant.xlsx") %>%
+  inner_join(dxa_whole) %>%
+  mutate(age = difftime(recruited, dob)/365.2425, 
+         bmi = weight/(height/100)^2) %>%
+  dplyr::select(participant, group, height, weight,age, sex,bmi, fat_rel) %>%
 
-  group_by(participant, sex, group) %>%
-  summarise(mass = mean(bodymass_manual.weight, na.rm = TRUE), 
-            stature = mean(height, na.rm = TRUE)) %>%
   group_by(group, sex) %>%
-  summarise(mass = mean(stature), 
-            stature = mean(stature), 
-            n = n()) %>%
+  summarise(n = n(), 
+            Age = mean(age, na.rm = TRUE), 
+            age.sd = sd(age, na.rm = TRUE),
+            mass = mean(weight), 
+            mass.sd = sd(weight),
+            stature = mean(height),
+            stature.sd = sd(height), 
+            BMI = mean(bmi), 
+            bmi.sd = sd(bmi), 
+            relfat = mean(fat_rel), 
+            relfat.sd = mean(fat_rel) ) %>%
+  ungroup() %>%
+  mutate(n = as.character(n), 
+         age = paste0(sprintf("%.1f", round(Age, 1)), " (", sprintf("%.1f", round(age.sd, 1)), ")"),
+         mass = paste0(sprintf("%.1f", round(mass, 1)), " (", sprintf("%.1f", round(mass.sd, 1)), ")"),
+         stature = paste0(sprintf("%.1f", round(stature, 1)), " (", sprintf("%.1f", round(stature.sd, 1)), ")"),
+         BMI = paste0(sprintf("%.1f", round(BMI, 1)), " (", sprintf("%.1f", round(bmi.sd, 1)), ")"),
+         fat = paste0(sprintf("%.1f", round(relfat, 1)), " (", sprintf("%.1f", round(relfat.sd, 1)), ")"))%>%
+  dplyr::select(group, sex, n, age, mass, stature, BMI, fat) %>%
+  pivot_longer(names_to = "variable", values_to = "values", n:fat) %>%
+  pivot_wider(names_from = c(sex,group), values_from = values) %>%
 
-  print()
+  mutate(variable =  factor(variable,
+                          levels = c("n", "age", "mass", "stature", "BMI", "fat"), 
+                          labels = c("n", "Age (yrs)", "Body mass, (kg)", "Stature (cm)", "Body mass index (kg/m<sup>2</sup>)", "Body fat (%)"))) %>%
+
+  dplyr::select(variable, F_experiment, M_experiment, F_control, M_control) %>%
+  flextable() %>% 
+  italic(i = ~variable == "n", j = 1, italic = TRUE) %>%
+  compose(part = "body", i = 5, j = 1,
+         value = as_paragraph("Body mass index (kg m",
+                              as_sup("-2"), ")")) %>%
+
+  set_header_labels(variable = "",
+    F_experiment = "Female", M_experiment = "Male",
+    F_control = "Female", M_control = "Male") %>%
+
+   add_header_row(values = c("", "Experimental group", "Experimental group", "Control group", "Control group"), top = TRUE) %>%
+  merge_at(i = 1, j = 2:3, part = "header") %>%
+  merge_at(i = 1, j = 4:5, part = "header") %>%
+  align(align = "center", part = "header") %>%
+    fontsize(size = 8, part = "body") %>%
+    fontsize(size = 10, part = "header") %>%
+  add_header_lines(values = "Table 1. Participant characteristics.") %>%
+  autofit()
 
 ```
+
+<br>
+<br>
+
+```{r table-2-primers}
+
+# Read qpcr data
+qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")
+
+primer_performance <- qdat %>%
+  group_by(target) %>%
+  summarise(mean.cq = mean(cq, na.rm = TRUE), 
+            sd.cq = sd(cq, na.rm = TRUE), 
+            mean.eff = mean(eff))
+
+merge_custom <- function(ft, x, columns){
+  z <- rle(x)
+  rows_at <- cumsum(z$lengths) - z$lengths + 1
+
+  for(i in seq_along(rows_at)){
+    for(j in columns)
+      ft <- merge_at(x = ft, i = seq( rows_at[i], rows_at[i] + z$lengths[i] - 1), j = j)
+  }
+
+  ft
+}
+
+# Read primer data 
+primer_table <- read_excel("./data/tr010_primers.xlsx", sheet = 1) %>%
+  dplyr::select(target = primer_id, forward, reverse, full_name, symbol_ext) %>%
+  inner_join(primer_performance) %>%
+  mutate(performance = paste0(sprintf("%.1f", round(mean.cq,1)), " (", sprintf("%.1f",        round(sd.cq,1)), "), E = ", sprintf("%.2f", round(mean.eff,2))), 
+         forward = paste0("F: 5'-", forward, "-3'"), 
+         reverse = paste0("R: 3'-", reverse, "-5'")) %>%
+
+  dplyr::select(symbol_ext, full_name, forward, reverse, performance) %>%
+  pivot_longer(names_to = "primer", values_to = "seq", forward:reverse) %>%
+dplyr::select(Symbol = symbol_ext, 'Transcript name' = full_name, Sequence = seq, 'Mean Cq (SD) and efficiency' = performance)
+
+primer_table %>%
+  flextable() %>%
+  merge_custom(x = primer_table$Symbol, columns = c(1, 2, 4)) %>%
+  add_header_lines(values = "Table 2. Primer sequences and average performance.") %>%
+      fontsize(size = 8, part = "body") %>%
+    fontsize(size = 10, part = "header") %>%
+  autofit()
+
+```

### 2020-08-27 Figures/Figure-rrna-qpcr.R

@@ -17,6 +17,16 @@
 # Sequences can be found : https://www.ncbi.nlm.nih.gov/nuccore/NR_146144.1
 # Add primer sequences to plot and annotate mature segments
 
+# Primer locations
+primers <- read_excel("./data/tr010_primers.xlsx") %>%
+  filter(primer_id %in% c("47S F1R1", 
+                          "45S F5R5", 
+                          "45SITS F12R12", 
+                          "5.8S F2R2", 
+                          "28S F2R2", 
+                          "18S F2R2")) %>%
+  print()
+
 
 
 data.frame(x = c(1, 13500), 
@@ -30,7 +40,18 @@ data.frame(x = c(1, 13500),
   # 5.8S segment
   annotate("segment", x = 6601, xend = 6757, y = 1.5, yend = 1.5, size = 2.5) +
   # 28S segment
-  annotate("segment", x = 7925, xend = 12990, y = 1.5, yend = 1.5, size = 2.5) 
+  annotate("segment", x = 7925, xend = 12990, y = 1.5, yend = 1.5, size = 2.5) +
+  
+  geom_segment(data = primers, aes(y = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56), 
+                                   yend = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56), 
+                                   x= on45S_start, xend = on45S_end), 
+               size = 4) +
+  scale_y_continuous(limits = c(1.48, 1.58)) +
+  plot_theme() + 
+  theme(axis.title = element_blank(), 
+          axis.text = element_blank(), 
+        axis.line = element_blank(),
+        axis.ticks = element_blank())
 
 
 

### 2020-08-27 manuscript.Rmd
Figure 1 complete draft

@@
 title: "Ribosome accumulation during early phase resistance training"
-author: Daniel Hammarström^1,2,£^&ast;,Sjur Øfsteng^2,£^  Bent R. Rønnestad^2^, Stian Ellefsen^1,3^
+author: Daniel Hammarström^1,2,£^&ast;,Sjur Øfsteng^2,£^, Lise Koll^3^,   Bent R. Rønnestad^2^, Stian Ellefsen^1,3^
 date: ""
@@
 # Results
+```{r training-data}
+
+load.stats <- read_excel("./data/tr010_training.xlsx", sheet = 1, na = "NA") %>%
+  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+  filter(exercise == "legext") %>%
+  mutate(set.load = repetitions * load, 
+         week = if_else(session %in% c(1:4), "W1", if_else(session %in% c(5:8), "W2", "W3"))) %>%
+  group_by(participant, cond, session, week) %>%
+  summarise(load = mean(load, na.rm = TRUE)) %>%
+  mutate(week = factor(week, levels = c("W1", "W2", "W3"))) %>%
+  print()
+
+load.m1 <- lme(log(load) ~ week + week:cond, random = list(participant = ~1), 
+               data = load.stats, method = "ML")
+load.m0 <- lme(log(load) ~ week, random = list(participant = ~1), 
+               data = load.stats, method = "ML")
+
+load.anova <- anova(load.m0, load.m1)
+
+load.intervals <- data.frame(exp(intervals(load.m0)$fixed) - 1)
+
+load.inr.w2 <- paste0(sprintf("%.0f",load.intervals[2, 2]*100), 
+                      "%, 95% CI: [",
+                      sprintf("%.0f",load.intervals[2, 1]*100), 
+                      ", ",
+                      sprintf("%.0f",load.intervals[2, 3]*100), 
+                      "]")
+load.inr.w3 <- paste0(sprintf("%.0f",load.intervals[3, 2]*100), 
+                      "%, 95% CI: [",
+                      sprintf("%.0f",load.intervals[3, 1]*100), 
+                      ", ",
+                      sprintf("%.0f",load.intervals[3, 3]*100), 
+                      "]")
+
+```
+
+All participants successfully completed their prescribed sessions.
+The different training protocols resulted in diverging loading profiles (load &#215; repetitions) over the course of the study (Figure 1B).
+The intensity of exercise (load at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`)  four-session block, respectively.
+Compared to the non-training control group, isokinetic strength increased to Session 12 in the experimental group, with an attenuated difference seen after de-training (Figure 1C).
+Average isometric strength showed the same general pattern although with considerable more uncertainty indicated by wider 95% CI (Figure 1C).
+Muscle thickness of the *m.
+vastus lateralis* increased in the experimental group compared to the control group both after Session 12 and after eight days of de-training (Figure 1D).
+There were no indications of differences between training conditions in strength or muscle thickness changes.
 

### 2020-08-27 Figures/Figure1.R

@@ -8,7 +8,7 @@
 
 
 source("./R/figure-source.R")
-
+source("./R/libs.R")
 
 
 
@@ -207,7 +207,95 @@ design_full_biopsies <- plot_grid(design_exp,
                                   rel_heights = c(1, 0.6)) 
 
 
-# Panel B #############################
+# Panel X. Training data plot ##########################
+
+
+
+### Exploratory graphs #
+training_sum <-  read_excel("./data/tr010_training.xlsx", sheet = 1, na = "NA") %>%
+
+  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+  filter(exercise == "legext") %>%
+ 
+  mutate(set.load = repetitions * load) %>%
+  group_by(participant, leg, cond, session) %>%
+  summarise(total.load = sum(set.load)) %>%
+  group_by(cond, session) %>%
+  summarise(total.load = mean(total.load)) 
+
+
+
+training_load <- read_excel("./data/tr010_training.xlsx", sheet = 1, na = "NA") %>%
+  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+  filter(exercise == "legext") %>%
+  mutate(set.load = repetitions * load) %>%
+  group_by(participant, leg, cond, session) %>%
+  summarise(total.load = sum(set.load)) %>%
+  ggplot(aes(session, total.load, group = paste0(participant, leg), color = cond)) + 
+  geom_line(alpha = 0.6) + 
+  geom_line(data = training_sum, 
+            aes(session, total.load, group = cond, color = cond), 
+            size = 1.5) +
+  scale_color_manual(values = c(color.scale[2], color.scale[3], color.scale[4])) +
+  geom_text(data = data.frame(session = c(1, 1),
+                              total.load = c(5000, 5000), 
+                              
+                              lab = c("Constant\nvolume","Variable\nvolume"),
+                              cond = factor(c("const", "var"), 
+                                            levels = c("const", "var"))), 
+            aes(label = lab, group = NULL, color = NULL), 
+            size = textsize + 0.5, 
+            hjust = 0) +
+  
+  scale_y_continuous(limits = c(0, 6000), breaks = c(0, 2000, 4000, 6000), expand = c(0, 0)) +
[…]

### 2020-08-27 R/scraps/ecss_figures.R
Figure 1 complete draft

@@ -1,0 +1,2301 @@
+######## ECSS conference presentation figures ####################
+
+# Filename: ./figures/ecss2019_figures.R
+# Author: Daniel H
+# Notes:
+
+# Purpose: Prepare plots for ecss presentation
+
+# Output: Plots saved with ggsave
+
+######## -----------------------####################
+
+source("./R/libraries.R")
+
+
+dpi <- 230
+
+#### Theme for presenatation
+
+# Changes plot background and text color
+
+presentation_theme <- function() {
+  
+    theme(axis.text = element_text(size = 7, color = "black"),
+          axis.title =  element_text(size = 7, color = "black"), 
+          plot.background = element_rect(color = "white", fill = "white"))
+  
+}
+
+
+
+### Background figure on Total RNA and protein synthesis
+
+millward <- read_excel("./presentationData/litrev.xlsx", sheet = "millward1973") 
+
+millward1973_fig1 <- millward %>%
+  mutate(group = if_else(group == "A", "Protein fed", "Protein starved")) %>%
+  filter(group != "Protein fed") %>%
+  ggplot(aes(RNA, proteinSynthesis, fill = group, color = group)) +
+  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) + 
+  geom_point(shape = 21, size = 3, alpha = 0.4) +
+  scale_x_continuous(name = TeX("$RNA\\;(\\mu g \\; mg^{-1} \\, protein)$"), 
+                     breaks = c(0, 4, 8, 12, 16),
+                     minor_breaks = c(2, 6, 10, 14),
+                     limits = c(0, 16), 
+                     expand = c(0, 0)) +
+  scale_y_continuous(name = TeX("$Protein\\,synthesis\\,rate\\;(days^{-1})"), 
+                     breaks = c(0, 0.04, 0.08, 0.12, 0.16, 0.2),
+                     minor_breaks = c(0.02, 0.06, 0.10, 0.14, 0.18, 0.22),
+                     limits = c(0, 0.22), 
+                     expand = c(0, 0)) + 
+  publr_theme() +
+  presentation_theme() +
+  theme(legend.position = c(0.2, 0.8), 
+        legend.title = element_blank(), 
+        legend.background = element_rect(color = "white")) +
+  scale_fill_manual(values = c("black")) +
+  scale_color_manual(values = c("black")) +
[…]

### 2020-08-28 Figures/Figure1.R
chaos in the modeling, figure 1

@@ -302,17 +302,186 @@ average_intensity <- load.stats %>%
 
 # Panel X: Strength tests -- ######################################
 
+
+# Statistical treatment of training effects (strength and muscle thickness) #
+# Mixed effects models are fitted within experimental to evaluate condition effects
+# and between intervention and control to evaluate effects of training. 
+# Random effects are reduced from maximal models to the most parsimonious 
+# based on lme4::rePCA and LRT tests of random effects.
+
+# Between-group comparisons (intervention vs. control) are done in two separate models 
+# where post values in intervention (post/post1w) are compared to controls (post).
+
+# Data is saved/plotted as estimated mean changes within groups (emmeans package) and
+# interaction effects. 
+
 strength_long <- read_excel("./data/tr010_humac.xlsx") %>%
   inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
   mutate(time = if_else(timepoint %in% c("B1", "B2", "fam"), "baseline", timepoint), 
          time = if_else(time == "post_ctrl", "post", time),
          cond = if_else(cond == "ctrl_leg", "ctrl", cond)) %>%
+
   dplyr::select(participant, sex, time, leg, group,cond, isokinetic_torque, isometric_torque) %>%
-  group_by(participant,sex, time, group, cond) %>%
+
+  group_by(participant,sex,leg, time, group, cond) %>%
   summarise(isok = mean(isokinetic_torque, na.rm = TRUE), 
             isom = mean(isometric_torque, na.rm = TRUE)) %>%
-  pivot_longer(names_to = "type", values_to= "torque", cols = isok:isom)
+  pivot_longer(names_to = "type", values_to= "torque", cols = isok:isom) %>%
+  pivot_wider(names_from = time, values_from = torque) %>%
+  mutate(post1w = if_else(group == "con", post, post1w)) %>%
+  pivot_longer(names_to = "time", values_to = "torque", cols = baseline:post1w) %>%
+  print()
 
+# Between condition differences in strength increases ######
+
+# Isokinetic data 
+
+
+ m1 <- lmer(torque ~ time + time:cond + (time|participant), 
+            data = strength_long[strength_long$type == "isok" & strength_long$group == "int",])
+ 
+ summary(rePCA(m1))
+# Singular, all variation is explained at the intercept, removing correlation reducing slopes
+
[…]

### 2020-08-28 R/libs.R

@@ -14,6 +14,7 @@ library(ggrepel)
 library(nlme)
 library(emmeans)
 library(mgcv)
+library(multcomp)
 
 
 

### 2020-08-29 Figures/Figure1.R
updated multcomp solutions

@@ -415,7 +415,7 @@ summary(rePCA(m1))
 # the fit is singular, most of the variance is contained in the 
 # intercept, simplify by removing correlation
 
-m2 <- lmer(torque ~  time_pp * group  + group:time_pp:detrain + (time_pp||participant), 
+m2 <- lmer(torque ~  time_pp * group  + group:time_pp:detrain + (dummy(time_pp, "post")||participant), 
            data = strength_long2[strength_long2$type == "isok", ])
 
 
@@ -429,6 +429,40 @@ m3 <- lmer(torque ~  time_pp * group + group:time_pp:detrain + (1|participant),
 
 # Check if random effects structure matters
 anova(m1, m2, m3) # No, the simplest model is the best
+plot(m3)
+
+## Isometric models 
+
+m1.isom <- lmer(log(torque) ~  time_pp * group + group:time_pp:detrain  + (1 + time_pp|participant), 
+           data = strength_long2[strength_long2$type == "isom", ])
+
+summary(rePCA(m1.isom)) 
+plot(m1.isom)
+# Not a singular fit, seeing if simplification may improve the model.
+
+m2.isom <- lmer(log(torque) ~  time_pp * group  + group:time_pp:detrain + (dummy(time_pp, "post")||participant), 
+           data = strength_long2[strength_long2$type == "isom", ])
+
+
+summary(rePCA(m2.isom)) # Some of the variance is contained in the uncorrelated random slope
+# Fitting a simpler model
+m3.isom <- lmer(torque ~  time_pp * group + group:time_pp:detrain + (1|participant), 
+           data = strength_long2[strength_long2$type == "isom", ])
+
+# The model is rank deficient (not all combinations of coefficients exists in the data)
+# it is OK.
+
+# Check if random effects structure matters to model fit
+anova(m1.isom, m2.isom, m3.isom) 
+# No, more elaborate models does not improve fit but som of the variance could be allocated to 
+# random slopes. 
+
+plot(m3.isom)
+
+summary(m1.isom)
+# The random effects structure does not change the inference, but the simplest model has the lowest AIC
+# and no improvement in fit with more elaborate models. The simplest model is preferred.
+
 
 
 # Combine data for plotting 
[…]

### 2020-08-29 R/libs.R
updated multcomp solutions

@@ -12,6 +12,7 @@ library(ggrepel)
 
 # Modeling 
 library(nlme)
+library(lme4)
 library(emmeans)
 library(mgcv)
 library(multcomp)

### 2020-08-29 Figures/Figure-rrna-qpcr.R
started figure 2

@@ -1,60 +1,0 @@
-
-
-#### Figure X RNA/qPCR ###################
-
-
-
-## Panels
-# A: rRNA primers 
-
-
-
-
-
-### rRNA and primers #########
-
-
-# Sequences can be found : https://www.ncbi.nlm.nih.gov/nuccore/NR_146144.1
-# Add primer sequences to plot and annotate mature segments
-
-# Primer locations
-primers <- read_excel("./data/tr010_primers.xlsx") %>%
-  filter(primer_id %in% c("47S F1R1", 
-                          "45S F5R5", 
-                          "45SITS F12R12", 
-                          "5.8S F2R2", 
-                          "28S F2R2", 
-                          "18S F2R2")) %>%
-  print()
-
-
-
-data.frame(x = c(1, 13500), 
-           y = c(1, 2)) %>%
-  ggplot(aes(x, y)) +
-  
-  # Full 45S segment 
-  annotate("segment", x = 1, xend = 13351, y = 1.5, yend = 1.5, size = 0.5) +
-  # 18S segment
-  annotate("segment", x = 3655, xend = 5523, y = 1.5, yend = 1.5, size = 2.5) +
-  # 5.8S segment
-  annotate("segment", x = 6601, xend = 6757, y = 1.5, yend = 1.5, size = 2.5) +
-  # 28S segment
-  annotate("segment", x = 7925, xend = 12990, y = 1.5, yend = 1.5, size = 2.5) +
-  
-  geom_segment(data = primers, aes(y = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56), 
-                                   yend = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56), 
-                                   x= on45S_start, xend = on45S_end), 
-               size = 4) +
-  scale_y_continuous(limits = c(1.48, 1.58)) +
-  plot_theme() + 
-  theme(axis.title = element_blank(), 
-          axis.text = element_blank(), 
-        axis.line = element_blank(),
-        axis.ticks = element_blank())
-
-
-
-
-
-

### 2020-08-29 Figures/Figure2.R

@@ -1,0 +1,261 @@
+############# Figure 1 ##################################
+
+
+## Panels
+# A: Study design
+# B: Performed volume per condition
+# C/D: Strength and body composition changes. 
+
+
+source("./R/figure-source.R")
+source("./R/libs.R")
+
+
+# 45S primers graphics #############################
+
+### rRNA and primers #########
+
+
+# Sequences can be found : https://www.ncbi.nlm.nih.gov/nuccore/NR_146144.1
+# Add primer sequences to plot and annotate mature segments
+
+# Primer locations
+primers <- read_excel("./data/tr010_primers.xlsx") %>%
+  filter(primer_id %in% c("rRNA47S F1R1", 
+                          "rRNA45S F5R5", 
+                          "rRNA45SITS F12R12", 
+                          "rRNA5.8S F2R2", 
+                          "rRNA28S F2R2", 
+                          "rRNA18S F2R2")) %>%
+  print()
+
+
+
+primers_fig <- data.frame(x = c(1, 13500), 
+           y = c(1, 2)) %>%
+  ggplot(aes(x, y)) +
+  
+  # Full 45S segment 
+  annotate("segment", x = 1, xend = 13351, y = 1.5, yend = 1.5, size = 0.5) +
+  # 18S segment
+  annotate("segment", x = 3655, xend = 5523, y = 1.5, yend = 1.5, size = 2.5) +
+  # 5.8S segment
+  annotate("segment", x = 6601, xend = 6757, y = 1.5, yend = 1.5, size = 2.5) +
+  # 28S segment
+  annotate("segment", x = 7925, xend = 12990, y = 1.5, yend = 1.5, size = 2.5) +
+  
+  geom_segment(data = primers, aes(y = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56), 
+                                   yend = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56), 
+                                   x= on45S_start, xend = on45S_end), 
+               size = 4) +
+  
+  geom_text_repel(data = primers, aes(y = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56),
+                                      x= on45S_start, label = symbol_ext), 
+                  size = 2.2, 
+                  nudge_y = 0.01) +
+  
+  scale_y_continuous(limits = c(1.48, 1.58)) +
+  plot_theme() + 
+  theme(axis.title = element_blank(), 
+        axis.text = element_blank(), 
+        axis.line = element_blank(),
[…]

### 2020-08-29 R/total-rna-compile.R

@@ -31,7 +31,10 @@ for(i in 1:length(files)){
 }
 
 
-
+tot_rna <- bind_rows(results) %>%
+  dplyr::select(Sample, dilution = `Dilution factor`, concentration = ...7) %>%
+  mutate(concentration = concentration * dilution) %>%
+  dplyr::select(-dilution)
 
 
 
@@ -45,7 +48,8 @@ p18_samples <- tot_rna %>%
                         sample = c("mRNA1","mRNA1", "mRNA2", "mRNA2", "mRNA2", "mRNA1" ),
                         time = c("postctrl", "S0", "S0", "postctrl", "S0", "S0"), 
                         leg = c("L", "R", "L", "L", "R", "L"))) %>%
-  mutate(rna = concentration * 20) %>%
+  mutate(concentration = ((concentration/3)/(50*(10/0.51))) * (40 *(10/0.51)) ,
+         rna = concentration * 20) %>%
   inner_join(read_excel("./data/tissue/tr010_tissue.xlsx", na = "NA") %>%
                filter(participant == "P18", sample != "prot")) %>%
   mutate(series = 1) %>%
@@ -67,7 +71,7 @@ tot_rna_round2 <- tot_rna %>%
   # RNA = Absorbance * 40 * (10/0.51) * dilutionfactor (= 3)
   # [The above was mistakenly calculated as DNA in raw data:
   # DNA = Absorbance * 50 * (10/0.51)]
-  mutate(concentration = ((concentration/3)/(50*(10/0.51))) * (40 *(10/0.51)) * 3,
+  mutate(concentration = ((concentration/3)/(50*(10/0.51))) * (40 *(10/0.51)) ,
          rna = concentration * 25) %>%
   dplyr::select(participant, series, sample, leg, time, tissue_weight, rna) %>%
   print()
@@ -137,16 +141,16 @@ rna$pred<- predict(l.mod)
 
 # plot predicted vs. standardized residuals
 rna %>%
-  mutate(outlier = if_else(resid < -2.5| resid > 2.5, "out", "in")) %>%
+  mutate(outlier = if_else(resid < -2| resid > 2, "out", "in")) %>%
   # filter(resid > 3) %>%
-  ggplot(aes(log(tissue_weight), log(rna), color = outlier)) + geom_point() 
+  ggplot(aes(log(tissue_weight), log(rna), color = time)) + geom_point() 
 
 
 
 ### Save data 
 
[…]

### 2020-08-29 Figures/Figure1.R
started figure 2

@@ -264,41 +264,6 @@ training_load <- read_excel("./data/tr010_training.xlsx", sheet = 1, na = "NA") 
 
 
 
-average_intensity <- load.stats %>%
-  group_by(cond, week) %>%
-  summarise(m = mean(load, na.rm = TRUE), 
-            s = sd(load, na.rm = TRUE)) %>%
-  mutate(week = factor(week, levels = c("W1", "W2", "W3"), labels = c("Week 1", "Week 2", "Week 3"))) %>%
-  ggplot(aes(week, m, fill = cond)) +
-  geom_errorbar(aes(ymin = m - s, ymax = m + s), position = position_dodge(width = 0.6), width = 0.2) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.6), width = 0.5, color = "black", size = 0.2) +
-  geom_segment(aes(x = 1, xend = 1.95, y = 50, yend = 50), size = line_size) +
-  geom_segment(aes(x = 2.05, xend = 3, y = 56, yend = 56), size = line_size) +
-  geom_segment(aes(x = 0.8, xend = 1.2, y = 42.5, yend = 42.5), size = line_size) +
-  geom_segment(aes(x = 1.8, xend = 2.2, y = 47.5, yend = 47.5), size = line_size) +
-  geom_segment(aes(x = 2.8, xend = 3.2, y = 53.2, yend = 53.2), size = line_size) +
-  geom_segment(aes(x = 3, xend = 3, y = 56, yend = 53.2), size = line_size) +
-  geom_segment(aes(x = 2.05, xend = 2.05, y = 56, yend = 47.5), size = line_size) +
-  geom_segment(aes(x = 1.95, xend = 1.95, y = 50, yend = 47.5), size = line_size) +
-  geom_segment(aes(x = 1, xend = 1, y = 50, yend = 42.5), size = line_size) +
-
-  
-  
-  
-
-  scale_y_continuous(breaks = c(0, 20, 40, 60), expand = c(0, 0), limits = c(0, 60)) + 
-  scale_fill_manual(values = c(const_color, var_color)) +
-  ylab("Average 10RM (kg)") +
-  theme(axis.title.x = element_blank(), 
-        legend.position = "none") +
-  plot_theme() +
-  labs(caption = "Mean + SD")
-
-
-
-
-
-
 
 # Panel X: Strength tests -- ######################################
 
[…]

### 2020-08-29 R/total-rna-analysis.R

@@ -92,13 +92,13 @@ pr %>%
 
 # No splines
 m0 <- gam(rna ~ tw + cond + time.c +  s(participant, bs = "re"), 
-          data = temp, method = "ML")
+          data = time_course, method = "ML")
 # Splines but not per group
 m1 <- gam(rna ~ tw + cond +  s(time.c,  k = 7) +  s(participant, bs = "re"), 
-          data = temp, method = "ML")
+          data = time_course, method = "ML")
 # Splines per group
 m2 <- gam(rna ~ tw + cond + s(time.c, by = cond,  k = 7) +  s(participant, bs = "re"), 
-          data = temp, method = "ML")
+          data = time_course, method = "ML")
 
 # Splines improves fit (as with polynomials above)
 anova(m0, m1, m2, test = "Chisq")
@@ -108,13 +108,14 @@ AIC(m1, m2)
 
 # REML estimation full model
 m3 <- gam(rna ~ tw +  s(time.c, k = 7,  by = cond) + cond +  s(participant, bs = "re"), 
-          data = temp, method = "REML", select = TRUE)
+          data = time_course, method = "REML", select = TRUE)
 
 
+library(ggeffects)
 ## Prediction from the full model 
 pr <- ggpredict(m3, c("time.c", "cond"), type = "re")
 
-plot(pr)
+str(pr)
 
 # Prediction from the reduced model 
 pr <- ggpredict(m2, c("time.c"), type = "re")
@@ -194,7 +195,7 @@ gam.check(m2)
 
 plot(m1, pages = 1, unconditional = FALSE)
 
-pr <- ggpredict(m1, c("time.c", "cond"), type = "re")
+pr <- ggpredict(m2, c("time.c", "cond"), type = "re")
 
 plot(pr)
 

### 2020-09-02 Figures/Figure1.R
Finnish draft of fig 1

@@ -1,6 +1,15 @@
 ############# Figure 1 ##################################
 
 
+
+# Note: To select estimates set bayes to TRU for Bayesian inference and FALSE
+# for frequentist inference
+
+bayes <- TRUE
+
+
+
+
 ## Panels
 # A: Study design
 # B: Performed volume per condition
@@ -267,291 +276,17 @@ training_load <- read_excel("./data/tr010_training.xlsx", sheet = 1, na = "NA") 
 
 # Panel X: Strength tests -- ######################################
 
-
-# Statistical treatment of training effects (strength and muscle thickness) #
-# Mixed effects models are fitted within experimental to evaluate condition effects
-# and between intervention and control to evaluate effects of training. 
-# Random effects are reduced from maximal models to the most parsimonious 
-# based on lme4::rePCA and LRT tests of random effects.
-
-# Between-group comparisons (intervention vs. control) are done in two separate models 
-# where post values in intervention (post/post1w) are compared to controls (post).
-
-# Data is saved/plotted as estimated mean changes within groups (emmeans package) and
-# interaction effects. 
-
-strength_long <- read_excel("./data/tr010_humac.xlsx") %>%
-  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
-  mutate(time = if_else(timepoint %in% c("B1", "B2", "fam"), "baseline", timepoint), 
-         time = if_else(time == "post_ctrl", "post", time),
-         cond = if_else(cond == "ctrl_leg", "ctrl", cond)) %>%
-
-  dplyr::select(participant, 
-                sex, 
-                time,
-                leg, 
-                group,
-                cond, 
-                isok = isokinetic_torque,
-                isom = isometric_torque) %>%
-  print()
-
-
-
-# Between condition differences in strength increases ######
-
-# Isokinetic data 
-
-
- m1 <- lmer(isok ~ time + time:cond + (time|participant), 
-            data = strength_long[strength_long$group == "int",])
- 
- summary(rePCA(m1))
-# Singular, all variation is explained at the intercept, removing correlation reducing slopes
-
- m2 <- lmer(isok ~ time + time:cond + (dummy(time, "post1w")||participant), 
-            data = strength_long[strength_long$group == "int",])
- 
- summary(rePCA(m2))
-
- anova(m1, m2)
-# Does not improve fit, simplify further, removing random slopes
- m3 <- lmer(isok ~  time + time:cond + (1|participant), 
[…]

### 2020-09-02 R/strength-freq.R

@@ -195,7 +195,7 @@ em.isom <- emmeans(m3.isom, specs = ~ tx)
 # object.  
 em.isok
 
-# To create contrats, each levels can be specified by a indicator vector. 
+# To create contrasts, each levels can be specified by a indicator vector. 
 
 baseline_con_train  =  c(1, 0, 0, 0, 0)
 post_con_train      =  c(0, 1, 0, 0, 0) 

### 2020-09-02 R/strength-bayes.R
Finnish draft of fig 1

@@ -47,6 +47,7 @@ strength_long <- read_excel("./data/tr010_humac.xlsx") %>%
 # vs. change in the control group). 
 
 
+
 strength_long2 <- strength_long %>%
   # Time coefficients are set as baseline/post
   mutate(time_pp = if_else(time == "post1w", "post", time),
@@ -64,53 +65,75 @@ strength_long2 <- strength_long %>%
   print()
 
 
-
 m1.isok <- brm(isok ~  tx + (1|participant),
                
-               warmup = 500, # number of samples before sampling
-               iter = 3000,  # number of mcmc iterations 
+               warmup = 1000, # number of samples before sampling
+               iter = 4000,  # number of mcmc iterations 
                cores = 4, # number of cores used in sampling
                chains = 4, # number of chains
                seed = 123, # seed for reproducible results
                
                data = strength_long2)
 
+# m2.isok <- brm(isok ~  tx + (tx||participant),
+#                
+#                warmup = 1000, # number of samples before sampling
+#                iter = 4000,  # number of mcmc iterations 
+#                cores = 4, # number of cores used in sampling
+#                chains = 4, # number of chains
+#                seed = 123, # seed for reproducible results
+#                
+#                data = strength_long2)
+# 
+# To check what model is better, use loo. 
+loo(m1.isok, m2.isok)
 
+# Model 2 which would involve slightly more complex 
+# random effect does not fit better. Keep model 1
+
+rm(m2.isok)
+
+
+# launch_shinystan(m1.isok) # to check diganstics in shiny interface.
 ## Model diagnostics -- Posterior predictive Check
 
 # Plots data and predictive values
-pp_check(m1.isok)
+pp_check(m1.isok) 
+
+# Predictive values match the data... good. 
+
+# plot chains
+plot(m1.isok)
+          
+summary(m1.isok)
 
 
-pp_check(m1.isok, type = "stat_grouped", stat = "mean", group = "tx")
-pp_check(m1.isok, type = "stat_2d", stat = c("max", "min"))
+saveRDS(m1.isok, "./data/derivedData/strength-bayes/m1.isok.RDS") 
 
-pp_check(m1.isok, type = "ribbon_grouped",  group = "tx") 
-           
+
 
 
 m1.isom <- brm(isom ~  tx + (1|participant),
-               
+              
                warmup = 500, # number of samples before sampling
[…]

### 2020-09-02 R/us-freq-bayes.R

@@ -1,0 +1,238 @@
+# Ultrasound Bayesian/frequentist estimates ###############################
+
+library(brms)
+library(tidybayes)
+library(readxl)
+library(tidyverse)
+library(lme4)
+
+
+
+# Principally the same model is used for US data. Lesser data points are 
+# available in this data set as pre testing is only done once. 
+
+
+# Combine data sets 
+us_data <- rbind(read_excel("./data/ultrasound/ultrasound_data.xlsx") %>%
+                   inner_join(read_csv("./data/ultrasound/ultrasound_codekey.csv")) %>%
+                   mutate(leg = gsub("VL", "", leg)) %>%
+                   inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+                   dplyr::select(participant, time, leg, sex, cond, code, length) %>%
+                   mutate(group = if_else(participant %in% paste("P", c(1:7,19:23), sep = ""), "experiment", "control")) %>%
+                   group_by(participant, time, leg, sex, cond, group) %>%
+                   summarise(thickness = mean(length, na.rm = TRUE)) %>%
+                   ungroup(), 
+                 read_excel("./data/ultrasound/ultrasound_data_2019.xlsx") %>%
+                   inner_join(read_csv("./data/ultrasound/ultrasound_codekey_2019.csv")) %>%
+                   mutate(leg = gsub("VL", "", leg)) %>%
+                   inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+                   dplyr::select(participant, time, leg, sex, cond, code, length) %>%
+                   mutate(group = if_else(participant %in% paste("P", c(1:7, 19:23), sep = ""), "experiment", "control")) %>%
+                   group_by(participant, time, leg, sex, cond, group) %>%
+                   summarise(thickness = mean(length, na.rm = TRUE)) %>%
+                   ungroup()) %>%
+  
+  mutate(time_pp = if_else(time == "post1w", "post", time),
+         # The de-training period get its own coefficient
+         detrain = if_else(time == "post1w" & group == "experiment", "detrain", "train"),
+         # The effect of de training will be added to the model --within-- the intervention group 
+         detrain = factor(detrain, levels = c("train", "detrain")), 
+         time = factor(time, levels = c("pre", "post", "post1w")), 
+         time_pp = factor(time_pp, levels = c("pre", "post"))) %>%
+  
+  print()
+
+
+
+
+## Model between conditions differences 
+# 
[…]

### 2020-09-03 Figures/Figure2.R
updates to fig 1 and 2

@@ -3,6 +3,8 @@
 
 ## Panels
 
+# Use Bayesian estimates?
+bayes <- TRUE
 
 
 source("./R/figure-source.R")
@@ -42,15 +44,15 @@ primers_fig <- data.frame(x = c(1, 13500),
   # 28S segment
   annotate("segment", x = 7925, xend = 12990, y = 1.5, yend = 1.5, size = 2.5) +
   
-  geom_segment(data = primers, aes(y = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56), 
-                                   yend = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56), 
+  geom_segment(data = primers, aes(y = c(1.54, 1.52, 1.53, 1.54, 1.55, 1.56), 
+                                   yend = c(1.54, 1.52, 1.53, 1.54, 1.55, 1.56), 
                                    x= on45S_start, xend = on45S_end), 
                size = 4) +
   
-  geom_text_repel(data = primers, aes(y = c(1.51, 1.52, 1.53, 1.54, 1.55, 1.56),
+  geom_text_repel(data = primers, aes(y = c(1.54, 1.52, 1.53, 1.54, 1.55, 1.56),
                                       x= on45S_start, label = symbol_ext), 
                   size = 2.2, 
-                  nudge_y = 0.01) +
+                  nudge_x = 100) +
   
   scale_y_continuous(limits = c(1.48, 1.58)) +
   plot_theme() + 
@@ -64,239 +66,208 @@ primers_fig <- data.frame(x = c(1, 13500),
 
 # Total RNA analysis and plot #################################################
 
-# load full data set
-rna <- readRDS("./data/derivedData/tot-rna/tot-rna.RDS")
+# load model comparisons between 
 
-
-rna_complete <- rna %>%
- # filter(outlier == "in") %>% # removes outlier based on RNA to tissue weight relationship
-  dplyr::select(participant, series, sample, leg, time, cond, tissue_weight, rna) %>%
-  mutate(time.c = gsub("S", "", time), 
-         detrain = if_else(time.c == "post1w", "detrain", "train"),
-         time.c = if_else(time.c == "post1w", "12", time.c),
-         time.c = if_else(time.c == "postctrl", "12", time.c),
-         time.c = gsub("c", "", time.c), 
-         time.c = as.numeric(time.c), 
-         time = factor(time, levels = c("S0", "S1","S1c", 
-                                        "S4", "S5", "S8", 
[…]

### 2020-09-03 R/qpcr-analysis-bayes.R

@@ -12,13 +12,6 @@ library(tidyverse)
 
 
 
-# brms offers several 
-
-
-
-
-
-
 # qdat is the compiled data frame containing sample information and 
 # qpcr estimates. Created in qpcr-compile.R. Use this for downstream analyses
 
@@ -66,6 +59,7 @@ nf <- qdat %>%
 
 #### rRNA per tissue weight analysis #### 
 
+# THIW IS WHERE IM AT ###############
 
 qdat.rrna  <- qdat %>%
   group_by(target) %>%
@@ -244,14 +238,60 @@ qpcr_res_int_con <- data.frame(target = rep(c("rRNA18SF2R2"      ,
 
 
 
+#### Model 47S/45S per mature rRNA ##############################
 
 
 
 
+qdat_rrna2 <- qdat.rrna %>%
+  filter(cond != "ctrl_leg") %>%
+  dplyr::select(participant, leg, time, time.c, cond, technical, biological, target,Ra, counts) %>%
+  
+  filter(target %in% c("rRNA18S F2R2", 
+                       "rRNA28S F2R2", 
+                       "rRNA5.8S F2R2", 
+                       "rRNA45S F5R5", 
+                       "rRNA47S F1R1")) %>%
+  
+  mutate(type = if_else(target %in% c("rRNA18S F2R2", 
+                                      "rRNA28S F2R2", 
+                                      "rRNA5.8S F2R2"), 
+                        "mature", "pre")) %>%
+
+  filter(time != "post1w") %>%
+
+  print()
 
 
 
 
+library(mgcv)
+
+qdat_rrna2 %>%
+  filter(rel.counts < 0.001) %>%
+  ggplot(aes(time.c, rel.counts)) + geom_point()
+
+
+m1 <- brm(bf(counts ~ s(time.c, by = target, k = 7) + (1|participant) + (1|technical), 
+             sigma ~ target),
+
+          warmup = 1000, # number of samples before sampling
+          iter = 6000,  # number of mcmc iterations 
+          cores = 4, # number of cores used in sampling
+          chains = 4, # number of chains
+          seed = 123, # seed for reproducible results
+          thin = 5,
+          data = qdat_rrna2)
+
+saveRDS(m1, "./data/derivedData/temp/qpcr_smooth.RDS")
+m1 <- readRDS("./data/derivedData/temp/qpcr_smooth.RDS")
+
+conditional_effects(m1)
+
+
+
+
+?brm
 
 
 

### 2020-09-03 manuscript.Rmd
updates to fig 1 and 2

@@
   summarise(load = mean(load, na.rm = TRUE)) %>%
-  mutate(week = factor(week, levels = c("W1", "W2", "W3"))) %>%
-  print()
+  mutate(week = factor(week, levels = c("W1", "W2", "W3")))
 
@@
 
+```{r total-rna}
+
+# Segmented model analysis in text 
+# Read the segmented model 
+tc.m2 <- readRDS("./data/derivedData/total-rna-analysis/tc.m2.RDS")
+
+time1.hyp <- hypothesis(tc.m2, "time1 = 0")
+time2.hyp <- hypothesis(tc.m2, c("time2 + time1 = 0")) # The slope (on log scale) between session 4 and 8
+time3.hyp <- hypothesis(tc.m2, c("time2 + time1 +time3 = 0")) # The slope on log scale between session 8 and 12
+
+slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,2]) - 1)),
+                           "%, 95% CI: [",
+                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,2]) - 1)),
+                           "%, [",
+                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,2]) - 1)),
+                           "%, [",
+                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+```
+
+Compared to control, a single session of RT led to robust increase in 45S and 47S pre-rRNA (Figure 2B) expressed per unit tissue weight.
+These were also increased compared to control after twelve sessions together with the 45S ITS and mature rRNA species 28S and 18S.
+After eight days of rest, only 18S remained elevated above control group  (Figure 2B).
+This general pattern of rRNA expression was reflected also in total RNA expressed per unit tissue weight were a robust increase compared to contral was seen after session twelve (Figure 2B).
+Within the experimental group, no differences were seen between volume conditions.
+Most of the increase was seen in after the first four sessions where the average increase was `r slope0_4` per session in the intervention group as a whole.
+Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not different from zero (`r slope4_8` and `r slope8_12`).
+
 # Discussion

### 2020-09-03 R/figure-source.R

@@ -7,7 +7,7 @@
 library(tidyverse)
 library(cowplot)
 library(ggpattern)
-
+library(ggtext)
 # Color scale 
 color.scale <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                  "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

### 2020-09-03 R/total-rna-analysis.R

@@ -1,13 +1,20 @@
 ######## Total RNA analysis ########################
 
 # Load libraries and data 
-source("./R/libs.R")
+
+library(tidyverse)
+library(tidybayes)
+library(lme4)
+library(brms)
+library(readxl)
+library(emmeans)
+library(gamm4)
 
 rna <- readRDS("./data/derivedData/tot-rna/tot-rna.RDS")
 
 
 
-### 
+
 
 
 rna_complete <- rna %>%
@@ -27,6 +34,135 @@ rna_complete <- rna %>%
 
 
 
+# Control vs. intervention in total RNA per mg tissue
+
+# The same setup for the data set is used as with other variables (qPCR, US, strength).
+# The tx variable is used to model combinations of time, group and detrain factors. 
+
+rna_data2 <- rna_complete %>%
+  filter(time %in% c("S0", "S1", "S1c",  "S12", "postctrl", "post1w")) %>%
+  mutate(Time = if_else(time == "S1c", "S1", 
+                        if_else(time %in% c("post1w", "postctrl", "S12"), 
+                                "post", as.character(time))), 
+         tw = tissue_weight - mean(tissue_weight), 
+         time = factor(Time, levels = c("S0", "S1", "post")), 
+         group = if_else(cond == "ctrl_leg", "con", "int"), 
+         tx = paste0(time, "_", group, "_", detrain), 
+         tx = factor(tx, levels = c("S0_con_train", 
+                                    "S1_con_train", 
+                                    "post_con_train", 
+                                    "S0_int_train", 
+                                    "S1_int_train", 
+                                    "post_int_train", 
+                                    "post_int_detrain"))) %>%
+
+  # Values are averaged over leg/time-point
+  
+  
+    group_by(participant, time, leg, group, tx) %>%
+  summarise(rna = mean(rna, na.rm = TRUE), 
+            tissue_weight = mean(tissue_weight)) %>%
+
+  print()
+  
+
+
+
+
+m1 <- lmer(log(rna/tissue_weight) ~ tx + (1|participant), 
+           data = rna_data2)
+
+
+# Combine data for plotting 
+
+# Emmeans can be used as the contrats specified are combinations of factors. 
+# emmeans::emmeans is used to create a emmGrid object:
+
+em <- emmeans(m1, specs =  ~ tx)
+
[…]

### 2020-09-03 R/total-rna-compile.R

@@ -143,19 +143,32 @@ rna$pred<- predict(l.mod)
 rna %>%
   mutate(outlier = if_else(resid < -2| resid > 2, "out", "in")) %>%
   # filter(resid > 3) %>%
-  ggplot(aes(log(tissue_weight), log(rna), color = time)) + geom_point() 
+  ggplot(aes(log(tissue_weight), log(rna), color = outlier)) + geom_point() 
 
 
 
 ### Save data 
 
 rna.save <- rna %>%
-#  mutate(outlier = if_else(resid < -2.5| resid > 2.5, "out", "in")) %>%
+  mutate(outlier = if_else(resid < -2| resid > 2, "out", "in")) %>%
   inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
   print()
   
 
 saveRDS(rna.save, "./data/derivedData/tot-rna/tot-rna.RDS")
+
+
+
+
+
+
+
+
+
+
+
+
+
 
 
 

### 2020-09-04 R/qpcr-analysis-bayes.R
misc updates qpcr modelling

@@ -18,55 +18,58 @@ library(tidyverse)
 qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")
 
 
-# Remove bad reaction prior to modelling. 
-# Bad reactions are no amplification or estimatyed to cq < 5
+# Remove bad reactions prior to modeling. 
 
-
+# Explorative plots
 qdat %>%
   ggplot(aes(cq, color = target)) + geom_histogram() + 
   facet_wrap(~ target)
 
 qdat %>%
   group_by(target) %>%
-  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 5 * sd(cq, na.rm = TRUE)|
-                             cq < mean(cq, na.rm = TRUE) - 5 * sd(cq, na.rm = TRUE), 
+  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 2 * sd(cq, na.rm = TRUE)|
+                             cq < mean(cq, na.rm = TRUE) - 2 * sd(cq, na.rm = TRUE), 
                            "outlier", "in")) %>%
   
-  ggplot(aes(paste(participant, leg, time), cq, color = target, shape = outlier)) + 
-  geom_point() + theme_minimal() 
+  ggplot(aes(paste(participant, leg, time), cq, 
+             color = target, 
+             shape = outlier, 
+             alpha = outlier)) + 
+  geom_point() + theme_minimal()  + scale_alpha_manual(values = c(0.2, 0.8))
 
-# 5 sd's away from target mean seems to capture the worst reactions.. 
-
-
+# 3 sd's away from target mean seems to capture the worst reactions.. 
 
 
 
 ### Lambda normalization ####
 
-nf <- qdat %>%
+nf <- qdat %>% group_by(target) %>%
+  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 2 * sd(cq, na.rm = TRUE)|
+                             cq < mean(cq, na.rm = TRUE) - 2 * sd(cq, na.rm = TRUE), 
+                           "outlier", "in")) %>%
   filter(target == "Lambda KIT") %>%
-  filter(cq > 5) %>% # Removes bad reactions
+ filter(outlier == "in") %>% # Removes bad reactions
   mutate(nf.w = (eff ^ -cq) * tissue_weight) %>%
   dplyr::select(participant, leg, time, cdna, nf.w) %>%
[…]

### 2020-09-04 R/reference-genes.R

@@ -1,0 +1,118 @@
+#### Reference gene stability in time-course data set ##########
+
+# library(devtools)
+# install_github("dhammarstrom/generefer")
+
+library(tidyverse)
+library(generefer)
+
+
+# Load data
+qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")
+
+
+
+qdat2 <- qdat %>%
+  group_by(target) %>%
+  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 3 * sd(cq, na.rm = TRUE)|
+                             cq < mean(cq, na.rm = TRUE) - 3 * sd(cq, na.rm = TRUE), 
+                           "outlier", "in")) %>%
+  filter(outlier == "in", target != "Lambda KIT") %>%
+  group_by(participant, leg, time, sex, cond, cdna, target) %>%
+  summarise(cq = mean(cq, na.rm = TRUE), 
+            eff = mean(eff, na.rm = TRUE), 
+            tissue_weight = mean(tissue_weight, na.rm = TRUE)) %>%
+  mutate(Ra = -cq * log(eff), # Relative abundance
+         counts = as.integer(round(eff ^ (39-cq), 0)), # counts, ref Matz et al.
+         technical = paste0(participant, leg, time, cdna), 
+         biological = paste0(participant, leg, time), 
+         leg.unique = paste0(participant, leg)) %>%
+  mutate(time.c = gsub("S", "", time), 
+         detrain = if_else(time.c == "post1w", "detrain", "train"),
+         time.c = if_else(time.c == "post1w", "12", time.c),
+         time.c = if_else(time.c == "postctrl", "12", time.c),
+         time.c = gsub("c", "", time.c), 
+         time.c = as.numeric(time.c), 
+         time = factor(time, levels = c("S0", "S1","S1c", 
+                                        "S4", "S5", "S8", 
+                                        "S9", "S12", "post1w", "postctrl"))) %>%
+  #print()
+  ungroup() %>%
+
+  print()
+
+
+qdat2$target
+
+combinations <- combn(unique(data.frame(qdat2)[,7 ]), 2)
+
+
+
+
+
+
+
+temp <- qdat2 %>%
+  filter(cond != "ctrl_leg", 
+         time != "post1w", 
+         !(target %in% c("rRNA45S F1R1", 
+                         "rRNA45S F5R5", 
+                         "rRNA45SITS F12R12", 
[…]

### 2020-09-10 manuscript.Rmd
updates to western and qpcr

@@
 
+```{r western-data}
+
+```
+
+Protein levels of RPS6 and UBF increased only after twelve sessions in the training group compared to control (Figure 3A) and stayed elevated after eight days of rest.
+There were no differences between volume-conditions.
+The average increase in expression levels over twelve sessions were linear in both and estimated to 4 and 8% per session for RPS6 and UBF, respectively.
+
 # Discussion

### 2020-09-10 Figures/Figure2.R

@@ -97,9 +97,12 @@ complete_rrna_comp <- comp_rna %>%
   dplyr::select(target, comparison, estimate:upper.CL) %>%
   rbind(qpcr_res_int_con %>%
           dplyr::select(target, comparison, estimate = Estimate, lower.CL = CI.Lower, upper.CL = CI.Upper)) %>%
+  filter(!(target %in% c( "UBTFF4R4",
+                        "UBTFF6R6", 
+                        "rpS6F2R2" ))) %>%
   print()
 
-
+ 
 
 # Create an annotation data frame
 

### 2020-09-10 R/qpcr-analysis-bayes.R

@@ -63,7 +63,6 @@ nf <- qdat %>% group_by(target) %>%
 
 #### rRNA per tissue weight analysis #### 
 
-
 qdat.rrna  <- qdat %>%
   group_by(target) %>%
   mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 2 * sd(cq, na.rm = TRUE)|
@@ -78,7 +77,9 @@ qdat.rrna  <- qdat %>%
   filter(target %in% c("rRNA18S F2R2",      "rRNA28S F2R2",     
                        "rRNA5.8S F2R2",       
                        "rRNA45S F5R5",      "rRNA45SITS F12R12", 
-                       "rRNA47S F1R1",      "rRNA5S F3R3")) %>%
+                       "rRNA47S F1R1",      "rRNA5S F3R3", 
+                       "UBTF F4R4", "UBTF F6R6", 
+                       "rpS6 F2R2")) %>%
   mutate(Ra = -cq * log(eff), # Relative abundance
          counts = as.integer(round(eff ^ (39-cq), 0)), # counts, ref Matz et al.
          technical = paste0(participant, leg, time, cdna), 
@@ -129,7 +130,8 @@ qdat.ctrl <- qdat.rrna %>%
 
 
 
-m1 <- brm(bf(counts ~ 0 + target + target:tx + (1|participant) + (1|technical) + offset(nf.w), 
+m1 <- brm(bf(counts ~ 0 + target + target:tx + (1|participant) + (1|technical) + 
+               offset(nf.w), 
              shape ~ target),
           family = negbinomial(),
           warmup = 1000, # number of samples before sampling
@@ -143,7 +145,10 @@ m1 <- brm(bf(counts ~ 0 + target + target:tx + (1|participant) + (1|technical) +
 # Save model 
 saveRDS(m1, "./data/derivedData/qpcr-analysis-bayes/ctrl_vs_int.RDS")
 
-pp_check(m1)
+m1 <- readRDS("./data/derivedData/qpcr-analysis-bayes/ctrl_vs_int.RDS")
+
+
+pp_check(m1, type = "stat", stat = "median")
 summary(m1)
 # combine factors for hypotheses
 
@@ -157,6 +162,11 @@ h <- hypothesis(m1,
           "targetrRNA28SF2R2:txS1_int_train -     targetrRNA28SF2R2:txpre_int_train =      targetrRNA28SF2R2:txS1_con_train",
[…]

### 2020-09-10 R/wester_analysis.R

@@ -1,0 +1,529 @@
+### Western analaysis 
+
+library(tidyverse)
+library(brms)
+library(bayesplot)
+library(lme4)
+
+western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
+
+
+
+
+west.dat <- western_data %>%
+ filter(time %in% c("S0", "S1c", "S1", "S12", "postctrl", "post1w")) %>%
+  mutate(group = if_else(cond == "ctrl_leg", "con", "int"), 
+         detrain = if_else(time == "post1w", "detrain", "train"), 
+         time = if_else(time == "S0", "pre", 
+                        if_else(time == "S1c", "S1", 
+                                if_else(time %in% c("postctrl", "post1w", "S12"), "post", time))), 
+         time = factor(time, levels = c("pre", "S1", "post")), 
+         tx = paste(time, group, detrain, sep = "_"), 
+         tx = factor(tx, levels = c("pre_con_train",
+                                    "S1_con_train",
+                                    "post_con_train",
+                                    "pre_int_train",
+                                    "S1_int_train",
+                                    "post_int_train", 
+                                    "post_int_detrain")), 
+         sample2 = paste0(participant, leg, time)) %>%
+  group_by(participant, tx, target) %>%
+  summarise(expression = mean(expression, na.rm = TRUE)) %>%
+  mutate(log.expr = log(expression)) %>%
+
+  print()
+  
+
+
+
+s6.m1 <- brm(bf(log.expr ~ tx + (1|participant)),
+
+          warmup = 1000, # number of samples before sampling
+          iter = 4000,  # number of mcmc iterations 
+          cores = 4, # number of cores used in sampling
+          chains = 4, # number of chains
+          seed = 5, # seed for reproducible results
+      #    control = list(adapt_delta = 0.95), 
+          data = west.dat[west.dat$target == "t-s6",])
+
+
+ubf.m1 <- brm(bf(log.expr ~ tx + (1|participant)),
+             
+             warmup = 1000, # number of samples before sampling
+             iter = 4000,  # number of mcmc iterations 
+             cores = 4, # number of cores used in sampling
+             chains = 4, # number of chains
+             seed = 5, # seed for reproducible results
[…]

### 2020-09-10 R/wester_compile.R

@@ -1,0 +1,96 @@
+######## Western blot analysis analysis ####################
+
+# Filename: ./R/wester_analysis.R
+# Author: Daniel H
+# Notes:
+
+# Purpose: Analyze western blot data
+
+# Output: Model estimates
+
+######## -----------------------####################
+
+## Packages ####
+source("./R/libs.R")
+
+
+### Load data  round 1 ####
+
+western_data_round1 <- read_excel("./data/wetlab/western/tr010_western_round1.xlsx", na = "NA", sheet = "total_protein") %>%
+  filter(participant != "LADDER") %>%
+  dplyr::select(participant, ExtractionNR, round, gel, well, mean.gray1:total.protein2) %>%
+  inner_join(read_excel("./data/wetlab/tr010_mRNASamples_round1.xlsx", na = "NA") %>%
+               dplyr::select(participant, leg, time, ExtractionNR)) %>%
+  inner_join(read_excel("./data/wetlab/western/tr010_western_round1.xlsx", na = "NA", sheet = "ecl") %>%
+               dplyr::select(round, gel, well, target, signal)) %>%
+  mutate(gel = paste0(round,"_", gel), 
+         sample = paste0(participant, "_", ExtractionNR)) %>%
+  rowwise() %>%
+  ### memcode stain - background = total protein ### 
+  mutate(total.protein = mean(c(total.protein1, total.protein2)) - mean(c(mean.gray1, mean.gray2, mean.gray3))) %>%
+  ungroup() %>%
+  dplyr::select(participant,sample, leg, time, gel, well, target, total.protein, signal, round) %>%
+  group_by(gel) %>%
+  ### Normalizes the total protein stain per 
+  mutate(tp.factor = total.protein) %>%
+  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+  dplyr::select(participant,sample, leg, cond, sex, time, gel, well, target, tp.factor, signal, round) %>%
+  ungroup() %>%
+  mutate(expression = signal / tp.factor) %>%
+  filter(target %in% c("t-s6", "t-UBF")) %>%
+  print()
+
+
+western_data_round2 <- read_excel("./data/wetlab/western/tr010_western_round2.xlsx", na = "NA", sheet = "total_protein") %>%
+
+  dplyr::select(participant, ser, num, round, gel, well, total.protein_1:meangray_4) %>%
+  mutate(num = as.character(num)) %>%
[…]

### 2020-09-11 Figures/Figure2.R
updates in qpcr modeling

@@ -87,8 +87,7 @@ if(bayes == FALSE){
 
 ###### Panel X Intervention vs. control qPCR rRNA + total RNA  ##################
 
-
-qpcr_res_int_con <- readRDS("./data/derivedData/qpcr-analysis-bayes/qpcr_res_int_con.RDS")
+qpcr_res_int_con <- readRDS("./data/derivedData/qpcr-analysis-bayes2/qpcr_res_int_con.RDS")
 
 
 complete_rrna_comp <- comp_rna %>%
@@ -96,16 +95,18 @@ complete_rrna_comp <- comp_rna %>%
          target = "totalRNA") %>%
   dplyr::select(target, comparison, estimate:upper.CL) %>%
   rbind(qpcr_res_int_con %>%
-          dplyr::select(target, comparison, estimate = Estimate, lower.CL = CI.Lower, upper.CL = CI.Upper)) %>%
+          filter(model == "tissue") %>%
+          dplyr::select(target, comparison = contrast, estimate, lower.CL , upper.CL)) %>%
 
-  filter(!(target %in% c( "UBTFF4R4",
-                        "UBTFF6R6", 
-                        "rpS6F2R2" ))) %>%
+  filter(!(target %in% c("UBTF F4R4",
+                        "UBTF F6R6", 
+                        "rpS6 F2R2", 
+                        "MyHC1 F1R1", 
+                        "MyHC2A F5R5", 
+                        "MyHC2X F5R5"))) %>%
   print()
 
  
-
-  print()
 
 
 
@@ -114,23 +115,23 @@ complete_rrna_comp <- comp_rna %>%
 
 anno.df <- data.frame(target = unique(complete_rrna_comp$target), 
            label = c("Total RNA", 
-                     "rRNA 18S", 
-                     "rRNA 45S ETS", 
-                     "rRNA 47S ETS", 
-                     "rRNA 5S", 
-                     "rRNA 28S", 
+                     "rRNA 18S",
+                     "rRNA 28S",  
+                     "rRNA 45S ETS",
                      "rRNA 45S ITS", 
-                     "rRNA 5.8S"), 
+                     "rRNA 47S ETS", 
+                     "rRNA 5.8S",
+                     "rRNA 5S" ), 
            time = factor("S1", levels = c("S1", "post")), 
            estimate = 3.5) %>%
   mutate(target = factor(target, levels = c("totalRNA", 
-                                            "rRNA18SF2R2",
-                                            "rRNA5.8SF2R2",
-                                            "rRNA28SF2R2",
[…]

### 2020-09-11 R/qpcr-analysis-bayes.R

@@ -339,7 +339,7 @@ h <- hypothesis(m1,
           
           "targetUBTFF4R4:txpost_con_train = 0",
           "targetUBTFF6R6:txpost_con_train = 0", 
-          "targetrpS6F2R2:txpost_con_train = 0" 
+          "targetrpS6F2R2:txpost_con_train = 0", 
           
           
 

### 2020-09-11 R/qpcr-analysis-bayes2.R

@@ -1,0 +1,575 @@
+
+# mcmcglmm modeling of qpcr
+#
+#
+#
+
+
+# Notes:
+
+# Matz suggested using a poisson mixed-effects model to model qpcr-data 
+# without the use of reference genes as random effects on technical duplictaes
+# could be used to control for technical variation (Matz el al. 2013).
+
+# An implementation of these models is done in the mcmc.qpcr package. Althogh 
+# this is a nice implementation we would like to have more control in the 
+# fitting process and downstream comparisons. To this end we use the 
+# mcmcglmm package directly. 
+
+# Two models are used, in model 1 estimates are retrieved from an "unadjusted model"
+# modeling transcript counts per total RNA (as this is the biological input). Model 2 
+# uses the external reference gene as a normalization factor modeling counts as a rate
+# per tissue weight. The normalization factor is entered as a fixed effect with strong priors
+# with the interpretation being counts per normalization factor. 
+# See discussion on offset in MCMCglmm:
+# https://hannahdugdale.wordpress.com/2016/03/16/why-you-shouldnt-use-the-offset-function-in-mcmcglmm/
+# https://stat.ethz.ch/pipermail/r-sig-mixed-models/2011q3/016817.html
+
+
+# Load packages 
+library(MCMCglmm)
+library(tidybayes)
+library(readxl)
+library(tidyverse)
+library(parallel)
+library(coda)
+
+
+
+# qdat is the compiled data frame containing sample information and 
+# qpcr estimates. Created in qpcr-compile.R. Use this for downstream analyses
+
+qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")
+
+
+
+# Remove bad reactions prior to modeling. 
+
+qdat %>%
+  ggplot(aes(cq, color = target)) + geom_histogram() + 
+  facet_wrap(~ target)
+
+qdat %>%
+  group_by(target) %>%
+  
+  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 2 * sd(cq, na.rm = TRUE)|
+                             cq < mean(cq, na.rm = TRUE) - 2 * sd(cq, na.rm = TRUE), 
+                           "outlier", "in")) %>%
+  
+  ggplot(aes(paste(participant, leg, time), cq, 
+             color = target, 
+             shape = outlier, 
+             alpha = outlier)) + 
[…]

### 2020-09-11 R/wester_analysis.R

@@ -10,6 +10,7 @@ western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
 
 
 
+
 west.dat <- western_data %>%
  filter(time %in% c("S0", "S1c", "S1", "S12", "postctrl", "post1w")) %>%
   mutate(group = if_else(cond == "ctrl_leg", "con", "int"), 

### 2020-09-12 manuscript.Rmd
misc updates

@@
 
+Compared to control, a single session of RT led to robust increase in 45S and 47S pre-rRNA (Figure 2B) expressed per unit tissue weight.
+These were also increased compared to control after twelve sessions together with the 45S ITS and mature rRNA species 28S and 18S.
+After eight days of rest, only 18S remained elevated above control group  (Figure 2B).
+This general pattern of rRNA expression was reflected also in total RNA expressed per unit tissue weight were a robust increase compared to contral was seen after session twelve (Figure 2B).
+Within the experimental group, no differences were seen between volume conditions.
+Most of the increase was seen in after the first four sessions where the average increase was `r slope0_4` per session in the intervention group as a whole.
+Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not different from zero (`r slope4_8` and `r slope8_12`).
+
 # Discussion
@@
 <div id="refs"></div>
+
+# Figure legends
+
+**Figure 1.** Study design (A) training load (B) and training outcomes (muscle strength, C; muscle thickness D).
+
+**Figure 2.**  Total RNA and ribosomal RNA subspecies in response to trainin.
+Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (A).
+There were no robust differences between different volume conditions in the training group (A).
+B shows primer locations targeting different ribosomal RNA subspecies.
+Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and other after the training period (C).
+Error bars shows 95% CI.
+Asterisk in C indicates robust differences between volume conditions (a 95% CI of pairwise differences not containing 0).

### 2020-09-12 Figures/Figure2.R

@@ -64,39 +64,17 @@ primers_fig <- data.frame(x = c(1, 13500),
 
 
 
-# Total RNA analysis and plot #################################################
-
-# load model comparisons between 
-
-if(bayes == TRUE){
-  
-comp_rna <- readRDS("./data/derivedData/total-rna-analysis/comp_rna_bayes.RDS")
-  
-}
-
-if(bayes == FALSE){
-  
-  comp_rna <- readRDS("./data/derivedData/total-rna-analysis/comp_rna_freq.RDS")
-  
-}
 
 
 
-
-
-
-###### Panel X Intervention vs. control qPCR rRNA + total RNA  ##################
+###### Panel X Intervention vs. control qPCR rRNA   ##################
 
 qpcr_res_int_con <- readRDS("./data/derivedData/qpcr-analysis-bayes2/qpcr_res_int_con.RDS")
 
 
-complete_rrna_comp <- comp_rna %>%
-  mutate(comparison = time_group, 
-         target = "totalRNA") %>%
-  dplyr::select(target, comparison, estimate:upper.CL) %>%
-  rbind(qpcr_res_int_con %>%
+complete_rrna_comp <- qpcr_res_int_con %>%
           filter(model == "tissue") %>%
-          dplyr::select(target, comparison = contrast, estimate, lower.CL , upper.CL)) %>%
+          dplyr::select(target, comparison = contrast, estimate, lower.CL , upper.CL) %>%
 
   filter(!(target %in% c("UBTF F4R4",
                         "UBTF F6R6", 
@@ -114,8 +92,7 @@ complete_rrna_comp <- comp_rna %>%
 # Create an annotation data frame
 
 anno.df <- data.frame(target = unique(complete_rrna_comp$target), 
-           label = c("Total RNA", 
-                     "rRNA 18S",
+           label = c("rRNA 18S",
                      "rRNA 28S",  
                      "rRNA 45S ETS",
                      "rRNA 45S ITS", 
@@ -124,8 +101,7 @@ anno.df <- data.frame(target = unique(complete_rrna_comp$target),
                      "rRNA 5S" ), 
            time = factor("S1", levels = c("S1", "post")), 
            estimate = 3.5) %>%
-  mutate(target = factor(target, levels = c("totalRNA", 
-                                            "rRNA18S F2R2",
+  mutate(target = factor(target, levels = c("rRNA18S F2R2",
                                             "rRNA5.8S F2R2",
[…]

### 2020-09-12 R/qpcr-analysis-bayes2.R

@@ -418,6 +418,16 @@ saveRDS(ctrl_vs_int, "./data/derivedData/qpcr-analysis-bayes2/qpcr_res_int_con.R
 
 qdat_int <- qdat.rrna %>%
   filter(cond != "ctrl_leg") %>%
+  
+  mutate(time = factor(time, 
+                       levels = c("S0", 
+                                  "S1", 
+                                  "S4", 
+                                  "S5", 
+                                  "S8", 
+                                  "S9", 
+                                  "S12", 
+                                  "post1w"))) %>%
   print()
 
 
@@ -446,15 +456,15 @@ clusterExport(cl, "qdat_int")
 # use parLapply() to execute 10 runs of MCMCglmm(), each with nitt=100000
 model3 <- parLapply(cl=cl, 1:4, function(i) {
   
-  MCMCglmm(fixed = counts ~ 0 + target + target:time + target:cond + target:time:cond,
+  MCMCglmm(fixed = counts ~ 0 + target + target:time + target:time:cond,
            random = ~participant + technical + idh(target):participant, 
            prior = prior3, 
            family = "poisson", 
            data = qdat_int, 
            rcov = ~idh(target):units,
            thin = 10,
-           burnin = 5000,
-           nitt = 50000)
+           burnin = 10000,
+           nitt = 65000)
   
 }
 )
@@ -481,7 +491,6 @@ prior4 <- list(R = list(V = diag(13), # n genes/targets
 prior4$B$V[1,1] <- 1e-5 # replace the offset variance
 
 
-
 setCores <- 4 # use detectCores() by itself if you want all CPUs
 
 # make the cluster
@@ -496,21 +505,95 @@ clusterExport(cl, "qdat_int")
 # use parLapply() to execute 10 runs of MCMCglmm(), each with nitt=100000
 model4 <- parLapply(cl=cl, 1:4, function(i) {
   
-  MCMCglmm(fixed = counts ~ 0 + log(nf.w) + target + target:time + target:cond + target:time:cond,
+  MCMCglmm(fixed = counts ~ 0 + log(nf.w) + target + target:time + target:time:cond,
            random = ~participant + technical + idh(target):participant, 
            prior = prior4, 
            family = "poisson", 
            data = qdat_int, 
            rcov = ~idh(target):units,
            thin = 10,
-           burnin = 5000,
-           nitt = 50000)
[…]

### 2020-09-13 manuscript.Rmd
updates to figures and results

@@
 title: "Ribosome accumulation during early phase resistance training"
-author: Daniel Hammarström^1,2,£^&ast;,Sjur Øfsteng^2,£^, Lise Koll^3^,   Bent R. Rønnestad^2^, Stian Ellefsen^1,3^
+author: Daniel Hammarström^1,2,£^&ast;,Sjur Øfsteng^2,£^, Bent R. Rønnestad^2^, Stian Ellefsen^1,3^
 date: ""
@@
 
-## Immunohistochemistry
-
 ## Statistics and data analysis
@@
 The intensity of exercise (load at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) four-session block, respectively.
-Compared to the non-training control group, isokinetic strength increased to Session 12 in the experimental group, with an attenuated difference seen after de-training (Figure 1C).
+Isokinetic strength increased to Session 12 in the experimental group compared to the non-training control group, with an attenuated difference seen after de-training (Figure 1C).
 Average isometric strength showed the same general pattern although with considerable more uncertainty indicated by wider 95% CI (Figure 1C).
@@
 
-Compared to control, a single session of RT led to robust increase in 45S and 47S pre-rRNA (Figure 2B) expressed per unit tissue weight.
-These were also increased compared to control after twelve sessions together with the 45S ITS and mature rRNA species 28S and 18S.
-After eight days of rest, only 18S remained elevated above control group (Figure 2B).
-This general pattern of rRNA expression was reflected also in total RNA expressed per unit tissue weight were a robust increase compared to contral was seen after session twelve (Figure 2B).
-Within the experimental group, no differences were seen between volume conditions.
+A single session of RT led to robust increase in 47S and 45S pre-rRNA expressed per unit tissue weight within the training group (Figure 2B) and compared to control (Figure 2C).
+These were also increased compared to control after twelve sessions together with the 45S ITS and mature rRNA species 28S and 18S (Figure 2C).
+After eight days of rest, only 18S and 28S remained elevated above levels measured in the control group (Figure 2B).
+This general pattern of rRNA expression was reflected also in total RNA expressed per unit tissue weight were a robust increase compared to control was seen after session twelve (Figure 2E and F).
+Within the experimental group, only small differences were seen in rRNA subspecies as 45S ETS was robustly higher after 12 sessions in the variable volume condition after an imbalance between conditions prior to training (Figure 2D).
+Total RNA increased similarly in both volume conditions throughout the training period (Figure 2G).
 Most of the increase was seen in after the first four sessions where the average increase was `r slope0_4` per session in the intervention group as a whole.
-Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not different from zero (`r slope4_8` and `r slope8_12`).
+Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not robustly different from zero (`r slope4_8` and `r slope8_12`).
 
 ```{r western-data}
+segmented_protein_results <- readRDS("./data/derivedData/western-analysis/segmented_results.RDS")
+
+s6_slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[6,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[6,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[6,7]) - 1)),
+                   "]")
+
+s6_slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[7,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[7,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[7,7]) - 1)),
+                   "]")
+
+s6_slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[8,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[8,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[8,7]) - 1)),
+                   "]")
+
+ubf_slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[14,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[14,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[14,7]) - 1)),
+                   "]")
+
+ubf_slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[15,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[15,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[15,7]) - 1)),
+                   "]")
+
+ubf_slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[16,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[16,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[16,7]) - 1)),
+                   "]")
 
@@
 
-Protein levels of RPS6 and UBF increased only after twelve sessions in the training group compared to control (Figure 3A) and stayed elevated after eight days of rest.
-There were no differences between volume-conditions.
-The average increase in expression levels over twelve sessions were linear in both and estimated to 4 and 8% per session for RPS6 and UBF, respectively.
-
-Compared to control, a single session of RT led to robust increase in 45S and 47S pre-rRNA (Figure 2B) expressed per unit tissue weight.
-These were also increased compared to control after twelve sessions together with the 45S ITS and mature rRNA species 28S and 18S.
-After eight days of rest, only 18S remained elevated above control group  (Figure 2B).
-This general pattern of rRNA expression was reflected also in total RNA expressed per unit tissue weight were a robust increase compared to contral was seen after session twelve (Figure 2B).
-Within the experimental group, no differences were seen between volume conditions.
-Most of the increase was seen in after the first four sessions where the average increase was `r slope0_4` per session in the intervention group as a whole.
-Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not different from zero (`r slope4_8` and `r slope8_12`).
+Protein levels of RPS6 and UBF increased only after twelve sessions in the training group compared to control  and stayed elevated after eight days of rest (Figure 3A and B).
+The average increase in expression levels over twelve sessions near to linear in both and estimated to `r s6_slope0_4`, `r s6_slope4_8` and `r s6_slope8_12` in RPS6, and `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12` in UBF per session between session 1-4, 4-8 and 8-12, respectively.
 

### 2020-09-13 Figures/Figure2.R

@@ -31,6 +31,9 @@ primers <- read_excel("./data/tr010_primers.xlsx") %>%
 
 
 
+
+
+
 primers_fig <- data.frame(x = c(1, 13500), 
            y = c(1, 2)) %>%
   ggplot(aes(x, y)) +
@@ -44,26 +47,36 @@ primers_fig <- data.frame(x = c(1, 13500),
   # 28S segment
   annotate("segment", x = 7925, xend = 12990, y = 1.5, yend = 1.5, size = 2.5) +
   
-  geom_segment(data = primers, aes(y = c(1.54, 1.52, 1.53, 1.54, 1.55, 1.56), 
-                                   yend = c(1.54, 1.52, 1.53, 1.54, 1.55, 1.56), 
+  geom_segment(data = primers, aes(y = c(1.54, 1.52, 1.528, 1.54, 1.55, 1.56), 
+                                   yend = c(1.54, 1.52, 1.528, 1.54, 1.55, 1.56), 
                                    x= on45S_start, xend = on45S_end), 
                size = 4) +
   
-  geom_text_repel(data = primers, aes(y = c(1.54, 1.52, 1.53, 1.54, 1.55, 1.56),
-                                      x= on45S_start, label = symbol_ext), 
+  geom_text(data = primers, aes(y = c(1.54, 1.52, 1.528, 1.544, 1.55, 1.56),
+                                      x= on45S_end, label = symbol_ext), 
                   size = 2.2, 
-                  nudge_x = 100) +
+          hjust = 0,
+            position = position_nudge(x = 200)) +
   
-  scale_y_continuous(limits = c(1.48, 1.58)) +
+  # Annotations 
+  annotate("text", x = c(3655/2, 5523 - (5523-3655)/2, 
+                         6757 - (6757-6601)/2,
+                         12990 - (12990-7925)/2), 
+                   y = rep(1.48, 4), 
+           color = "gray50",
+           label = c("47/45S ETS", "18S", "5.8S", "28S"), 
+           fontface = "italic",
+  
+           size = 2.2) +
+  
+  
+  
+  scale_y_continuous(limits = c(1.46, 1.58)) +
   plot_theme() + 
   theme(axis.title = element_blank(), 
         axis.text = element_blank(), 
         axis.line = element_blank(),
         axis.ticks = element_blank())
-
-
-
-
 
 
 
@@ -107,7 +120,12 @@ anno.df <- data.frame(target = unique(complete_rrna_comp$target),
[…]

### 2020-09-13 Figures/Figure3.R

@@ -1,0 +1,188 @@
+##-------------------------------------
+## Figure3.R
+##
+## Title: Figure 3
+## Purpose: Make file for Fig 3
+## Author: DH
+##
+##
+##
+##-------------------------------------
+## Notes: 
+# Figure 3 contains wetsern blot data 
+# and corresponding mRNA data (rps6 and UBF)
+#
+#
+#
+#
+#
+#
+#
+## ------------------------------------
+
+
+source("./R/figure-source.R")
+source("./R/libs.R")
+
+
+
+
+
+s6.m1 <- readRDS( "./data/derivedData/western-analysis/s6_tx_m1.RDS")
+ubf.m1 <- readRDS( "./data/derivedData/western-analysis/ubf_tx_m1.RDS")
+
+
+
+tx_contrasts <-  c("txS1_int_train -     txpre_int_train =      txS1_con_train",
+                   # Interaction effects post-training
+                   "txpost_int_train -     txpre_int_train =      txpost_con_train",
+                   # Interaction effects detraining (post1w)
+                   "txpost_int_detrain -     txpre_int_train =      txpost_con_train",
+                   # Whitin group fold change
+                   "txS1_int_train =     txpre_int_train",     
+                   "txpost_int_train =     txpre_int_train",     
+                   "txpost_int_detrain =     txpre_int_train",     
+                   "txS1_con_train = 0",
+                   "txpost_con_train = 0")
+
+
+
+h.s6 <- hypothesis(s6.m1, tx_contrasts)
+h.ubf <- hypothesis(ubf.m1, tx_contrasts)
+
+# Create a data frame for the results 
+tx_results_western <- data.frame(target = rep(c("rpS6","UBF"), each = 8), 
+                                 comparison = rep(c("inter:S1",
+                                                    "inter:post", 
+                                                    "inter:post1w", 
+                                                    
+                                                    "int_S1", 
+                                                    "int_post", 
+                                                    "int_post1w", 
+                                                    
+                                                    "con_S1",
+                                                    "con_post"), 2)) %>%
+  cbind(data.frame(rbind(h.s6$hypothesis[,-1], 
[…]

### 2020-09-13 R/qpcr-analysis-bayes.R

@@ -67,7 +67,7 @@ nf <- qdat %>% group_by(target) %>%
                              cq < mean(cq, na.rm = TRUE) - 2 * sd(cq, na.rm = TRUE), 
                            "outlier", "in")) %>%
   filter(target == "Lambda KIT") %>%
- filter(outlier == "in") %>% # Removes bad reactions
+ filter(outlier == "in")  # Removes bad reactions
 
 nf <- qdat %>%
   filter(target == "Lambda KIT") %>%
@@ -104,7 +104,7 @@ qdat.rrna  <- qdat %>%
 
 #### rRNA per tissue weight analysis #### 
 
-# THIW IS WHERE IM AT ###############
+
 
 qdat.rrna  <- qdat %>%
   group_by(target) %>%
@@ -126,7 +126,6 @@ qdat.rrna  <- qdat %>%
                        "UBTF F4R4", "UBTF F6R6", 
                        "rpS6 F2R2")) %>%
 
-                       "rRNA47S F1R1",      "rRNA5S F3R3")) %>%
 
   mutate(Ra = -cq * log(eff), # Relative abundance
          counts = as.integer(round(eff ^ (39-cq), 0)), # counts, ref Matz et al.

### 2020-09-13 R/wester_analysis.R

@@ -67,6 +67,10 @@ saveRDS(ubf.m1, "./data/derivedData/western-analysis/ubf_tx_m1.RDS")
 
 
 s6.m1 <- readRDS( "./data/derivedData/western-analysis/s6_tx_m1.RDS")
+ubf.m1 <- readRDS( "./data/derivedData/western-analysis/ubf_tx_m1.RDS")
+
+
+
 loo(s6.m1)
 
 # Model checks
@@ -85,84 +89,6 @@ loo(ubf.m1)
 
 summary(s6.m1)
 # combine factors for hypotheses
-
-tx_contrasts <-  c("txS1_int_train -     txpre_int_train =      txS1_con_train",
-                   # Interaction effects post-training
-                   "txpost_int_train -     txpre_int_train =      txpost_con_train",
-                   # Interaction effects detraining (post1w)
-                   "txpost_int_detrain -     txpre_int_train =      txpost_con_train",
-                   # Whitin group fold change
-                   "txS1_int_train =     txpre_int_train",     
-                   "txpost_int_train =     txpre_int_train",     
-                   "txpost_int_detrain =     txpre_int_train",     
-                   "txS1_con_train = 0",
-                   "txpost_con_train = 0")
-
-
-
-h.s6 <- hypothesis(s6.m1, tx_contrasts)
-h.ubf <- hypothesis(ubf.m1, tx_contrasts)
-
-# Create a data frame for the results 
-tx_results_western <- data.frame(target = rep(c("rpS6","UBF"), each = 8), 
-                               comparison = rep(c("inter:S1",
-                                              "inter:post", 
-                                              "inter:post1w", 
-                                      
-                                              "int_S1", 
-                                              "int_post", 
-                                              "int_post1w", 
-                                      
-                                              "con_S1",
-                                              "con_post"), 2)) %>%
-  cbind(data.frame(rbind(h.s6$hypothesis[,-1], 
-                   h.ubf$hypothesis[,-1]) )) %>%
-  print()
-
-
-#### Fold change plot rpS6 and UBF 
-
-
-tx_results_western %>%
[…]

### 2020-09-14 Figures/Figure2.R
misc updates

@@ -272,8 +272,15 @@ rrna_data <- readRDS("./data/derivedData/qpcr-analysis-bayes2/time_course_qpcr.R
 
 
 rrna_timecourse <- rrna_data$estimated_means %>%
-  filter(model == "tissue") %>%
-
+  filter(model == "tissue", 
+         target %in% c("rRNA18S F2R2",
+                       "rRNA5.8S F2R2",
+                       "rRNA28S F2R2",
+                       "rRNA5S F3R3", 
+                       "rRNA45SITS F12R12",
+                       "rRNA45S F5R5",     
+                       "rRNA47S F1R1")) %>%
+  
   mutate(target = factor(target, levels = c("rRNA18S F2R2",
                                             "rRNA5.8S F2R2",
                                             "rRNA28S F2R2",
@@ -340,10 +347,11 @@ for(i in 1:length(targets)) {
     
     scale_color_manual(values = c(color.scale[1], color.scale[2])) +
     
-    scale_y_continuous(limits = c(floor(min(dat$emmean)), ceiling(max(dat$emmean))),
-                       breaks = c(floor(min(dat$emmean)),
-                                  ceiling(max(dat$emmean)) + ((floor(min(dat$emmean)) - ceiling(max(dat$emmean)))/2), 
-                       ceiling(max(dat$emmean)))) +
+    scale_y_continuous(limits = c(floor(min(dat$emmean, na.rm = TRUE)), 
+                                  ceiling(max(dat$emmean, na.rm = TRUE))),
+                       breaks = c(floor(min(dat$emmean, na.rm = TRUE)),
+                                  ceiling(max(dat$emmean, na.rm = TRUE)) + ((floor(min(dat$emmean, na.rm = TRUE)) - ceiling(max(dat$emmean, na.rm = TRUE)))/2), 
+                       ceiling(max(dat$emmean, na.rm = TRUE)))) +
     
     labs(x = "Session", 
          y = "Estimated log-abundance per tissue weight (AU)")  +

### 2020-09-14 Figures/Figure3.R

@@ -155,14 +155,112 @@ prot_interaction <- tx_results_western %>%
 
 
 
-
-############## Combine panels  ###################
-
+### mRNA rpS6 and UBF ###########################
 
 
 
-figure3 <- plot_grid(prot_fold_change, prot_interaction, 
-                     ncol = 2, rel_widths = c(0.5, 0.5)) +
+ctrl_vs_int <- readRDS("./data/derivedData/qpcr-analysis-bayes2/qpcr_res_int_con.RDS")
+
+
+mrna_vs_ctrl <- ctrl_vs_int %>%
+  filter(target %in% c("UBTF F4R4", "UBTF F6R6", "rpS6 F2R2")) %>%
+  
+  filter(contrast %in% c("inter:post", "inter:post1w", "inter:S1")) %>%
+  mutate(contrast = gsub("inter:", "", contrast), 
+         contrast = factor(contrast, levels = c("S1", 
+                                                "post", 
+                                                "post1w"), 
+                           labels = c("Session 1", 
+                                      "Post-training", 
+                                      "De-training")), 
+         robust = if_else(lower.CL > 0, "robust", "notrobust"), 
+         target = if_else(target == "rpS6 F2R2", "rpS6 mRNA", 
+                          if_else(target == "UBTF F4R4", "UBF mRNA[1,4]", "UBF mRNA[2,3]")), 
+         estimate = exp(estimate), 
+         lower.CL = exp(lower.CL), 
+         upper.CL = exp(upper.CL)) %>%
+  
+  ggplot(aes(contrast, estimate, shape = model, color = robust)) + 
+  
+  geom_hline(yintercept = 1, lty = 2, color = "gray50") +
+  
+  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
+                position = position_dodge(width = 0.2), 
+                width = 0) +
+  geom_point(position = position_dodge(width = 0.2))  +
+  
+  scale_color_manual(values = c("gray50", "gray10")) +
+  scale_y_continuous(limits = c(0.5, 3.1), 
+                     breaks = c(1, 2, 3)) +
+  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
+  labs(y = "Fold change compared to Control") +
+  
+  plot_theme() +
+  theme(strip.background = element_rect(color = "white", fill = "white"), 
[…]

### 2020-09-14 R/qpcr-analysis-bayes2.R

@@ -39,10 +39,14 @@ library(coda)
 # qdat is the compiled data frame containing sample information and 
 # qpcr estimates. Created in qpcr-compile.R. Use this for downstream analyses
 
-qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")
+qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS") %>%
+  # problems with S1/R in P3, remove these for now until diagnosed
+  filter(!(participant == "P3" & leg == "R" &  time == "S1"))
 
 
 
+  
+  
 # Remove bad reactions prior to modeling. 
 
 qdat %>%
@@ -52,9 +56,10 @@ qdat %>%
 qdat %>%
   group_by(target) %>%
   
-  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 2 * sd(cq, na.rm = TRUE)|
-                             cq < mean(cq, na.rm = TRUE) - 2 * sd(cq, na.rm = TRUE), 
+  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 2.5 * sd(cq, na.rm = TRUE)|
+                             cq < mean(cq, na.rm = TRUE) - 2.5 * sd(cq, na.rm = TRUE), 
                            "outlier", "in")) %>%
+
   
   ggplot(aes(paste(participant, leg, time), cq, 
              color = target, 
@@ -69,8 +74,8 @@ qdat %>%
 
 
 nf <- qdat %>% group_by(target) %>%
-  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 2 * sd(cq, na.rm = TRUE)|
-                             cq < mean(cq, na.rm = TRUE) - 2 * sd(cq, na.rm = TRUE), 
+  mutate(outlier = if_else(cq > mean(cq, na.rm = TRUE) + 2.5 * sd(cq, na.rm = TRUE)|
+                             cq < mean(cq, na.rm = TRUE) - 2.5 * sd(cq, na.rm = TRUE), 
                            "outlier", "in")) %>%
   filter(target == "Lambda KIT") %>%
   filter(outlier == "in") %>% # Removes bad reactions
@@ -99,8 +104,8 @@ nf <- qdat %>% group_by(target) %>%
 
 qdat.rrna  <- qdat %>%
   group_by(target) %>%
[…]

### 2020-09-14 manuscript.Rmd
misc updates
figure 3 in draft shape

@@
 
+UBF levels determine rRNA gene activity throgh regulation of chromatin[@RN2556] <!--  Sanij 2008 -->
+
 # References
@@
 Asterisk in C indicates robust differences between volume conditions (a 95% CI of pairwise differences not containing 0).
+
+**Figure 3.** mRNA (A and B) and protein abundances (C and D) of ribosomal protein S6 and Upstream bindning factor (UBF).
+mRNA abundaces are expressed per total RNA (round points) or per muscle weight (triangles).
+Black points and error bars represents statistically robust results (a 95% CI not containing 0).
+E shows western-blots and total protein stains from a representative participant.

### 2020-09-15 manuscript.Rmd
update method

--- 2020-09-13 manuscript.Rmd
@@
 title: "Ribosome accumulation during early phase resistance training"
-author: Daniel Hammarström^1,2,£^&ast;,Sjur Øfsteng^2,£^, Bent R. Rønnestad^2^, Stian Ellefsen^1,3^
+author: Daniel Hammarström^1,2,£^&ast;,Sjur J. Øfsteng^2,£^, Bent R. Rønnestad^2^, Stian Ellefsen^1,3^
 date: ""
@@
 
-Muscle thickness was assessed using (...) ultra sound. DXA.
+For determination of body composition participants were scanned using DXA Lunar Prodigy densitometer (GE Healthcare, Madison, WI, USA) with the standard scanning mode (13-25 cm).
+Participants were lying supine within the scanning bed reference lines, with a strap secured around the ankles to ensure standardised body position in each of the scans.
+The scans were conducted with the participants in a fasted state, with empty bladder, wearing under-wear only, between 07.00-10.00 AM.
+Prior to each scanning a phantom scanning was run to prevent baseline drifting from affecting analyses.
+The same technician was used at each time point.
+Analyses was performed using GE enCORE version 17.0 software (GE Healthcare).
+Region of interest was customized for covering upper thigh, marked with a sqaure from pubic symphysis to lateral part of tuberculum major, and distal to art.
+genu.
+
+Muscle thickness (MT) was measured using a B-mode ultra sound unit (SmartUS EXT-1M, Telemed, Vilnius, Lithuania).
+The participant lay supine in a relaxed position for 20 min before scan, with their feet strapped to a standardized box (15x15cm).
+A mark was set on the line 60% of the distance between Spinia Iliac Anterior Superior and the lateral femur condyle.
+MT of *m.vastus lateralis* was measured applying a water-soluble transmission gel (Aquasonic 100 Ultrasound Transmission Gel; Parker Laboratories Inc., Fairfield, NJ, USA), and a 39 mm 12 MHz ultrasound probe was placed perpendicular to the site of interest without depressing the skin.
+When the quality of the image was satisfactory, evident as distinct upper and lower muscle fascia, three images were captured, where the probe was relocated to the same position between each image.
+Position of the probe was marked on the skin and subsequently marked on a transparent paper to ensure similar probe placement for both the right and left *m.vastus lateralis* at the subsequent assessments.
+Analyses were done in Jamovi (REF).
+Pictures were cropped and coded for blinding of the assessor.
+The average MT of the three images at each assesment was used for further analyses.
+
+<!--Schindelin J, Arganda-Carreras I, Frise E, et al. Fiji: an open-source platform for biological-image analysis. Nat
+1222 Methods. 2012;28(9(7)):676-682-->
 
@@
 
+UBF levels determine rRNA gene activity throgh regulation of chromatin[@RN2556] <!--  Sanij 2008 -->
+
+Per total RNA 
+Per muscle tissue
+
 # References

### 2020-09-15 R/wester_analysis.R
misc updates western models

@@ -61,6 +61,7 @@ ubf.m1 <- brm(bf(log.expr ~ tx + (1|participant)),
 
 
 
+
 # Save model 
 saveRDS(s6.m1, "./data/derivedData/western-analysis/s6_tx_m1.RDS")
 saveRDS(ubf.m1, "./data/derivedData/western-analysis/ubf_tx_m1.RDS")
@@ -222,5 +223,70 @@ segmented_protein_results <- cbind(data.frame(target = rep(c("rpS6", "UBF"), eac
 
 ### Save compiled results 
 saveRDS(segmented_protein_results, "./data/derivedData/western-analysis/segmented_results.RDS")
+
+################ De-training model comparing cond and var ################
+
+
+# A de-training model is used to estimate differences between 
+# volume conditions 
+
+west_detrain <- western_data %>%
+  filter(cond != "ctrl_leg", 
+         time %in% c("S12", "post1w")) %>%
+  mutate(participant = factor(participant), 
+         cond = factor(cond), 
+         target = factor(target),
+         time = factor(time, levels = c("S12", "post1w"))) %>%
+  
+  group_by(target, participant, cond, time) %>%
+  summarise(expression = mean(expression)) %>%
+  mutate(log.expr = log(expression)) %>%
+  filter(!is.na(expression)) %>%
+  
+  print()
+
+
+s6.m3 <- brm(bf(log.expr ~ time * cond + (1|participant)),
+             
+             warmup = 2000, # number of samples before sampling
+             iter = 10000,  # number of mcmc iterations 
+             cores = 4, # number of cores used in sampling
+             chains = 4, # number of chains
+             seed = 5, # seed for reproducible results
+             thin = 10,
+             control = list(adapt_delta = 0.95), 
+             data = west_detrain[west_detrain$target == "t-s6",])
+
+
+ubf.m3 <- brm(bf(log.expr ~ time * cond + (1|participant)),
+              
+              warmup = 2000, # number of samples before sampling
+              iter = 10000,  # number of mcmc iterations 
+              cores = 4, # number of cores used in sampling
+              chains = 4, # number of chains
+              seed = 5, # seed for reproducible results
+              thin = 10,
+              control = list(adapt_delta = 0.95), 
+              data = west_detrain[west_detrain$target == "t-UBF",])
+
+
+
[…]

### 2020-09-15 R/wester_compile.R

@@ -82,10 +82,11 @@ western_data <- rbind(western_data_round1, western_data_round2) %>%
   mutate(tp.factor = tp.factor / max(tp.factor, na.rm = TRUE), 
          signal = signal / max(signal, na.rm = TRUE)) %>%
   ungroup() %>%
-  mutate(expression = signal / tp.factor) %>%
-  dplyr::select(participant,sample, leg, cond, time, gel, target, expression) %>%
+  mutate(expression = signal / tp.factor ) %>%
+  group_by(gel, target) %>%
+  mutate(expression.sd = (expression - mean(expression))/sd(expression)) %>%
+  dplyr::select(participant,sample, leg, cond, time, gel, target, expression, expression.scaled = expression.sd) %>%
   print()
-
 
 
 

### 2020-09-15 manuscript.Rmd
misc updates

@@
 
+summary(tc.m2)
+
 time1.hyp <- hypothesis(tc.m2, "time1 = 0")
@@
 
+## Effect of detraining
+
+detrain.hyp <- hypothesis(tc.m2, "detraindetrain = 0")
+
+detrain_rna <- paste0(sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,2]) - 1)),
+                           "%, [",
+                      sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
 ```
@@
 Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not robustly different from zero (`r slope4_8` and `r slope8_12`).
+After the de-training period Total-RNA decreased by `r detrain_rna` from levels measured after Session 12.
 
@@
 
+s6.m3 <- readRDS("./data/derivedData/western-analysis/s6_detrain.RDS")
+ubf.m3 <- readRDS("./data/derivedData/western-analysis/ubf_detrain.RDS")
+
+ubf.sum <- summary(ubf.m3)$fixed
+
+var_s12 <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[3,1]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[3,3]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[3,4]) - 1)),
+                   "]")
+
+const_detrain <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[2,1]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[2,3]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[2,4]) - 1)),
+                   "]")
+
+var_const_interaction <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[4,1]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[4,3]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[4,4]) - 1)),
+                   "]")
+
+h <- hypothesis(ubf.m3, "condvar + timepost1w:condvar = 0")$hypothesis %>%
+  data.frame()
+
+var_change <- paste0(sprintf("%.1f", 100 * (exp(h[1,2]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(h[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(h[1,5]) - 1)),
+                   "]")
+
 ```
@@
 Protein levels of RPS6 and UBF increased only after twelve sessions in the training group compared to control and stayed elevated after eight days of rest (Figure 3A and B).
-The average increase in expression levels over twelve sessions near to linear in both and estimated to `r s6_slope0_4`, `r s6_slope4_8` and `r s6_slope8_12` in RPS6, and `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12` in UBF per session between session 1-4, 4-8 and 8-12, respectively.
+The average increase in protein levels over twelve sessions was near linear in both targets and estimated to `r s6_slope0_4`, `r s6_slope4_8` and `r s6_slope8_12` in RPS6, and `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12` in UBF per session between session 1-4, 4-8 and 8-12, respectively.
+Slopes did not differ between volume-conditions but there was a tendency towards lower levels of UBF in the variable condition at session 12 (`r var_s12`).
+After the de-training period the constant volume condition tended to show decreased levels of UBF (`r const_detrain`) while the variable volume condition remaind at a similar level compare to session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
+Levels of rpS6 did not change in the de-training period and were similar between volume conditions.
+
+```{r combined-analysis}
+
+# Save model 
+comb.m1 <- summary(readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS"))$fixed
+
+ubf_cor_tot_rna <- paste0(sprintf("%.1f", 100 * (exp(comb.m1[2,1]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(comb.m1[2,3]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(comb.m1[2,4]) - 1)),
+                   "]")
+
+```
+
+UBF levels robustly predicted total RNA values with a 1 standard deviation change in UBF levels increasing Total RNA by `r ubf_cor_tot_rna` (Table X).
+
+```{r predict-muscle-growth}
+
+```
 

### 2020-09-15 R/total-rna-analysis.R

@@ -170,7 +170,8 @@ saveRDS(comp_rna, "./data/derivedData/total-rna-analysis/comp_rna_freq.RDS")
   
 time_course <- rna_complete %>%
   filter(cond != "ctrl_leg") %>%
-  filter(time != "post1w") %>%
+  
+  
   
   group_by(participant,leg, time,time.c, cond, ) %>%
   summarise(rna = mean(rna, na.rm = TRUE), 
@@ -180,12 +181,15 @@ time_course <- rna_complete %>%
          cond = factor(cond), 
          rna.tissue = rna/tissue_weight, 
          tw = tissue_weight - mean(tissue_weight), 
-         time.c.cent = time.c - 4) %>%
+         time.c.cent = time.c - 4, 
+         detrain = factor(if_else(time == "post1w", "detrain", "train"), 
+                          levels = c("train", "detrain"))) %>%
   # Create dummy variables for segmented model 
   
   mutate(time1 = time.c,
          time2 = if_else(time %in% c("S0", "S1", "S4"), 0, time.c - 4),
-         time3 = if_else(time %in% c("S0", "S1", "S4", "S5", "S8"), 0, time.c - 8)) 
+         time3 = if_else(time %in% c("S0", "S1", "S4", "S5", "S8"), 0, time.c - 8))  %>%
+  print()
   
   
   
@@ -212,7 +216,7 @@ tc.m1 <- brm(bf(log(rna.tissue) ~ cond  + s(time.c, by = cond, k = 7) + (1|parti
 pp_check(tc.m1, type = "ecdf_overlay")
 summary(tc.m1)
 
-tc.m2 <- brm(bf(log(rna.tissue) ~ (time1 + time2 + time3) + (1|participant)), 
+tc.m2 <- brm(bf(log(rna.tissue) ~ (time1 + time2 + time3) + detrain + (1|participant)), 
              data = time_course, 
              warmup = 1000, # number of samples before sampling
              iter = 4000,  # number of mcmc iterations 

### 2020-09-15 R/ubf-tot-rna-model.R

@@ -1,0 +1,331 @@
+##-------------------------------------
+## ubf_tot_rna_model.R
+##
+## Title: 
+## Purpose: Model total RNA over time and influence of UBF levels
+## Author:
+##
+##
+##
+##-------------------------------------
+## Notes:
+#
+#
+#
+#
+#
+#
+#
+#
+## ------------------------------------
+
+# Load packages
+library(tidyverse)
+library(brms)
+library(bayesplot)
+library(lme4)
+
+
+
+
+
+# Load RNA data 
+
+rna <- readRDS("./data/derivedData/tot-rna/tot-rna.RDS")
+
+
+
+rna_complete <- rna %>%
+  filter(outlier == "in") %>% # removes outlier based on RNA to tissue weight relationship
+  dplyr::select(participant, series, sample, leg, time, cond, tissue_weight, rna) %>%
+  mutate(time.c = gsub("S", "", time), 
+         detrain = if_else(time.c == "post1w", "detrain", "train"),
+         time.c = if_else(time.c == "post1w", "12", time.c),
+         time.c = if_else(time.c == "postctrl", "12", time.c),
+         time.c = gsub("c", "", time.c), 
+         time.c = as.numeric(time.c), 
+         time = factor(time, levels = c("S0", "S1","S1c", 
+                                        "S4", "S5", "S8", 
+                                        "S9", "S12", "post1w", "postctrl")), 
+         
+         # Create segments for segmented model
+         
+         time1 = time.c,
+         time2 = if_else(time %in% c("S0", "S1", "S4"), 0, time.c - 4),
+         time3 = if_else(time %in% c("S0", "S1", "S4", "S5", "S8"), 0, time.c - 8) ) %>%
+  print()
+
+
+# Load western blod data 
+
+western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
+
+
+west_cont <- western_data %>%
+  
+  mutate(time.c = gsub("S", "", time), 
+         detrain = if_else(time.c == "post1w", "detrain", "train"),
+         time.c = if_else(time.c == "post1w", "12", time.c),
+         time.c = if_else(time.c == "postctrl", "12", time.c),
+         time.c = gsub("c", "", time.c), 
+         time.c = as.numeric(time.c), 
+         time = factor(time, levels = c("S0", "S1","S1c", 
+                                        "S4", "S5", "S8", 
[…]

### 2020-09-15 R/wester_compile.R

@@ -83,10 +83,11 @@ western_data <- rbind(western_data_round1, western_data_round2) %>%
          signal = signal / max(signal, na.rm = TRUE)) %>%
   ungroup() %>%
   mutate(expression = signal / tp.factor ) %>%
-  group_by(gel, target) %>%
-  mutate(expression.sd = (expression - mean(expression))/sd(expression)) %>%
+  group_by(target) %>%
+  mutate(expression.sd = (expression - mean(expression, na.rm = TRUE))/sd(expression, na.rm = TRUE)) %>%
   dplyr::select(participant,sample, leg, cond, time, gel, target, expression, expression.scaled = expression.sd) %>%
   print()
+
 
 
 

### 2020-09-15 manuscript.Rmd
notes in manuscripts

@@
 
+Time course data 
+UBF and S6 protein levels may be differently regulated in men and rats as more consecutive bouts of electrical stimulation in rats does not prtoduce the same pattern seen in our study 
+<!-- Kotani 2019 10.1152/japplphysiol.00665.2018 increase in t UBF at 6 h post ex but not 48 despite 2 weeks training (3 x week).
+S6 increase at 48 h and 6 bouts but return to sed values at 18 bouts.
+18 and 28 S increased over sed at 48 h post ex with no further increase beyond 3 bouts 45S maximal at three bouts -->
+
+<!-- Goodman 2011 UBF increase due to synergist ablation is rapamycin insensitive https://doi.org/10.1113/jphysiol.2011.218255 -->
+
+<!-- Contrfary to goodman Nader found rapamycin sensitivity in UBF increase 10.1152/ajpcell.00165.2005-->
+
+<!-- Hannan 1996 10.1074/jbc.271.6.3213 increased rRNA transcription is associated with increased levels of UBF mRNA and protein in cardiomyocutes -->
+
+<!-- In cell lines UBF is proportional to the rat e of rDNA transcription ref 25/26 in Hannan-->
+
+<!-- phospho of UBF shows it has a role in recruioting SL1 to the promotot  10.1152/ajpcell.00176.2006 -->
+
 Per total RNA 

### 2020-09-16 manuscript.Rmd
updates results and figures

@@
 All participants successfully completed their prescribed sessions.
-The different training protocols resulted in diverging loading profiles (load &#215; repetitions) over the course of the study (Figure 1B).
+Different volume conditions resulted in diverging loading profiles (load &#215; repetitions) over the course of the study (Figure 1B).
 The intensity of exercise (load at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) four-session block, respectively.
@@
 There were no indications of differences between training conditions in strength or muscle thickness changes.
-
-```{r total-rna}
-
-# Segmented model analysis in text 
-# Read the segmented model 
-tc.m2 <- readRDS("./data/derivedData/total-rna-analysis/tc.m2.RDS")
-
-summary(tc.m2)
-
-time1.hyp <- hypothesis(tc.m2, "time1 = 0")
-time2.hyp <- hypothesis(tc.m2, c("time2 + time1 = 0")) # The slope (on log scale) between session 4 and 8
-time3.hyp <- hypothesis(tc.m2, c("time2 + time1 +time3 = 0")) # The slope on log scale between session 8 and 12
-
-slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,2]) - 1)),
-                           "%, 95% CI: [",
-                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,4]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,5]) - 1)),
-                   "]")
-
-slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,2]) - 1)),
-                           "%, [",
-                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,4]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,5]) - 1)),
-                   "]")
-
-slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,2]) - 1)),
-                           "%, [",
-                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,4]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,5]) - 1)),
-                   "]")
-
-## Effect of detraining
-
-detrain.hyp <- hypothesis(tc.m2, "detraindetrain = 0")
-
-detrain_rna <- paste0(sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,2]) - 1)),
-                           "%, [",
-                      sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,4]) - 1)), 
-                   ", ",
-                       sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,5]) - 1)),
-                   "]")
-
-```
-
-A single session of RT led to robust increase in 47S and 45S pre-rRNA expressed per unit tissue weight within the training group (Figure 2B) and compared to control (Figure 2C).
-These were also increased compared to control after twelve sessions together with the 45S ITS and mature rRNA species 28S and 18S (Figure 2C).
-After eight days of rest, only 18S and 28S remained elevated above levels measured in the control group (Figure 2B).
-This general pattern of rRNA expression was reflected also in total RNA expressed per unit tissue weight were a robust increase compared to control was seen after session twelve (Figure 2E and F).
-Within the experimental group, only small differences were seen in rRNA subspecies as 45S ETS was robustly higher after 12 sessions in the variable volume condition after an imbalance between conditions prior to training (Figure 2D).
-Total RNA increased similarly in both volume conditions throughout the training period (Figure 2G).
-Most of the increase was seen in after the first four sessions where the average increase was `r slope0_4` per session in the intervention group as a whole.
-Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not robustly different from zero (`r slope4_8` and `r slope8_12`).
-After the de-training period Total-RNA decreased by `r detrain_rna` from levels measured after Session 12.
 
@@
 
-Protein levels of RPS6 and UBF increased only after twelve sessions in the training group compared to control and stayed elevated after eight days of rest (Figure 3A and B).
+Protein levels of RPS6 and UBF increased only after twelve sessions in the training group compared to control and stayed elevated after eight days of rest (Figure 2A and B).
 The average increase in protein levels over twelve sessions was near linear in both targets and estimated to `r s6_slope0_4`, `r s6_slope4_8` and `r s6_slope8_12` in RPS6, and `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12` in UBF per session between session 1-4, 4-8 and 8-12, respectively.
 Slopes did not differ between volume-conditions but there was a tendency towards lower levels of UBF in the variable condition at session 12 (`r var_s12`).
-After the de-training period the constant volume condition tended to show decreased levels of UBF (`r const_detrain`) while the variable volume condition remaind at a similar level compare to session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
+After the de-training period the constant volume condition tended to show decreased levels of UBF (`r const_detrain`) while the variable volume condition remained at a similar level compare to session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
 Levels of rpS6 did not change in the de-training period and were similar between volume conditions.
+
+```{r total-rna}
+
+# Segmented model analysis in text 
+# Read the segmented model 
+tc.m2 <- readRDS("./data/derivedData/total-rna-analysis/tc.m2.RDS")
+
+summary(tc.m2)
+
+time1.hyp <- hypothesis(tc.m2, "time1 = 0")
+time2.hyp <- hypothesis(tc.m2, c("time2 + time1 = 0")) # The slope (on log scale) between session 4 and 8
+time3.hyp <- hypothesis(tc.m2, c("time2 + time1 +time3 = 0")) # The slope on log scale between session 8 and 12
+
+slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,2]) - 1)),
+                           "%, 95% CI: [",
+                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,2]) - 1)),
+                           "%, [",
+                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,2]) - 1)),
+                           "%, [",
+                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+## Effect of detraining
+
+detrain.hyp <- hypothesis(tc.m2, "detraindetrain = 0")
+
+detrain_rna <- paste0(sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,2]) - 1)),
+                           "%, [",
+                      sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+```
+
+A single session of RT led to robust increase in 47S and 45S pre-rRNA expressed per unit tissue weight within the training group (Figure 3B) and compared to control (Figure 3C).
+These transcripts were also increased compared to control after twelve sessions together with the 45S ITS and mature rRNA species 28S and 18S (Figure 3C).
+After eight days of rest, only 18S and 28S remained elevated above levels measured in the control group (Figure 3B).
+This general pattern of rRNA expression was reflected also in total RNA expressed per unit tissue weight were a robust increase compared to control was seen after session twelve (Figure 3E and F).
+Within the experimental group, only small differences were seen in rRNA subspecies as 45S ETS was robustly higher after 12 sessions in the variable volume condition after an imbalance between conditions prior to training (Figure 3D).
+Total RNA increased similarly in both volume conditions throughout the training period (Figure 3G).
+Most of the increase was seen in the first four sessions where the average increase was `r slope0_4` per session in the intervention group as a whole.
+Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not robustly different from zero (`r slope4_8` and `r slope8_12`).
+After the de-training period Total-RNA decreased by `r detrain_rna` from levels measured after Session 12.
 
@@
 
-UBF levels robustly predicted total RNA values with a 1 standard deviation change in UBF levels increasing Total RNA by `r ubf_cor_tot_rna` (Table X).
+UBF levels robustly predicted total RNA values as a standard deviation change in UBF levels led to a `r ubf_cor_tot_rna` increase in total RNA per unit tissue weight (Table X).
+To estimate the influence of total RNA on muscle growth the average linear increase in total RNA was calculated for each leg together with the average total RNA abundance over the course of the training intervention in the training group.<!-- (estimated as the intercept in a regression model with number of sessions centered at session 6).
+move to methods -->
+The relationship between muscle growth and characteristics of individual total RNA profiles were estimated in a regression model containing the increase in total RNA concentrations, expressed as percentage increase per session, and average total RNA abundance at Session 6, expressed as standard deviations from the mean.
+The model also contained sex as predictor to control for sex differences in muscle growth.
+Total RNA increase in response to training together with the average level of total RNA robustly predicted muscle growth (Table 4, Figure 4A and B).
+To assess the robustness of the model, each data point was iteratively removed from the data set and the model was refitted using new estimates of the total RNA abundance and increase.
+This analysis indicated that the relationship between Total RNA increase and muscle growth was preserved in all iterations but the upper limit of the 95% CI spanned 0 in some instances regarding the relationship between average RNA and muscle growth (Figure 4D).
+We further assessed the model by iteratively removing one participant from the data set, similarly to removing a single data point this showed that estimates of the total RNA increase effect on muscle growth was robust but average total RNA estimates spanned 0 in some cases (Participant 8 and 3 in Figure 4D).
 
@@
 
+<!-- Hannan 1996 10.1073/pnas.93.16.8750 over expression of UBF leads to rRNA transcription in cardiomyocytes -->
+
+<!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1.
+The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF  -->
+
 <!-- In cell lines UBF is proportional to the rat e of rDNA transcription ref 25/26 in Hannan-->
@@
 <!-- phospho of UBF shows it has a role in recruioting SL1 to the promotot  10.1152/ajpcell.00176.2006 -->
+
+<!-- Pootinga 2004 10.1038/sj.emboj.7600335 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
 

### 2020-09-16 tables.Rmd
updates results and figures

@@
 title: "Tables"
-output: rmarkdown::word_document
+output:
+  word_document:
+    reference_docx: resources/tables_manuscript_word_template.docx
 editor_options: 
@@
 source("./R/libs.R")
-library(flextable)
+library(flextable); library(officer)
 
@@
 
-<br>
-<br>
+\pagebreak
 
@@
 ```
+
+\pagebreak
+
+```{r table4-regression-models-muscle-thickness}
+# Save model 
+hyp.pre.m1 <- readRDS( "./data/derivedData/ubf-tot-rna-model/hyp_pre_m1.RDS")
+hyp.pre.m2 <- readRDS( "./data/derivedData/ubf-tot-rna-model/hyp_pre_m2.RDS")
+hyp.pre.m3 <- readRDS( "./data/derivedData/ubf-tot-rna-model/hyp_pre_m3.RDS")
+
+m1.sum <- summary(hyp.pre.m1)
+m2.sum <- summary(hyp.pre.m2)
+m3.sum <- summary(hyp.pre.m3)
+
+# combine output from all models
+
+reg_results <- 
+  cbind(rbind(m3.sum$fixed,
+      m3.sum$random$participant, 
+      m3.sum$random$`participant:leg`, 
+      m3.sum$spec_pars), 
+      data.frame(coef = c("Intercept", 
+                          "Sex (Male)", 
+                          "Mean Total RNA",
+                          "Mean Total RNA increase per session", 
+                          "Between Participant variation", 
+                          "Between Participant:Leg variation", 
+                          "Residual SD"))) %>%
+  remove_rownames() %>%
+  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)
+
+std_border = fp_border(color="gray")
+
+reg_results %>%
+
+  mutate(Estimate = round(Estimate, 2), 
+         Est.Error = round(Est.Error, 2), 
+         lwr = round(lwr, 2), 
+         upr = round(upr, 2)) %>%
+
+  flextable() %>%
+  set_header_labels(coef = "Coefficient", 
+                    Est.Error = "SD", 
+                    lwr = "Lower 95% CI", 
+                    upr = "Upper 95% CI") %>%
+    add_header_lines(values = "Table 3. Total RNA as a predictor of muscle growth.") %>%
+
+ # bold(i = c(5, 6, 13, 14, 20, 21), j = c(2:6), bold = TRUE) %>%
+
+ # hline(i = c(9, 17), part="body", border = std_border ) %>%
+      fontsize(size = 8, part = "body") %>%
+    fontsize(size = 10, part = "header") %>%
+
+  footnote( i = 2, j = 2,
+            value = as_paragraph(
+              c("The dependent variable is \U0394 Muscle thickness (mm)")
+            ),
+            ref_symbols = c("a"),
+            part = "header") %>%
+
+  autofit()
+
+```

### 2020-09-16 Figures/Figure4.R

@@ -1,0 +1,276 @@
+##-------------------------------------
+## figure4.R
+##
+## Title: Figure 4
+## Purpose: Show relationship between Total RNA increase and muscle hypertrophy
+## Author:
+##
+##
+##
+##-------------------------------------
+## Notes:
+
+# For analysis see ubf-tot-rna-model.R
+#
+#
+#
+#
+#
+#
+#
+## ------------------------------------
+
+
+source("./R/figure-source.R")
+library(brms)
+
+
+#### Figure 4B Leave one out analysis #########################
+
+
+
+loo.results <- readRDS("./data/derivedData/ubf-tot-rna-model/leave-one-out.RDS")
+
+res <- loo.results$loo.sample %>%
+  
+  mutate(Participant = factor(participant, 
+                              levels = paste0("P", c(seq(1:7), 19, 21, 22, 23)), 
+                              labels = paste0("P", seq(1:11))), 
+         Participant = fct_rev(Participant)) %>%
+
+  
+  pivot_longer(names_to = "stat", values_to = "estimates", estimate:upr) %>%
+  print()
+  
+res2 <- loo.results$loo.participant %>%
+  
+  mutate(Participant = factor(participant, 
+                              levels = paste0("P", c(seq(1:7), 19, 21, 22, 23)), 
+                              labels = paste0("P", seq(1:11))), 
+         Participant = fct_rev(Participant)) %>%
+  
+  
+  filter(model == "m3", coef != "sex") %>%
+  mutate(coef = factor(coef, levels = c("slope", "intercept"), 
+                       labels = c("Total RNA increase\nper session", 
+                                  "Average total RNA\n(Session 6)"))) %>%
+    print()
+  
+
+
+loo_panel <- res %>%
+  filter(stat != "t_val", model == "m3", coef != "sex") %>%
+  
+  mutate(coef = factor(coef, levels = c("slope", "intercept"), 
+                       labels = c("Total RNA increase\nper session", 
+                                  "Average total RNA\n(Session 6)"))) %>%
+  
+  ggplot(aes(estimates, Participant, fill = stat)) +
+
+  
+  geom_vline(xintercept = 0, lty = 2, color = "gray50") +
+    
+  geom_density_ridges(color = "white") +
+  facet_wrap( ~ coef, scales = "free") +
+  
+
+  
+  
+  
+
[…]

### 2020-09-16 R/figure-source.R

@@ -7,6 +7,8 @@
 library(tidyverse)
 library(cowplot)
 library(ggpattern)
+library(ggridges)
+library(ggrepel)
 library(ggtext)
 # Color scale 
 color.scale <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 

### 2020-09-16 R/ubf-tot-rna-model.R

@@ -2,19 +2,25 @@
 ## ubf_tot_rna_model.R
 ##
 ## Title: 
-## Purpose: Model total RNA over time and influence of UBF levels
+## Purpose: Model total RNA over time and influence of UBF levels and effect of rRNA on muscle hypertrophy
 ## Author:
 ##
 ##
 ##
 ##-------------------------------------
 ## Notes:
+# UBF is believed to be a contributor to rDNA transcription by recruiting 
+# the SL1 transcription factor to the rDNA promoter. mTOR increases UBF 
+# phosphorylation (rapamycin sensitive) increasing its activity and c-Myc induces
+# UBTF transcription increasing protein levels (rapamycin insensitive). 
+# 
+# Below we want to test if rRNA levels are assiciated with UBF levels. Both 
+# are increasing with time and therefore, the model controls for the effect of training (time).
 #
-#
-#
-#
-#
-#
+# This script also contain a model of muscle growth were we estimate the effect of increase in rRNA
+# on muscle growth (muscle thicknes change pre- to post training). Post training values are 
+# from both S12 and post 1w (de-training). rRNA transcription is estimated as the average linear increase over 
+# all sessions for each individual and leg. 
 #
 #
 ## ------------------------------------
@@ -24,7 +30,7 @@ library(tidyverse)
 library(brms)
 library(bayesplot)
 library(lme4)
-
+library(readxl)
 
 
 
@@ -56,7 +62,7 @@ rna_complete <- rna %>%
   print()
 
 
-# Load western blod data 
+# Load western blot data 
 
 western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
 
@@ -88,12 +94,8 @@ west_cont <- western_data %>%
   
   print()
 
-west_cont %>%
-  filter(participant == "P1", cond == "const") %>%
-  dplyr::select(time, time1, time2, time3)
 
 # Combined data from both Total-RNA and western blot analysis
-
 
 combined_df  <- west_cont %>%
   group_by(target, participant, cond, time) %>%
@@ -112,10 +114,19 @@ combined_df  <- west_cont %>%
   print()
 
 
+saveRDS(combined_df, "./data/derivedData/ubf-tot-rna-model/combined_df.RDS")
 
 
 
 ### Combined df in brms models ##############################
+
[…]

### 2020-09-17 manuscript.Rmd
manuscript updates

@@
 title: "Ribosome accumulation during early phase resistance training"
-author: Daniel Hammarström^1,2,£^&ast;,Sjur J. Øfsteng^2,£^, Bent R. Rønnestad^2^, Stian Ellefsen^1,3^
+author: Daniel Hammarström^1,2,£^&ast;,Sjur J. Øfsteng^2,£^, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^,  Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
 date: ""
@@
 
-Our skeletal muscles not only constitutes our locomotive apparatus but also serve as our primary storage site of amino-acids.
+Skeletal muscles not only constitutes our locomotive apparatus but also serve as our primary storage site of amino-acids.
 This makes the skeletal muscle system a critical target for interventions that promotes health across the lifespan [@RN763]<!-- Wolfe 2006 -->.
-The beneficial effects of resistance-training (RT) are in large part mediated through exercise-induced changes in the balance between muscular protein breakdown and synthesis, with successful RT leading to net protein accretion and muscle hypertrophy.
+Resistance training can promote beneficial effects on skeletal muscles and does so through exercise-induced changes in the balance between muscular protein breakdown and synthesis, with successful RT leading to net protein accretion and muscle hypertrophy.
 A traditional view of exercise-induced muscle hypertrophy has been that muscle protein accumulates as a consequence of repeated episodes of elevated post-exercise protein synthesis in the acute phase, resulting in a net-positive protein balance over time.
-Indeed, one bout of resistance exercise acutely up-regulates skeletal muscle protein synthesis up to 48-hr after exercise [@RN786]<!-- Phillips 1997 -->
+Indeed, one bout of resistance exercise acutely up-regulates skeletal muscle protein synthesis up to 48 hr after exercise [@RN786]<!-- Phillips 1997 -->
 and subsequently, repeated bouts would lead to accumulation of muscle protein over time
@@
 [@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->@RN1755; <!-- Stec Ribosome bio higher in high responders -->
-@Hammarstrm2020]<!-- Volume dependent RNA accum relates to volume depend hyper -->.
-Furthermore, inhibition of ribosomal RNA (rRNA) transcription or inhibition of up-stream transcription factors leads to halted biosynthesis of ribosomes resulting in diminished cellular growth
+@RN2358]<!-- Volume dependent RNA accum relates to volume depend hyper -->.
+Furthermore, inhibition of ribosomal RNA (rRNA) transcription or inhibition of up-stream transcription factors leads to halted biosynthesis of ribosomes and diminished cellular growth
 [@RN1755<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->;
 @RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
-@vonWalden2016]<!-- Inhibition of Pol 1 rDNA tracsription and myotube growth -->.
+@RN1810]<!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->.
 
 Formation of new ribosomes is a complex and energy demanding process involving the whole transcriptional apparatus
-[@RN1920<!-- Warner review yeast -->,
+[@RN1920;<!-- Warner review yeast -->,
 @RN1820;<!-- Moss review -->
-@vonWalden2016]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
+@RN1810]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
 In skeletal muscle, a single bout of resistance exercise elevates markers of ribosome biogenesis in skeletal muscle evident from studies investigating immature forms of rRNA in response to exercise in the untrained state
-[@REFNEEDED<!-- References on acute 45S changes -->;
+[@RN1656<!-- A single session upregulates 45 S 24 h post ex -->;
 @RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->].
@@
 @RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
-@Hammarstrom2020; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
+@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
 @RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
@@
 We have previously reported peak RNA levels 48 h after the fourth training session with a tendency towards normalization after 31 sessions
-[@Hammarstrom2020].
-This is similar to what was reported by Brook *et al.* who reported peak values after nine sessions followed by a slight decrease to after 18 sessions
+[@RN2358].
+This is similar to what was reported by Brook *et al.* where peak values were seen after nine sessions followed by a slight decrease to after 18 sessions
 [@RN1809].
@@
 Eligible participants had to be non-smokers, between 18 and 35 years of age, report a training history of less than one RT session per week during the six months leading up to the study and not consume dietary supplements or medication with known effects on muscle metabolism.
-Potential participants were further screened for injuries causing impaired strength and/or affecting their ability to perform RT, symptoms or history of disease or known adverse reactions to local anaesthetics.
-Included participants were allocated to either an experimental group (EXP, *n* = 11) or a non-training control group (CTRL, *n* = 8; see Table 1 for participant characteristics).
-EXP performed a 12 session RT protocol consisting unilateral knee-extension to allow for within-participant comparison exercise volume.
-In EXP, one leg performed constant volume (CONST, 6 sets per session) and the other leg performed variable volume over the course of the study (VAR, four sessions each with 6, 3 and 9 sets per session; Figure 1A).
+Potential participants were further screened for injuries causing impaired strength and/or affecting their ability to perform RT, symptoms or history of disease or known adverse reactions to local anesthetics.
+Included participants were allocated to either an training group (TR, *n* = 11) or a non-training control group (CTRL, *n* = 8; see Table 1 for participant characteristics).
+TR performed a 12 session RT protocol consisting unilateral knee-extension to allow for within-participant comparison exercise volume.
+In TR, one leg performed constant volume (CONST, 6 sets per session) and the other leg performed variable volume over the course of the study (VAR, four sessions each with 6, 3 and 9 sets per session; Figure 1A).
 CTRL did not partake in RT and were instructed to continue their everyday activities.
-Muscle biopsies were sampled bilaterally in EXP before and 48 h after the first session and 48 h after the fourth, fifth, eight, ninth and twelfth session as wells as after a seven day de-training period.
+Muscle biopsies were sampled bilaterally in TR before and 48 h after the first session and 48 h after the fourth, fifth, eight, ninth and twelfth session as wells as after a seven day de-training period.
 Muscle biopsies were obtained from CTRL at three occasions with 48 h separating the first two and 2-4 weeks separating second and third sampling.
-EXP and CTRL performed strength assessments at least seven days prior to the first biopsy sampling and 72 h after the twelwth session (EXP) and 24 h after the last biopsy (EXP and CTRL).
+TR and CTRL performed strength assessments at least seven days prior to the first biopsy sampling and 72 h after the twelfth session (EXP) and 24 h after the last biopsy (TR and CTRL).
 Appendicular lean mass (Dual-energy X-ray absorptiometry, DXA) and muscle thickness of *m.
-vastus lateralis* were assessed prior to the first biopsy (EXP and CTRL) as wells as before the second to last (EXP) and last (EXP and CTRL) biopsy (Figure 1A).
+vastus lateralis* were assessed prior to the first biopsy (TR and CTRL) as wells as before the second to last (TR) and last (TR and CTRL) biopsy (Figure 1A).
 
 ## RT protocol
-Prior to all RT-sessions, participants performed a standardized warm-up consisting of 5 min ergometer cycling (RPE 12-14), followed by ten repetitions each of push-ups, sit-ups and back-extensions.
-After warm-up participants performed unilateral knee-extension with the prescribed number of sets.
+Prior to all RT-sessions, participants performed a standardized warm-up consisting of 5 min ergometer cycling (rating of perceived exertion (RPE): 12-14), followed by ten repetitions each of push-ups, sit-ups and back-extensions.
+After warm-up, participants performed unilateral knee-extension with the prescribed number of sets.
 Each set was prescribed with 10 repetitions maximum (RM).
@@
 The second session of each four-session block (session 2, 6 and 10) were performed at a sub-maximal resistance (~90% of the previous session) with the same number of repetitions.
-In each session, after the knee-extension exercise, participants completed two sets each of three upper-body exercises (bench press, lateral pull-down and shoulder press).
-Prior to each session, participants were asked to rate muscle soreness on a scale from 0 to 10 (0 = no soreness, 10 = hurting very bad, e.g.
+In each session, after the knee-extension exercise, participants completed two sets each of three upper-body exercises (bench press, lateral pull-down and shoulder press).<!--Prior to each session, participants were asked to rate muscle soreness on a scale from 0 to 10 (0 = no soreness, 10 = hurting very bad, e.g.
 not being able to walk properly) [@Vickers2010].
-After each session, participants were asked to rate their perceived exertion (Borg CR10).
-After the completion of each session participants were given a standardized drink to aid recovery (0.15 g × kg-1 protein, 11.2 g × kg-1 carbohydrates and 0.5 g × kg-1 fat).
+After each session, participants were asked to rate their perceived exertion (Borg CR10).-->
+After the completion of each session participants were given a standardized drink to aid recovery (0.15 g kg<sup>-1</sup> protein, 11.2 g kg<sup>-1</sup> carbohydrates and 0.5 g kg<sup>-1</sup> fat).
 
@@
 After an additional 15-sec restitution period, participants were instructed to push against the lever with maximal effort.
-During the same assessment session, participants remained seated in the dynamometer for meassurement from both legs.
-The first measurement was alternated between legs and sessions.
+Within the same assessment session, participants remained seated in the dynamometer for measurement perfomed on both legs.
+The first measurement was alternated between legs every other session.
 For statistical treatment of the data, all successful attempts were used.
-Successful attempts were decided based on participants (...).
-The last strength assessment at baseline was performed at least **seven** days prior to the first biopsy sampling.
-At least one strength was performed on separate days with two sessions allowed perform on the same day with a short rest between sessions.
-Post training assessments were performed approximately 48 hours after the last session and after seven days of no training.
+The last strength assessment at baseline was performed at least seven days prior to the first biopsy sampling.
+At least one of the baseline strength tests was performed on separate day with two sessions allowed to be perform on the same day with a short rest between assessments.
+Post training assessments were performed 48 hours and eight days after the last session.
 
-For determination of body composition participants were scanned using DXA Lunar Prodigy densitometer (GE Healthcare, Madison, WI, USA) with the standard scanning mode (13-25 cm).
-Participants were lying supine within the scanning bed reference lines, with a strap secured around the ankles to ensure standardised body position in each of the scans.
-The scans were conducted with the participants in a fasted state, with empty bladder, wearing under-wear only, between 07.00-10.00 AM.
-Prior to each scanning a phantom scanning was run to prevent baseline drifting from affecting analyses.
+For determination of body composition participants were scanned using DXA (Lunar Prodigy densitometer, GE Healthcare, Madison, WI, USA) with the standard scanning mode (13-25 cm).
+Participants were lying supine within the scanning bed reference lines, with a strap secured around the ankles to ensure a standardized body position in each scan.
+The scans were conducted with participants in a fasted state between 07.00-10.00 AM, with empty bladder and wearing only under-wear.
+Prior to each scan, a phantom scan was run to prevent baseline drifting from affecting analyses.
 The same technician was used at each time point.
@@
 Muscle thickness (MT) was measured using a B-mode ultra sound unit (SmartUS EXT-1M, Telemed, Vilnius, Lithuania).
-The participant lay supine in a relaxed position for 20 min before scan, with their feet strapped to a standardized box (15x15cm).
+Participants lay supine in a relaxed position for 20 min before assessments, with their feet strapped in a standardized position.
 A mark was set on the line 60% of the distance between Spinia Iliac Anterior Superior and the lateral femur condyle.
-MT of *m.vastus lateralis* was measured applying a water-soluble transmission gel (Aquasonic 100 Ultrasound Transmission Gel; Parker Laboratories Inc., Fairfield, NJ, USA), and a 39 mm 12 MHz ultrasound probe was placed perpendicular to the site of interest without depressing the skin.
+MT of *m.vastus lateralis* was measured applying a water-soluble transmission gel (Aquasonic 100 Ultrasound Transmission Gel; Parker Laboratories Inc., Fairfield, NJ, USA), and a 39 mm 12 MHz ultrasound probe was placed perpendicular to the site of interest without pressing the skin.
 When the quality of the image was satisfactory, evident as distinct upper and lower muscle fascia, three images were captured, where the probe was relocated to the same position between each image.
-Position of the probe was marked on the skin and subsequently marked on a transparent paper to ensure similar probe placement for both the right and left *m.vastus lateralis* at the subsequent assessments.
-Analyses were done in Jamovi (REF).
-Pictures were cropped and coded for blinding of the assessor.
-The average MT of the three images at each assesment was used for further analyses.
-
-<!--Schindelin J, Arganda-Carreras I, Frise E, et al. Fiji: an open-source platform for biological-image analysis. Nat
-1222 Methods. 2012;28(9(7)):676-682-->
+Position of the probe was marked on the skin and subsequently marked on a transparent paper to ensure similar probe placement for both the right and left *m.vastus lateralis* at subsequent assessments.
+Analyses were done in ImageJ Fiji [@RN2561] with images cropped and coded to ensure blinding of the assessor.
 
@@
 Muscle specimens were sampled bilaterally from *m.
-vastus laterlis* under local anesthesia (lidocaion %, Manufacturer, Oslo, Norway) using a disposable needle (12-14 gauge, Universal plus, Medax, Italy), operated with a spring loaded device (Bard Magnum).
+vastus laterlis* under local anesthesia (Lidokain 10 mg ml<sup>-1</sup>, Mylan, Mylan Ireland Limited, Dublin, Ireland) using a disposable needle (12-14 gauge, Universal plus, Medax, Poggio Rusco, Italy), operated with a spring loaded device (Bard Magnum, Bard Norway, Rud, Norway).
 Two to four passes were made to get sufficient material.
-One to two aliquots of the samples were quickly dissected free from connective and fat tissue, weighed and frozen in isopentan chilled to -80&deg;C and stored at -80&deg;C until further processing.
-At baseline, after twelve weeks of training and, after the de-training period, a further aliquot from each sample was selected for immunohistochemistry analysis and subsequently immersed in 4% formalin for fixation.
+One to two aliquots of the samples were quickly dissected free from connective and fat tissue, weighed and frozen in isopentan chilled to -80&deg;C and stored at -80&deg;C until further processing.<!-- At baseline, after twelve weeks of training and, after the de-training period, a further aliquot from each sample was selected for immunohistochemistry analysis and subsequently immersed in 4% formalin for fixation.-->
 
 ## RNA and protein extraction
-Frozen muscle tissue was homogenized in 1 ml of Trizol (ThermoFisher Scientific, Oslo, Norway) spiked with 0.04 ng of an external, non-mammalian, RNA spike-in (Lambda, Takara Bioscience).
+Frozen muscle tissue was homogenized in 1 ml of Trizol (ThermoFisher Scientific, Oslo, Norway) spiked with 0.04 ng of an external, non-mammalian, RNA spike-in (Lambda PolyA External Standard Kit, Takara Bio Europe, Saint-Germain-en-Laye, France).
 Mechanical disruption of the samples was achieved using Zirconium Oxide Beads (0.5 mm, Next Advance, Inc., New York, USA) and a bead mill (Bullet blender, Next Advance).
 Chloroform (200 &#956;l) was added prior to centrifugation (12000 g, 15 min at 4&deg;C) to achieve phase separation.
-Four houndred fifty &#956;l of the upper aqueous phase was transferred to a fresh tube and 500 &#956;l of isopropanol was added to precipitate the RNA.
+Four hundred fifty &#956;l of the upper aqueous phase was transferred to a fresh tube and 500 &#956;l of isopropanol was added to precipitate the RNA.
 After a 10 min incubation at room temperature, samples were centrifuged (12000 g, 10 min at 4&deg;C), after which a pellet formed.
@@
 
-Protein was extracted from Trizol preparations according to manufacturers instructions and [@KOPEC] with modifications.
-The remaining aqueous phase was removed and DNA was precipitated by the addition of XXX &#956;l of absolute ethanol followed by gentle centrifugation (2000 g, 5 min at room temperature).
+Protein was extracted from Trizol preparations according to manufacturers instructions and [@RN2050]<!-- Kopec trizol protocol --> with modifications.
+The remaining aqueous phase was removed and DNA was precipitated by the addition of 300 &#956;l of absolute ethanol followed by gentle centrifugation (2000 g, 5 min at room temperature).
 An aliquot of the phenol-ethanol phase, corresponding to ~1.75 mg of tissue, was transferred to to a fresh tube.
@@
 The pellet was washed three times in 95% ethanol with each wash separated by centrifugation (5000g, 5 min at room temperature).
-After the last wash all liquid was removed and 45 &#956;l of Kopec buffer was added (4% SDS, 300 mM Tris ..., pH; containing protease and phosphatase inhibitors).
+After the last wash all liquid was removed and 45 &#956;l of Kopec buffer [@RN2050] was added (5% SDS, 10 mM Tris, 140 mM NaCl and 20 mM EDTA, pH 8; containing protease and phosphatase inhibitors).
 Pellets were incubated at 50&deg;C for three hours after which the majority of samples were dissolved.
@@
 ## Quantitative polymerase chain reaction (qPCR)
-Complementary DNA (cDNA) was synthesized in technical duplicates from 500 ng of total RNA using random hexemer and anchored Oligo-dT primers (Thermo Scientific) together with Superscript IV (ThermoFisher Scientific) according to manufacturer's instruction.
-qPCR reactions were performed with diluted cDNA (2 &#956;l, 1:25 dilution), a SYBR-green based commercial master mix (PowerUp™ SYBR™ Green Master Mix, Thermo Fisher) and, target-specific primers (500 nM) in 10 &#956;l reaction volumes using a real-time detection system (QuantStudio 5 Real-Time PCR System, Thermo Fisher).
+Complementary DNA (cDNA) was synthesized in technical duplicates from 500 ng of total RNA using random hexemer and anchored Oligo-dT primers (Thermo Fisher Scientific) together with Superscript IV (Thermo Fisher Scientific) according to manufacturer's instruction.
+qPCR reactions were performed with diluted cDNA (2 &#956;l, 1:25 dilution), a SYBR-green based commercial master mix (PowerUp™ SYBR™ Green Master Mix, Thermo Fisher) and, target-specific primers (500 nM) in 10 &#956;l reaction volumes using a real-time detection system (QuantStudio 5 Real-Time PCR System, Thermo Fisher Scientific).
 Fast cycling was used (1 sec denaturing, 30 sec annealing) after UNG (2 min, 50&deg;C) and polymerase  (2 min, 95&deg;C) activation.
 Melt curves were collected from all reactions to confirm single product amplification.
-Primers were further evaluated by agarose gel electrophoresis an non-template control experiments.
+Primers were further evaluated by agarose gel electrophoresis which confirmed primer sizes and non-template control experiments confirming no amplification without template.
 Primer sequences and their respective average performances are shown in Table 2.
@@
 ```
-
-Raw fluorescence data was exported from the QuantStudio software and estimates of quantification cycles (Cq) and amplification efficiencies were derived using the qpcR package[qpcR-package].
-Relative abundance data was linearized using average efficiencies per primer pair and reaction specific Cq-values before any further statistical treatment.
+Raw fluorescence data was exported from the QuantStudio software and estimates of quantification cycle (Cq) and amplification efficiency was derived for each reaction using the qpcR package[@RN1768].
 
@@
 Protein samples (20 &#956;g) were separated on 4-20% Tris-Glycin gels (Criterion TGX Precast Gels, Bio-Rad) at 250 V for 45 min using the recommended running buffer (25 mM Tris, 192 mM Glycin, 0.1% SDS).
-All samples from the same participant were run on the same gel and all samples were run in at leat duplicates.
+All samples from the same participant were run on the same gel and all samples were run in at least duplicates.
 Separated samples were transferred to &#956m PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 Glycin, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h.
-Membranes were then stained to confirm transfer and enable total protein quantification using a reversible protein stain (Pierce Reversible Protein Stain, ThermoFisher Scientiﬁc).
-Primary antibodies were acquired to detect UBF (F-9, sc-13125, Santa-Cruz Biotechnology) and ribosomal protein S6 (54D2, #2317, Cell signaling technology).
-After blocking (Tris-buffered saline blocking buffer, 20 mM Tris, 150 mM NaCl, 5% fat-free milk, 0.1% Tween-20), membranes were incubated over-night with primary antibodies diluted in blocking buffer (UBF, 1:200; S6, 1:1000) followed by incubation with a secondary antibody conjugated to horseradish peroxidase (Anti-mouse IgG, #7076, Cell Singlaing Technology, 1:10000).
+Membranes were then stained to confirm transfer and enable total protein quantification using a reversible protein stain (Pierce Reversible Protein Stain, Thermo Fisher Scientiﬁc).
+Primary antibodies were acquired to detect UBF (F-9, sc-13125, Santa-Cruz Biotechnology, Dallas, Texas, USA) and ribosomal protein S6 (54D2, #2317, Cell Signaling Technology, Danvers, MA, USA).
+After blocking (Tris-buffered saline blocking buffer, 20 mM Tris, 150 mM NaCl, 5% fat-free milk, 0.1% Tween-20), membranes were incubated over-night with primary antibodies diluted in blocking buffer (UBF, 1:200; S6, 1:1000) followed by incubation with a secondary antibody conjugated to horseradish peroxidase (Anti-mouse IgG, #7076, Cell Signaling Technology, 1:10000).
 Membranes were washed 6 &#215; 5 min after incubation with primary antibodies and 8 &#215; 5 min after incubation with the secondary antibody.
 All incubation and washing steps were performed at 4&deg;C using an automatic membrane processor (BlotCycler, Precision Biosystems, Mansﬁeld, MA, USA).
-Chemiluminescent signals from membranes were detected after 5 min incubation in substrate (Super Signal West Femto Maximum Sensitivity Substrate, ThermoFisher Scientiﬁc) using a documentation system ().
-Total protein content was quantified whole membrane images defined as the mean gray value of the whole lane.
+Chemiluminescent signals from membranes were detected after 5 min incubation in substrate (Super Signal West Femto Maximum Sensitivity Substrate, Thermo Fisher Scientiﬁc) using a documentation system.
+Total protein content was quantified from whole membrane images and defined as the mean gray value of the whole lane.
 Between-lane gray values were used as background subtracted from protein values.
-Total protein quantification was done using ImageJ [REF].
+Total protein quantification was done using ImageJ Fiji [@RN2561].
 Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA).
@@
 ## Statistics and data analysis
+Descriptive data are presented as mean and standard deviation (SD).
+The effect of training on muscle strength, muscle thickness, protein, total RNA and gene abundances was assessed in mixed effects regression models with time and group (TR vs.
+CTRL) information combined in a single factor as population effects and leg nested within participant included as group effects.
+Interaction effects were calculated from contrasting changes ($\Delta$) within each group to each other (i.e.
+$\Delta$ TR - $\Delta$ CTRL).
+The effects of different volume conditions and general time-course patterns were assessed from data using all pairwise observations from the TR group.
+Based on graphical evaluation of protein and total RNA data, segmented regression models were used to estimate slopes in three segments (session 1-4, 4-8 and 8-12; corresponding to blocks of different volume prescription in the training group).
+When no robust effects of volume conditions were detected, group averages are presented.
+Segmented models were fitted with time and volume condition as population effects and legs nested within participants as group level effects.
+Muscle strength, muscle thickness, protein and total RNA data was modeled after log transformation.
+Gene abundance data was fitted with time as a categorical variable in comparing volume conditions and Cq values converted to counts as suggested by Matz et al.
+[@RN1964].
+A Poisson-lognormal model was used to fit the count data using all genes in the data set and including group level effects for each technical duplicate controlling for technical errors in sample preparation.
+An offset consisting of a normalization factor proportional to the amount of muscle used to prepare cDNA was used to model gene abundance per tissue weight.
+The external reference gene was used to calculate the normalization factor (External reference counts $\times$ muscle weight (mg) in each Trizol preparation).
+The offset was specified as a predictor with the coefficient fixed to 1.
+All models were fitted in a Bayesian framework using either the brms [@RN2562] or MCMCglmm [@RN1992] package written for R [@Rcore].
+Inference about effects of interest was done based on point estimates and their 95% credible intervals (CI).
+Credible intervals not containing null effects were interpreted as robust.
+Models were fitted with default priors which also makes CI analogous to frequentist confidence intervals but with the interpretation being that the CI contains the true population value with the specified certainty (95%) given the data.
+Fitting performance was assessed by confirming convergence of at least four different chains of MCMC samples (graphically assessed and confirmed with $\hat{R}\approx 1$).
+Model performance was assessed from comparing simulated data from each model to observed data graphically (posterior predictive checks).
 
@@
 
-summary(tc.m2)
-
 time1.hyp <- hypothesis(tc.m2, "time1 = 0")
@@
 
-UBF levels determine rRNA gene activity throgh regulation of chromatin[@RN2556] <!--  Sanij 2008 -->
+A main objective of this study was to establish a time course for training induced ribosomal biogenesis.
+We first set out to confirmed that markers of ribosome density were actually training induced and could do so as total RNA, ribosomal RNA subspecies and ribosomal protein S6 mRNA and protein increased in the training group compared to the non-training control group.
+Total RNA has been used in many previous studies as an approximation of ribosomal density as the majority of total RNA is ribosomal [@RN2142]<!-- Figueriedo --> and total RNA concentrations predicts protein synthesis [@RN2145]<!-- Millward 1973 -->.
+Previous studies have indicated that the training induced rise in total RNA reaches a peak after which a plateau or decrease is observed
+[@RN1809;<!-- brook 2016 -->
+@RN2358].<!-- Hammarström -->
+In the present study we observed a rapid increase in the first four sessions after which a plateau was observed.
+The observed plateau could be a consequence of the unit of measure as total RNA is estimated per unit tissue weight and that ribosome biogenesis occurs simultaneously as protein accretion leading to a dilution effect [@RN2142]<!-- Figueriedo  -->.
+We could confirm muscle hypertrophy both after session 12 and after a period of de-training indicating that the dilution effect is a reasonable assumption in interpreting the observed plateau.
+Further supporting this view is that both rpS6 and UBF increased over the whole training period.
+From a mechanistic perspective, UBF is an important factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA polymerase I
+[@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
+Activation of UBF is thought to be related to signaling through the mechanosensitive mTOR pathway as rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
+[@RN1632; <!-- Nader rapamycin blocj UBF availability  -->
+@RN2564]<!-- Hannan 2003 -->
+<!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1.
+The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF -->
+Evidence from human exercise studies confirms training induced activation of UBF through phosphorylation
+[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
+@RN1644].<!-- Figueriedo ubf increase total and active form (ser388) --> 
+In addition to exercise induced activation of UBF, mechanical loading leads to increases of total levels of UBF
+[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
+@RN1644]. Increases in UBF was determined to be rapamycin insensitive after synergist ablation in mice
+[@RN1072]<!-- Goodman 2011 UBF increase due to synergist ablation is rapamycin insensitive https://doi.org/10.1113/jphysiol.2011.218255 --> 
+pointing to an effect observed in cell models where c-Myc induces UBF mRNA transcription [@RN2565].<!-- Pootinga 2004 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
+Interestingly the avalability of UBF *per se* has been shown to regulate rRNA transcription
+[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->  through control of rDNA gene activity
+[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->
+In the present study we measured total UBF levels and observed that the relative amount of UBF was related to total RNA concentrations over the whole course of the study indicating its role in ribosomal biogenesis.
+This is in contrast to what was concluded from a recent study of electrical stimulation in rats [@RN2567].
+Comparing the time course in the present to that from electrical stimulation in rats highlights that regulation and time course of ribosome biogenesis in response to mechanical stress may differ between species.
+
+Exercise training volume, expressed in human exercise studies as the number of sets performed per session and muscle group is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
+[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
+@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
+@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
+total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
+and subsequent training outcomes [@RN1776; @RN2358].<!-- Schoenfeld 2016 meta and Hammarström -->
+We have previously shown different amounts of training volume in constant volume protocols leads to differences in total RNA and rRNA accumulation in the initial phase of RT [@RN2358].
+Based on the notion that ribosomal biogenesis is a complex tightly regulated and energy demanding process for any cell [@RN1820] we hypnotized that fluctuations in training load would reflect in markers of ribosomal biosynthesis reflecting the current need.
+Comparing the two volume conditions in the present study we only found one part of the 45S ETS to be differentially expressed at Session 12 in favor of VAR.
+Together with a tendency of a rescued UBF level in response to increased volume in the VAR protocol after de-training, these observations do not give much support for hypothesis.
+It is instead possible that after the initial four sessions a reduction of training load was sufficient to maximize ribosome biogenesis in this short time span.
+This leads us to suggest that ribsomal biogenesis is not sensitive to fluctuations in training volume during the initial phase of resistance training.
+
+This observation has consequences for the use of total RNA per unit tissue weight as a determinant muscle growth as the muscle growth itself potentially affects the measurement.
+Indeed many of the reported observational evidences for a determinant role of ribosome biogenesis in training induced muscle growth has reported measures from the early phase of RT.
+Figueriedo
+
+Brook et al. [@RN1809] reported a positive correlation between RNA to DNA ratio
 
@@
 18 and 28 S increased over sed at 48 h post ex with no further increase beyond 3 bouts 45S maximal at three bouts -->
-
-<!-- Goodman 2011 UBF increase due to synergist ablation is rapamycin insensitive https://doi.org/10.1113/jphysiol.2011.218255 -->
-
-<!-- Contrfary to goodman Nader found rapamycin sensitivity in UBF increase 10.1152/ajpcell.00165.2005-->
-
-<!-- Hannan 1996 10.1074/jbc.271.6.3213 increased rRNA transcription is associated with increased levels of UBF mRNA and protein in cardiomyocutes -->
-
-<!-- Hannan 1996 10.1073/pnas.93.16.8750 over expression of UBF leads to rRNA transcription in cardiomyocytes -->
-
-<!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1.
-The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF  -->
-
-<!-- In cell lines UBF is proportional to the rat e of rDNA transcription ref 25/26 in Hannan-->
-
-<!-- phospho of UBF shows it has a role in recruioting SL1 to the promotot  10.1152/ajpcell.00176.2006 -->
-
-<!-- Pootinga 2004 10.1038/sj.emboj.7600335 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
 

### 2020-09-17 manuscript.Rmd
manuscript and changed order of figures

@@
 
-Skeletal muscles not only constitutes our locomotive apparatus but also serve as our primary storage site of amino-acids.
+Skeletal muscles not only constitutes our locomotive apparatus but also serve as our primary storage site of amino acids.
 This makes the skeletal muscle system a critical target for interventions that promotes health across the lifespan [@RN763]<!-- Wolfe 2006 -->.
@@
 In recent years, this view has been supplemented with evidence showing that chronic RT also leads to an increased resting synthetic rate of muscle protein
-[@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> @RN1866;<!-- Wilkingson 2008 -->
+[@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> 
+@RN1866;<!-- Wilkingson 2008 -->
 @RN1897]<!-- Reidy 2017 Increased rested state protein synthesis -->.
@@
 a notion supported by observations relating exercise-induced changes in total RNA, a proxy marker for ribosome abundance, to muscle hypertrophy
-[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->@RN1755; <!-- Stec Ribosome bio higher in high responders -->
+[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
+@RN1755; <!-- Stec Ribosome bio higher in high responders -->
 @RN2358]<!-- Volume dependent RNA accum relates to volume depend hyper -->.
@@
 [@RN1809].
-These data suggests that ribosome accumulation per tissue weight, reaches a plateau in the early phase of RT.
-Interestingly, in the study by Figueiredo and colleagues, acute exercise in the trained state (after eight weeks of RT), rRNA and pre-rRNA rapidly decreases in response to acute exercise [@RN1644].
+These data suggests that ribosome accumulation, reaches a plateau in the early phase of RT.
+
 Based on these observations we are hypothesizing that (1) ribosome accumulation occurs in the early phase (3-4 weeks) of RT and (2) reaches a plateau during this phase in response to constant volume RT and finally (3) that fluctuations in ribosome abundance occurs in response to altered training volume and training cessation.
@@
 Descriptive data are presented as mean and standard deviation (SD).
-The effect of training on muscle strength, muscle thickness, protein, total RNA and gene abundances was assessed in mixed effects regression models with time and group (TR vs.
-CTRL) information combined in a single factor as population effects and leg nested within participant included as group effects.
-Interaction effects were calculated from contrasting changes ($\Delta$) within each group to each other (i.e.
-$\Delta$ TR - $\Delta$ CTRL).
+The effect of training on muscle strength, muscle thickness, UBF and rpS6 protein, total RNA and gene abundances was assessed in mixed effects regression models with time and group (TR vs.
+CTRL) combined in a single factor as population (fixed) effects and leg nested within participant included as group level (random) effects.
+Data from TR that matched CTRL was included in the comparisons.
+Changes ($\Delta$) within each group were contrasted to each other (i.e.
+$\Delta$ TR - $\Delta$ CTRL) to estimate interactions.
 The effects of different volume conditions and general time-course patterns were assessed from data using all pairwise observations from the TR group.
@@
 The offset was specified as a predictor with the coefficient fixed to 1.
+
+A linear model was used to estimate the increase per session and average total RNA for every leg in the TR group.
+These estimates were then used to estimate the effect of training induced increase per session and average total RNA abundance on muscle hypertrophy.
+For each leg, session was used as the independent variable centered on Session 6 and log transformed RNA per tissue weight as the dependent variable.
+Mean-centering of the independent variable was done to obtain an estimate of the average RNA concentration per leg.
+This also assured that the slope and intercept did not correlate, something that could lead to colinearity issues in subsequent modeling.
+A mixed effects model was subsequently fitted with differences in muscle thickness pre- to post-training as the dependent variable and estimated percentage per session increases in total RNA, the mean total RNA scaled as standard deviations from the mean and sex as independent variables.
+Leg nested within participant was used as group levels effects.
+The robustness of this model was assessed by leave-one-out analysis on the level of individual data points in the relationship between total RNA and sessions and on the level of participants (see Results).
+
 All models were fitted in a Bayesian framework using either the brms [@RN2562] or MCMCglmm [@RN1992] package written for R [@Rcore].
@@
 Credible intervals not containing null effects were interpreted as robust.
-Models were fitted with default priors which also makes CI analogous to frequentist confidence intervals but with the interpretation being that the CI contains the true population value with the specified certainty (95%) given the data.
+Models were fitted with default priors which also makes CI analogous to confidence intervals but with the interpretation being that the CI contains the true population value with the specified certainty (95%), given the data.
 Fitting performance was assessed by confirming convergence of at least four different chains of MCMC samples (graphically assessed and confirmed with $\hat{R}\approx 1$).
@@
 
-load.stats <- read_excel("./data/tr010_training.xlsx", sheet = 1, na = "NA") %>%
-  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
-  filter(exercise == "legext") %>%
-  mutate(set.load = repetitions * load, 
-         week = if_else(session %in% c(1:4), "W1", if_else(session %in% c(5:8), "W2", "W3"))) %>%
-  group_by(participant, cond, session, week) %>%
-  summarise(load = mean(load, na.rm = TRUE)) %>%
-  mutate(week = factor(week, levels = c("W1", "W2", "W3")))
+m2 <- readRDS("./data/derivedData/training-data/m2.RDS")
+# Model containing interaction 
+# readRDS("./data/derivedData/training-data/m1.RDS")
 
-load.m1 <- lme(log(load) ~ week + week:cond, random = list(participant = ~1), 
-               data = load.stats, method = "ML")
-load.m0 <- lme(log(load) ~ week, random = list(participant = ~1), 
-               data = load.stats, method = "ML")
+# Create a hypothesis and convert to percentages
 
-load.anova <- anova(load.m0, load.m1)
+h <- hypothesis(m2, c("(weekW2 / Intercept) * 100 = 0",
+                 "(weekW3 / Intercept) * 100 = 0"))$hypothesis
 
-load.intervals <- data.frame(exp(intervals(load.m0)$fixed) - 1)
-
-load.inr.w2 <- paste0(sprintf("%.0f",load.intervals[2, 2]*100), 
+load.inr.w2 <- paste0(sprintf("%.0f",h[1, 2]), 
                       "%, 95% CI: [",
-                      sprintf("%.0f",load.intervals[2, 1]*100), 
+                      sprintf("%.0f",h[1, 4]), 
                       ", ",
-                      sprintf("%.0f",load.intervals[2, 3]*100), 
+                      sprintf("%.0f",h[1, 5]), 
                       "]")
-load.inr.w3 <- paste0(sprintf("%.0f",load.intervals[3, 2]*100), 
-                      "%, 95% CI: [",
-                      sprintf("%.0f",load.intervals[3, 1]*100), 
+load.inr.w3 <- paste0(sprintf("%.0f",h[2, 2]), 
+                      "%, [",
+                      sprintf("%.0f",h[2, 4]), 
                       ", ",
-                      sprintf("%.0f",load.intervals[3, 3]*100), 
+                      sprintf("%.0f",h[2, 5]), 
                       "]")
@@
 This analysis indicated that the relationship between Total RNA increase and muscle growth was preserved in all iterations but the upper limit of the 95% CI spanned 0 in some instances regarding the relationship between average RNA and muscle growth (Figure 4D).
-We further assessed the model by iteratively removing one participant from the data set, similarly to removing a single data point this showed that estimates of the total RNA increase effect on muscle growth was robust but average total RNA estimates spanned 0 in some cases (Participant 8 and 3 in Figure 4D).
-
-```{r predict-muscle-growth}
-
-```
+We further assessed the model by iteratively removing one participant from the data set, similarly this showed that estimates of the total RNA increase effect on muscle growth was robust but average total RNA estimates spanned 0 in some cases (Participant 8 and 3 in Figure 4D).
 
@@
 A main objective of this study was to establish a time course for training induced ribosomal biogenesis.
-We first set out to confirmed that markers of ribosome density were actually training induced and could do so as total RNA, ribosomal RNA subspecies and ribosomal protein S6 mRNA and protein increased in the training group compared to the non-training control group.
+We first set out to confirmed that markers of ribosome density were actually training induced and could do so as total RNA, ribosomal RNA subspecies and rpS6 mRNA and protein increased in the training group compared to the non-training control group.
 Total RNA has been used in many previous studies as an approximation of ribosomal density as the majority of total RNA is ribosomal [@RN2142]<!-- Figueriedo --> and total RNA concentrations predicts protein synthesis [@RN2145]<!-- Millward 1973 -->.
@@
 We could confirm muscle hypertrophy both after session 12 and after a period of de-training indicating that the dilution effect is a reasonable assumption in interpreting the observed plateau.
-Further supporting this view is that both rpS6 and UBF increased over the whole training period.
+Further supporting this view is UBF increased over the whole training period indicating that the transcriptional apparatus continued to increase over the whole course of the training period.
 From a mechanistic perspective, UBF is an important factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA polymerase I
@@
 @RN1644].<!-- Figueriedo ubf increase total and active form (ser388) --> 
-In addition to exercise induced activation of UBF, mechanical loading leads to increases of total levels of UBF
+In addition to exercise induced activation of UBF, mechanical loading leads to increased levels of total UBF
 [@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
@@
 In the present study we measured total UBF levels and observed that the relative amount of UBF was related to total RNA concentrations over the whole course of the study indicating its role in ribosomal biogenesis.
-This is in contrast to what was concluded from a recent study of electrical stimulation in rats [@RN2567].
-Comparing the time course in the present to that from electrical stimulation in rats highlights that regulation and time course of ribosome biogenesis in response to mechanical stress may differ between species.
+Together the above suggests that rRNA transcription may not plateau during short term training but muscle hyertrophy leads to a dilution effect.
+This observation has consequences for the use of total RNA per unit tissue weight as a determinant muscle growth as the muscle growth itself affects the measurement.
 
-Exercise training volume, expressed in human exercise studies as the number of sets performed per session and muscle group is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
+After the short de-training period, total RNA levels decreased in the training group (check if this is correct robust?
+More than no change in control?).
+This occured at the same time as no additional muscle hypertrophy was observed.
+[To do: check de training effect on pre-rRNA and total rRNA].
+<!-- This is in contrast to what was concluded from a recent study of electrical stimulation in rats [@RN2567].
+Comparing the time course in the present to that from electrical stimulation in rats highlights that regulation and time course of ribosome biogenesis in response to mechanical stress may diffe between species.-->
+Resistance training volume, expressed in human exercise studies as the number of sets performed per session and muscle group is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
 [@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
@@
 total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
-and subsequent training outcomes [@RN1776; @RN2358].<!-- Schoenfeld 2016 meta and Hammarström -->
+and subsequent training outcomes [@RN1767; @RN2358].<!-- Schoenfeld 2016 meta and Hammarström -->
 We have previously shown different amounts of training volume in constant volume protocols leads to differences in total RNA and rRNA accumulation in the initial phase of RT [@RN2358].
@@
 Comparing the two volume conditions in the present study we only found one part of the 45S ETS to be differentially expressed at Session 12 in favor of VAR.
-Together with a tendency of a rescued UBF level in response to increased volume in the VAR protocol after de-training, these observations do not give much support for hypothesis.
+Together with a tendency of a rescued UBF level in response to increased volume in the VAR protocol after de-training, these observations do not give much support for this hypothesis.
 It is instead possible that after the initial four sessions a reduction of training load was sufficient to maximize ribosome biogenesis in this short time span.
@@
 
-This observation has consequences for the use of total RNA per unit tissue weight as a determinant muscle growth as the muscle growth itself potentially affects the measurement.
-Indeed many of the reported observational evidences for a determinant role of ribosome biogenesis in training induced muscle growth has reported measures from the early phase of RT.
-Figueriedo
+In conclusion...q
 
-Brook et al. [@RN1809] reported a positive correlation between RNA to DNA ratio
-
-Time course data 
-UBF and S6 protein levels may be differently regulated in men and rats as more consecutive bouts of electrical stimulation in rats does not prtoduce the same pattern seen in our study 
-<!-- Kotani 2019 10.1152/japplphysiol.00665.2018 increase in t UBF at 6 h post ex but not 48 despite 2 weeks training (3 x week).
+<!-- UBF and S6 protein levels may be differently regulated in men and rats as more consecutive bouts of electrical stimulation in rats does not prtoduce the same pattern seen in our study 
+ Kotani 2019 10.1152/japplphysiol.00665.2018 increase in t UBF at 6 h post ex but not 48 despite 2 weeks training (3 x week).
 S6 increase at 48 h and 6 bouts but return to sed values at 18 bouts.
 18 and 28 S increased over sed at 48 h post ex with no further increase beyond 3 bouts 45S maximal at three bouts -->
-
-Per total RNA 
-Per muscle tissue
 

### 2020-09-17 Figures/Figure2.R

@@ -1,617 +1,282 @@
-############# Figure 2 ##################################
-
-
-## Panels
-
-# Use Bayesian estimates?
-bayes <- TRUE
+##-------------------------------------
+## Figure2.R
+##
+## Title: Figure 2
+## Purpose: Make file for Fig 2
+## Author: DH
+##
+##
+##
+##-------------------------------------
+## Notes: 
+# Figure 2 contains wetsern blot data 
+# and corresponding mRNA data (rps6 and UBF)
+#
+#
+#
+#
+#
+#
+#
+## ------------------------------------
 
 
 source("./R/figure-source.R")
-source("./R/libs.R")
+# source("./R/libs.R")
 
 
-# 45S primers graphics #############################
-
-### rRNA and primers #########
 
 
-# Sequences can be found : https://www.ncbi.nlm.nih.gov/nuccore/NR_146144.1
-# Add primer sequences to plot and annotate mature segments
 
-# Primer locations
-primers <- read_excel("./data/tr010_primers.xlsx") %>%
-  filter(primer_id %in% c("rRNA47S F1R1", 
-                          "rRNA45S F5R5", 
-                          "rRNA45SITS F12R12", 
-                          "rRNA5.8S F2R2", 
-                          "rRNA28S F2R2", 
-                          "rRNA18S F2R2")) %>%
+s6.m1 <- readRDS( "./data/derivedData/western-analysis/s6_tx_m1.RDS")
+ubf.m1 <- readRDS( "./data/derivedData/western-analysis/ubf_tx_m1.RDS")
+
+
+
+tx_contrasts <-  c("txS1_int_train -     txpre_int_train =      txS1_con_train",
+                   # Interaction effects post-training
+                   "txpost_int_train -     txpre_int_train =      txpost_con_train",
+                   # Interaction effects detraining (post1w)
+                   "txpost_int_detrain -     txpre_int_train =      txpost_con_train",
+                   # Whitin group fold change
+                   "txS1_int_train =     txpre_int_train",     
+                   "txpost_int_train =     txpre_int_train",     
+                   "txpost_int_detrain =     txpre_int_train",     
+                   "txS1_con_train = 0",
+                   "txpost_con_train = 0")
+
+
+
+h.s6 <- hypothesis(s6.m1, tx_contrasts)
+h.ubf <- hypothesis(ubf.m1, tx_contrasts)
+
+# Create a data frame for the results 
+tx_results_western <- data.frame(target = rep(c("rpS6","UBF"), each = 8), 
[…]

### 2020-09-17 Figures/Figure3.R

@@ -1,282 +1,617 @@
-##-------------------------------------
-## Figure3.R
-##
-## Title: Figure 3
-## Purpose: Make file for Fig 3
-## Author: DH
-##
-##
-##
-##-------------------------------------
-## Notes: 
-# Figure 3 contains wetsern blot data 
-# and corresponding mRNA data (rps6 and UBF)
-#
-#
-#
-#
-#
-#
-#
-## ------------------------------------
+############# Figure 3 ##################################
+
+
+## Panels
+
+# Use Bayesian estimates?
+bayes <- TRUE
 
 
 source("./R/figure-source.R")
-source("./R/libs.R")
+# source("./R/libs.R")
 
 
+# 45S primers graphics #############################
+
+### rRNA and primers #########
 
 
+# Sequences can be found : https://www.ncbi.nlm.nih.gov/nuccore/NR_146144.1
+# Add primer sequences to plot and annotate mature segments
 
-s6.m1 <- readRDS( "./data/derivedData/western-analysis/s6_tx_m1.RDS")
-ubf.m1 <- readRDS( "./data/derivedData/western-analysis/ubf_tx_m1.RDS")
-
-
-
-tx_contrasts <-  c("txS1_int_train -     txpre_int_train =      txS1_con_train",
-                   # Interaction effects post-training
-                   "txpost_int_train -     txpre_int_train =      txpost_con_train",
-                   # Interaction effects detraining (post1w)
-                   "txpost_int_detrain -     txpre_int_train =      txpost_con_train",
-                   # Whitin group fold change
-                   "txS1_int_train =     txpre_int_train",     
-                   "txpost_int_train =     txpre_int_train",     
-                   "txpost_int_detrain =     txpre_int_train",     
-                   "txS1_con_train = 0",
-                   "txpost_con_train = 0")
-
-
-
-h.s6 <- hypothesis(s6.m1, tx_contrasts)
-h.ubf <- hypothesis(ubf.m1, tx_contrasts)
-
-# Create a data frame for the results 
-tx_results_western <- data.frame(target = rep(c("rpS6","UBF"), each = 8), 
-                                 comparison = rep(c("inter:S1",
-                                                    "inter:post", 
-                                                    "inter:post1w", 
-                                                    
-                                                    "int_S1", 
-                                                    "int_post", 
[…]

### 2020-09-17 R/figure-source.R

@@ -6,7 +6,6 @@
 
 library(tidyverse)
 library(cowplot)
-library(ggpattern)
 library(ggridges)
 library(ggrepel)
 library(ggtext)

### 2020-09-17 R/libs.R

@@ -6,7 +6,7 @@ library(readxl)
 library(qpcR)
 
 # devtools::install_github("dhammarstrom/qpcrpal")
-library(qpcrpal)
+# library(qpcrpal)
 library(ggrepel)
 
 
@@ -15,9 +15,9 @@ library(nlme)
 library(lme4)
 library(brms)
 library(ggmcmc)
-library(emmeans)
+
 library(mgcv)
-library(multcomp)
+
 
 
 

### 2020-09-17 R/training-data.R

@@ -1,0 +1,75 @@
+# Training load data 
+
+
+library(tidyverse)
+library(readxl)
+library(brms)
+library(lme4)
+
+
+
+load.stats <- read_excel("./data/tr010_training.xlsx", sheet = 1, na = "NA") %>%
+  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+  filter(exercise == "legext") %>%
+  mutate(set.load = repetitions * load, 
+         week = if_else(session %in% c(1:4), "W1", if_else(session %in% c(5:8), "W2", "W3"))) %>%
+  group_by(participant,leg,sex, cond, session, week) %>%
+  summarise(load = mean(load, na.rm = TRUE)) %>%
+  mutate(week = factor(week, levels = c("W1", "W2", "W3")))  %>%
+  print()
+
+  
+
+load.stats %>%
+  ggplot(aes(week, load, color = sex, shape = cond)) + geom_point(position = position_jitter(width = 0.2))
+
+
+# Is there differences between conditions?
+
+
+
+m1 <- brm(bf(load ~ week * cond + (1|participant)),
+               
+               warmup = 1000, # number of samples before sampling
+               iter = 4000,  # number of mcmc iterations 
+               cores = 4, # number of cores used in sampling
+               chains = 4, # number of chains
+               seed = 5, # seed for reproducible results
+               #    control = list(adapt_delta = 0.95), 
+               data = load.stats)
+
+pp_check(m1, type = "stat", stat = "median")
+# The data is in agreement with what the model predicts, good.
+# Leave one out statistics
+loo(m1)
+
+# No effects of condition, these are disregarded to produce 
+
+m2 <- brm(bf(load ~ week + (1|participant/leg)),
+          
+          warmup = 1000, # number of samples before sampling
+          iter = 4000,  # number of mcmc iterations 
+          cores = 4, # number of cores used in sampling
+          chains = 4, # number of chains
+          seed = 5, # seed for reproducible results
+          #    control = list(adapt_delta = 0.95), 
+          data = load.stats)
+
+pp_check(m2, type = "stat", stat = "median")
+# The data is in agreement with what the model predicts, good.
+# Leave one out statistics
+loo(m1)
+
+summary(m2)
+pp_check(m2, type = "ecdf_overlay")
+
+
+
+
+saveRDS(m2, "./data/derivedData/training-data/m2.RDS")
+saveRDS(m1, "./data/derivedData/training-data/m1.RDS")
+
+
+
+
+# All Pareto estimates are OK

### 2020-09-21 manuscript.Rmd
misc

@@
 
-Skeletal muscles not only constitutes our locomotive apparatus but also serve as our primary storage site of amino acids.
-This makes the skeletal muscle system a critical target for interventions that promotes health across the lifespan [@RN763]<!-- Wolfe 2006 -->.
-Resistance training can promote beneficial effects on skeletal muscles and does so through exercise-induced changes in the balance between muscular protein breakdown and synthesis, with successful RT leading to net protein accretion and muscle hypertrophy.
-A traditional view of exercise-induced muscle hypertrophy has been that muscle protein accumulates as a consequence of repeated episodes of elevated post-exercise protein synthesis in the acute phase, resulting in a net-positive protein balance over time.
-Indeed, one bout of resistance exercise acutely up-regulates skeletal muscle protein synthesis up to 48 hr after exercise [@RN786]<!-- Phillips 1997 -->
-and subsequently, repeated bouts would lead to accumulation of muscle protein over time
+Skeletal muscle is a critical target for interventions that promotes health across the lifespan [@RN763]<!-- Wolfe 2006 -->,
+with resistance training (RT) being the advocated remedy.  
+Such training promotes changes in the balance between muscular protein breakdown and synthesis, thereby leading to protein accretion and muscle hypertrophy, alongside a multitude of functional and health-related benefits.
+Exercise-induced muscle hypertrophy is traditionally viewed as the result of repeated episodes of elevated post-exercise protein synthesis, leading to net positive protein balance over time.
+Indeed, one bout of resistance exercise acutely increases skeletal muscle protein synthesis for up to 48 hrs after the exercise [@RN786],<!-- Phillips 1997 -->
+with subsequent repeated bouts leading to accumulation of muscle protein over time
 [@RN1272<!-- Phillips review 2014 -->;
 @RN1642]<!-- Brook 2015 experiment -->. 
-In recent years, this view has been supplemented with evidence showing that chronic RT also leads to an increased resting synthetic rate of muscle protein
+In recent years, this view has been supplemented<!-- supported? --> by evidence suggesting that chronic RT leads to increased resting synthetic rate of muscle protein
 [@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> 
 @RN1866;<!-- Wilkingson 2008 -->
-@RN1897]<!-- Reidy 2017 Increased rested state protein synthesis -->.
-An increase in the resting protein synthetic rate has been postulated to be related to an increased translational capacity, i.e. accumulation of ribosomes
+@RN1897],<!-- Reidy 2017 Increased rested state protein synthesis -->
+which has been postulated to be associated with increased translational capacity, i.e. accumulation of ribosomes
 [@RN1897;<!-- Reidy 2017 -->
-@RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->,
-a notion supported by observations relating exercise-induced changes in total RNA, a proxy marker for ribosome abundance, to muscle hypertrophy
+@RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->.
+This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, occuring alongside muscle hypertrophy
 [@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
 @RN1755; <!-- Stec Ribosome bio higher in high responders -->
-@RN2358]<!-- Volume dependent RNA accum relates to volume depend hyper -->.
-Furthermore, inhibition of ribosomal RNA (rRNA) transcription or inhibition of up-stream transcription factors leads to halted biosynthesis of ribosomes and diminished cellular growth
+@RN2358]<!-- Volume dependent RNA accum relates to volume depend hyper -->
+and the close relationship between RNA abundance and protein synthesis
+[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
+@RN2145].<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
+Conversely, inhibition of ribosomal RNA (rRNA) transcription and inhibition of its up-stream transcription factors act to diminish muscle cell growth
 [@RN1755<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->;
@@
 
-Formation of new ribosomes is a complex and energy demanding process involving the whole transcriptional apparatus
-[@RN1920;<!-- Warner review yeast -->,
+Biosynthesis of novel ribosomes is a complex, highly coordinated and energy demanding process that involves synthesis of both proteins and four different mature rRNA
+[@RN1920;<!-- Warner review yeast -->
 @RN1820;<!-- Moss review -->
 @RN1810]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
-In skeletal muscle, a single bout of resistance exercise elevates markers of ribosome biogenesis in skeletal muscle evident from studies investigating immature forms of rRNA in response to exercise in the untrained state
+This process is limited at the level of pre-rRNA transcription by RNA polymerase I (Pol I) which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
+[@RN1820]<!-- Moss review -->.
+Specifically, phosphorylation of the upstream bindning factor (UBF) is needed initiate transcription
+[@RN2582; <!-- Tuan UBF phosphorylation important for SL1-UBF interaction -->
+@RN2563]<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
+and this is related to signaling through the mechanosensitive mTOR pathway as its inhibition blocks UBF phosphorylation and subsequent rRNA transcription
+[@RN1632; <!-- Nader rapamycin block UBF availability  -->
+@RN2564].<!-- Hannan mTOR  -->
+Interestingly the availability of UBF *per se* has been shown to also regulate rRNA transcription
+[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->
+through control of rDNA gene activity
+[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->
+
+Resistance exercise is a potent stimuli for rRNA transcription as a single session leads to increases in pre-rRNA
 [@RN1656<!-- A single session upregulates 45 S 24 h post ex -->;
-@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->].
-Studies examining repeated bouts of resistance exercise suggests that RT leads to accumulation of ribosomes, measured as total RNA, with increased number of completed sessions
+@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->]
+and repeated bouts lead to accumulation of mature rRNA thus also total RNA and presumably functional ribosomes
 [@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
@@
 @RN1897<!-- 11% 36 sessions young -->].
-Time-course data is however sparse as only a limited number of studies have investigated exercise induced accumulation over multiple time-points.
-Bickel *et al*.
-showed that two subsequent bouts of a hypertrophic stimulus increased total RNA to the highest levels at 72 h after a second bout of RT [@RN1520]<!-- Total RNA increase 72 h after a second bout -->.
-We have previously reported peak RNA levels 48 h after the fourth training session with a tendency towards normalization after 31 sessions
+However, the true time course of ribosomal transcription and accumulation in response to RT remains largely unstudied, with only a few studies having investigated exercise-induced changes in rRNA over multiple time-points.
+As such, Bickel *et al*.
+showed that two consecutive bouts of hypertrophic stimuli were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout [@RN1520]<!-- Total RNA increase 72 h after a second bout -->.
+Brook *et al.*, reported peak values after nine sessions followed by a slight decrease to after 18 sessions
+[@RN1809],
+resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), whereupon a numerical lowering occurred to after the last training session of the 12 wk interventions (31 session).
 [@RN2358].
-This is similar to what was reported by Brook *et al.* where peak values were seen after nine sessions followed by a slight decrease to after 18 sessions
-[@RN1809].
-These data suggests that ribosome accumulation, reaches a plateau in the early phase of RT.
+Interestingly, total RNA accumulation in the initial phase was shown to be training volume sensitive as three sets per exercise in leg exercises led to increased total RNA and rRNA accumulation compared to a single set per exercise, coinciding with differences in muscle hypertrophy after week 12.
+[@RN2358].
+These data suggest that ribosome accumulation reaches a plateau in the early phase of RT and that increases are sensitive to training volume in constant volume protocols.
 
-Based on these observations we are hypothesizing that (1) ribosome accumulation occurs in the early phase (3-4 weeks) of RT and (2) reaches a plateau during this phase in response to constant volume RT and finally (3) that fluctuations in ribosome abundance occurs in response to altered training volume and training cessation.
-Subsequently the aim of this study is to characterize fluctuations in ribosomal biogenesis during the initial phase of RT, in response to altered training volumes and in response to a short-term de-training period.
+Based on these observations we hypothesize that (1) ribosome accumulation occurs during the early phase (3-4 weeks) of RT, within which this accumulation (2) reaches a plateau when RT volume is kept constant, (3) displays fluctuations in response to fluctuating training volume and (4) is partially reversed to one week after cessation of RT.
+In addition to addressing these hypotheses we aimed to relate RNA accumulation to total UBF levels and muscle growth.
 
@@
 ## Study overview
+
+```{r length-of-control-period}
+
+# Calculate the average of the control period (in weeks) for the CTRL group
+ctrl_days <- read_excel("./data/tissue/tr010_tissue.xlsx") %>%
+  filter(participant %in% paste0("P", 9:15)) %>%
+  mutate(date = as.Date(date, origin = "1899-12-30")) %>%
+  group_by(participant, time) %>%
+  summarise(date = mean(date)) %>%
+  pivot_wider(names_from = time, values_from = date) %>%
+  mutate(ctrl.period = postctrl - S0) %>%
+  ungroup() %>%
+  summarise(m = round(as.numeric(mean(ctrl.period))/7,1), 
+            s = round(as.numeric(sd(ctrl.period))/7, 1), 
+            min = round(as.numeric(min(ctrl.period))/7, 1), 
+            max =  round(as.numeric(max(ctrl.period))/7, 1))
+
+ctrl <- paste0(ctrl_days[1,1], " (", ctrl_days[1,2], ")") 
+```
+
 Eighteen volunteers were recruited to the study.
-Eligible participants had to be non-smokers, between 18 and 35 years of age, report a training history of less than one RT session per week during the six months leading up to the study and not consume dietary supplements or medication with known effects on muscle metabolism.
-Potential participants were further screened for injuries causing impaired strength and/or affecting their ability to perform RT, symptoms or history of disease or known adverse reactions to local anesthetics.
-Included participants were allocated to either an training group (TR, *n* = 11) or a non-training control group (CTRL, *n* = 8; see Table 1 for participant characteristics).
-TR performed a 12 session RT protocol consisting unilateral knee-extension to allow for within-participant comparison exercise volume.
-In TR, one leg performed constant volume (CONST, 6 sets per session) and the other leg performed variable volume over the course of the study (VAR, four sessions each with 6, 3 and 9 sets per session; Figure 1A).
+Eligible participants were non-smokers between 18 and 35 years of age with a training history of less than one RT session per week during the six months leading up to the study.
+Exclusion criteria were consumption of dietary supplements or medication with known effects on muscle metabolism, injuries causing impaired strength and/or affecting their ability to perform RT, symptoms or history of disease, and known adverse reactions to local anesthetics.
+Participants were allocated to either an training group (TRAIN, *n* = 11) or a non-training control group (CTRL, *n* = 8; see Table 1 for participant characteristics; see Figure 1A for overview of the intervention).
+TRAIN performed a 12 session RT protocol lasting for 3-4 weeks, consisting of 10 repetition maximum (RM) unilateral knee-extension, with the two legs conducting RT with different volume profiles, allowing within-participant comparison of the effects of volume regimes.
+In TRAIN, one leg conducted RT with constant volume throughout the intervention (CONST, 6 sets per session) and the other leg performed RT with variable volume (VAR, 3 blocks of four sessions with 6, 3 and 9 sets per session, respectively; Figure 1A).
 CTRL did not partake in RT and were instructed to continue their everyday activities.
-Muscle biopsies were sampled bilaterally in TR before and 48 h after the first session and 48 h after the fourth, fifth, eight, ninth and twelfth session as wells as after a seven day de-training period.
-Muscle biopsies were obtained from CTRL at three occasions with 48 h separating the first two and 2-4 weeks separating second and third sampling.
-TR and CTRL performed strength assessments at least seven days prior to the first biopsy sampling and 72 h after the twelfth session (EXP) and 24 h after the last biopsy (TR and CTRL).
+Muscle biopsies were sampled bilaterally in TRAIN before and 48 hrs after the first session, as well as 48 hrs after the fourth, fifth, eight, ninth and twelfth session, and after a seven days of de-training.
+Muscle biopsies were obtained from CTRL at three occasions; at baseline and 48 hrs and 3-5 weeks (average (SD) `r ctrl`) after the first sampling event.
+TRAIN and CTRL performed strength assessments > seven days prior to the first biopsy sampling (TRAIN; CTRL), 72 hrs after the twelfth session (TRAIN) and 24 h after the last biopsy (TRAIN, following de-training; CTRL).
 Appendicular lean mass (Dual-energy X-ray absorptiometry, DXA) and muscle thickness of *m.
-vastus lateralis* were assessed prior to the first biopsy (TR and CTRL) as wells as before the second to last (TR) and last (TR and CTRL) biopsy (Figure 1A).
+vastus lateralis* were assessed prior to the first biopsy (TRAIN and CTRL) as wells as before the second to last (TRAIN) and last (TRAIN and CTRL) biopsy.
 
 ## RT protocol
-Prior to all RT-sessions, participants performed a standardized warm-up consisting of 5 min ergometer cycling (rating of perceived exertion (RPE): 12-14), followed by ten repetitions each of push-ups, sit-ups and back-extensions.
+Prior to all RT sessions, participants performed a standardized warm-up consisting of 5 min ergometer cycling (rating of perceived exertion (RPE): 12-14), followed by ten repetitions of push-ups, sit-ups and back-extensions.
 After warm-up, participants performed unilateral knee-extension with the prescribed number of sets.
 Each set was prescribed with 10 repetitions maximum (RM).
-When sets was completed with either fewer (8) or more (12) repetitions than 10, the resistance was adjusted accordingly.
+When sets were completed with either fewer (8) or more (12) repetitions, the resistance was adjusted accordingly.
 Inter-set rest periods were 90 sec.
-Participants performed the first set alternating between the right and left leg every other session and the set of the contralateral leg was performed in the rest-period of the first leg.
-The second session of each four-session block (session 2, 6 and 10) were performed at a sub-maximal resistance (~90% of the previous session) with the same number of repetitions.
-In each session, after the knee-extension exercise, participants completed two sets each of three upper-body exercises (bench press, lateral pull-down and shoulder press).<!--Prior to each session, participants were asked to rate muscle soreness on a scale from 0 to 10 (0 = no soreness, 10 = hurting very bad, e.g.
+Throughout the intervention RT sessions were alternatingly initiated by training the right and left leg, changing every other session, with the contralateral leg being trained in the rest period between sets of the first leg.
+The second session of each four-session block (session 2, 6 and 10) was performed at a sub-maximal resistance (~90% of the previous session) with the same number of repetitions (10).
+Within each session, participants also conducted two sets of three upper-body exercises (bench press, lateral pull-down and shoulder press; 10RM).<!--Prior to each session, participants were asked to rate muscle soreness on a scale from 0 to 10 (0 = no soreness, 10 = hurting very bad, e.g.
 not being able to walk properly) [@Vickers2010].
 After each session, participants were asked to rate their perceived exertion (Borg CR10).-->
-After the completion of each session participants were given a standardized drink to aid recovery (0.15 g kg<sup>-1</sup> protein, 11.2 g kg<sup>-1</sup> carbohydrates and 0.5 g kg<sup>-1</sup> fat).
+After completion of each session participants were given a standardized drink to aid recovery (0.15 g kg<sup>-1</sup> protein, 11.2 g kg<sup>-1</sup> carbohydrates and 0.5 g kg<sup>-1</sup> fat).
 
@@
 Descriptive data are presented as mean and standard deviation (SD).
-The effect of training on muscle strength, muscle thickness, UBF and rpS6 protein, total RNA and gene abundances was assessed in mixed effects regression models with time and group (TR vs.
-CTRL) combined in a single factor as population (fixed) effects and leg nested within participant included as group level (random) effects.
-Data from TR that matched CTRL was included in the comparisons.
-Changes ($\Delta$) within each group were contrasted to each other (i.e.
-$\Delta$ TR - $\Delta$ CTRL) to estimate interactions.
-The effects of different volume conditions and general time-course patterns were assessed from data using all pairwise observations from the TR group.
-Based on graphical evaluation of protein and total RNA data, segmented regression models were used to estimate slopes in three segments (session 1-4, 4-8 and 8-12; corresponding to blocks of different volume prescription in the training group).
+The effect of training on muscle strength, muscle thickness, UBF/rpS6 protein, total RNA and gene abundances were assessed using mixed effects regression models.
+Time and group (TRAIN vs.
+CTRL) were treated as population (fixed) effects and leg nested within participant included as group level (random) effects.
+These analyses were performed on data with matching time points between TRAIN and CTRL with the exeption that all post-training data from TRAIN were included (post-training and de-training).
+Relative interactions between groups were estimated as $\Delta$ TRAIN - $\Delta$ CTRL.
+The effects of different volume conditions and general time-course patterns were assessed using all pairwise observations from the TRAIN group.
+For protein and total RNA data, segmented regression models were used to estimate changes over sessions in three segments (session 1-4, 4-8 and 8-12; corresponding to blocks of different volume prescription in TRAIN).
 When no robust effects of volume conditions were detected, group averages are presented.
 Segmented models were fitted with time and volume condition as population effects and legs nested within participants as group level effects.
+
 Muscle strength, muscle thickness, protein and total RNA data was modeled after log transformation.
-Gene abundance data was fitted with time as a categorical variable in comparing volume conditions and Cq values converted to counts as suggested by Matz et al.
-[@RN1964].
-A Poisson-lognormal model was used to fit the count data using all genes in the data set and including group level effects for each technical duplicate controlling for technical errors in sample preparation.
+Gene abundance data were fitted with number of sessions as a categorical variable in comparisons of volume conditions, and Cq values converted to counts as suggested by Matz et al.
+[-@RN1964].
+A Poisson-lognormal model was used to fit these count, using data from all genes and including group level effects for each technical duplicate, controlling for technical errors during sample preparation.
 An offset consisting of a normalization factor proportional to the amount of muscle used to prepare cDNA was used to model gene abundance per tissue weight.
@@
 A linear model was used to estimate the increase per session and average total RNA for every leg in the TR group.
-These estimates were then used to estimate the effect of training induced increase per session and average total RNA abundance on muscle hypertrophy.
+These estimates were then used to estimate the effect of training-induced increase per session and average total RNA abundance on muscle hypertrophy.
 For each leg, session was used as the independent variable centered on Session 6 and log transformed RNA per tissue weight as the dependent variable.
@@
 
-All models were fitted in a Bayesian framework using either the brms [@RN2562] or MCMCglmm [@RN1992] package written for R [@Rcore].
+All models were fitted using a Bayesian framework using either the brms [@RN2562] or MCMCglmm [@RN1992] package written for R [@Rcore].
 Inference about effects of interest was done based on point estimates and their 95% credible intervals (CI).
@@
 
-All participants successfully completed their prescribed sessions.
-Different volume conditions resulted in diverging loading profiles (load &#215; repetitions) over the course of the study (Figure 1B).
-The intensity of exercise (load at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) four-session block, respectively.
-Isokinetic strength increased to Session 12 in the experimental group compared to the non-training control group, with an attenuated difference seen after de-training (Figure 1C).
-Average isometric strength showed the same general pattern although with considerable more uncertainty indicated by wider 95% CI (Figure 1C).
-Muscle thickness of the *m.
-vastus lateralis* increased in the experimental group compared to the control group both after Session 12 and after eight days of de-training (Figure 1D).
-There were no indications of differences between training conditions in strength or muscle thickness changes.
+In TRAIN, all participants successfully completed their prescribed RT on both legs, with the two volume conditions resulting in diverging volume profiles (load &#215; repetitions) over the course of the study (Figure 1B).
+Exercise intensities (resistance at 10RM) increased similarly between conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with eacj block consisting of four training sessions.
+Isokinetic strength increased from baseline to after Session 12 in TRAIN compared to CTRL.
+This difference was still seen after de-training (Figure 1C).
+Isometric strength showed the same general pattern, though with considerably larger degrees of uncertainty, as indicated by  wider 95% CI (Figure 1C).
+Muscle thickness of *m.
+vastus lateralis* increased in TRAIN compared CTRL, measured both after Session 12 and after eight days of de-training (Figure 1D).
+There were no indications of differences between volume conditions in strength or muscle thickness changes.
 
@@
 
-Protein levels of RPS6 and UBF increased only after twelve sessions in the training group compared to control and stayed elevated after eight days of rest (Figure 2A and B).
-The average increase in protein levels over twelve sessions was near linear in both targets and estimated to `r s6_slope0_4`, `r s6_slope4_8` and `r s6_slope8_12` in RPS6, and `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12` in UBF per session between session 1-4, 4-8 and 8-12, respectively.
-Slopes did not differ between volume-conditions but there was a tendency towards lower levels of UBF in the variable condition at session 12 (`r var_s12`).
-After the de-training period the constant volume condition tended to show decreased levels of UBF (`r const_detrain`) while the variable volume condition remained at a similar level compare to session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
-Levels of rpS6 did not change in the de-training period and were similar between volume conditions.
+UBF and rpS6 protein levels were increased only after twelve sessions in the training group compared to control and stayed elevated after eight days of rest (Figure 2A and B), with no changes being observed after the first training session (48 hrs).
+For both rpS6 and UBF, protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12) and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
+Slopes did not differ between volume-conditions but there was a tendency towards lower levels of UBF in VAR after session 12 (`r var_s12`).
+After the de-training period CONST tended to show decreased levels of UBF (`r const_detrain`) while VAR remained at elevated level compare to after session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
+Levels of rpS6 did not change in response to the de-training period and remained similar between volume conditions.
 
@@
 slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,2]) - 1)),
-                           "%, 95% CI: [",
+                           "%, [",
                    sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,4]) - 1)), 
@@
 
-A single session of RT led to robust increase in 47S and 45S pre-rRNA expressed per unit tissue weight within the training group (Figure 3B) and compared to control (Figure 3C).
-These transcripts were also increased compared to control after twelve sessions together with the 45S ITS and mature rRNA species 28S and 18S (Figure 3C).
-After eight days of rest, only 18S and 28S remained elevated above levels measured in the control group (Figure 3B).
-This general pattern of rRNA expression was reflected also in total RNA expressed per unit tissue weight were a robust increase compared to control was seen after session twelve (Figure 3E and F).
-Within the experimental group, only small differences were seen in rRNA subspecies as 45S ETS was robustly higher after 12 sessions in the variable volume condition after an imbalance between conditions prior to training (Figure 3D).
-Total RNA increased similarly in both volume conditions throughout the training period (Figure 3G).
-Most of the increase was seen in the first four sessions where the average increase was `r slope0_4` per session in the intervention group as a whole.
-Between session four and eight, and eight to twelve changes in total RNA per tissue weight per session was not robustly different from zero (`r slope4_8` and `r slope8_12`).
-After the de-training period Total-RNA decreased by `r detrain_rna` from levels measured after Session 12.
+A single session of RT led to robust increases in pre-rRNA 47S and 45S abundance, measured as expression per unit tissue weight, both from baseline to 48 hrs after session 1 within TRAIN (Figure 3B) and compared to CTRL (Figure 3C).
+Similarly, 47S and 45S abundances were also increased after twelve sessions, together with 45S ITS and rRNA 28S and 18S species, measured both within-participant and compared to CTRL (Figure 3C).
+After eight days of rest, only 18S and 28S remained at elevated levels compared to CTRL (Figure 3B).
+This general pattern of rRNA expression was reflected by total RNA expression per unit tissue weight, which were robustly increase in TRAIN compared to CTRL after session twelve (Figure 3E and F), followed by decreased levels after de-training `r detrain_rna`.
+For both rRNA expression and total RNA levels, these training-associated increases in abundance predominately occured during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from session four to eight `r slope4_8` and from session eight to twelve `r slope8_12`.
+In TRAIN, the two volume conditions were associated with minor differences in rRNA subspecies abundance, with only 45S ETS showing differential expression between conditions, evident as robustly higher levels in VAR compared to CONST after the 12th session (Figure 3D), coinciding with increased volume in this condition towards the end of the intervention.
+No robust differences were seen between conditions for total RNA abundances (Figure 3G), increasing similarly in both conditions.
 
@@
 
-UBF levels robustly predicted total RNA values as a standard deviation change in UBF levels led to a `r ubf_cor_tot_rna` increase in total RNA per unit tissue weight (Table X).
-To estimate the influence of total RNA on muscle growth the average linear increase in total RNA was calculated for each leg together with the average total RNA abundance over the course of the training intervention in the training group.<!-- (estimated as the intercept in a regression model with number of sessions centered at session 6).
+In TRAIN (after for controlling for time), UBF levels robustly predicted total RNA levels with a one unit difference in UBF levels corresponding to one standard deviation leading to `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight (Table 3).
+To estimate the influence of total RNA on changes in muscle mass in TRAIN, measured as *m.vastus lateralis* thickness, the average linear increase in total RNA was calculated for each leg together with the average total RNA abundance over the course of the training intervention.<!-- (estimated as the intercept in a regression model with number of sessions centered at session 6).
 move to methods -->
-The relationship between muscle growth and characteristics of individual total RNA profiles were estimated in a regression model containing the increase in total RNA concentrations, expressed as percentage increase per session, and average total RNA abundance at Session 6, expressed as standard deviations from the mean.
-The model also contained sex as predictor to control for sex differences in muscle growth.
-Total RNA increase in response to training together with the average level of total RNA robustly predicted muscle growth (Table 4, Figure 4A and B).
-To assess the robustness of the model, each data point was iteratively removed from the data set and the model was refitted using new estimates of the total RNA abundance and increase.
-This analysis indicated that the relationship between Total RNA increase and muscle growth was preserved in all iterations but the upper limit of the 95% CI spanned 0 in some instances regarding the relationship between average RNA and muscle growth (Figure 4D).
-We further assessed the model by iteratively removing one participant from the data set, similarly this showed that estimates of the total RNA increase effect on muscle growth was robust but average total RNA estimates spanned 0 in some cases (Participant 8 and 3 in Figure 4D).
+The relationship between muscle growth and characteristics of individual total RNA profiles were estimated using a regression model containing the increase in total RNA concentrations, expressed as percentage increase per session, and average total RNA abundance at Session 6, expressed as standard deviations from the mean.
+The model also contained sex as predictor to account for sex differences in muscle growth.
+Total RNA increase in response to training RNA was robustly related to muscle growth (Table 4, Figure 4A).
+Average total RNA levels tended to negatively influence muscle growth (Table 4, Figure 4B).
+To assess the robustness of the model, individual relationships between sessions and total RNA levels (Figure 4C) were recalculated after removal of single data points.
+The model was refitted using new estimates of the total RNA abundance and increase per session.
+Effects of refitting the model is shown in Figure 4D.
+This analysis indicated that the relationship between Total RNA increase and muscle growth was preserved in all iterations (Figure 4D).
+We further assessed the model by iteratively removing one participant from the data set, similarly this showed that estimates of the total RNA increase effect on muscle growth was robust but the effect of average total RNA estimates were more variable (Participant 8 and 3 in Figure 4D).
 
@@
 Activation of UBF is thought to be related to signaling through the mechanosensitive mTOR pathway as rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
-[@RN1632; <!-- Nader rapamycin blocj UBF availability  -->
+[@RN1632; <!-- Nader rapamycin block UBF availability  -->
 @RN2564]<!-- Hannan 2003 -->

### 2020-09-21 tables.Rmd
misc

@@
 
+```{r table3-regression-models-muscle-thickness}
+# Save model
+
+comb.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")
+
+comb.m1 <- summary(comb.m1)
+
+reg_results <- 
+  cbind(rbind(comb.m1$fixed,
+      comb.m1$random$participant, 
+      comb.m1$random$`participant:leg`, 
+      comb.m1$spec_pars), 
+      data.frame(coef = c("Intercept", 
+                          "UBF protein levels (SD-units)", 
+                          "Session 1-4",
+                          "Session 4-8", 
+                          "Session 8-12", 
+                          "De-training",
+                          "Between Participant variation",
+                          "Between Participant:Leg variation", 
+                          "Residual SD"))) %>%
+  remove_rownames() %>%
+  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)
+
+std_border = fp_border(color="gray")
+
+reg_results %>%
+
+  mutate(Estimate = round(Estimate, 2), 
+         Est.Error = round(Est.Error, 2), 
+         lwr = round(lwr, 2), 
+         upr = round(upr, 2)) %>%
+
+  flextable() %>%
+  set_header_labels(coef = "Coefficient", 
+                    Est.Error = "SD", 
+                    lwr = "Lower 95% CI", 
+                    upr = "Upper 95% CI") %>%
+    add_header_lines(values = "Table 3. Effect of UBF levels, sessions and de-training on RNA-levels.") %>%
+
+ # bold(i = c(5, 6, 13, 14, 20, 21), j = c(2:6), bold = TRUE) %>%
+
+ # hline(i = c(9, 17), part="body", border = std_border ) %>%
+      fontsize(size = 8, part = "body") %>%
+    fontsize(size = 10, part = "header") %>%
+      fontsize(size = 7, part = "footer") %>%
+
+  footnote( i = 2, j = 2,
+            value = as_paragraph(
+              c("The dependent variable is total RNA levels (log)")
+            ),
+            ref_symbols = c("a"),
+            part = "header", inline = TRUE) %>%
+    footnote( i = c(3, 4, 5), j = 1,
+            value = as_paragraph(
+              c("Slope in response to session 1-4", 
+                "Change in slope in session 4-8", 
+                "Change in slope in session 8-12")
+            ),
+            ref_symbols = c("b","c","d"),
+            part = "body", inline = TRUE) %>%
+
+  autofit()
+
+```
+
+\pagebreak
+
 ```{r table4-regression-models-muscle-thickness}
 # Save model
-hyp.pre.m1 <- readRDS( "./data/derivedData/ubf-tot-rna-model/hyp_pre_m1.RDS")
-hyp.pre.m2 <- readRDS( "./data/derivedData/ubf-tot-rna-model/hyp_pre_m2.RDS")
+
 hyp.pre.m3 <- readRDS( "./data/derivedData/ubf-tot-rna-model/hyp_pre_m3.RDS")
 
-m1.sum <- summary(hyp.pre.m1)
-m2.sum <- summary(hyp.pre.m2)
 m3.sum <- summary(hyp.pre.m3)
@@
                     upr = "Upper 95% CI") %>%
-    add_header_lines(values = "Table 3. Total RNA as a predictor of muscle growth.") %>%
+    add_header_lines(values = "Table 4. Total RNA as a predictor of muscle growth.") %>%
 

### 2020-09-21 Figures/Figure4.R

@@ -101,8 +101,8 @@ loo_panel <- res %>%
 predict_df <- readRDS("./data/derivedData/ubf-tot-rna-model/predict_df.RDS")
 
 # get predictions from the model
-hyp.pre.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/hyp_pre_m1.RDS")
-
+hyp.pre.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/hyp_pre_m3.RDS")
+summary(hyp.pre.m1)
 effects_m1 <- conditional_effects(hyp.pre.m1)
 
 mean_diff <- (effects_m1$sex[2, 9] - effects_m1$sex[1, 9]) / 2

### 2020-09-21 R/ubf-tot-rna-model.R

@@ -99,10 +99,11 @@ west_cont <- western_data %>%
 
 combined_df  <- west_cont %>%
   group_by(target, participant, cond, time) %>%
-  summarise(ubf.scaled = mean(expression.scaled, na.rm = TRUE)) %>%
+  summarise(ubf.scaled = mean(expression.scaled, na.rm = TRUE), 
+            ubf = mean(expression)) %>%
   filter(target != "t-s6") %>%
   ungroup() %>%
-  dplyr::select(participant, cond, time, ubf.scaled) %>%
+  dplyr::select(participant, cond, time, ubf.scaled, ubf) %>%
   inner_join(rna_complete %>%
                mutate(rna.mg = rna/tissue_weight) %>%
                group_by(participant, leg, time,time.c, time1, time2, time3, cond) %>%
@@ -129,15 +130,89 @@ saveRDS(combined_df, "./data/derivedData/ubf-tot-rna-model/combined_df.RDS")
 # meaning that the effect of UBF levels are similar over the course of the study. 
 
 
-comb.m1 <- brm(bf(log.rna ~ ubf.scaled + (time1 +time2 + time3) + detrain + (1 + time1|participant)),
+comb.m1 <- brm(bf(log.rna ~ ubf.scaled + (time1 +time2 + time3) + detrain + (1|participant/leg)),
               
               warmup = 1000, # number of samples before sampling
-              iter = 4000,  # number of mcmc iterations 
+              iter = 5000,  # number of mcmc iterations 
               cores = 4, # number of cores used in sampling
               chains = 4, # number of chains
               seed = 5, # seed for reproducible results
               #    control = list(adapt_delta = 0.95), 
               data = combined_df)
+
+# A naive model not accounting for time 
+# comb.m1 <- brm(bf(log.rna ~ ubf.scaled + time + (1 + time1|participant)),
+#                
+#                warmup = 1000, # number of samples before sampling
+#                iter = 4000,  # number of mcmc iterations 
+#                cores = 4, # number of cores used in sampling
+#                chains = 4, # number of chains
+#                seed = 5, # seed for reproducible results
+#                #    control = list(adapt_delta = 0.95), 
+#                data = combined_df)
+# 
+
[…]

### 2020-09-21 manuscript.Rmd
discussion updates

@@
 
+ hypothesize that (1) ribosome accumulation occurs during the early phase (3-4 weeks) of RT, within which this accumulation (2) reaches a plateau when RT volume is kept constant, (3) displays fluctuations in response to fluctuating training volume and (4) is partially reversed to one week after cessation of RT.
+In addition to addressing these hypotheses we aimed to relate RNA accumulation to total UBF levels and muscle growth.
+
+Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, expression of ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
+We show that this increase is interconnected with increases in UBF protein abundance, suggesting a plausible manner for rRNA transcription regulation in response to RT.
+We did not find any convincing support for the notion that ribosomal biogenesis is tightly regulated by changes in training volume, however, training cessation may lead to halted ribosome biogenesis.
+
 A main objective of this study was to establish a time course for training induced ribosomal biogenesis.
+Previous data from human studies indicate that mechanical loading leads to increased total RNA concentrations
+<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
+[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
+@RN2005;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased -->
+
+@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
+@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
+@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
+@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
+@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
+@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
+@RN1897<!-- 11% 36 sessions young -->].
+Time course data to date have suggests that maximal levels are reached within four to nine sessions in young males and females
+[@RN1809;
+@RN2358].
+This was confirmed in the present study and
+
+In the present study we hypothesized that markers of ribosomal biogenesis would be sensitive to fluctuations in training volume.
+We based this hypothesis on a previous observation indicating that RT induced increase in total RNA, together with increases in rRNA are training volume dependent
+[@RN2358]. Additionally, ribosome biogenesis is known to be an energy demanding process
+[@RN1920]<!-- Warner, in cell models a high percentage  -->
+suggesting that it would be regulated on demand.
+It is important to note that RT in the current study was performed with constant volume in the first four sessions, something that could have maximized rRNA transcription in the initial phase.
+This is supported by the observation that pre-rRNA increased rapidly initially in both protocols with minimal changes in response to subsequent sessions, regardless of exercise volume.
+The CONST protocol in the present study corresponded to volumes used in the high volume condition in a previous study from our lab (three sets in two exercises activation knee extensor muscles) [@RN2358].
+Compared to a low volume, this led to higher levels of total RNA accumulation [@RN2358].
+This study was not designed to answers questions about the minimal RT dose required to maximize RT induced ribosomal accumulation, this remains to be determined.
+Interestingly, using a progressive volume protocol in well trained participants, increases in total RNA were seen throughout six weeks of training 
+[@RN2055].<!-- Mobley 2018 trend towards higher increases in high responders 32% high resp, 8% low resp 12 wk training -->
+Although intriguing, this observation was done without any constant volume control.
+
+[@RN1916]<!-- Bickel 2003 RNA lower in atrophied muscles from spinal cord injured -->
+
+The time course may be informative for future studies
+
 We first set out to confirmed that markers of ribosome density were actually training induced and could do so as total RNA, ribosomal RNA subspecies and rpS6 mRNA and protein increased in the training group compared to the non-training control group.

### 2020-09-22 manuscript.Rmd
discussion notes

@@
 
-In the present study we hypothesized that markers of ribosomal biogenesis would be sensitive to fluctuations in training volume.
-We based this hypothesis on a previous observation indicating that RT induced increase in total RNA, together with increases in rRNA are training volume dependent
-[@RN2358]. Additionally, ribosome biogenesis is known to be an energy demanding process
-[@RN1920]<!-- Warner, in cell models a high percentage  -->
-suggesting that it would be regulated on demand.
+Resistance training volume, expressed in human exercise studies as the number of sets performed per session and muscle group is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
+[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
+@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
+@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
+total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
+and subsequent training outcomes [@RN1767; @RN2358].<!-- Schoenfeld 2016 meta and Hammarström -->
+We have previously shown different amounts of training volume in constant volume protocols leads to differences in total RNA and rRNA accumulation in the initial phase of RT [@RN2358].
+Based on the notion that ribosomal biogenesis is a complex and tightly regulated, energy demanding process
+[@RN1920],<!-- Warner economics of ribosomal biogenesis -->
+we hypnotized that fluctuations in training load would reflect in markers of ribosomal biosynthesis reflecting the current need.
+Comparing the two volume conditions in the present study we only found one part of the 45S ETS to be differentially expressed at Session 12 in favor of VAR together with a tendency of a rescued UBF level after de-training in response to increased volume in the VAR protocol.
+These observations do not give much support for a clear effect of fluctuations in training volume in the short time span used in the present study although the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume.
 It is important to note that RT in the current study was performed with constant volume in the first four sessions, something that could have maximized rRNA transcription in the initial phase.
@@
 The CONST protocol in the present study corresponded to volumes used in the high volume condition in a previous study from our lab (three sets in two exercises activation knee extensor muscles) [@RN2358].
-Compared to a low volume, this led to higher levels of total RNA accumulation [@RN2358].
-This study was not designed to answers questions about the minimal RT dose required to maximize RT induced ribosomal accumulation, this remains to be determined.
-Interestingly, using a progressive volume protocol in well trained participants, increases in total RNA were seen throughout six weeks of training 
+Compared to a low volume protocol, this led to higher levels of total RNA accumulation [@RN2358].
+Interestingly, using a progressive volume protocol in well-trained participants, increases in total RNA were seen throughout six weeks of training
 [@RN2055].<!-- Mobley 2018 trend towards higher increases in high responders 32% high resp, 8% low resp 12 wk training -->
-Although intriguing, this observation was done without any constant volume control.
+Compared to studies using constant volume protocols
+[@RN1809;<!-- Brook 2016 -->
+@RN2358],<!-- Hammarström 2020 -->
+progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
+This observation was done in well-trained participants performing a high volume protocol without a comparator for volume periodization
 
@@
 Further supporting this view is UBF increased over the whole training period indicating that the transcriptional apparatus continued to increase over the whole course of the training period.
+
 From a mechanistic perspective, UBF is an important factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA polymerase I
 [@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
-Activation of UBF is thought to be related to signaling through the mechanosensitive mTOR pathway as rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
+
+Activation of UBF is thought to be controlled by signaling from the mechanosensitive mTOR, pathway as rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
 [@RN1632; <!-- Nader rapamycin block UBF availability  -->
@@
 The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF -->
-Evidence from human exercise studies confirms training induced activation of UBF through phosphorylation
+Evidence from human exercise studies confirms training associated activation of UBF through phosphorylation
 [@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
 @RN1644].<!-- Figueriedo ubf increase total and active form (ser388) --> 
-In addition to exercise induced activation of UBF, mechanical loading leads to increased levels of total UBF
+In addition to exercise-induced activation of UBF, mechanical loading also leads to increased levels of total UBF
 [@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
@@
 [@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->
-In the present study we measured total UBF levels and observed that the relative amount of UBF was related to total RNA concentrations over the whole course of the study indicating its role in ribosomal biogenesis.
+In the present study, we measured total UBF levels and observed that the relative amount of UBF was related to total RNA concentrations over the whole course of the study indicating its role in ribosomal biogenesis.
 Together the above suggests that rRNA transcription may not plateau during short term training but muscle hyertrophy leads to a dilution effect.
-This observation has consequences for the use of total RNA per unit tissue weight as a determinant muscle growth as the muscle growth itself affects the measurement.
+This observation has consequences for the use of total RNA per unit tissue weight as a determinant muscle growth as hypertrophy itself affects the measurement.
 
@@
 Comparing the time course in the present to that from electrical stimulation in rats highlights that regulation and time course of ribosome biogenesis in response to mechanical stress may diffe between species.-->
-Resistance training volume, expressed in human exercise studies as the number of sets performed per session and muscle group is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
-[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
-@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
-@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
-total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
-and subsequent training outcomes [@RN1767; @RN2358].<!-- Schoenfeld 2016 meta and Hammarström -->
-We have previously shown different amounts of training volume in constant volume protocols leads to differences in total RNA and rRNA accumulation in the initial phase of RT [@RN2358].
-Based on the notion that ribosomal biogenesis is a complex tightly regulated and energy demanding process for any cell [@RN1820] we hypnotized that fluctuations in training load would reflect in markers of ribosomal biosynthesis reflecting the current need.
-Comparing the two volume conditions in the present study we only found one part of the 45S ETS to be differentially expressed at Session 12 in favor of VAR.
-Together with a tendency of a rescued UBF level in response to increased volume in the VAR protocol after de-training, these observations do not give much support for this hypothesis.
-It is instead possible that after the initial four sessions a reduction of training load was sufficient to maximize ribosome biogenesis in this short time span.
-This leads us to suggest that ribsomal biogenesis is not sensitive to fluctuations in training volume during the initial phase of resistance training.
 

### 2020-09-22 manuscript.Rmd
discussion notes

@@
 title: "Ribosome accumulation during early phase resistance training"
-author: Daniel Hammarström^1,2,£^&ast;,Sjur J. Øfsteng^2,£^, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^,  Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
-date: ""
-header-includes:
-  - \usepackage{wasysym}
-csl: resources/experimental-physiology.csl
-bibliography: resources/bibliography.bib
+author: Daniel Hammarström^1,2,£^&ast;,Sjur J. Øfsteng^2,£^, Nicolai B. Jacobsen^1^,
+  Krister B. Flobergseter^1^,  Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
+date: ''
 output:
@@
     reference_docx: resources/manuscript_word_template.docx
+csl: resources/experimental-physiology.csl
+bibliography: resources/bibliography.bib
+header-includes: \usepackage{wasysym}
 editor_options:

### 2020-09-22 manuscript.Rmd
updates to discussion

@@
 [@RN1820]<!-- Moss review -->.
-Specifically, phosphorylation of the upstream bindning factor (UBF) is needed initiate transcription
+Specifically, phosphorylation of the upstream bindning factor (UBF) is needed to initiate transcription
 [@RN2582; <!-- Tuan UBF phosphorylation important for SL1-UBF interaction -->
@@
 In TRAIN, all participants successfully completed their prescribed RT on both legs, with the two volume conditions resulting in diverging volume profiles (load &#215; repetitions) over the course of the study (Figure 1B).
-Exercise intensities (resistance at 10RM) increased similarly between conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with eacj block consisting of four training sessions.
+Exercise intensities (resistance at 10RM) increased similarly between conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with each block consisting of four training sessions.
 Isokinetic strength increased from baseline to after Session 12 in TRAIN compared to CTRL.
@@
 Levels of rpS6 did not change in response to the de-training period and remained similar between volume conditions.
+UBF was robustly up-regulated at the mRNA level compared to CTRL after 48 hrs (Figure 2D) with no other robust differences seen between TRAIN and CTRL or between volume conditions in TRAIN in either UBF or rpS6 mRNA levels (Figure 2D and E).
 
@@
 In TRAIN (after for controlling for time), UBF levels robustly predicted total RNA levels with a one unit difference in UBF levels corresponding to one standard deviation leading to `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight (Table 3).
+In contrast, no evidence for a relationship between total RNA and rpS6 was found when controlling for time (Table 3).
 To estimate the influence of total RNA on changes in muscle mass in TRAIN, measured as *m.vastus lateralis* thickness, the average linear increase in total RNA was calculated for each leg together with the average total RNA abundance over the course of the training intervention.<!-- (estimated as the intercept in a regression model with number of sessions centered at session 6).
@@
 
-Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, expression of ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
+Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
 We show that this increase is interconnected with increases in UBF protein abundance, suggesting a plausible manner for rRNA transcription regulation in response to RT.
 We did not find any convincing support for the notion that ribosomal biogenesis is tightly regulated by changes in training volume, however, training cessation may lead to halted ribosome biogenesis.
+Finally, the rate of increase in total RNA abundance predicted the magnitude of muscle growth.
 
-A main objective of this study was to establish a time course for training induced ribosomal biogenesis.
-Previous data from human studies indicate that mechanical loading leads to increased total RNA concentrations
-<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
-[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
-@RN2005;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased -->
+Several studies have presented evidence of a determinant role for ribosomal biogenesis in RT induced muscle hypertrophy.
+Increases in total RNA has been shown to positively correlated with training induced hypertrophy
+[@RN1644;<!-- Figueriedo  -->
+@RN1897;
+@RN2055].<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
+<!-- Reidy correlation between total RNA and muscle hypertrophy-->
+and differs between high- and low-responders to RT
+[@RN1755].<!-- Stec 2016  -->
+Additionally, volume-dependent regulation of total RNA predicted beneficial effects of high- vs. low-volume RT
+[@RN2358]. In addition to observational evidence from human studies, blockage of ribosomal biogenesis halts muscle cellular growth *in vitro* in some
+[@RN1632;<!-- Nader 2005 blockage of mTOR halts ribo bio and growth -->
+@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
+@RN1755]<!-- Stec specific pol I inhibition -->
+but not all situations
+[@RN1929]<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
+Results from the present study corroborates the link between ribosomal biogenesis and muscle hypertrophy.
+However, when using all available data points we were able to robustly estimate both the rate of increase in total RNA and the average RNA levels over the course of the study and include both when modeling changes in muscle thickness.
+The results of this analysis showed that the rate, but not the absolute levels of total RNA predicted muscle growth.
+In other words, at a fixed relative rate of increase in total RNA, legs with higher levels of RNA did not grow more but at a fixed amount of total RNA, legs with higher increase per session did.
+This suggest that the absolute ribosomal density is not as important as the rate of increase in ribosomal density.
+In an homogeneous group, in response to RT, the rate of increase will be highly correlated with maximal values seen after RT, linking maximal rate to the resulting abundance.
+However, the same pattern revealed in our model is seen when comparing young and old muscle, where aged muscle display higher levels of total RNA at rest
+[@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
+and lower levels of increase in response to RT
+[@RN1809].<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
 
-@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
-@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
-@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
-@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
-@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
-@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
-@RN1897<!-- 11% 36 sessions young -->].
-Time course data to date have suggests that maximal levels are reached within four to nine sessions in young males and females
-[@RN1809;
-@RN2358].
-This was confirmed in the present study and
-
+Both training protocols utilized in the present study increased strength and induced muscle hypertrophy in the relative short time-span studied.
+This is in line with previous data utilizing similar protocols
+[@RN1642].<!-- Brook 2015 young only, strength and muscle thickness increased over control legs -->
+Fluctuations in training volume did not influence training outcomes, although volume is an important factor for increases in  muscle strength and mass
+[@RN1767;<!-- schoenfeld 2016 dose response  -->
+@RN2492]<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
+different organization of training loads over time is likely of minor importance when training volumes are equated
+[@RN2572].<!-- Grgic 2017 volume equated periodization schemes led to similar muscle hyeprtrophy -->
 Resistance training volume, expressed in human exercise studies as the number of sets performed per session and muscle group is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
@@
 progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
-This observation was done in well-trained participants performing a high volume protocol without a comparator for volume periodization
+This observation was done in well-trained participants performing a high volume protocol without a control group condition with constant volume.
 
-[@RN1916]<!-- Bickel 2003 RNA lower in atrophied muscles from spinal cord injured -->
-
-The time course may be informative for future studies
+A main objective in the present study was to establish a time course for training induced ribosomal biogenesis.
+Previous data from human studies indicate that mechanical loading leads to increased total RNA concentrations
+<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
+[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
+@RN2180;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --> 
+@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
+@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
+@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
+@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
+@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
+@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
+@RN1897<!-- 11% 36 sessions young -->], indicating expansion of the ribosomal pool as the majority of RNA is ribosomal
+[@RN2223].<!-- Young 1970  -->  
+Time course data to date have suggests that maximal levels are reached within four to nine sessions in young males and females
+[@RN1809;
+@RN2358].
+This was confirmed in the present study and
 
@@
 Further supporting this view is UBF increased over the whole training period indicating that the transcriptional apparatus continued to increase over the whole course of the training period.
+
+<!-- rpS6 dissconect from total RNA levels. -->
+In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6.
+Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period.
+Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
+[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
+but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
+[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
+Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for (Table 3).
+Together with a disconnect after the de-training period, this may suggest that regulation of rpS6 expression and transcription of ribosomal RNA displays different temporal characteristics resulting from RT.
+Additionally, ribosomal proteins may have extra-ribosomal functions
+affecting their expression
+[@RN2588].<!-- Warner 2009 General description of   -->
 
@@
 
-After the short de-training period, total RNA levels decreased in the training group (check if this is correct robust?
-More than no change in control?).
-This occured at the same time as no additional muscle hypertrophy was observed.
-[To do: check de training effect on pre-rRNA and total rRNA].
-<!-- This is in contrast to what was concluded from a recent study of electrical stimulation in rats [@RN2567].
-Comparing the time course in the present to that from electrical stimulation in rats highlights that regulation and time course of ribosome biogenesis in response to mechanical stress may diffe between species.-->
-
-In conclusion...q
-
 <!-- UBF and S6 protein levels may be differently regulated in men and rats as more consecutive bouts of electrical stimulation in rats does not prtoduce the same pattern seen in our study 
@@
 
-**Figure 2.**  Total RNA and ribosomal RNA subspecies in response to trainin.
+**Figure 2.** Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF.
+Black points and error bars represents statistically robust results (a 95% CI not containing 0).
+C shows western-blots and total protein stains from a representative participant.
+mRNA data is normalized per total RNA.
+
+**Figure 3.**  Total RNA and ribosomal RNA subspecies in response to trainin.
 Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (A).
@@
 
-**Figure 3.** mRNA (A and B) and protein abundances (C and D) of ribosomal protein S6 and Upstream bindning factor (UBF).
-mRNA abundaces are expressed per total RNA (round points) or per muscle weight (triangles).
-Black points and error bars represents statistically robust results (a 95% CI not containing 0).
-E shows western-blots and total protein stains from a representative participant.
+**Figure 4.**

### 2020-09-22 tables.Rmd
updates to discussion

@@
 
+# Model for rpS6 predicting total RNA
+comb.s6.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")
+
 comb.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")
@@
 comb.m1 <- summary(comb.m1)
+comb.s6.m1 <- summary(comb.s6.m1)
 
@@
 
-std_border = fp_border(color="gray")
+reg_results_s6 <- 
+  cbind(rbind(comb.s6.m1$fixed,
+      comb.s6.m1$random$participant, 
+      comb.s6.m1$random$`participant:leg`, 
+      comb.s6.m1$spec_pars), 
+      data.frame(coef = c("Intercept", 
+                          "rpS6 protein levels (SD-units)", 
+                          "Session 1-4",
+                          "Session 4-8", 
+                          "Session 8-12", 
+                          "De-training",
+                          "Between Participant variation",
+                          "Between Participant:Leg variation", 
+                          "Residual SD"))) %>%
+  remove_rownames() %>%
+  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)
 
-reg_results %>%
+std_border = fp_border(color="black", width = 2)
+
+  rbind(reg_results, reg_results_s6) %>%
 
@@
                     upr = "Upper 95% CI") %>%
-    add_header_lines(values = "Table 3. Effect of UBF levels, sessions and de-training on RNA-levels.") %>%
+    add_header_lines(values = "Table 3. Effect of UBF and rpS6 levels, sessions and de-training on RNA-levels.") %>%
 
@@
             part = "header", inline = TRUE) %>%
-    footnote( i = c(3, 4, 5), j = 1,
+    footnote( i = c(3, 4, 5, 12, 13, 14), j = 1,
             value = as_paragraph(
@@
             part = "body", inline = TRUE) %>%
+    hline(i = 9, border = std_border) %>%
 

### 2020-09-22 Figures/Figure1.R

@@ -28,8 +28,6 @@ source("./R/libs.R")
 
 control_color <- "white"
 
-const_color <- "#56B4E9"
-var_color <- "#009E73"
 experiment_color <- "#4abdcb"
 
 
@@ -66,7 +64,7 @@ design_exp <- expand.grid(cond = c("Variable volume", "Constant volume"),
   
   scale_y_continuous(breaks = c(0, 3, 6, 9, 12), limits = c(0,12), expand = c(0,0), 
                      labels = c("", "3", "6", "9", "")) +
-  scale_fill_manual(values = c(color.scale[2], color.scale[3])) +
+  scale_fill_manual(values = c(color.scale[2], color.scale[1])) +
 
   annotate("text", x = c(0, 0, -2), y = c(9.1, 8.2, 6), 
            label = c("Muscle biopsy", "Muscle thickness", "Strength tests"), 
@@ -245,7 +243,7 @@ training_load <- read_excel("./data/tr010_training.xlsx", sheet = 1, na = "NA") 
   geom_line(data = training_sum, 
             aes(session, total.load, group = cond, color = cond), 
             size = 1.5) +
-  scale_color_manual(values = c(color.scale[2], color.scale[3], color.scale[4])) +
+  scale_color_manual(values = c(color.scale[1], color.scale[2], color.scale[4])) +
   geom_text(data = data.frame(session = c(1, 1),
                               total.load = c(5000, 5000), 
                               

### 2020-09-22 Figures/Figure2.R

@@ -182,6 +182,8 @@ mrna_vs_ctrl <- ctrl_vs_int %>%
          Normalization = if_else(model == "total_rna", "Total RNA", "Muscle weight"), 
          Normalization = factor(Normalization, levels = c("Total RNA", "Muscle weight"))) %>%
   
+  filter(Normalization != "Muscle weight") %>%
+  
   ggplot(aes(contrast, estimate, shape = Normalization, color = robust)) + 
   
   geom_hline(yintercept = 1, lty = 2, color = "gray50") +
@@ -265,18 +267,21 @@ reps_west <-  ggdraw() + draw_image("./data/wetlab/western/representative_img/ro
 
 
 figure2 <- plot_grid(
-   plot_grid(mrna_vs_ctrl, volume_diff_mrna, ncol = 2),
+
     
   plot_grid(prot_fold_change, prot_interaction, 
                      ncol = 2, rel_widths = c(0.5, 0.5)), 
   reps_west,
-  ncol = 1, rel_heights = c(0.75/2, 0.75/2, 0.25)) +
+  
+  plot_grid(mrna_vs_ctrl, volume_diff_mrna, ncol = 2),
+  
+  ncol = 1, rel_heights = c(0.75/2, 0.25, 0.75/2)) +
   
 
   
   draw_plot_label(label=c("A",  "B", "C", "D", "E"),
-                  x =   c(0.02, 0.51, 0.02, 0.51, 0.02), 
-                  y =   c(0.98, 0.98, 0.6, 0.6, 0.25),
+                  x =   c(0.02, 0.51, 0.08, 0.02, 0.51), 
+                  y =   c(0.98, 0.98, 0.6, 0.35, 0.35),
                   hjust=.5, vjust=.5, size = label.size)
 
 

### 2020-09-22 Figures/Figure3.R

@@ -574,8 +574,8 @@ rna_tc_fig <-  cond_eff_rna_tc %>%
                       breaks = c(250, 300, 350, 400, 450, 500, 550, 600, 650), 
                       labels = c("",  300, "",  400, "",  500, "" , 600, "")) +
   
-  scale_color_manual(values = c(color.scale[1], color.scale[2]), guide = NULL) +
-  scale_fill_manual(values = c(color.scale[1], color.scale[2])) +
+  scale_color_manual(values = c(color.scale[2], color.scale[1]), guide = NULL) +
+  scale_fill_manual(values = c(color.scale[2], color.scale[1])) +
    
    plot_theme() +
    theme(legend.position = c(0.7, 0.15), 

### 2020-09-22 R/ubf-tot-rna-model.R

@@ -99,11 +99,14 @@ west_cont <- western_data %>%
 
 combined_df  <- west_cont %>%
   group_by(target, participant, cond, time) %>%
-  summarise(ubf.scaled = mean(expression.scaled, na.rm = TRUE), 
-            ubf = mean(expression)) %>%
-  filter(target != "t-s6") %>%
+  summarise(scaled = mean(expression.scaled, na.rm = TRUE), 
+            raw = mean(expression)) %>%
+  mutate(target = if_else(target == "t-s6", "rps6", "ubf")) %>%
+  
+  pivot_wider(names_from = target, values_from = c(scaled, raw)) %>%
+
   ungroup() %>%
-  dplyr::select(participant, cond, time, ubf.scaled, ubf) %>%
+  dplyr::select(participant, cond, time, scaled_rps6:raw_ubf) %>%
   inner_join(rna_complete %>%
                mutate(rna.mg = rna/tissue_weight) %>%
                group_by(participant, leg, time,time.c, time1, time2, time3, cond) %>%
@@ -129,8 +132,8 @@ saveRDS(combined_df, "./data/derivedData/ubf-tot-rna-model/combined_df.RDS")
 # Preliminary model did not reveal any interaction effect between UBF and time or detraining
 # meaning that the effect of UBF levels are similar over the course of the study. 
 
-
-comb.m1 <- brm(bf(log.rna ~ ubf.scaled + (time1 +time2 + time3) + detrain + (1|participant/leg)),
+### The ubf model
+comb.m1 <- brm(bf(log.rna ~ scaled_ubf + (time1 +time2 + time3) + detrain + (1|participant/leg)),
               
               warmup = 1000, # number of samples before sampling
               iter = 5000,  # number of mcmc iterations 
@@ -139,6 +142,21 @@ comb.m1 <- brm(bf(log.rna ~ ubf.scaled + (time1 +time2 + time3) + detrain + (1|p
               seed = 5, # seed for reproducible results
               #    control = list(adapt_delta = 0.95), 
               data = combined_df)
+
+#### The rpS6 model
+
+
+comb.s6.m1 <- brm(bf(log.rna ~ scaled_rps6 + (time1 +time2 + time3) + detrain + (1|participant/leg)),
+               
+               warmup = 1000, # number of samples before sampling
+               iter = 5000,  # number of mcmc iterations 
[…]

### 2020-09-22 manuscript.Rmd
updates to discussion and figures

@@
 with resistance training (RT) being the advocated remedy.  
-Such training promotes changes in the balance between muscular protein breakdown and synthesis, thereby leading to protein accretion and muscle hypertrophy, alongside a multitude of functional and health-related benefits.
-Exercise-induced muscle hypertrophy is traditionally viewed as the result of repeated episodes of elevated post-exercise protein synthesis, leading to net positive protein balance over time.
+RT promotes changes in the balance between muscular protein breakdown and synthesis through repeated episodes of elevated post-exercise protein synthesis, leading to net positive protein balance over time.
 Indeed, one bout of resistance exercise acutely increases skeletal muscle protein synthesis for up to 48 hrs after the exercise [@RN786],<!-- Phillips 1997 -->
@@
 @RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->.
-This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, occuring alongside muscle hypertrophy
+This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, occuring in proportion to muscle hypertrophy
 [@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
@@
 As such, Bickel *et al*.
-showed that two consecutive bouts of hypertrophic stimuli were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout [@RN1520]<!-- Total RNA increase 72 h after a second bout -->.
-Brook *et al.*, reported peak values after nine sessions followed by a slight decrease to after 18 sessions
+showed that two consecutive bouts of electrically induced muscle contractions were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout
+[@RN1520]<!-- Total RNA increase 72 h after a second bout -->. 
+Brook *et al.*, reported peak values after nine sessions followed by a slight decrease to after 18 sessions of RT
 [@RN1809],
-resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), whereupon a numerical lowering occurred to after the last training session of the 12 wk interventions (31 session).
+resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), whereupon a numerical lowering occurred to after the last training session of the 12 wk interventions (31 sessions).
 [@RN2358].
-Interestingly, total RNA accumulation in the initial phase was shown to be training volume sensitive as three sets per exercise in leg exercises led to increased total RNA and rRNA accumulation compared to a single set per exercise, coinciding with differences in muscle hypertrophy after week 12.
+Interestingly, total RNA accumulation in the initial phase was shown to be training volume sensitive as three sets per exercise in leg exercises led to increased total RNA and rRNA accumulation compared to a single set per exercise, coinciding with differences in muscle hypertrophy after 12 weeks of RT.
 [@RN2358].
@@
 slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,2]) - 1)),
-                           "%, [",
+                           "% [",
                    sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,4]) - 1)), 
@@
 slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,2]) - 1)),
-                           "%, [",
+                           "% [",
                    sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,4]) - 1)), 
@@
 slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,2]) - 1)),
-                           "%, [",
+                           "% [",
                    sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,4]) - 1)), 
@@
                    "]")
+
+# Estimate changes to end of periods
+
+sessions <- hypothesis(tc.m2, c("time1 * 4 = 0", 
+                                     "(time1 * 8) + (time2 * 4)= 0", 
+                                     "(time1 * 12) + (time2 * 8) + (time3 * 4)= 0"))
 
@@
 
+fourth.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,5]) - 1)),
+                   "]")
+
+eight.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,5]) - 1)),
+                   "]")
+
+twelth.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,5]) - 1)),
+                   "]")
+
 ```
@@
 This general pattern of rRNA expression was reflected by total RNA expression per unit tissue weight, which were robustly increase in TRAIN compared to CTRL after session twelve (Figure 3E and F), followed by decreased levels after de-training `r detrain_rna`.
-For both rRNA expression and total RNA levels, these training-associated increases in abundance predominately occured during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from session four to eight `r slope4_8` and from session eight to twelve `r slope8_12`.
+For both rRNA expression and total RNA levels, these training-associated increases in abundance predominately occured during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from sessions four to eight `r slope4_8` and from sessions eight to twelve `r slope8_12`.
+This corresponded to an increase from baseline by `r fourth.session`, `r eight.session` and `r twelth.session` to 48 hrs after session four, eight and twelve, respectively.
 In TRAIN, the two volume conditions were associated with minor differences in rRNA subspecies abundance, with only 45S ETS showing differential expression between conditions, evident as robustly higher levels in VAR compared to CONST after the 12th session (Figure 3D), coinciding with increased volume in this condition towards the end of the intervention.
@@
 This analysis indicated that the relationship between Total RNA increase and muscle growth was preserved in all iterations (Figure 4D).
-We further assessed the model by iteratively removing one participant from the data set, similarly this showed that estimates of the total RNA increase effect on muscle growth was robust but the effect of average total RNA estimates were more variable (Participant 8 and 3 in Figure 4D).
+We further assessed the model by iteratively removing one participant from the data set, similarly this showed that estimates of the effect of total RNA increase on muscle growth was robust but the effect of average total RNA estimates were more variable (Participant 8 and 3 in Figure 4D).
 
 # Discussion
-
- hypothesize that (1) ribosome accumulation occurs during the early phase (3-4 weeks) of RT, within which this accumulation (2) reaches a plateau when RT volume is kept constant, (3) displays fluctuations in response to fluctuating training volume and (4) is partially reversed to one week after cessation of RT.
-In addition to addressing these hypotheses we aimed to relate RNA accumulation to total UBF levels and muscle growth.
-
 Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
@@
 Several studies have presented evidence of a determinant role for ribosomal biogenesis in RT induced muscle hypertrophy.
-Increases in total RNA has been shown to positively correlated with training induced hypertrophy
+Increases in total RNA has been shown to positively correlate with RT induced hypertrophy
 [@RN1644;<!-- Figueriedo  -->
@@
 Results from the present study corroborates the link between ribosomal biogenesis and muscle hypertrophy.
-However, when using all available data points we were able to robustly estimate both the rate of increase in total RNA and the average RNA levels over the course of the study and include both when modeling changes in muscle thickness.
-The results of this analysis showed that the rate, but not the absolute levels of total RNA predicted muscle growth.
+When using all available data points we were able to robustly estimate both the rate of increase in total RNA and the average RNA levels over the course of the study and include both when modeling changes in muscle thickness.
+The results of this analysis showed that the rate of RNA accumulation determined muscle growth when average levels of total RNA was controlled for.
 In other words, at a fixed relative rate of increase in total RNA, legs with higher levels of RNA did not grow more but at a fixed amount of total RNA, legs with higher increase per session did.
 This suggest that the absolute ribosomal density is not as important as the rate of increase in ribosomal density.
-In an homogeneous group, in response to RT, the rate of increase will be highly correlated with maximal values seen after RT, linking maximal rate to the resulting abundance.
-However, the same pattern revealed in our model is seen when comparing young and old muscle, where aged muscle display higher levels of total RNA at rest
+In an homogeneous group, in response to RT, the rate of increase will be highly correlated with maximal values seen after RT, linking maximal rate to the resulting abundance of ribosomes.
+The same pattern revealed in our model is seen when comparing young and old muscle, where aged muscle display higher levels of total RNA at rest
 [@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
@@
 [@RN1809].<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
+Together with our results this underlines the importance of understanding RT induced ribosomal synthesis in relation to manipulation of training loads for specific populations.
 
-Both training protocols utilized in the present study increased strength and induced muscle hypertrophy in the relative short time-span studied.
-This is in line with previous data utilizing similar protocols
-[@RN1642].<!-- Brook 2015 young only, strength and muscle thickness increased over control legs -->
-Fluctuations in training volume did not influence training outcomes, although volume is an important factor for increases in  muscle strength and mass
-[@RN1767;<!-- schoenfeld 2016 dose response  -->
-@RN2492]<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
-different organization of training loads over time is likely of minor importance when training volumes are equated
-[@RN2572].<!-- Grgic 2017 volume equated periodization schemes led to similar muscle hyeprtrophy -->
-Resistance training volume, expressed in human exercise studies as the number of sets performed per session and muscle group is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
-[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
-@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
-@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
-total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
-and subsequent training outcomes [@RN1767; @RN2358].<!-- Schoenfeld 2016 meta and Hammarström -->
-We have previously shown different amounts of training volume in constant volume protocols leads to differences in total RNA and rRNA accumulation in the initial phase of RT [@RN2358].
-Based on the notion that ribosomal biogenesis is a complex and tightly regulated, energy demanding process
-[@RN1920],<!-- Warner economics of ribosomal biogenesis -->
-we hypnotized that fluctuations in training load would reflect in markers of ribosomal biosynthesis reflecting the current need.
-Comparing the two volume conditions in the present study we only found one part of the 45S ETS to be differentially expressed at Session 12 in favor of VAR together with a tendency of a rescued UBF level after de-training in response to increased volume in the VAR protocol.
-These observations do not give much support for a clear effect of fluctuations in training volume in the short time span used in the present study although the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume.
-It is important to note that RT in the current study was performed with constant volume in the first four sessions, something that could have maximized rRNA transcription in the initial phase.
-This is supported by the observation that pre-rRNA increased rapidly initially in both protocols with minimal changes in response to subsequent sessions, regardless of exercise volume.
-The CONST protocol in the present study corresponded to volumes used in the high volume condition in a previous study from our lab (three sets in two exercises activation knee extensor muscles) [@RN2358].
-Compared to a low volume protocol, this led to higher levels of total RNA accumulation [@RN2358].
-Interestingly, using a progressive volume protocol in well-trained participants, increases in total RNA were seen throughout six weeks of training
-[@RN2055].<!-- Mobley 2018 trend towards higher increases in high responders 32% high resp, 8% low resp 12 wk training -->
-Compared to studies using constant volume protocols
-[@RN1809;<!-- Brook 2016 -->
-@RN2358],<!-- Hammarström 2020 -->
-progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
-This observation was done in well-trained participants performing a high volume protocol without a control group condition with constant volume.
+<!-- rpS6 dissconect from total RNA levels. -->
+In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6.
+Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period.
+Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
+[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
+but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
+[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
+Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for.
+Together with a disconnect after the de-training period, this may suggest that regulation of rpS6 expression and transcription of ribosomal RNA displays different temporal characteristics resulting from RT.
+Additionally, ribosomal proteins may have extra-ribosomal functions
+affecting their expression
+[@RN2588].<!-- Warner 2009 General description of   -->
 
-A main objective in the present study was to establish a time course for training induced ribosomal biogenesis.
-Previous data from human studies indicate that mechanical loading leads to increased total RNA concentrations
+Total RNA is the most studied proxy marker of ribosomal density as most of the RNA is assumed to be ribosomal [@RN2223].<!-- Young  -->
+This marker has been shown to be sensitive to RT in numerous studies 
 <!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
@@
 @RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
-@RN1897<!-- 11% 36 sessions young -->], indicating expansion of the ribosomal pool as the majority of RNA is ribosomal
-[@RN2223].<!-- Young 1970  -->  
-Time course data to date have suggests that maximal levels are reached within four to nine sessions in young males and females
+@RN1897<!-- 11% 36 sessions young -->].
+However, a detailed time course has not been mapped.
+To date, available data suggests that maximal levels are reached within four to nine sessions in young males and females
 [@RN1809;
 @RN2358].
-This was confirmed in the present study and
+We did not detect any meaningful further increases after 48 hrs after the eight session were peack the peack average increase was 50% from baseline.
+This confirmed a plateau when total RNA was expressed per unit tissue weight.
+A possible reason for the plateau could be attenuated ribosomal biogenesis.
+However, pre-rRNA abundances per unit tissue weight remained at elevated levels indicating retained transcriptional activity after the twelfth session together with maximal levels of UBF protein.
+Another possibility is a dilution effect
+[@RN2142]<!-- Figueiredo review details the concept of rna concentration -->
+due to muscle hypertrophy and thus protein accretion.
+After the twelfth session, muscle hypertrophy was apparent and did not change in response to the de-training period.
+As muscle hypertrophy did not change from after session twelve to after the de-training period, the apparent decrease in total RNA was not driven by dilution but by attenuated rRNA transcription.
+This notion is supported by normalization of pre-rRNA levels and possibly by lowered levels of UBF protein, although this apparent effect was not confirmed as statistically robust.
 
-We first set out to confirmed that markers of ribosome density were actually training induced and could do so as total RNA, ribosomal RNA subspecies and rpS6 mRNA and protein increased in the training group compared to the non-training control group.
-Total RNA has been used in many previous studies as an approximation of ribosomal density as the majority of total RNA is ribosomal [@RN2142]<!-- Figueriedo --> and total RNA concentrations predicts protein synthesis [@RN2145]<!-- Millward 1973 -->.
-Previous studies have indicated that the training induced rise in total RNA reaches a peak after which a plateau or decrease is observed
-[@RN1809;<!-- brook 2016 -->
-@RN2358].<!-- Hammarström -->
-In the present study we observed a rapid increase in the first four sessions after which a plateau was observed.
-The observed plateau could be a consequence of the unit of measure as total RNA is estimated per unit tissue weight and that ribosome biogenesis occurs simultaneously as protein accretion leading to a dilution effect [@RN2142]<!-- Figueriedo  -->.
-We could confirm muscle hypertrophy both after session 12 and after a period of de-training indicating that the dilution effect is a reasonable assumption in interpreting the observed plateau.
-Further supporting this view is UBF increased over the whole training period indicating that the transcriptional apparatus continued to increase over the whole course of the training period.
-
-<!-- rpS6 dissconect from total RNA levels. -->
-In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6.
-Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period.
-Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
-[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
-but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
-[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
-Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for (Table 3).
-Together with a disconnect after the de-training period, this may suggest that regulation of rpS6 expression and transcription of ribosomal RNA displays different temporal characteristics resulting from RT.
-Additionally, ribosomal proteins may have extra-ribosomal functions
-affecting their expression
-[@RN2588].<!-- Warner 2009 General description of   -->
-
+Over the full training period UBF levels did robustly explain total RNA levels after accounting for time, something that otherwise would lead to biased estimates as both the dependent variable (total RNA) and the covariate (UBF levels) varies with time.
 From a mechanistic perspective, UBF is an important factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA polymerase I
 [@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
-
-Activation of UBF is thought to be controlled by signaling from the mechanosensitive mTOR, pathway as rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
+Activation of UBF is controlled by signaling from the mechanosensitive mTOR, pathway as rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
 [@RN1632; <!-- Nader rapamycin block UBF availability  -->
@@
 The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF -->
-Evidence from human exercise studies confirms training associated activation of UBF through phosphorylation
+Evidence from human exercise studies confirms training-induced activation of UBF through phosphorylation
 [@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
@@
 [@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->
-In the present study, we measured total UBF levels and observed that the relative amount of UBF was related to total RNA concentrations over the whole course of the study indicating its role in ribosomal biogenesis.
-Together the above suggests that rRNA transcription may not plateau during short term training but muscle hyertrophy leads to a dilution effect.
-This observation has consequences for the use of total RNA per unit tissue weight as a determinant muscle growth as hypertrophy itself affects the measurement.
+Together with our observations, this underlines the importance of UBF as a regulator of RT-induced ribosomal biogenesis.
 
+Resistance training volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
+[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
+@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
+@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
+subsequent total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
+and subsequent training outcomes [@RN1767; @RN2358].<!-- Schoenfeld 2016 meta and Hammarström -->
+We have previously shown that different amounts of training volume in constant volume protocols leads to differences in total RNA and rRNA accumulation in the initial phase of RT [@RN2358].
+Based on the notion that ribosomal biogenesis is a complex and energy demanding process
+[@RN1920],<!-- Warner economics of ribosomal biogenesis -->
+we hypnotized that fluctuations in training load would reflect in markers of ribosomal biogenesis.
+When comparing the two volume conditions in the present study we only found one part of the 45S ETS to be differentially expressed at Session 12 in favor of VAR together with a tendency of a rescued UBF level after de-training in response to increased volume in the VAR protocol.
+These observations do not give much support for a clear effect of fluctuations in training volume in the short time span used in the present study although the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume.
+Indeed, both training protocols utilized in the present study also increased strength and induced muscle hypertrophy to a similar degree.
+From a general perspective, although volume is an important factor for increases in muscle strength and mass
+[@RN1767;<!-- schoenfeld 2016 dose response  -->
+@RN2492]<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
+different organization of training loads over time is likely of minor importance when training volumes are equated
+[@RN2572].<!-- Grgic 2017 volume equated periodization schemes led to similar muscle hyeprtrophy -->
+It is important to note that RT in the current study was performed with constant volume in the first four sessions, something that could have been more than enough to maximize rRNA transcription in previously untrained individuals.
+This is supported by the observation that pre-rRNA increased rapidly initially in both protocols with minimal changes in response to subsequent sessions, regardless of exercise volume.
+The CONST protocol in the present study corresponded to volumes used in the high volume condition in a previous study from our lab (three sets in two exercises activating knee extensor muscles) [@RN2358].
+There, higher levels of total RNA accumulation were observed compared to a low volume protocol [@RN2358].
+Interestingly, using a progressive volume protocol in well-trained participants, increases in total RNA have been reported throughout six weeks of training
+[@RN2055].<!-- Mobley 2018 trend towards higher increases in high responders 32% high resp, 8% low resp 12 wk training -->
+Compared to the present study and other using constant volume protocols
+[@RN1809;<!-- Brook 2016 -->
+@RN2358],<!-- Hammarström 2020 -->
+progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
+This observation was done in well-trained participants performing a high volume protocol without a control group condition with constant volume.
+
+In conclusion, RT induced ribosome accumulation reaches peak values in the initial phase of RT (8-12 sessions) affected by total levels of UBF.
+Fluctuations in training volume does not transfer to fluctuations in ribosomal biogenesis a moderate volume initiation of RT but training cessation does lead to attenuated ribosomal biogenesis.
+The rate of total RNA accumulation predicts RT induced muscle hypertrophy.
+
+<!-- NOTES -->
 <!-- UBF and S6 protein levels may be differently regulated in men and rats as more consecutive bouts of electrical stimulation in rats does not prtoduce the same pattern seen in our study 
@@
 **Figure 4.** 
+Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 4).
+Values are averaged over values from men and women.
+Individual plots of estimates total RNA increases over time is shown in C together with results from leave-one-out analysis (D).
+Leave-one-out analysis shows the effect of removing a single participant (black point and error-bars) and individual values from the total RNA per time estimates where green points represents bounds of the 95% CI and yellow points represents mean estimates.

### 2020-09-22 Figures/Figure4.R

@@ -23,6 +23,7 @@
 
 source("./R/figure-source.R")
 library(brms)
+library(readxl)
 
 
 #### Figure 4B Leave one out analysis #########################
@@ -31,7 +32,11 @@ library(brms)
 
 loo.results <- readRDS("./data/derivedData/ubf-tot-rna-model/leave-one-out.RDS")
 
+leg <- read_excel("./data/leg_randomization.xlsx")
+
+
 res <- loo.results$loo.sample %>%
+  inner_join(leg) %>%
   
   mutate(Participant = factor(participant, 
                               levels = paste0("P", c(seq(1:7), 19, 21, 22, 23)), 
@@ -43,7 +48,7 @@ res <- loo.results$loo.sample %>%
   print()
   
 res2 <- loo.results$loo.participant %>%
-  
+  inner_join(leg) %>%
   mutate(Participant = factor(participant, 
                               levels = paste0("P", c(seq(1:7), 19, 21, 22, 23)), 
                               labels = paste0("P", seq(1:11))), 
@@ -70,18 +75,18 @@ loo_panel <- res %>%
   
   geom_vline(xintercept = 0, lty = 2, color = "gray50") +
     
-  geom_density_ridges(color = "white") +
+#  geom_density_ridges(color = "white") +
+  
+  geom_point(position = position_nudge(y = 0.2), shape = 21, alpha = 0.4) +
+  
   facet_wrap( ~ coef, scales = "free") +
   
-
-  
-  
-  
+  scale_fill_manual(values = c(color.scale[5], color.scale[4], color.scale[4])) +
 
   geom_point(data = res2, aes(estimate, Participant, fill = NULL)) +
   geom_errorbarh(data = res2, aes(fill = NULL, x = estimate, y = Participant, 
                                   xmax =  upr, xmin = lwr), 
-                 height = 0.5) + 
+                 height = 0) + 
  
    labs(x = "Estimated change in muscle thickness (mm)\nper one unit change in predictor", 
         y = "Participant") +
@@ -109,8 +114,13 @@ mean_diff <- (effects_m1$sex[2, 9] - effects_m1$sex[1, 9]) / 2
 
 
 prediction_slope <- predict_df %>%
+  
+  inner_join(leg) %>%
+  mutate(cond = factor(cond, levels = c("const", "var"), 
+                       labels = c("Constant volume", 
+                                  "Variable volume")))  %>%
 
-  group_by(participant, leg) %>%
+  group_by(participant, leg, cond) %>%
[…]

### 2020-09-22 R/total-rna-analysis.R

@@ -192,10 +192,7 @@ time_course <- rna_complete %>%
   print()
   
   
-  
-  print()
  
-
 
 
 # gam models fitted in brms to visualize general patterns in the data. 
@@ -244,18 +241,23 @@ saveRDS(tc.m1, "./data/derivedData/total-rna-analysis/tc.m1.RDS")
 pp_check(tc.m2, type = "ecdf_overlay")
 summary(tc.m2)
 
+### A model for "descriptive" analysis of avaiable data points
 
-# Get predictions from model 1 (gam model)
 
-ce.m1 <- conditional_effects(tc.m1)
 
-cond_eff_rna_tc <- ce.m1$`time.c:cond` %>%
-  dplyr::select(time.c, cond, estimate = estimate__, lower = lower__, upper = upper__) %>%
-  mutate(estimate = exp(estimate), 
-         lower = exp(lower),
-         upper = exp(upper)) %>%
-  print()
-  
+tc.m3 <- brm(bf(log(rna.tissue) ~ (time1 + time2 + time3) + detrain + (1|participant)), 
+             data = time_course, 
+             warmup = 1000, # number of samples before sampling
+             iter = 4000,  # number of mcmc iterations 
+             cores = 4, # number of cores used in sampling
+             chains = 4, # number of chains
+             seed = 123, # seed for reproducible results
+             save_all_pars = TRUE, 
+             control = list(adapt_delta = 0.99))
+
+
+
+
 
 saveRDS(cond_eff_rna_tc, "./data/derivedData/total-rna-analysis/cond_eff_rna_tc.RDS")
 

### 2020-09-23 manuscript.Rmd
finnished draft

@@
 
-<div custom-style="runningtitle_style">**Running title:** Ribosome biogenesis in early phase resistace training</div>
+<div custom-style="runningtitle_style">**Running title:** Ribosome biogenesis in early phase resistance training</div>
 
@@
 <div custom-style="abstract_style">
-<!-- Word limit 200 -->
-Abstract...
+
+Increased ribosomal density is a phenomena observed in response to resistance training (RT) supporting elevated.
+A time course of RT-induced accumulation of ribosomes has not been described.
+We mapped markers of ribosomal biogenesis (total RNA, ribosomal RNA (rRNA), ribosomal protein S6 (rpS6) and upstream binding factor (UBF) protein content) in response to twelve RT sessions.
+Additionally we investigated the effect of fluctuations in training volume on these markers and relationship between RNA accumulation and muscle growth measured as muscle thickness (ultra sound).
+Eighteen participants were allocated to either a training (TRAIN, *n*=11) or control (CTRL, *n*=7) group.
+In TRAIN, participants performed unilateral knee extension with either constant (CONST, 6 sets) or a variable volume (VAR, 6, 3 and 9 sets in sessions 1-4, 5-8 and 9-12, respectively).
+Muscle biopsies were sampled from *m vastus lateralis* in TRAIN before and 48 hrs after the first session as well as 48 hrs after session 4, 5, 8, 9 and 12 and after eight days of de-training.
+Control biopsies were sampled in CTRL at baseline and after 48 hrs and 3-5 weeks.
+RT led to muscle growth in TRAIN compared to CTRL and concomitant increases in total RNA, rRNA, including precursor rRNA as well as UBF and rpS6 protein through gradual increase throughout the training period, confirmed in comparison to CTRL.
+Total RNA increased in a curve-linear fashion, most rapidly in response to the first four sessions  (8.6%, 95% CI: [5.5, 11.7] per session), followed by a plateau and peak values (49.5% [34.2, 66.5] above baseline) after Session 8.
+UBF protein levels explained total RNA levels after controlling for time and increases in total RNA levels predicted RT induced muscle hypertrophy.
+After de-training, total RNA and specific rRNA species decreased without changes in muscle mass indicating reduced concentrations and biosynthesis of ribosomes.
+These result underlines a determinant role for ribosomal biogenesis in RT-induced muscle hypertrophy and that ribosomal biogenesis is sensitive to training cessation.
 
@@
 with resistance training (RT) being the advocated remedy.  
-RT promotes changes in the balance between muscular protein breakdown and synthesis through repeated episodes of elevated post-exercise protein synthesis, leading to net positive protein balance over time.
-Indeed, one bout of resistance exercise acutely increases skeletal muscle protein synthesis for up to 48 hrs after the exercise [@RN786],<!-- Phillips 1997 -->
+Prolonged RT leads to changes in the balance between muscle protein breakdown and synthesis, with one bout of resistance exercise acutely increasing protein synthesis for up to 48 hrs after exercise [@RN786],<!-- Phillips 1997 -->
 with subsequent repeated bouts leading to accumulation of muscle protein over time
@@
 @RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->.
-This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, occuring in proportion to muscle hypertrophy
+This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, which is closely connected to protein synthesis
+[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
+@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
+and muscle hypertrophy
 [@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
 @RN1755; <!-- Stec Ribosome bio higher in high responders -->
-@RN2358]<!-- Volume dependent RNA accum relates to volume depend hyper -->
-and the close relationship between RNA abundance and protein synthesis
-[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
-@RN2145].<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
+@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->
 Conversely, inhibition of ribosomal RNA (rRNA) transcription and inhibition of its up-stream transcription factors act to diminish muscle cell growth
@@
 
-Biosynthesis of novel ribosomes is a complex, highly coordinated and energy demanding process that involves synthesis of both proteins and four different mature rRNA
+Biosynthesis of novel ribosomes is a complex, highly coordinated and energy demanding process that involves synthesis of both ribosomal proteins and the four mature rRNA transcripts
 [@RN1920;<!-- Warner review yeast -->
@@
 @RN1810]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
-This process is limited at the level of pre-rRNA transcription by RNA polymerase I (Pol I) which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
+Ribosomal accumulation is believed to be determined by the rates of pre-rRNA transcription by RNA polymerase I (Pol I), which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
 [@RN1820]<!-- Moss review -->.
-Specifically, phosphorylation of the upstream bindning factor (UBF) is needed to initiate transcription
+Specifically, activation of the of the upstream binding factor (UBF) through phosphorylation is needed to initiate transcription
 [@RN2582; <!-- Tuan UBF phosphorylation important for SL1-UBF interaction -->
-@RN2563]<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
-and this is related to signaling through the mechanosensitive mTOR pathway as its inhibition blocks UBF phosphorylation and subsequent rRNA transcription
+@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
+Such activation is at least partly controlled by the mechanosensitive mTOR pathway, with its inhibition being associated with blocked UBF phosphorylation and subsequent rRNA transcription
 [@RN1632; <!-- Nader rapamycin block UBF availability  -->
 @RN2564].<!-- Hannan mTOR  -->
-Interestingly the availability of UBF *per se* has been shown to also regulate rRNA transcription
+Interestingly the availability of UBF *per se* has been shown to be a determinant of rRNA transcription
 [@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->
@@
 @RN1897<!-- 11% 36 sessions young -->].
-However, the true time course of ribosomal transcription and accumulation in response to RT remains largely unstudied, with only a few studies having investigated exercise-induced changes in rRNA over multiple time-points.
-As such, Bickel *et al*.
-showed that two consecutive bouts of electrically induced muscle contractions were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout
+However, the true time course of ribosomal transcription and accumulation in response to RT remains largely unstudied, with a mere few studies having investigated exercise-induced changes in rRNA over multiple time-points, all of which are either limited to a selected few time-points or a limited time frame.
+For example, two consecutive bouts of electrically evoked muscle contractions were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout
 [@RN1520]<!-- Total RNA increase 72 h after a second bout -->. 
-Brook *et al.*, reported peak values after nine sessions followed by a slight decrease to after 18 sessions of RT
-[@RN1809],
-resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), whereupon a numerical lowering occurred to after the last training session of the 12 wk interventions (31 sessions).
+Using voluntary contractions, peak values were reported after nine sessions, followed by a slight decrease to after 18 sessions [@RN1809],
+resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), whereby a numerical lowering occurred to after the last training session of the 12 wk interventions (31 sessions).
 [@RN2358].
-Interestingly, total RNA accumulation in the initial phase was shown to be training volume sensitive as three sets per exercise in leg exercises led to increased total RNA and rRNA accumulation compared to a single set per exercise, coinciding with differences in muscle hypertrophy after 12 weeks of RT.
+Interestingly, during the initial phase of RT, total RNA accumulation seems to be volume-dependent, as three sets per exercise in leg exercises led to augmented total RNA and rRNA levels compared to one set per exercise, coinciding with the differences in muscle hypertrophy seen after 12 weeks of RT.
 [@RN2358].
@@
 load.inr.w3 <- paste0(sprintf("%.0f",h[2, 2]), 
-                      "%, [",
+                      "% [",
                       sprintf("%.0f",h[2, 4]), 
@@
 
+# Approximate percentages increase in strength and muscle mass
+
+m1.isom <- readRDS("./data/derivedData/strength-bayes/m1.isom.RDS") 
+m1.isok <- readRDS("./data/derivedData/strength-bayes/m1.isok.RDS")
+
+# Change in percentage-points between groups 
+isok.h <- hypothesis(m1.isok, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"
+                                ))$hypothesis
+
+isom.h <- hypothesis(m1.isom, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"
+                                ))$hypothesis
+
+isok_pre_post <-  paste0(sprintf("%.1f",isok.h[1, 2]), 
+                      "% [",
+                      sprintf("%.1f",isok.h[1, 4]), 
+                      ", ",
+                      sprintf("%.1f",isok.h[1, 5]), 
+                      "]")
+
+isok_pre_post_aprox <- paste0(sprintf("%.1f",isok.h[1, 2]))
+
+isok_pre_detrain <-  paste0(sprintf("%.1f",isok.h[2, 2]), 
+                      "% [",
+                      sprintf("%.1f",isok.h[2, 4]), 
+                      ", ",
+                      sprintf("%.1f",isok.h[2, 5]), 
+                      "]")
+
+isok_pre_detrain_approx <- paste0(sprintf("%.1f",isok.h[2, 2]))
+
+isom_pre_post <-  paste0(sprintf("%.1f",isom.h[1, 2]), 
+                      "% [",
+                      sprintf("%.1f",isom.h[1, 4]), 
+                      ", ",
+                      sprintf("%.1f",isom.h[1, 5]), 
+                      "]")
+isom_pre_detrain <-  paste0(sprintf("%.1f",isom.h[2, 2]), 
+                      "% [",
+                      sprintf("%.1f",isom.h[2, 4]), 
+                      ", ",
+                      sprintf("%.1f",isom.h[2, 5]), 
+                      "]")
+
+isom_pre_post_approx <- paste0(sprintf("%.1f",isom.h[1, 2]))
+isom_pre_detrain_approx <- paste0(sprintf("%.1f",isom.h[2, 2]))
+
+## Muscle thickness
+
+m1.us <- readRDS("./data/derivedData/us-freq-bayes/m1us.RDS")
+
+us.h <- hypothesis(m1.us, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"))$hypothesis
+
+us_pre_post_approx <- paste0(sprintf("%.1f",us.h[1, 2]))
+us_pre_detrain_approx <- paste0(sprintf("%.1f",us.h[2, 2]))
+
 ```
@@
 In TRAIN, all participants successfully completed their prescribed RT on both legs, with the two volume conditions resulting in diverging volume profiles (load &#215; repetitions) over the course of the study (Figure 1B).
-Exercise intensities (resistance at 10RM) increased similarly between conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with each block consisting of four training sessions.
-Isokinetic strength increased from baseline to after Session 12 in TRAIN compared to CTRL.
-This difference was still seen after de-training (Figure 1C).
-Isometric strength showed the same general pattern, though with considerably larger degrees of uncertainty, as indicated by  wider 95% CI (Figure 1C).
-Muscle thickness of *m.
-vastus lateralis* increased in TRAIN compared CTRL, measured both after Session 12 and after eight days of de-training (Figure 1D).
-There were no indications of differences between volume conditions in strength or muscle thickness changes.
+Exercise intensities (resistance at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with each block consisting of four training sessions.
+Concomitantly, in TRAIN, isokinetic strength and thickness of *m.
+vastus lateralis* increased from baseline to after Session 12 compared to CTRL (isokinetic strength $\sim$ `r isok_pre_post_aprox`%-point difference; muscle thickness $\sim$ `r us_pre_post_approx`%-point difference, Figure 1C and D), a difference that was sustained to after eight days of de-training ($\sim$ `r isok_pre_detrain_approx`%-point and $\sim$ `r us_pre_detrain_approx`%-point difference in change in isokinetic strength and muscle thickness, respectively; Figure 1C and D).
+Isometric strength showed the same general pattern to after Session 12 ($\sim$ `r us_pre_detrain_approx`%-point difference), though with considerably larger degrees of uncertainty, as indicated by wider 95% CI normalization compared to CTRL after de-training ($\sim$ `r isom_pre_detrain_approx`%-point; Figure 1C).
+No differences were observed between volume conditions for either strength or muscle thickness.
 
@@
 
-UBF and rpS6 protein levels were increased only after twelve sessions in the training group compared to control and stayed elevated after eight days of rest (Figure 2A and B), with no changes being observed after the first training session (48 hrs).
-For both rpS6 and UBF, protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12) and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
-Slopes did not differ between volume-conditions but there was a tendency towards lower levels of UBF in VAR after session 12 (`r var_s12`).
-After the de-training period CONST tended to show decreased levels of UBF (`r const_detrain`) while VAR remained at elevated level compare to after session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
-Levels of rpS6 did not change in response to the de-training period and remained similar between volume conditions.
-UBF was robustly up-regulated at the mRNA level compared to CTRL after 48 hrs (Figure 2D) with no other robust differences seen between TRAIN and CTRL or between volume conditions in TRAIN in either UBF or rpS6 mRNA levels (Figure 2D and E).
+For both rpS6 and UBF, protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
+This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12 and elevated after eight days of rest (Figure 2A and B), however with no differences being observed after the first training session (48 hrs).
+Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`).
+After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
+For rpS6, de-training affect protein levels, which remained similar between volume conditions period and remained similar between volume conditions.
+At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point.
+No differences were observed between volume conditions for either transcripts (Figure 2D and E).
 
@@
 
-A single session of RT led to robust increases in pre-rRNA 47S and 45S abundance, measured as expression per unit tissue weight, both from baseline to 48 hrs after session 1 within TRAIN (Figure 3B) and compared to CTRL (Figure 3C).
-Similarly, 47S and 45S abundances were also increased after twelve sessions, together with 45S ITS and rRNA 28S and 18S species, measured both within-participant and compared to CTRL (Figure 3C).
-After eight days of rest, only 18S and 28S remained at elevated levels compared to CTRL (Figure 3B).
-This general pattern of rRNA expression was reflected by total RNA expression per unit tissue weight, which were robustly increase in TRAIN compared to CTRL after session twelve (Figure 3E and F), followed by decreased levels after de-training `r detrain_rna`.
-For both rRNA expression and total RNA levels, these training-associated increases in abundance predominately occured during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from sessions four to eight `r slope4_8` and from sessions eight to twelve `r slope8_12`.
-This corresponded to an increase from baseline by `r fourth.session`, `r eight.session` and `r twelth.session` to 48 hrs after session four, eight and twelve, respectively.
-In TRAIN, the two volume conditions were associated with minor differences in rRNA subspecies abundance, with only 45S ETS showing differential expression between conditions, evident as robustly higher levels in VAR compared to CONST after the 12th session (Figure 3D), coinciding with increased volume in this condition towards the end of the intervention.
-No robust differences were seen between conditions for total RNA abundances (Figure 3G), increasing similarly in both conditions.
+A single session of RT (Session 1) led to robust increases in pre-rRNA 47S ETS and 45S ETS abundance per unit tissue weight, measured as changes from baseline to 48 hrs after exercise within TRAIN (Figure 3B), as well as compared to CTRL (Figure 3C).
+After Session 1, pre-rRNA 47S ETS and 45S ETS levels remained at similar levels at all measured time-points in TRAIN (Figure 3D), confirmed in comparison to CTRL after Session 12 (Figure 3C).
+Other rRNA transcripts showed increases in response to training with slightly different temporal patterns with exception of rRNA 5S which did not change and rRNA 5.8 which tended to follow other mature transcript spliced from pre-rRNA 45S ETS, but without statistical robustness (Figure 3C and D).
+After eight days of rest, 18S and 28S remained at elevated levels compared to CTRL (Figure 3B).
+This general pattern of rRNA expression was reflected by total RNA expression per unit tissue weight, which increased robustly and steadily in TRAIN throughout the initial part of the intervention (Figure 3E and G), leading to robust increase compared to CTRL after Session 12 (Figure 3F), followed by decreased levels after de-training (`r detrain_rna`).
+For both rRNA expression and total RNA levels, the training-associated increases in abundances occurred predominately during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from sessions four to eight (`r slope4_8` increase per session) and from sessions eight to twelve `r slope8_12`, corresponding to `r fourth.session`, `r eight.session` and `r twelth.session` increases from baseline to 48 hrs after session four, eight and twelve, respectively.
+In TRAIN, the two volume conditions led to similar changes for most variables (Figure 3D and G), with 45S ETS abundance only showing differential expression, evident as robustly higher levels in VAR compared to CONST after the 12th session (Figure 3D), coinciding with the increased training volume towards the end of the intervention for this condition.
 
@@
 
-In TRAIN (after for controlling for time), UBF levels robustly predicted total RNA levels with a one unit difference in UBF levels corresponding to one standard deviation leading to `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight (Table 3).
-In contrast, no evidence for a relationship between total RNA and rpS6 was found when controlling for time (Table 3).
-To estimate the influence of total RNA on changes in muscle mass in TRAIN, measured as *m.vastus lateralis* thickness, the average linear increase in total RNA was calculated for each leg together with the average total RNA abundance over the course of the training intervention.<!-- (estimated as the intercept in a regression model with number of sessions centered at session 6).
-move to methods -->
-The relationship between muscle growth and characteristics of individual total RNA profiles were estimated using a regression model containing the increase in total RNA concentrations, expressed as percentage increase per session, and average total RNA abundance at Session 6, expressed as standard deviations from the mean.
-The model also contained sex as predictor to account for sex differences in muscle growth.
-Total RNA increase in response to training RNA was robustly related to muscle growth (Table 4, Figure 4A).
-Average total RNA levels tended to negatively influence muscle growth (Table 4, Figure 4B).
-To assess the robustness of the model, individual relationships between sessions and total RNA levels (Figure 4C) were recalculated after removal of single data points.
-The model was refitted using new estimates of the total RNA abundance and increase per session.
-Effects of refitting the model is shown in Figure 4D.
-This analysis indicated that the relationship between Total RNA increase and muscle growth was preserved in all iterations (Figure 4D).
-We further assessed the model by iteratively removing one participant from the data set, similarly this showed that estimates of the effect of total RNA increase on muscle growth was robust but the effect of average total RNA estimates were more variable (Participant 8 and 3 in Figure 4D).
+In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coninciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 3).
+In contrast, no evidence was found for a relationship between total RNA and rpS6 protein levels (Table 3).
+
+In TRAIN, there was a robust positive relationship between rates of increase in total RNA in response to training and muscle growth measured as increases in *m.
+vastus lateralis* thickness (Table 4, Figure 4A), with changes in total RNA over the course of the training intervention being estimated in each leg using a regression model containing number of sessions as the independent variable.
+Conversely, there was a tendency towards a negative relationship between average total RNA levels at Session 6 and changes in muscle thickness (Table 4, Figure 4B), with the average total RNA levels estimated as the predicted value at Session 6 (estimated as the intercept-term) from the model used to estimate the rate of total RNA increase per session.
+
+To assess the robustness of the model for predicting muscle growth, individual relationships between sessions and total RNA levels (Figure 4C) were recalculated after removal of single data points from each participant.
+The model predicting muscle growth was refitted using new estimates of changes in total RNA abundances and increases thereof per session.
+Each refitted model resulted in slightly different estimates (displayed as means and 95% CI in Figure 4D).
+No single data point influenced the r4esults in any meaningful way.
+Next we assessed the robustness by iteratively removing one participant from the data set, similarly this showed that estimates of the effect of total RNA increase on muscle growth was robust but the effect of average total RNA estimates were more variable (Participant 8 and 3 in Figure 4D).
 
@@
 Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
-We show that this increase is interconnected with increases in UBF protein abundance, suggesting a plausible manner for rRNA transcription regulation in response to RT.
-We did not find any convincing support for the notion that ribosomal biogenesis is tightly regulated by changes in training volume, however, training cessation may lead to halted ribosome biogenesis.
-Finally, the rate of increase in total RNA abundance predicted the magnitude of muscle growth.
+These markers accumulated progressively during the initial part of the intervention before it leveled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
+This increase in total RNA was interconnected with increases in UBF protein abundance, suggesting UBF levels *per se* to play a role in regulation of rRNA transcription regulation in response to RT.
+Total RNA increases was not affected by weekly fluctuations in training volume, however, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
+Finally, individual rates of increases in total RNA abundance predicted the magnitude of muscle growth, confirming the likely link between ribosomal biogenesis and muscle mass accretion 
+[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
+@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
+and muscle hypertrophy
+[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
+@RN1755; <!-- Stec Ribosome bio higher in high responders -->
+@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->
 
-Several studies have presented evidence of a determinant role for ribosomal biogenesis in RT induced muscle hypertrophy.
-Increases in total RNA has been shown to positively correlate with RT induced hypertrophy
+Total RNA seems to be a valid proxy marker of ribosomal density, as most of the RNA is assumed to be ribosomal RNA
+[@RN2223],<!-- Young  -->
+which in turn is a valid marker of translational capacity
+[@RN2145].
+Several studies have shown that total RNA content is altered by RT
+<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
+[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
+@RN2180;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --> 
+@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
+@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
+@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
+@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
+@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
+@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
+@RN1897<!-- 11% 36 sessions young -->],
+as was also the case in the present data set.
+However, the time course of total RNA/rRNA changes in response to RT has so far remained speculative, with no study investigating responses to prolonged interventions with multiple sampling time points.
+In the present data, RT led to a clear session-to-session increase in total RNA per unit tissue weight in response to the first four session, whereupon the changes gradually leveled out before peaking after the 8^th^ session, with the peak increase from baseline being $\sim$ 50%, defining an accumulation phase.
+This corroborates well with previous suggestions of peak values being reached within four to nine sessions in young males and females
+[@RN1809;
+@RN2358], and may be essential for preparing muscle fibers for subsequent growth
+[@RN1809;<!-- Correlation between increase and muscle growth -->
+@RN1644; <!--Figueiredo correlation between incr and muscle hypertrophy -->
+@RN2358].<!-- Early accumulation of RNA predicts volume depdent growth -->
+After the 8^th^ session, no meaningful increase or decrease were observed for total RNA/rRNA content within the training period, suggesting a plateau phase with attenuated net synthesis of novel ribosomes.
+Within this last part of the intervention, synthesis of novel rRNA still seemed to be elevated per weight unit muscle tissue compared to baseline, as suggested by sustained elevation of pre-rRNA transcripts, coinciding with peak values of UBF protein levels.
+This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth [@RN2142].<!-- Figueiredo review details the concept of rna concentration -->
+
+The observed rates of RNA accumulation over the entirety of the intervention were found to be a determinant of changes in muscle thickness (after controlling for average total RNA levels).
+Individuals with higher rates of accumulation showed larger accretion of muscle mass.
+This supports the notion that ribosomal biogenesis is an important determinant of RT-induced muscle hypertrophy, with previous studies showing that increases in total RNA are positively correlated with increases in muscle mass
 [@RN1644;<!-- Figueriedo  -->
 @RN1897;
-@RN2055].<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
+@RN2055],<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
 <!-- Reidy correlation between total RNA and muscle hypertrophy-->
-and differs between high- and low-responders to RT
-[@RN1755].<!-- Stec 2016  -->
-Additionally, volume-dependent regulation of total RNA predicted beneficial effects of high- vs. low-volume RT
-[@RN2358]. In addition to observational evidence from human studies, blockage of ribosomal biogenesis halts muscle cellular growth *in vitro* in some
+differs between individuals displaying low vs. high levels of muscle hypertrophy in response to RT
+[@RN1755]<!-- Stec 2016  -->
+an contribute to explain RT volume-dependent changes in muscle mass and strength
+[@RN2358].
+In addition, supression of ribosomal biogenesis in *in vitro* models leads to halted muscle cellular growth in some 
 [@RN1632;<!-- Nader 2005 blockage of mTOR halts ribo bio and growth -->
@@
 @RN1755]<!-- Stec specific pol I inhibition -->
-but not all situations
-[@RN1929]<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
-Results from the present study corroborates the link between ribosomal biogenesis and muscle hypertrophy.
-When using all available data points we were able to robustly estimate both the rate of increase in total RNA and the average RNA levels over the course of the study and include both when modeling changes in muscle thickness.
-The results of this analysis showed that the rate of RNA accumulation determined muscle growth when average levels of total RNA was controlled for.
-In other words, at a fixed relative rate of increase in total RNA, legs with higher levels of RNA did not grow more but at a fixed amount of total RNA, legs with higher increase per session did.
-This suggest that the absolute ribosomal density is not as important as the rate of increase in ribosomal density.
-In an homogeneous group, in response to RT, the rate of increase will be highly correlated with maximal values seen after RT, linking maximal rate to the resulting abundance of ribosomes.
-The same pattern revealed in our model is seen when comparing young and old muscle, where aged muscle display higher levels of total RNA at rest
+but not all studies
+[@RN1929].<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
+Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA was instead associated with a tendency towards lowered muscle growth.
+Overall, the rate of increases in ribosomal density thus seems to be a better predictor of individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
+Interestingly, the interaction between rRNA synthesis rate and muscle mass accretion (but not between ribosomal content and muscle mass accretion) may shed light on observed differences in muscular responses to RT between young and old individuals.
+Whereas aged muscle display higher levels of total RNA at rest 
 [@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
-and lower levels of increase in response to RT
-[@RN1809].<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
-Together with our results this underlines the importance of understanding RT induced ribosomal synthesis in relation to manipulation of training loads for specific populations.
+they show reduced changes in total RNA levels in response to RT
+[@RN1809],<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
+potentially explaining their alleged poorer overall hypertrophic responses [@RN1809].
+Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
+Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.
 
@@
 Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for.
-Together with a disconnect after the de-training period, this may suggest that regulation of rpS6 expression and transcription of ribosomal RNA displays different temporal characteristics resulting from RT.
+Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT.
 Additionally, ribosomal proteins may have extra-ribosomal functions
@@
 
-Total RNA is the most studied proxy marker of ribosomal density as most of the RNA is assumed to be ribosomal [@RN2223].<!-- Young  -->
-This marker has been shown to be sensitive to RT in numerous studies 
-<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
-[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
-@RN2180;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --> 
-@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
-@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
-@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
-@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
-@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
-@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
-@RN1897<!-- 11% 36 sessions young -->].
-However, a detailed time course has not been mapped.
-To date, available data suggests that maximal levels are reached within four to nine sessions in young males and females
-[@RN1809;
-@RN2358].
-We did not detect any meaningful further increases after 48 hrs after the eight session were peack the peack average increase was 50% from baseline.
-This confirmed a plateau when total RNA was expressed per unit tissue weight.
-A possible reason for the plateau could be attenuated ribosomal biogenesis.
-However, pre-rRNA abundances per unit tissue weight remained at elevated levels indicating retained transcriptional activity after the twelfth session together with maximal levels of UBF protein.
-Another possibility is a dilution effect
-[@RN2142]<!-- Figueiredo review details the concept of rna concentration -->
-due to muscle hypertrophy and thus protein accretion.
-After the twelfth session, muscle hypertrophy was apparent and did not change in response to the de-training period.
-As muscle hypertrophy did not change from after session twelve to after the de-training period, the apparent decrease in total RNA was not driven by dilution but by attenuated rRNA transcription.
-This notion is supported by normalization of pre-rRNA levels and possibly by lowered levels of UBF protein, although this apparent effect was not confirmed as statistically robust.
-
-Over the full training period UBF levels did robustly explain total RNA levels after accounting for time, something that otherwise would lead to biased estimates as both the dependent variable (total RNA) and the covariate (UBF levels) varies with time.
-From a mechanistic perspective, UBF is an important factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA polymerase I
+UBF levels robustly explained total RNA levels over the entire course of the intervention.
+These analyses were done with the number of sessions accounted for, allowing unbiased estimates.
+Unrealistically strong relationships could have been otherwise expected as both the dependent variable (total RNA) and the covariate (UBF levels) varies with the number of sessions.
+From a mechanistic perspective, UBF is an important transcription factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA Pol I
 [@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
-Activation of UBF is controlled by signaling from the mechanosensitive mTOR, pathway as rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
+Activation of UBF is controlled by the mechanosensitive mTOR pathway, and rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
 [@RN1632; <!-- Nader rapamycin block UBF availability  -->
-@RN2564]<!-- Hannan 2003 -->
+@RN2564].<!-- Hannan 2003 -->
 <!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1.
@@
 pointing to an effect observed in cell models where c-Myc induces UBF mRNA transcription [@RN2565].<!-- Pootinga 2004 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
-Interestingly the avalability of UBF *per se* has been shown to regulate rRNA transcription
+Interestingly the availability of UBF *per se* has been shown to regulate rRNA transcription
 [@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->  through control of rDNA gene activity
@@
 
-Resistance training volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including c-Myc induction, mTOR activation
+After eight days of de-training, total RNA and rRNA levels per weight unit muscle tissue returned toward baseline levels, though without concomitant reversal of muscle thickness, which remained at elevate levels.
+This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
+This supports the idea that ribosomal biogenesis is a cellular activity on demand, possibly relating to its relative expense [@RN1920] also in muscle tissue.
+Based on this notion, and the fact that RT volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including induction of c-Myc expression, mTOR activation
 [@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
@@
 subsequent total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
-and subsequent training outcomes [@RN1767; @RN2358].<!-- Schoenfeld 2016 meta and Hammarström -->
-We have previously shown that different amounts of training volume in constant volume protocols leads to differences in total RNA and rRNA accumulation in the initial phase of RT [@RN2358].
-Based on the notion that ribosomal biogenesis is a complex and energy demanding process
-[@RN1920],<!-- Warner economics of ribosomal biogenesis -->
-we hypnotized that fluctuations in training load would reflect in markers of ribosomal biogenesis.
-When comparing the two volume conditions in the present study we only found one part of the 45S ETS to be differentially expressed at Session 12 in favor of VAR together with a tendency of a rescued UBF level after de-training in response to increased volume in the VAR protocol.
-These observations do not give much support for a clear effect of fluctuations in training volume in the short time span used in the present study although the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume.
-Indeed, both training protocols utilized in the present study also increased strength and induced muscle hypertrophy to a similar degree.
-From a general perspective, although volume is an important factor for increases in muscle strength and mass
+and subsequent training outcomes [@RN1767; @RN2358], we hypothesized that fluctuations in training volume would be reflected in markers of ribosomal biogenesis.
+When comparing VAR to CONST in the present study we found only one part of the pre-rRNA, 45S ETS, to be differentially expressed and only so after Session 12 in favor of VAR together with a tendency towards rescued UBF levels after de-training in response to increased volume in the VAR but not CONST protocol.
+These observations do not give support to a clear effect of fluctuations in training volume on total RNA levels or rRNA expression within a relatively short and training-intensive intervention, though it should be noted that the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume.
+Indeed, both training protocols utilized in the present study increased muscle strength and induced muscle hypertrophy to a similar degree.
+From a general perspective, albeit volume is an important determinant of increases in muscle strength and mass
 [@RN1767;<!-- schoenfeld 2016 dose response  -->
-@RN2492]<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
-different organization of training loads over time is likely of minor importance when training volumes are equated
+@RN2492],<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
+differences in organization of training loads is likely of minor importance when training volumes are equated over time
 [@RN2572].<!-- Grgic 2017 volume equated periodization schemes led to similar muscle hyeprtrophy -->
-It is important to note that RT in the current study was performed with constant volume in the first four sessions, something that could have been more than enough to maximize rRNA transcription in previously untrained individuals.
+It is important to note that RT in the current study was performed with the same volume in the first four sessions, something that could have been more than enough to maximize rRNA transcription in previously untrained individuals.
 This is supported by the observation that pre-rRNA increased rapidly initially in both protocols with minimal changes in response to subsequent sessions, regardless of exercise volume.
-The CONST protocol in the present study corresponded to volumes used in the high volume condition in a previous study from our lab (three sets in two exercises activating knee extensor muscles) [@RN2358].
-There, higher levels of total RNA accumulation were observed compared to a low volume protocol [@RN2358].
+The CONST protocol in the present study corresponded to volumes used in the moderate volume condition in a previous study from our lab (three sets in two exercises activating knee extensor muscles) [@RN2358].
+There, higher levels of total RNA were observed after four sessions in the moderate compared to a low volume protocol [@RN2358].
 Interestingly, using a progressive volume protocol in well-trained participants, increases in total RNA have been reported throughout six weeks of training
-[@RN2055].<!-- Mobley 2018 trend towards higher increases in high responders 32% high resp, 8% low resp 12 wk training -->
-Compared to the present study and other using constant volume protocols
+[@RN2225].<!-- Haun total RNA increase througout a 6 week intervention -->
+Altough this observation was done in well-trained participants performing a high volume protocol without a control condition with constant volume, compared to constant volume protocols
 [@RN1809;<!-- Brook 2016 -->
-@RN2358],<!-- Hammarström 2020 -->
+@RN2358; and the present study],<!-- Hammarström 2020 -->
 progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
-This observation was done in well-trained participants performing a high volume protocol without a control group condition with constant volume.
 
-In conclusion, RT induced ribosome accumulation reaches peak values in the initial phase of RT (8-12 sessions) affected by total levels of UBF.
-Fluctuations in training volume does not transfer to fluctuations in ribosomal biogenesis a moderate volume initiation of RT but training cessation does lead to attenuated ribosomal biogenesis.
-The rate of total RNA accumulation predicts RT induced muscle hypertrophy.
+In conclusion, RT-induced ribosome accumulation reached peak values in the initial phase of RT (8 sessions) and was interconnected with increases in UBF protein levels.
+The rate of total RNA accumulation predicted RT-induced muscle hypertrophy.
+Fluctuations in training volume did not transfer to fluctuations in ribosomal biogenesis, but training cessation led to decreased ribosomal content.
 
@@
 
-**Figure 1.** Study design (A) training load (B) and training outcomes (muscle strength, C; muscle thickness D).
+**Figure 1.** Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).
 

### 2020-09-23 Figures/Figure1.R

@@ -576,6 +576,10 @@ figure1 <- plot_grid(design_full_biopsies,
 
 
 
+### Save figure for plotting 
+
+saveRDS(figure1, "./Figures/rds/figure1.RDS")
+
 
 
 

### 2020-09-23 Figures/Figure2.R

@@ -287,6 +287,7 @@ figure2 <- plot_grid(
 
 
 
+saveRDS(figure2, "./Figures/rds/figure2.RDS")
 
 
 # Width of figure = 1x columns 8.9 cm

### 2020-09-23 Figures/Figure3.R

@@ -618,6 +618,7 @@ figure3 <- plot_grid(
 
 
 
+saveRDS(figure3, "./Figures/rds/figure3.RDS")
 
 # Width of figure = 1x columns 8.9 cm
 # height of figure = full page = 23 cm

### 2020-09-23 Figures/Figure4.R

@@ -277,6 +277,7 @@ figure4 <- plot_grid(
 
 
 
+saveRDS(figure4, "./Figures/rds/figure4.RDS")
 
 
 # Width of figure = 1x columns 8.9 cm

### 2020-09-23 R/ubf-tot-rna-model.R

@@ -265,7 +265,7 @@ conditional_effects(comb.m1)
 
 # Save model 
 saveRDS(comb.m1, "./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")
-saveRDS(comb.s6.m1, "./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")
+saveRDS(comb.s6.m1, "./data/derivedData/ubf-tot-rna-model/comb_m1_s6.RDS")
 
 
 

### 2021-04-12 manuscript.Rmd
added small changes..

@@
 title: "Ribosome accumulation during early phase resistance training"
-author: Daniel Hammarström^1,2,£^&ast;,Sjur J. Øfsteng^2,£^, Nicolai B. Jacobsen^1^,
-  Krister B. Flobergseter^1^,  Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
+author:
+  - Daniel Hammarström^1,2,£^*
+  - Sjur J. Øfsteng^2,£^
+  - Nicolai B. Jacobsen^1^
+  - Krister B. Flobergseter^1^
+  - Bent R. Rønnestad^1^
+  - Stian Ellefsen^1,3^
 date: ''
 output:
+  pdf_document:
+    latex_engine: xelatex
   word_document:
@@
 bibliography: resources/bibliography.bib
-header-includes: \usepackage{wasysym}
+header-includes: 
+  \usepackage{wasysym}
 editor_options:
@@
 source("./R/libs.R")
+library(knitr); library(kableExtra)
 # Settings for chunk options
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = "cairo_pdf")
 
@@
 ^1^ Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway.
-
 ^2^ Swedish School of Sport and Health Sciences, Stockholm, Sweden.
-
-^3^ Innlandet Hospital Trust, Lillehammer, Norway. </div>
+^3^ Innlandet Hospital Trust, Lillehammer, Norway.
+</div>
 
@@
 
-\pagebreak
-
 # Abstract
@@
 
-Increased ribosomal density is a phenomena observed in response to resistance training (RT) supporting elevated.
+Increased ribosomal density is a phenomenon observed in response to resistance training (RT) supporting elevated.
 A time course of RT-induced accumulation of ribosomes has not been described.
-We mapped markers of ribosomal biogenesis (total RNA, ribosomal RNA (rRNA), ribosomal protein S6 (rpS6) and upstream binding factor (UBF) protein content) in response to twelve RT sessions.
-Additionally we investigated the effect of fluctuations in training volume on these markers and relationship between RNA accumulation and muscle growth measured as muscle thickness (ultra sound).
-Eighteen participants were allocated to either a training (TRAIN, *n*=11) or control (CTRL, *n*=7) group.
-In TRAIN, participants performed unilateral knee extension with either constant (CONST, 6 sets) or a variable volume (VAR, 6, 3 and 9 sets in sessions 1-4, 5-8 and 9-12, respectively).
-Muscle biopsies were sampled from *m vastus lateralis* in TRAIN before and 48 hrs after the first session as well as 48 hrs after session 4, 5, 8, 9 and 12 and after eight days of de-training.
+We mapped markers of ribosomal biogenesis (total RNA, ribosomal RNA (rRNA), ribosomal protein S6 (rpS6), and upstream binding factor (UBF) protein content) in response to twelve RT sessions.
+Additionally, we investigated the effect of fluctuations in training volume on these markers and the relationship between RNA accumulation and muscle growth measured as muscle thickness (ultrasound).
+Eighteen participants were allocated to either a training (TRAIN, *n*=11) or a control (CTRL, *n*=7) group.
+In TRAIN, participants performed unilateral knee extension with either constant (CONST, 6 sets) or a variable volume (VAR, 6, 3, and 9 sets in sessions 1-4, 5-8, and 9-12, respectively).
+Muscle biopsies were sampled from *m vastus lateralis* in TRAIN before and 48 hrs after the first session and 48 hrs after sessions 4, 5, 8, 9, and 12, and after eight days of de-training.
 Control biopsies were sampled in CTRL at baseline and after 48 hrs and 3-5 weeks.
@@
 Total RNA increased in a curve-linear fashion, most rapidly in response to the first four sessions  (8.6%, 95% CI: [5.5, 11.7] per session), followed by a plateau and peak values (49.5% [34.2, 66.5] above baseline) after Session 8.
-UBF protein levels explained total RNA levels after controlling for time and increases in total RNA levels predicted RT induced muscle hypertrophy.
-After de-training, total RNA and specific rRNA species decreased without changes in muscle mass indicating reduced concentrations and biosynthesis of ribosomes.
-These result underlines a determinant role for ribosomal biogenesis in RT-induced muscle hypertrophy and that ribosomal biogenesis is sensitive to training cessation.
+UBF protein levels explained total RNA levels after controlling for time, and increases in total RNA levels predicted RT-induced muscle hypertrophy.
+After de-training, total RNA and specific rRNA species decreased without muscle mass changes indicating reduced concentrations and biosynthesis of ribosomes.
+These results underline a determinant role for ribosomal biogenesis in RT-induced muscle hypertrophy and that ribosomal biogenesis is sensitive to training cessation.
 
@@
 vastus lateralis* were assessed prior to the first biopsy (TRAIN and CTRL) as wells as before the second to last (TRAIN) and last (TRAIN and CTRL) biopsy.
+
+```{r table1}
+
+# Read dxa data
+
+dxa_whole <- read_excel("./data/tr010_dxa_standarddefinitions.xlsx") %>%
+  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:`HeleVev%fett`) %>%
+  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:`HeleVev%fett`) %>%
+  mutate(value = gsub("kg", "", value), 
+         value = gsub("[[:space:]]", "", value, fixed = FALSE), 
+         value = as.numeric(value)) %>%
+
+  separate(Etternavn, into = c("Study", "participant")) %>%
+  dplyr::select(participant, Date = Målingsdato, variable, value) %>%
+  filter(!(participant %in% c("P08", "P20", "P17"))) %>%
+  group_by(participant) %>%
+  mutate(time = if_else(Date == min(Date), "S0", 
+                        if_else(Date == max(Date), "post1w", 
+                                "S12")), 
+         leg = if_else(variable == "BeinhøyreMagermasse", "R", "L")) %>%
+  filter(variable %in% c("HeleBeinmasse", 
+                         "HeleFettmasse", 
+                         "HeleVevmasse", 
+                         "HeleMagermasse" ,
+                         "Vekt", 
+                         "HeleTotalmasse")) %>%
+  pivot_wider(names_from = variable, values_from = value) %>%
+  mutate(fat_rel = 100*(HeleFettmasse/HeleTotalmasse)) %>%
+  dplyr::select(participant, time, Vekt,HeleTotalmasse, fat_rel) %>%
+  filter(time == "S0")
+
+# DXA weight stature
+
+read_excel("./data/tr010_age_participant.xlsx") %>%
+  inner_join(dxa_whole) %>%
+  mutate(age = difftime(recruited, dob)/365.2425, 
+         bmi = weight/(height/100)^2) %>%
+  dplyr::select(participant, group, height, weight,age, sex,bmi, fat_rel) %>%
+
+  group_by(group, sex) %>%
+  summarise(n = n(), 
+            Age = mean(age, na.rm = TRUE), 
+            age.sd = sd(age, na.rm = TRUE),
+            mass = mean(weight), 
+            mass.sd = sd(weight),
+            stature = mean(height),
+            stature.sd = sd(height), 
+            BMI = mean(bmi), 
+            bmi.sd = sd(bmi), 
+            relfat = mean(fat_rel), 
+            relfat.sd = mean(fat_rel) ) %>%
+  ungroup() %>%
+  mutate(n = as.character(n), 
+         age = paste0(sprintf("%.1f", round(Age, 1)), " (", sprintf("%.1f", round(age.sd, 1)), ")"),
+         mass = paste0(sprintf("%.1f", round(mass, 1)), " (", sprintf("%.1f", round(mass.sd, 1)), ")"),
+         stature = paste0(sprintf("%.1f", round(stature, 1)), " (", sprintf("%.1f", round(stature.sd, 1)), ")"),
+         BMI = paste0(sprintf("%.1f", round(BMI, 1)), " (", sprintf("%.1f", round(bmi.sd, 1)), ")"),
+         fat = paste0(sprintf("%.1f", round(relfat, 1)), " (", sprintf("%.1f", round(relfat.sd, 1)), ")"))%>%
+  dplyr::select(group, sex, n, age, mass, stature, BMI, fat) %>%
+  pivot_longer(names_to = "variable", values_to = "values", n:fat) %>%
+  pivot_wider(names_from = c(sex,group), values_from = values) %>%
+
+  mutate(variable =  factor(variable,
+                          levels = c("n", "age", "mass", "stature", "BMI", "fat"), 
+                          labels = c("n", "Age (yrs)", "Body mass, (kg)", "Stature (cm)", "Body mass index (kg/m\\textsuperscript{2})", "Body fat (\\%)"))) %>%
+
+  dplyr::select(variable, F_experiment, M_experiment, F_control, M_control) %>%
+
+  kable(caption = "Participant characteristics.", 
+        booktabs = TRUE, 
+        col.names = c("", "Female", "Male", "Female", "Male"),
+        escape = FALSE,
+        format = "latex") %>%
+  add_header_above(c(" ", "Experimental group" = 2, "Control group" = 2))
+
+```
 
@@
 Raw fluorescence data was exported from the QuantStudio software and estimates of quantification cycle (Cq) and amplification efficiency was derived for each reaction using the qpcR package[@RN1768].
+
+```{r primerstable}
+# Read qpcr data
+qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")
+
+primer_performance <- qdat %>%
+  group_by(target) %>%
+  summarise(mean.cq = mean(cq, na.rm = TRUE), 
+            sd.cq = sd(cq, na.rm = TRUE), 
+            mean.eff = mean(eff))
+
+merge_custom <- function(ft, x, columns){
+  z <- rle(x)
+  rows_at <- cumsum(z$lengths) - z$lengths + 1
+
+  for(i in seq_along(rows_at)){
+    for(j in columns)
+      ft <- merge_at(x = ft, i = seq( rows_at[i], rows_at[i] + z$lengths[i] - 1), j = j)
+  }
+
+  ft
+}
+
+# Read primer data 
+primer_table <- read_excel("./data/tr010_primers.xlsx", sheet = 1) %>%
+  dplyr::select(target = primer_id, forward, reverse, full_name, symbol_ext) %>%
+  inner_join(primer_performance) %>%
+  mutate(performance = paste0(sprintf("%.1f", round(mean.cq,1)), " (", sprintf("%.1f",        round(sd.cq,1)), "), E = ", sprintf("%.2f", round(mean.eff,2))), 
+         forward = paste0("F: 5'-", forward, "-3'"), 
+         reverse = paste0("R: 3'-", reverse, "-5'")) %>%
+
+  dplyr::select(symbol_ext, full_name, forward, reverse, performance) %>%
+  pivot_longer(names_to = "primer", values_to = "seq", forward:reverse) %>%
+dplyr::select(Symbol = symbol_ext, 'Transcript name' = full_name, Sequence = seq, 'Mean Cq (SD) and efficiency' = performance) %>%
+    filter(!(Symbol %in% c("RPL32", "MYH7", "MYH2", "MYH1")))
+
+primer_table %>%
+  kable(format = "latex", 
+        caption = "Primer sequences and average performance.",
+        booktabs = TRUE) %>%
+  collapse_rows(columns = c(1, 2, 4)) %>%
+  landscape()
+
+# 
+# flextable() %>%
+# merge_custom(x = primer_table$Symbol, columns = c(1, 2, 4)) %>%
+# add_header_lines(values = "Table 2. Primer sequences and average performance.") %>%
+#     fontsize(size = 8, part = "body") %>%
+#   fontsize(size = 10, part = "header") %>%
+# autofit()
+#
+
+```
 
@@
 
+```{r figure1, fig.height=(23*0.75)/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).", echo=FALSE}
+
+figure1 <- readRDS("./Figures/rds/figure1.RDS")
+
+figure1
+
+```
+
 ```{r western-data}
@@
 
+```{r figure2, fig.height=(23*0.75)/2.54, fig.width=8.9/2.54, fig.align="center", fig.cap="Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF.
+Black points and error bars represents statistically robust results (a 95\\% CI not containing 0).
+C shows western-blots and total protein stains from a representative participant.
+mRNA data is normalized per total RNA.", echo=FALSE}
+
+figure2 <- readRDS("./Figures/rds/figure2.RDS")
+
+figure2
+
+```
+
 ```{r total-rna}
@@
 
+```{r figure3, fig.height=(23*0.75)/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Total RNA and ribosomal RNA subspecies in response to trainin.
+Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (A).
+There were no robust differences between different volume conditions in the training group (A).
+B shows primer locations targeting different ribosomal RNA subspecies.
+Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and other after the training period (C).
+Error bars shows 95\\% CI.
+Asterisk in C indicates robust differences between volume conditions (a 95\\% CI of pairwise differences not containing 0).", echo=FALSE}
+
+figure3 <- readRDS("./Figures/rds/figure3.RDS")
+
+figure3
+
+```
+
+```{r table3-regression-models-muscle-thickness}
+# Save model
+
+# Model for rpS6 predicting total RNA
+comb.s6.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1_s6.RDS")
+
+comb.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")
+
+comb.m1 <- summary(comb.m1)
+comb.s6.m1 <- summary(comb.s6.m1)
+
+reg_results <- 
+  cbind(rbind(comb.m1$fixed,
+      comb.m1$random$participant, 
+      comb.m1$random$`participant:leg`, 
+      comb.m1$spec_pars), 
+      data.frame(coef = c("Intercept", 
+                          "UBF protein levels (SD-units)", 
+                          "Session 1-4\\textsuperscript{b}",
+                          "Session 4-8\\textsuperscript{c}", 
+                          "Session 8-12\\textsuperscript{d}", 
+                          "De-training",
+                          "Between Participant variation",
+                          "Between Participant:Leg variation", 
+                          "Residual SD"))) %>%
+  remove_rownames() %>%
+  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)
+
+reg_results_s6 <- 
+  cbind(rbind(comb.s6.m1$fixed,
+      comb.s6.m1$random$participant, 
+      comb.s6.m1$random$`participant:leg`, 
+      comb.s6.m1$spec_pars), 
+      data.frame(coef = c("Intercept", 
+                          "rpS6 protein levels (SD-units)", 
+                          "Session 1-4\\textsuperscript{b}",
+                          "Session 4-8\\textsuperscript{c}", 
+                          "Session 8-12\\textsuperscript{d}", 
+                          "De-training",
+                          "Between Participant variation",
+                          "Between Participant:Leg variation", 
+                          "Residual SD"))) %>%
+  remove_rownames() %>%
+  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)
+
+  rbind(reg_results, reg_results_s6) %>%
+
+  mutate(Estimate = round(Estimate, 2), 
+         Est.Error = round(Est.Error, 2), 
+         lwr = round(lwr, 2), 
+         upr = round(upr, 2)) %>%
+
+    kable(col.names = c("Coefficient", 
+                        "Estimate\\textsuperscript{a}", 
+                        "SD", 
+                        "Lower 95\\% CI", 
+                        "Upper 95\\% CI"), 
+          caption = "Effect of UBF and rpS6 levels, sessions and de-training on RNA-levels.",
+          format = "latex", 
+          escape = FALSE,
+          booktabs = TRUE) %>%
+    row_spec(9, hline_after = TRUE) %>%
+    footnote(alphabet = c("The dependent variable is total RNA levels (log).",
+    "Slope in response to session 1-4.", 
+    "Change in slope in session 4-8.",
+    "Change in slope in session 8-12."), 
+             footnote_as_chunk = TRUE,
+             threeparttable = TRUE, 
+             escape = FALSE,
+             general_title = "")
+
+```
+
 ```{r combined-analysis}
@@
                    "]")
+
+```
+
+```{r table4-regression-models-muscle-thickness}
+# Save model
+
+hyp.pre.m3 <- readRDS( "./data/derivedData/ubf-tot-rna-model/hyp_pre_m3.RDS")
+
+m3.sum <- summary(hyp.pre.m3)
+
+# combine output from all models
+
+reg_results <- 
+  cbind(rbind(m3.sum$fixed,
+      m3.sum$random$participant, 
+      m3.sum$random$`participant:leg`, 
+      m3.sum$spec_pars), 
+      data.frame(coef = c("Intercept", 
+                          "Sex (Male)", 
+                          "Mean Total RNA",
+                          "Mean Total RNA increase per session", 
+                          "Between Participant variation", 
+                          "Between Participant:Leg variation", 
+                          "Residual SD"))) %>%
+  remove_rownames() %>%
+  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)
+
+reg_results %>%
+
+  mutate(Estimate = round(Estimate, 2), 
+         Est.Error = round(Est.Error, 2), 
+         lwr = round(lwr, 2), 
+         upr = round(upr, 2)) %>%
+
+      kable(col.names = c("Coefficient", 
+                        "Estimate\\textsuperscript{a}", 
+                        "SD", 
+                        "Lower 95\\% CI", 
+                        "Upper 95\\% CI"), 
+          caption = "Total RNA as a predictor of muscle growth.",
+          format = "latex", 
+          escape = FALSE,
+          booktabs = TRUE) %>%
+
+   footnote(alphabet = c("The dependent variable is $\\\\Delta$ Muscle thickness (mm)"),
+
+             threeparttable = TRUE, 
+             escape = FALSE,
+             general_title = "")
 
@@
 Next we assessed the robustness by iteratively removing one participant from the data set, similarly this showed that estimates of the effect of total RNA increase on muscle growth was robust but the effect of average total RNA estimates were more variable (Participant 8 and 3 in Figure 4D).
+
+```{r figure4, fig.height=17.25/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 4).
+Values are averaged over values from men and women.
+Individual plots of estimates total RNA increases over time is shown in C together with results from leave-one-out analysis (D).
+Leave-one-out analysis shows the effect of removing a single participant (black point and error-bars) and individual values from the total RNA per time estimates where green points represents bounds of the 95\\% CI and yellow points represents mean estimates.", echo=FALSE}
+
+figure4 <- readRDS("./Figures/rds/figure4_x.RDS")
+
+figure4
+
+```
 
@@
 <div id="refs"></div>
-
-# Figure legends
-
-**Figure 1.** Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).
-
-**Figure 2.** Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF.
-Black points and error bars represents statistically robust results (a 95% CI not containing 0).
-C shows western-blots and total protein stains from a representative participant.
-mRNA data is normalized per total RNA.
-
-**Figure 3.**  Total RNA and ribosomal RNA subspecies in response to trainin.
-Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (A).
-There were no robust differences between different volume conditions in the training group (A).
-B shows primer locations targeting different ribosomal RNA subspecies.
-Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and other after the training period (C).
-Error bars shows 95% CI.
-Asterisk in C indicates robust differences between volume conditions (a 95% CI of pairwise differences not containing 0).
-
-**Figure 4.** 
-Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 4).
-Values are averaged over values from men and women.
-Individual plots of estimates total RNA increases over time is shown in C together with results from leave-one-out analysis (D).
-Leave-one-out analysis shows the effect of removing a single participant (black point and error-bars) and individual values from the total RNA per time estimates where green points represents bounds of the 95% CI and yellow points represents mean estimates.

### 2021-04-12 Figures/Figure3.R

@@ -565,7 +565,7 @@ rna_tc_fig <-  cond_eff_rna_tc %>%
    geom_ribbon(aes(ymin = lower, ymax = upper, color = NULL), alpha = 0.2) +
    
    labs(x = "Session", 
-        y = "RNA ng \U00D7 mg<sup>-1</sup>") +
+        y = bquote('Total RNA (mg'~'ng'^-1*')')) +
    
    scale_x_continuous(limits = c(0, 12), expand = c(0,0), 
                       breaks = c(0, 3, 6, 9, 12)) +
@@ -580,8 +580,7 @@ rna_tc_fig <-  cond_eff_rna_tc %>%
    plot_theme() +
    theme(legend.position = c(0.7, 0.15), 
          legend.key.size = unit(0.3, "cm"), 
-         axis.title.y = element_markdown(), 
-         ) 
+         axis.title.y = element_text() ) 
    
 
    

### 2021-04-12 Figures/Figure4.R

@@ -94,7 +94,7 @@ loo_panel <- res %>%
   plot_theme() + 
   theme(strip.background = element_rect(color = "white", fill = "white"),
         strip.text = element_text(size = 8, hjust = 0),
-        axis.title.y = element_markdown(size = 7), 
+        axis.title.y = element_text(size = 7), 
         legend.position = "none")
 
 
@@ -147,7 +147,7 @@ prediction_slope <- predict_df %>%
   labs(x = "Total RNA increase per session (%)", 
        y = "Muscle thickness change (mm)") +
   plot_theme() + 
-  theme(axis.title.y = element_markdown(size = 7), 
+  theme(axis.title.y = element_text(size = 7), 
         legend.position = c(0.2, 0.90), 
         legend.text = element_text(size = 7))
 
@@ -187,7 +187,7 @@ prediction_intercept <- predict_df %>%
                      breaks = c(-3, -2, -1, 0, 1, 2, 3, 4, 5), 
                      labels = c("", -2, "", 0, "", 2, "", 4, "")) +
   plot_theme() + 
-  theme(axis.title.y = element_markdown(size = 7), 
+  theme(axis.title.y = element_text(size = 7), 
         legend.position = "none")
 
 
@@ -230,13 +230,13 @@ rna_to_time_estimates <- combined_df %>%
   scale_color_manual(values = c(color.scale[1], color.scale[2])) +
   
   
-  labs(y = "Total RNA (ng \U00D7 mg <sup>-1</sup>)", 
+  labs(y = bquote('Total RNA (mg'~'ng'^-1*')'), 
        x = "Sessions") +
   
   plot_theme() + 
   theme(strip.background = element_rect(color = "white", fill = "white"),
         strip.text = element_text(size = 8, hjust = 0),
-        axis.title.y = element_markdown(size = 7), 
+        axis.title.y = element_text(size = 7), 
         legend.position = "none")
 
 
@@ -273,11 +273,11 @@ figure4 <- plot_grid(
 #                   x =   c(0.02, 0.02, 0.02, 0.02), 
 #                   y =   c(0.98, 0.73, 0.48, 0.23),
 #                   hjust=.5, vjust=.5, size = label.size)
+#
 
 
 
-
-saveRDS(figure4, "./Figures/rds/figure4.RDS")
+saveRDS(figure4, "./Figures/rds/figure4_x.RDS")
 
 
 # Width of figure = 1x columns 8.9 cm

### 2021-04-12 R/ubf-tot-rna-model.R

@@ -615,6 +615,11 @@ samples <- combined_df %>%
 # Save results
 results <- list()
 
+
+
+
+
+
 for(i in 1:length(samples)) {
   
   

### 2021-12-07 manuscript.Rmd
update after internal revision

@@
     reference_docx: resources/manuscript_word_template.docx
-csl: resources/experimental-physiology.csl
+csl: resources/acta-physiologica.csl
 bibliography: resources/bibliography.bib

### 2021-12-07 Figures/Figure3.R

@@ -8,6 +8,8 @@ bayes <- TRUE
 
 
 source("./R/figure-source.R")
+library(readxl)
+library(tidybayes)
 # source("./R/libs.R")
 
 
@@ -551,23 +553,91 @@ tot_rna_fold_change <- comp_rna %>%
 
 # Time course in the intervention group #####
 
-cond_eff_rna_tc <- readRDS("./data/derivedData/total-rna-analysis/cond_eff_rna_tc.RDS")
+
+
+#### NEW PIECE-WISE FIGURE FROM POSTERIOR DISTRIBUTION ###########
+
+
+tc.m1 <- readRDS( "./data/derivedData/total-rna-analysis/tc.m1.RDS")
 
 
 
-rna_tc_fig <-  cond_eff_rna_tc %>%
-  mutate(cond = factor(cond, levels = c("const", "var"), 
-                       labels = c("Constant volume", 
-                                  "Variable volume")), 
-         cond = fct_rev(cond)) %>%
-  ggplot(aes(time.c, estimate, color = cond, fill = cond)) +
-  geom_line() + 
-   geom_ribbon(aes(ymin = lower, ymax = upper, color = NULL), alpha = 0.2) +
+tc.m1.predictions  <- tc.m1 %>%
+  spread_draws(b_Intercept, b_condvar,
+               b_time1, 
+               b_time2, 
+               b_time3,
+               `b_condvar:time1`,
+               `b_condvar:time2`,
+               `b_condvar:time3`,
+               b_detraindetrain) %>%
+  mutate(
+    s0_const  = b_Intercept, 
+    s1_const  = b_Intercept + (b_time1 * 1), 
+    s2_const  = b_Intercept + (b_time1 * 2), 
+    s3_const  = b_Intercept + (b_time1 * 3),
+    s4_const  = b_Intercept + (b_time1 * 4), 
+    s5_const  = b_Intercept + (b_time1 * 5) + (b_time2 * 1), 
+    s6_const  = b_Intercept + (b_time1 * 6) + (b_time2 * 2), 
+    s7_const  = b_Intercept + (b_time1 * 7) + (b_time2 * 3), 
+    s8_const  = b_Intercept + (b_time1 * 8) + (b_time2 * 4), 
+    s9_const  = b_Intercept + (b_time1 * 9) + (b_time2 * 5) + (b_time3 * 1),
[…]

### 2021-12-07 Figures/Figure4.R

@@ -39,8 +39,8 @@ res <- loo.results$loo.sample %>%
   inner_join(leg) %>%
   
   mutate(Participant = factor(participant, 
-                              levels = paste0("P", c(seq(1:7), 19, 21, 22, 23)), 
-                              labels = paste0("P", seq(1:11))), 
+                              levels = c("Full\nmodel", paste0("P", c(seq(1:7), 19, 21, 22, 23))), 
+                              labels = c("Full\nmodel", paste0("P", seq(1:11)))), 
          Participant = fct_rev(Participant)) %>%
 
   
@@ -50,21 +50,55 @@ res <- loo.results$loo.sample %>%
 res2 <- loo.results$loo.participant %>%
   inner_join(leg) %>%
   mutate(Participant = factor(participant, 
-                              levels = paste0("P", c(seq(1:7), 19, 21, 22, 23)), 
-                              labels = paste0("P", seq(1:11))), 
+                              levels = c("Full\nmodel", paste0("P", c(seq(1:7), 19, 21, 22, 23))), 
+                              labels = c("Full\nmodel", paste0("P", seq(1:11)))), 
          Participant = fct_rev(Participant)) %>%
   
   
-  filter(model == "m3", coef != "sex") %>%
+  filter(model == "m3", !(coef %in% c("sex", "pre"))) %>%
   mutate(coef = factor(coef, levels = c("slope", "intercept"), 
                        labels = c("Total RNA increase\nper session", 
                                   "Average total RNA\n(Session 6)"))) %>%
     print()
   
 
+## Predictions from the model
+
+# get predictions from the model
+hyp.pre.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/hyp_pre_m3.RDS")
+
+m.coefs <- summary(hyp.pre.m1)
+
+full_model_estimates <- data.frame(Participant = "Full\nmodel",
+                                   stat = c("estimate", 
+                                            "lwr", 
+                                            "upr", 
+                                            "estimate", 
+                                            "lwr", 
+                                            "upr"),
+                                    coef = c(rep("slope",3), 
+                                             rep("intercept", 3)), 
+                                    estimates = c(m.coefs$fixed[5, 1], 
+                                                  m.coefs$fixed[5, 3], 
+                                                  m.coefs$fixed[5, 4], 
[…]

### 2021-12-07 R/total-rna-analysis.R

@@ -198,8 +198,8 @@ time_course <- rna_complete %>%
 # gam models fitted in brms to visualize general patterns in the data. 
 # a piecewise model is fitted to estimate slopes in different parts of the time course.   
 
-# GAM model
-tc.m1 <- brm(bf(log(rna.tissue) ~ cond  + s(time.c, by = cond, k = 7) + (1|participant)), 
+# Piecewise model "full" model
+tc.m1 <- brm(bf(log(rna.tissue) ~ cond  * (time1 + time2 + time3) + detrain + (1|participant)), 
      data = time_course, 
      warmup = 1000, # number of samples before sampling
      iter = 4000,  # number of mcmc iterations 
@@ -223,6 +223,7 @@ tc.m2 <- brm(bf(log(rna.tissue) ~ (time1 + time2 + time3) + detrain + (1|partici
              save_all_pars = TRUE, 
              control = list(adapt_delta = 0.99))
 
+
 # No interaction effect between any segment and cond. Reduce model to 
 # only containing time effects. 
 # Slopes in the different segments can be calculated through the hypothesis 
@@ -241,12 +242,18 @@ saveRDS(tc.m1, "./data/derivedData/total-rna-analysis/tc.m1.RDS")
 pp_check(tc.m2, type = "ecdf_overlay")
 summary(tc.m2)
 
-### A model for "descriptive" analysis of avaiable data points
+
+exp(0.05)
+
+### A model for "descriptive" analysis of available data points including other studies using priors
+prior_summary(tc.m3)
+prior_m3 <- set_prior("normal(0.0168, 0.0426)", class = "b", coef = "time1")
 
 
 
 tc.m3 <- brm(bf(log(rna.tissue) ~ (time1 + time2 + time3) + detrain + (1|participant)), 
              data = time_course, 
+             prior = prior_m3,
              warmup = 1000, # number of samples before sampling
              iter = 4000,  # number of mcmc iterations 
              cores = 4, # number of cores used in sampling
@@ -257,7 +264,66 @@ tc.m3 <- brm(bf(log(rna.tissue) ~ (time1 + time2 + time3) + detrain + (1|partici
 
 
 
[…]

### 2021-12-07 R/ubf-tot-rna-model.R

@@ -14,11 +14,11 @@
 # phosphorylation (rapamycin sensitive) increasing its activity and c-Myc induces
 # UBTF transcription increasing protein levels (rapamycin insensitive). 
 # 
-# Below we want to test if rRNA levels are assiciated with UBF levels. Both 
+# Below we want to test if rRNA levels are associated with UBF levels. Both 
 # are increasing with time and therefore, the model controls for the effect of training (time).
 #
 # This script also contain a model of muscle growth were we estimate the effect of increase in rRNA
-# on muscle growth (muscle thicknes change pre- to post training). Post training values are 
+# on muscle growth (muscle thickness change pre- to post training). Post training values are 
 # from both S12 and post 1w (de-training). rRNA transcription is estimated as the average linear increase over 
 # all sessions for each individual and leg. 
 #
@@ -29,8 +29,6 @@
 library(tidyverse)
 library(brms)
 library(bayesplot)
-library(lme4)
-library(readxl)
 
 
 
@@ -62,13 +60,17 @@ rna_complete <- rna %>%
   print()
 
 
+
+
 # Load western blot data 
 
-western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
+western_data <- readRDS("./data/derivedData/western-compile/western_data_calibrated.RDS")
+
+
 
 
 west_cont <- western_data %>%
-  
+
   mutate(time.c = gsub("S", "", time), 
          detrain = if_else(time.c == "post1w", "detrain", "train"),
          time.c = if_else(time.c == "post1w", "12", time.c),
@@ -84,29 +86,30 @@ west_cont <- western_data %>%
          time1 = time.c,
          time2 = if_else(time %in% c("S0", "S1", "S4"), 0, time.c - 4),
          time3 = if_else(time %in% c("S0", "S1", "S4", "S5", "S8"), 0, time.c - 8) ) %>%
-  
-  group_by(target, participant, time.c, cond, time, time1, time2, time3) %>%
- # Summarise duplicate values (if any) keep scaled expression for modeling
-   summarise(expression = mean(expression), 
-            expression.scaled = mean(expression.scaled)) %>%
 
-  filter(!is.na(expression)) %>%
+  
+  group_by(target, participant, cond, time.c,  time, time1, time2, time3) %>%
+ # Summarise duplicate values (if any) keep scaled expression for modeling
[…]

### 2021-12-07 R/wester_analysis.R

@@ -1,292 +1,0 @@
-### Western analaysis 
-
-library(tidyverse)
-library(brms)
-library(bayesplot)
-library(lme4)
-
-western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
-
-
-
-
-
-west.dat <- western_data %>%
- filter(time %in% c("S0", "S1c", "S1", "S12", "postctrl", "post1w")) %>%
-  mutate(group = if_else(cond == "ctrl_leg", "con", "int"), 
-         detrain = if_else(time == "post1w", "detrain", "train"), 
-         time = if_else(time == "S0", "pre", 
-                        if_else(time == "S1c", "S1", 
-                                if_else(time %in% c("postctrl", "post1w", "S12"), "post", time))), 
-         time = factor(time, levels = c("pre", "S1", "post")), 
-         tx = paste(time, group, detrain, sep = "_"), 
-         tx = factor(tx, levels = c("pre_con_train",
-                                    "S1_con_train",
-                                    "post_con_train",
-                                    "pre_int_train",
-                                    "S1_int_train",
-                                    "post_int_train", 
-                                    "post_int_detrain")), 
-         sample2 = paste0(participant, leg, time)) %>%
-  group_by(participant, tx, target) %>%
-  summarise(expression = mean(expression, na.rm = TRUE)) %>%
-  mutate(log.expr = log(expression)) %>%
-
-  print()
-  
-
-
-
-s6.m1 <- brm(bf(log.expr ~ tx + (1|participant)),
-
-          warmup = 1000, # number of samples before sampling
-          iter = 4000,  # number of mcmc iterations 
-          cores = 4, # number of cores used in sampling
-          chains = 4, # number of chains
-          seed = 5, # seed for reproducible results
-      #    control = list(adapt_delta = 0.95), 
-          data = west.dat[west.dat$target == "t-s6",])
-
-
-ubf.m1 <- brm(bf(log.expr ~ tx + (1|participant)),
-             
-             warmup = 1000, # number of samples before sampling
-             iter = 4000,  # number of mcmc iterations 
-             cores = 4, # number of cores used in sampling
-             chains = 4, # number of chains
-             seed = 5, # seed for reproducible results
-             #    control = list(adapt_delta = 0.95), 
-             data = west.dat[west.dat$target == "t-UBF",])
-
-
-
[…]

### 2021-12-07 R/wester_compile.R

@@ -1,98 +1,0 @@
-######## Western blot analysis analysis ####################
-
-# Filename: ./R/wester_analysis.R
-# Author: Daniel H
-# Notes:
-
-# Purpose: Analyze western blot data
-
-# Output: Model estimates
-
-######## -----------------------####################
-
-## Packages ####
-source("./R/libs.R")
-
-
-### Load data  round 1 ####
-
-western_data_round1 <- read_excel("./data/wetlab/western/tr010_western_round1.xlsx", na = "NA", sheet = "total_protein") %>%
-  filter(participant != "LADDER") %>%
-  dplyr::select(participant, ExtractionNR, round, gel, well, mean.gray1:total.protein2) %>%
-  inner_join(read_excel("./data/wetlab/tr010_mRNASamples_round1.xlsx", na = "NA") %>%
-               dplyr::select(participant, leg, time, ExtractionNR)) %>%
-  inner_join(read_excel("./data/wetlab/western/tr010_western_round1.xlsx", na = "NA", sheet = "ecl") %>%
-               dplyr::select(round, gel, well, target, signal)) %>%
-  mutate(gel = paste0(round,"_", gel), 
-         sample = paste0(participant, "_", ExtractionNR)) %>%
-  rowwise() %>%
-  ### memcode stain - background = total protein ### 
-  mutate(total.protein = mean(c(total.protein1, total.protein2)) - mean(c(mean.gray1, mean.gray2, mean.gray3))) %>%
-  ungroup() %>%
-  dplyr::select(participant,sample, leg, time, gel, well, target, total.protein, signal, round) %>%
-  group_by(gel) %>%
-  ### Normalizes the total protein stain per 
-  mutate(tp.factor = total.protein) %>%
-  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
-  dplyr::select(participant,sample, leg, cond, sex, time, gel, well, target, tp.factor, signal, round) %>%
-  ungroup() %>%
-  mutate(expression = signal / tp.factor) %>%
-  filter(target %in% c("t-s6", "t-UBF")) %>%
-  print()
-
-
-western_data_round2 <- read_excel("./data/wetlab/western/tr010_western_round2.xlsx", na = "NA", sheet = "total_protein") %>%
-
-  dplyr::select(participant, ser, num, round, gel, well, total.protein_1:meangray_4) %>%
-  mutate(num = as.character(num)) %>%
[…]

### 2021-12-07 R/western-analysis.R

@@ -1,0 +1,319 @@
+### Western blot analysis  ######
+
+# Load packages ###########
+
+library(tidyverse)
+library(brms)
+library(bayesplot)
+
+
+
+# Western blot data are compiled in western_compile.R and stored
+# gel information is included in the raw data but averaged over all gels
+# below. 
+
+
+western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
+
+# Re-format time variables to match comparison between control and 
+# experimental groups. The "tx" factor enables these comparisons. The 
+# tx factor is needed to avoid "rank deficiency" in the comparison.
+
+west.dat <- western_data %>%
+ filter(time %in% c("S0", "S1c", "S1", "S12", "postctrl", "post1w")) %>%
+  mutate(group = if_else(cond == "ctrl_leg", "con", "int"), 
+         detrain = if_else(time == "post1w", "detrain", "train"), 
+         time = if_else(time == "S0", "pre", 
+                        if_else(time == "S1c", "S1", 
+                                if_else(time %in% c("postctrl", "post1w", "S12"), "post", time))), 
+         time = factor(time, levels = c("pre", "S1", "post")), 
+         tx = paste(time, group, detrain, sep = "_"), 
+         tx = factor(tx, levels = c("pre_con_train",
+                                    "S1_con_train",
+                                    "post_con_train",
+                                    "pre_int_train",
+                                    "S1_int_train",
+                                    "post_int_train", 
+                                    "post_int_detrain")), 
+         sample2 = paste0(participant, leg, time)) %>%
+  group_by(participant, tx, target) %>%
+  summarise(expression = mean(expression, na.rm = TRUE)) %>%
+  mutate(log.expr = log(expression)) %>%
+  filter(!is.na(expression)) %>%
+  print()
+  
+
+s6.m1 <- brm(bf(log.expr ~ tx + (1|participant)),
+
+          warmup = 1000, # number of samples before sampling
+          iter = 4000,  # number of mcmc iterations 
+          cores = 4, # number of cores used in sampling
+          chains = 4, # number of chains
+          seed = 5, # seed for reproducible results
+          save_pars = save_pars(all = TRUE),
+      #    control = list(adapt_delta = 0.95), 
+          data = west.dat[west.dat$target == "t-s6",])
+
+
[…]

### 2021-12-07 R/western-compile.R

@@ -1,0 +1,207 @@
+######## Western blot analysis analysis ####################
+
+# Filename: ./R/wester_analysis.R
+# Author: Daniel H
+# Notes:
+
+# Purpose: Analyze western blot data
+
+# Output: Western blot data sets
+
+######## -----------------------####################
+
+## Packages ####
+source("./R/libs.R")
+
+
+### Load data  round 1 ####
+
+western_data_round1 <- read_excel("./data/wetlab/western/tr010_western_round1.xlsx", na = "NA", sheet = "total_protein") %>%
+  filter(participant != "LADDER") %>%
+  dplyr::select(participant, ExtractionNR, round, gel, well, mean.gray1:total.protein2) %>%
+  inner_join(read_excel("./data/wetlab/tr010_mRNASamples_round1.xlsx", na = "NA") %>%
+               dplyr::select(participant, leg, time, ExtractionNR)) %>%
+  inner_join(read_excel("./data/wetlab/western/tr010_western_round1.xlsx", na = "NA", sheet = "ecl") %>%
+               dplyr::select(round, gel, well, target, signal)) %>%
+  mutate(gel = paste0(round,"_", gel), 
+         sample = paste0(participant, "_", ExtractionNR)) %>%
+  rowwise() %>%
+  ### memcode stain - background = total protein ### 
+  mutate(total.protein = mean(c(total.protein1, total.protein2)) - mean(c(mean.gray1, mean.gray2, mean.gray3))) %>%
+  ungroup() %>%
+  dplyr::select(participant,sample, leg, time, gel, well, target, total.protein, signal, round) %>%
+  group_by(gel) %>%
+  ### Normalizes the total protein stain per 
+  mutate(tp.factor = total.protein) %>%
+  inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+  dplyr::select(participant,sample, leg, cond, sex, time, gel, well, target, tp.factor, signal, round) %>%
+  ungroup() %>%
+  mutate(expression = signal / tp.factor) %>%
+  filter(target %in% c("t-s6", "t-UBF")) %>%
+  print()
+
+
+western_data_round2 <- read_excel("./data/wetlab/western/tr010_western_round2.xlsx", na = "NA", sheet = "total_protein") %>%
+
+  dplyr::select(participant, ser, num, round, gel, well, total.protein_1:meangray_4) %>%
+  mutate(num = as.character(num)) %>%
+  inner_join(read_csv2("./data/written/extraction_numbers_round2.csv", na = "NA") %>%
[…]

### 2021-12-09 manuscript.Rmd
rewrite abstract

@@
 <div custom-style="abstract_style">
+
+**Aim**
+To describe the time course of ribosome accumulation in response to resistance training with constant and variable volume and relate RNA accumulation to total UBF levels and muscle growth. 
+**Methods**
+Participants allocated to a training group (n=11) performed unilateral knee extension with constant (6 sets per session) and variable volume (6, 3, and 9 sets in sessions 1-4, 5-8, and 9-12, respectively) allocated to either leg.
+Muscle biopsies (m.
+vastus lateralis) were sampled in the training group before and 48 hrs after the first session and 48 hrs after sessions 4, 5, 8, 9, and 12, and after eight days of de-training.
+Muscle samples were also obtained from a non-training control group (n=7) at baseline and after 48 hrs and 3-5 weeks.
+Muscle specimens were assayed for total RNA, ribosomal RNA (rRNA), ribosomal protein S6 (rpS6), and upstream binding factor (UBF) protein content.
+Muscle thickness (ultrasound) and strength (isokinetic and isometric knee extension) were measured before and after resistance training or control periods.
+**Results**
+Resistance training led to muscle growth (0.84 mm, 95% CI: [0.05, 1.64] vs.
+control) with concomitant increases in total RNA, rRNA, UBF and rpS6 protein.
+Total RNA increased most rapidly in response to the first four sessions (8.6% [5.5, 11.7] per session), followed by a plateau and peak values (49.5% [34.2, 66.5] above baseline) after session 8.
+UBF protein levels were associated with total RNA levels after controlling for time, and increases in total RNA levels predicted RT-induced muscle hypertrophy.
+After de-training, total RNA and rRNA decreased without muscle mass changes indicating halted biosynthesis of ribosomes.
+In conclusion, ribosomes accumulate in the initial phase of RT and is sensitive to training cessation.
+The rate of accumulation predicts muscle hypertrophy
 

### 2021-12-09 R/us-freq-bayes.R

@@ -171,14 +171,14 @@ m1.us_freq <- lmer(thickness ~  tx + (1|participant/leg),
 anova(m0.us_freq, m1.us_freq)
 # Favors inclusion of random intercept per leg. 
 
-em <- emmeans(m2.us_freq, specs = ~ time_pp * group  + time_pp:group:detrain)
+
 
 # Combine data for plotting 
 
 # Emmeans can be used as the contrats specified are combinations of factors. 
 # emmeans::emmeans is used to create a emmGrid object:
 
-em <- emmeans(m1.us_freq, specs =  ~ tx)
+em <- emmeans::emmeans(m1.us_freq, specs =  ~ tx)
 
 # To create contrasts, each levels can be specified by a indicator vector. 
 

### 2021-12-10 manuscript.Rmd
last updates before submission

--- 2021-12-07 manuscript.Rmd
@@
 ---
-title: "Ribosome accumulation during early phase resistance training"
-author:
-  - Daniel Hammarström^1,2,£^*
-  - Sjur J. Øfsteng^2,£^
-  - Nicolai B. Jacobsen^1^
-  - Krister B. Flobergseter^1^
-  - Bent R. Rønnestad^1^
-  - Stian Ellefsen^1,3^
+title: "Ribosome accumulation during early phase resistance training in humans"
+author: Daniel Hammarström^1,2,£^*, Sjur J. Øfsteng^2,£^*, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^, Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
 date: ''
 output:
-  pdf_document:
-    latex_engine: xelatex
-  word_document:
+  officedown::rdocx_document:
     reference_docx: resources/manuscript_word_template.docx
@@
 source("./R/libs.R")
-library(knitr); library(kableExtra)
+library(knitr); library(kableExtra); library(officedown)
 # Settings for chunk options
@@
 Daniel Hammarström, Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway.
-Email: daniel.hammarstrom@inn.no  </div>
+Email: daniel.hammarstrom@inn.no
+
+Sjur Johansen Øfsteng, Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway.
+Email: sjur.johansen.ofsteng@inn.no
+</div>
 
 <div custom-style="runningtitle_style">**Running title:** Ribosome biogenesis in early phase resistance training</div>
+
+\pagebreak
 
@@
 
-Increased ribosomal density is a phenomenon observed in response to resistance training (RT) supporting elevated.
-A time course of RT-induced accumulation of ribosomes has not been described.
-We mapped markers of ribosomal biogenesis (total RNA, ribosomal RNA (rRNA), ribosomal protein S6 (rpS6), and upstream binding factor (UBF) protein content) in response to twelve RT sessions.
-Additionally, we investigated the effect of fluctuations in training volume on these markers and the relationship between RNA accumulation and muscle growth measured as muscle thickness (ultrasound).
-Eighteen participants were allocated to either a training (TRAIN, *n*=11) or a control (CTRL, *n*=7) group.
-In TRAIN, participants performed unilateral knee extension with either constant (CONST, 6 sets) or a variable volume (VAR, 6, 3, and 9 sets in sessions 1-4, 5-8, and 9-12, respectively).
-Muscle biopsies were sampled from *m vastus lateralis* in TRAIN before and 48 hrs after the first session and 48 hrs after sessions 4, 5, 8, 9, and 12, and after eight days of de-training.
-Control biopsies were sampled in CTRL at baseline and after 48 hrs and 3-5 weeks.
-RT led to muscle growth in TRAIN compared to CTRL and concomitant increases in total RNA, rRNA, including precursor rRNA as well as UBF and rpS6 protein through gradual increase throughout the training period, confirmed in comparison to CTRL.
-Total RNA increased in a curve-linear fashion, most rapidly in response to the first four sessions  (8.6%, 95% CI: [5.5, 11.7] per session), followed by a plateau and peak values (49.5% [34.2, 66.5] above baseline) after Session 8.
-UBF protein levels explained total RNA levels after controlling for time, and increases in total RNA levels predicted RT-induced muscle hypertrophy.
-After de-training, total RNA and specific rRNA species decreased without muscle mass changes indicating reduced concentrations and biosynthesis of ribosomes.
-These results underline a determinant role for ribosomal biogenesis in RT-induced muscle hypertrophy and that ribosomal biogenesis is sensitive to training cessation.
-
-**Keywords:** Resistance training, ribosome biogenesis
+```{r}
+
+### Data in abstract ###########
+
+# Ultra sound #################
+
+# Model data
+comp_us <- readRDS("./data/derivedData/us-freq-bayes/comp_us_bayes.RDS") %>%
+   dplyr::select(time_group, estimate, lower.CL, upper.CL)
+
+# Express as percentage of average
+library(tidyverse); library(readxl)
+
+us_data <- rbind(read_excel("./data/ultrasound/ultrasound_data.xlsx") %>%
+                   inner_join(read_csv("./data/ultrasound/ultrasound_codekey.csv")) %>%
+                   mutate(leg = gsub("VL", "", leg)) %>%
+                   inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+                   dplyr::select(participant, time, leg, sex, cond, code, length) %>%
+                   mutate(group = if_else(participant %in% paste("P", 1:7, sep = ""), "experiment", "control")) %>%
+                   group_by(participant, time, leg, sex, cond, group) %>%
+                   summarise(thickness = mean(length, na.rm = TRUE)) %>%
+                   ungroup(), 
+                 read_excel("./data/ultrasound/ultrasound_data_2019.xlsx") %>%
+                   inner_join(read_csv("./data/ultrasound/ultrasound_codekey_2019.csv")) %>%
+                   mutate(leg = gsub("VL", "", leg)) %>%
+                   inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
+                   dplyr::select(participant, time, leg, sex, cond, code, length) %>%
+                   mutate(group = if_else(participant %in% paste("P", c(1:7, 19:23), sep = ""), "experiment", "control")) %>%
+                   group_by(participant, time, leg, sex, cond, group) %>%
+                   summarise(thickness = mean(length, na.rm = TRUE)) %>%
+                   ungroup()) %>%
+
+  mutate(time_pp = if_else(time == "post1w", "post", time),
+         # The de-training period get its own coefficient
+         detrain = if_else(time == "post1w" & group == "experiment", "detrain", "train"),
+         # The effect of de training will be added to the model --within-- the intervention group 
+         detrain = factor(detrain, levels = c("train", "detrain")), 
+         time = factor(time, levels = c("pre", "post", "post1w")), 
+         time_pp = factor(time_pp, levels = c("pre", "post"))) %>%
+  group_by(time) %>%
+  summarise(m = mean(thickness, na.rm = TRUE)) %>%
+  data.frame()
+
+hypertrophy <- paste0(sprintf("%.1f", 100 * (comp_us[4,2] / us_data[1,2])), 
+       "% over baseline values, 95% CrI: [",
+       sprintf("%.1f", 100 * (comp_us[4,3] / us_data[1,2])), 
+       ", ",
+       sprintf("%.1f", 100 * (comp_us[4,4] / us_data[1,2])),
+       "]" )
+
+## Total RNA increase ######################
+
+# Segmented model analysis in text 
+# Read the segmented model
+
+library(brms)
+
+tc.m2 <- readRDS("./data/derivedData/total-rna-analysis/tc.m2.RDS")
+
+time1.hyp <- hypothesis(tc.m2, "time1 = 0")
+
+time2.hyp <- hypothesis(tc.m2, c("time2 + time1 = 0")) # The slope (on log scale) between session 4 and 8
+time3.hyp <- hypothesis(tc.m2, c("time2 + time1 +time3 = 0")) # The slope on log scale between session 8 and 12
+
+slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,2]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,2]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,2]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+# Estimate changes to end of periods
+
+sessions <- hypothesis(tc.m2, c("time1 * 4 = 0", 
+                                     "(time1 * 8) + (time2 * 4)= 0", 
+                                     "(time1 * 12) + (time2 * 8) + (time3 * 4)= 0"))
+
+## Effect of detraining
+
+detrain.hyp <- hypothesis(tc.m2, "detraindetrain = 0")
+
+detrain_rna <- paste0(sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+fourth.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,5]) - 1)),
+                   "]")
+
+eight.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,5]) - 1)),
+                   "]")
+
+twelth.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,5]) - 1)),
+                   "]")
+
+# UBF effect on total RNA ############
+comb.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")
+
+ubf_model <- paste0(sprintf("%.1f", (exp(summary(comb.m1)$fixed[2,1])-1) * 100), 
+       "% [",
+       sprintf("%.1f", (exp(summary(comb.m1)$fixed[2,3])-1) * 100), 
+       ", ",
+       sprintf("%.1f", (exp(summary(comb.m1)$fixed[2,4])-1) * 100),
+       "]" )
+
+# Rate of increase in total rna on muscle hypertrophy ################
+
+hyp.pre.m3 <- readRDS("./data/derivedData/ubf-tot-rna-model/hyp_pre_m3.RDS")
+
+rna_hypertrophy <- paste0(sprintf("%.1f", summary(hyp.pre.m3)$fixed[5,1]), 
+       " mm [",
+       sprintf("%.1f", summary(hyp.pre.m3)$fixed[5,3]), 
+       ", ",
+       sprintf("%.1f", summary(hyp.pre.m3)$fixed[5,4]),
+       "]" )
+
+```
+
+**Aim:** To describe the time course of ribosome accumulation during resistance training, its relation to training volume, UBF protein levels, and muscle growth.
+
+**Methods:** A training group (n=11) performed unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg.
+Vastus lateralis biopsies obtained at rest and 48 hrs after selected training sessions and after eight days of de-training and from a non-training control group (n=7) were assayed for ribosome abundance and biogenesis markers.
+In addition, muscle thickness was measured before and after training and control periods.
+
+**Results:** Training led to muscle growth (`r hypertrophy` vs.
+control) with concomitant increases in total RNA, rRNA, UBF, and rpS6 with no differences between volume conditions.
+Total RNA increased most rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline).
+After controlling for time, total RNA abundance was associated with UBF protein levels (`r ubf_model` per unit UBF), and the rate of increase in total RNA levels predicted hypertrophy (`r rna_hypertrophy` per %-point increase in total RNA per session).
+After de-training, total RNA decreased (`r detrain_rna`) without muscle mass changes indicating halted biosynthesis of ribosomes.
+
+**Conclusion:** Ribosomes accumulate in the initial phase of resistance training with abundances sensitive to training cessation and associated with UBF protein levels.
+The average accumulation rate predicts muscle training-induced hypertrophy.
+
+**Keywords:** muscle hypertrophy, resistance training, ribosome biogenesis
 
@@
 In addition to addressing these hypotheses we aimed to relate RNA accumulation to total UBF levels and muscle growth.
-
-# Methods
+We utilized a within-participant unilateral training model where one leg was assigned constant volume (CONST, 6 sets per session) and the contra lateral leg variable volume (VAR, 6, 3, and 9 sets in sessions 1-4, 5-8, and 9-12, respectively).
+Effects of training *per se* were assessed by comparison to a non-training control group (CTRL).
+
+# Results
+```{r training-data}
+
+m2 <- readRDS("./data/derivedData/training-data/m2.RDS")
+# Model containing interaction 
+# readRDS("./data/derivedData/training-data/m1.RDS")
+
+# Create a hypothesis and convert to percentages
+
+h <- hypothesis(m2, c("(weekW2 / Intercept) * 100 = 0",
+                 "(weekW3 / Intercept) * 100 = 0"))$hypothesis
+
+load.inr.w2 <- paste0(sprintf("%.0f",h[1, 2]), 
+                      "%, 95% credible interval (CI): [",
+                      sprintf("%.0f",h[1, 4]), 
+                      ", ",
+                      sprintf("%.0f",h[1, 5]), 
+                      "]")
+load.inr.w3 <- paste0(sprintf("%.0f",h[2, 2]), 
+                      "% [",
+                      sprintf("%.0f",h[2, 4]), 
+                      ", ",
+                      sprintf("%.0f",h[2, 5]), 
+                      "]")
+
+# Approximate percentages increase in strength and muscle mass
+
+m1.isom <- readRDS("./data/derivedData/strength-bayes/m1.isom.RDS") 
+m1.isok <- readRDS("./data/derivedData/strength-bayes/m1.isok.RDS")
+
+# Change in percentage-points between groups 
+isok.h <- hypothesis(m1.isok, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"
+                                ))$hypothesis
+
+isom.h <- hypothesis(m1.isom, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"
+                                ))$hypothesis
+
+isok_pre_post <-  paste0(sprintf("%.1f",isok.h[1, 2]), 
+                      "% [",
+                      sprintf("%.1f",isok.h[1, 4]), 
+                      ", ",
+                      sprintf("%.1f",isok.h[1, 5]), 
+                      "]")
+
+isok_pre_post_aprox <- paste0(sprintf("%.1f",isok.h[1, 2]))
+
+isok_pre_detrain <-  paste0(sprintf("%.1f",isok.h[2, 2]), 
+                      "% [",
+                      sprintf("%.1f",isok.h[2, 4]), 
+                      ", ",
+                      sprintf("%.1f",isok.h[2, 5]), 
+                      "]")
+
+isok_pre_detrain_approx <- paste0(sprintf("%.1f",isok.h[2, 2]))
+
+isom_pre_post <-  paste0(sprintf("%.1f",isom.h[1, 2]), 
+                      "% [",
+                      sprintf("%.1f",isom.h[1, 4]), 
+                      ", ",
+                      sprintf("%.1f",isom.h[1, 5]), 
+                      "]")
+isom_pre_detrain <-  paste0(sprintf("%.1f",isom.h[2, 2]), 
+                      "% [",
+                      sprintf("%.1f",isom.h[2, 4]), 
+                      ", ",
+                      sprintf("%.1f",isom.h[2, 5]), 
+                      "]")
+
+isom_pre_post_approx <- paste0(sprintf("%.1f",isom.h[1, 2]))
+isom_pre_detrain_approx <- paste0(sprintf("%.1f",isom.h[2, 2]))
+
+## Muscle thickness
+
+m1.us <- readRDS("./data/derivedData/us-freq-bayes/m1us.RDS")
+
+us.h <- hypothesis(m1.us, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"))$hypothesis
+
+us_pre_post_approx <- paste0(sprintf("%.1f",us.h[1, 2]))
+us_pre_detrain_approx <- paste0(sprintf("%.1f",us.h[2, 2]))
+
+```
+
+All participants allocated to TRAIN successfully completed their prescribed RT on both legs, with the two volume conditions resulting in diverging volume profiles (load &#215; repetitions) over the course of the study (Figure 1B).
+Exercise intensities (resistance at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with each block consisting of four training sessions.
+Concomitantly, in TRAIN, isokinetic strength and thickness of *m.
+vastus lateralis* increased from baseline to after Session 12 compared to CTRL (isokinetic strength $\sim$ `r isok_pre_post_aprox`%-point difference; muscle thickness $\sim$ `r us_pre_post_approx`%-point difference, Figure 1C and D), a difference that was sustained to after eight days of de-training ($\sim$ `r isok_pre_detrain_approx`%-point and $\sim$ `r us_pre_detrain_approx`%-point difference in change in isokinetic strength and muscle thickness, respectively; Figure 1C and D).
+Isometric strength showed the same general pattern to after Session 12 ($\sim$ `r us_pre_detrain_approx`%-point difference), though with considerably larger degrees of uncertainty, as indicated by wider 95% CI normalization compared to CTRL after de-training ($\sim$ `r isom_pre_detrain_approx`%-point; Figure 1C).
+No differences were observed between volume conditions for either strength or muscle thickness.
+
+```{r western-data}
+segmented_protein_results <- readRDS("./data/derivedData/western-analysis/segmented_results.RDS")
+
+s6_slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[6,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[6,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[6,7]) - 1)),
+                   "]")
+
+s6_slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[7,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[7,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[7,7]) - 1)),
+                   "]")
+
+s6_slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[8,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[8,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[8,7]) - 1)),
+                   "]")
+
+ubf_slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[14,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[14,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[14,7]) - 1)),
+                   "]")
+
+ubf_slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[15,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[15,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[15,7]) - 1)),
+                   "]")
+
+ubf_slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[16,4]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[16,6]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(segmented_protein_results[16,7]) - 1)),
+                   "]")
+
+s6.m3 <- readRDS("./data/derivedData/western-analysis/s6_detrain.RDS")
+ubf.m3 <- readRDS("./data/derivedData/western-analysis/ubf_detrain.RDS")
+
+ubf.sum <- summary(ubf.m3)$fixed
+
+var_s12 <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[3,1]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[3,3]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[3,4]) - 1)),
+                   "]")
+
+const_detrain <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[2,1]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[2,3]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[2,4]) - 1)),
+                   "]")
+
+var_const_interaction <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[4,1]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[4,3]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(ubf.sum[4,4]) - 1)),
+                   "]")
+
+h <- hypothesis(ubf.m3, "condvar + timepost1w:condvar = 0")$hypothesis %>%
+  data.frame()
+
+var_change <- paste0(sprintf("%.1f", 100 * (exp(h[1,2]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(h[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(h[1,5]) - 1)),
+                   "]")
+
+```
+
+For both ribosomal protein S6 (rpS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
+This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12 and elevated after eight days of rest (Figure 2A and B), however with no robust differences being observed after the first training session (48 hrs).
+Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`).
+After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
+For rpS6, de-training did not affect protein levels, which remained similar between volume conditions.
+At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point.
+No differences were observed between volume conditions for either transcripts (Figure 2D and E).
+
+```{r total-rna}
+
+# Segmented model analysis in text 
+# Read the segmented model 
+tc.m2 <- readRDS("./data/derivedData/total-rna-analysis/tc.m2.RDS")
+
+time1.hyp <- hypothesis(tc.m2, "time1 = 0")
+time2.hyp <- hypothesis(tc.m2, c("time2 + time1 = 0")) # The slope (on log scale) between session 4 and 8
+time3.hyp <- hypothesis(tc.m2, c("time2 + time1 +time3 = 0")) # The slope on log scale between session 8 and 12
+
+slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,2]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,2]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,2]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+# Estimate changes to end of periods
+
+sessions <- hypothesis(tc.m2, c("time1 * 4 = 0", 
+                                     "(time1 * 8) + (time2 * 4)= 0", 
+                                     "(time1 * 12) + (time2 * 8) + (time3 * 4)= 0"))
+
+## Effect of detraining
+
+detrain.hyp <- hypothesis(tc.m2, "detraindetrain = 0")
+
+detrain_rna <- paste0(sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,2]) - 1)),
+                           "%, [",
+                      sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,5]) - 1)),
+                   "]")
+
+fourth.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,5]) - 1)),
+                   "]")
+
+eight.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,5]) - 1)),
+                   "]")
+
+twelth.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,2]) - 1)),
+                           "% [",
+                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,4]) - 1)), 
+                   ", ",
+                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,5]) - 1)),
+                   "]")
+
+```
+
+A single session of RT (Session 1) led to robust increases in precursor ribosomal RNA (pre-rRNA 47S ETS and 45S ETS) abundance per unit tissue weight, measured as changes from baseline to 48 hrs after exercise within TRAIN (Figure 3B), as well as compared to CTRL (Figure 3C).
+After Session 1, pre-rRNA 47S ETS and 45S ETS levels remained at similar levels at all measured time-points in TRAIN (Figure 3D), confirmed in comparison to CTRL after Session 12 (Figure 3C).
+Other rRNA transcripts showed increases in response to training with slightly different temporal patterns with exception of rRNA 5S which did not change and rRNA 5.8 which tended to follow other mature transcript spliced from pre-rRNA 45S ETS, but without statistical robustness (Figure 3C and D).
+After eight days of rest, 18S and 28S remained at elevated levels compared to CTRL (Figure 3B).
+This general pattern of rRNA expression was reflected by total RNA abundance per unit tissue weight, which increased robustly and steadily in TRAIN throughout the initial part of the intervention (Figure 3E and G), leading to robust increase compared to CTRL after Session 12 (Figure 3F), followed by decreased levels after de-training (`r detrain_rna`).
+For both rRNA expression and total RNA levels, the training-associated increases in abundances occurred predominately during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from sessions four to eight (`r slope4_8` increase per session) and from sessions eight to twelve `r slope8_12`, corresponding to `r fourth.session`, `r eight.session` and `r twelth.session` increases from baseline to 48 hrs after session four, eight and twelve, respectively.
+In TRAIN, the two volume conditions led to similar changes for most variables (Figure 3D and G), with 45S ETS abundance only showing differential expression, evident as robustly higher levels in VAR compared to CONST after the 12th session (Figure 3D), coinciding with the increased training volume towards the end of the intervention for this condition.
+
+```{r combined-analysis}
+
+# Save model 
+comb.m1 <- summary(readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS"))$fixed
+
+ubf_cor_tot_rna <- paste0(sprintf("%.1f", 100 * (exp(comb.m1[2,1]) - 1)),
+                           "% [",
+                   sprintf("%.1f", 100 * (exp(comb.m1[2,3]) - 1)), 
+                   ", ",
+                   sprintf("%.1f", 100 * (exp(comb.m1[2,4]) - 1)),
+                   "]")
+
+```
+
+In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coinciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 1).
+In contrast, no evidence was found for a relationship between total RNA and rpS6 protein levels (Table 1).
+
+In TRAIN, there was a robust positive relationship between rates of increase in total RNA in response to training and muscle growth measured as increases in *m.
+vastus lateralis* thickness (Table 2, Figure 4A), with changes in total RNA over the course of the training intervention being estimated in each leg using a regression model containing number of sessions as the independent variable.
+Conversely, there was a tendency towards a negative relationship between average total RNA levels at Session 6 and changes in muscle thickness (Table 2, Figure 4B), with the average total RNA levels estimated as the predicted value at Session 6 (estimated as the intercept-term) from the model used to estimate the rate of total RNA increase per session.
+
+To assess the robustness of the model for predicting muscle growth, individual relationships between sessions and total RNA levels (Figure 4C) were recalculated after removal of single data points from each participant.
+The model predicting muscle growth was refitted using new estimates of changes in total RNA abundances and increases thereof per session.
+Each refitted model resulted in slightly different estimates (displayed as means and 95% CI in Figure 4D).
+No single data point influenced the results in any meaningful way.
+Next we assessed the robustness by iteratively removing one participant from the data set, similarly this showed that estimates of the effect of total RNA increase on muscle growth was robust but the effect of average total RNA estimates were more variable (e.g.
+Participant 11 and 3 in Figure 4D).
+
+# Discussion
+Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
+These markers accumulated progressively during the initial part of the intervention before it leveled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
+This increase in total RNA was interconnected with increases in UBF protein abundance, suggesting UBF levels *per se* to play a role in regulation of rRNA transcription regulation in response to RT.
+Total RNA increases was not affected by weekly fluctuations in training volume, however, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
+Finally, individual rates of increases in total RNA abundance predicted the magnitude of muscle growth, confirming the likely link between ribosomal biogenesis and muscle protein accretion 
+[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
+@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
+and muscle hypertrophy
+[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
+@RN1755; <!-- Stec Ribosome bio higher in high responders -->
+@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->
+
+Total RNA seems to be a valid proxy marker of ribosomal density, as most of the RNA is assumed to be ribosomal RNA
+[@RN2223],<!-- Young  -->
+which in turn is a valid marker of translational capacity
+[@RN2145].
+Several studies have shown that total RNA content is altered by RT
+<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
+[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
+@RN2180;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --> 
+@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
+@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
+@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
+@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
+@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
+@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
+@RN1897<!-- 11% 36 sessions young -->],
+as was also the case in the present data set.
+However, the time course of total RNA/rRNA changes in response to RT has so far remained speculative, with no study investigating responses to prolonged interventions with multiple sampling time points.
+In the present data, RT led to a clear session-to-session increase in total RNA per unit tissue weight in response to the first four session, whereupon the changes gradually leveled out before peaking after the 8^th^ session, with the peak increase from baseline being $\sim$ 50%, defining an accumulation phase.
+This corroborates well with previous suggestions of peak values being reached within four to nine sessions in young males and females
+[@RN1809;
+@RN2358], and may be essential for preparing muscle fibers for subsequent growth
+[@RN1809;<!-- Correlation between increase and muscle growth -->
+@RN1644; <!--Figueiredo correlation between incr and muscle hypertrophy -->
+@RN2358].<!-- Early accumulation of RNA predicts volume depdent growth -->
+After the 8^th^ session, no meaningful increase or decrease were observed for total RNA or rRNA content within the training period, suggesting a plateau phase with attenuated net synthesis of novel ribosomes.
+Within this last part of the intervention, synthesis of novel rRNA still seemed to be elevated per weight unit muscle tissue compared to baseline, as suggested by sustained elevation of pre-rRNA transcripts, coinciding with peak values of UBF protein levels.
+This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth [@RN2142].<!-- Figueiredo review details the concept of rna concentration -->
+
+The observed rates of RNA accumulation over the entirety of the intervention were found to be a determinant of changes in muscle thickness (after controlling for average total RNA levels).
+Individuals with higher rates of accumulation showed larger accretion of muscle mass.
+This supports the notion that ribosomal biogenesis is an important determinant of RT-induced muscle hypertrophy, with previous studies showing that increases in total RNA are positively correlated with increases in muscle mass
+[@RN1644;<!-- Figueriedo  -->
+@RN1897;
+@RN2055],<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
+<!-- Reidy correlation between total RNA and muscle hypertrophy-->
+differs between individuals displaying low vs. high levels of muscle hypertrophy in response to RT
+[@RN1755]<!-- Stec 2016  -->
+an contribute to explain RT volume-dependent changes in muscle mass and strength
+[@RN2358].
+In addition, supression of ribosomal biogenesis in *in vitro* models leads to halted muscle cellular growth in some 
+[@RN1632;<!-- Nader 2005 blockage of mTOR halts ribo bio and growth -->
+@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
+@RN1755]<!-- Stec specific pol I inhibition -->
+but not all studies
+[@RN1929].<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
+Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA was instead associated with a tendency towards lowered muscle growth.
+Overall, the rate of increases in ribosomal density thus seems to be a better predictor of individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
+Interestingly, the interaction between rRNA synthesis rate and muscle mass accretion (but not between ribosomal content and muscle mass accretion) may shed light on observed differences in muscular responses to RT between young and old individuals.
+Whereas aged muscle display higher levels of total RNA at rest 
+[@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
+they show reduced changes in total RNA levels in response to RT
+[@RN1809],<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
+potentially explaining their alleged poorer overall hypertrophic responses [@RN1809].
+Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
+Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.
+
+<!-- rpS6 dissconect from total RNA levels. -->
+In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6.
+Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period.
+Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
+[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
+but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
+[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
+Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for.
+Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT.
+Additionally, ribosomal proteins may have extra-ribosomal functions
+affecting their expression
+[@RN2588].<!-- Warner 2009 General description of   -->
+
+UBF levels robustly explained total RNA levels over the entire course of the intervention.
+These analyses were done with the number of sessions accounted for, allowing unbiased estimates.
+Unrealistically strong relationships could have been otherwise expected as both the dependent variable (total RNA) and the covariate (UBF levels) varies with the number of sessions.
+From a mechanistic perspective, UBF is an important transcription factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA Pol I
+[@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
+Activation of UBF is controlled by the mechanosensitive mTOR pathway, and rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
+[@RN1632; <!-- Nader rapamycin block UBF availability  -->
+@RN2564].<!-- Hannan 2003 -->
+<!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1.
+The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF -->
+Evidence from human exercise studies confirms training-induced activation of UBF through phosphorylation
+[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
+@RN1644].<!-- Figueriedo ubf increase total and active form (ser388) --> 
+In addition to exercise-induced activation of UBF, mechanical loading also leads to increased levels of total UBF
+[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
+@RN1644]. Increases in UBF was determined to be rapamycin insensitive after synergist ablation in mice
+[@RN1072]<!-- Goodman 2011 UBF increase due to synergist ablation is rapamycin insensitive https://doi.org/10.1113/jphysiol.2011.218255 --> 
+pointing to an effect observed in cell models where c-Myc induces UBF mRNA transcription [@RN2565].<!-- Pootinga 2004 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
+Interestingly the availability of UBF *per se* has been shown to regulate rRNA transcription
+[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->  through control of rDNA gene activity
+[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->
+Together with our observations, this underlines the importance of UBF as a regulator of RT-induced ribosomal biogenesis.
+
+After eight days of de-training, total RNA and rRNA levels per weight unit muscle tissue returned toward baseline levels, though without concomitant reversal of muscle thickness, which remained at elevate levels.
+This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
+This supports the idea that ribosomal biogenesis is a cellular activity on demand, possibly relating to its relative expense [@RN1920] also in muscle tissue.
+Based on this notion, and the fact that RT volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including induction of c-Myc expression, mTOR activation
+[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
+@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
+@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
+subsequent total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
+and subsequent training outcomes [@RN1767; @RN2358], we hypothesized that fluctuations in training volume would be reflected in markers of ribosomal biogenesis.
+When comparing VAR to CONST in the present study we found only one part of the pre-rRNA, 45S ETS, to be differentially expressed and only so after Session 12 in favor of VAR together with a tendency towards rescued UBF levels after de-training in response to increased volume in the VAR but not CONST protocol.
+These observations do not give support to a clear effect of fluctuations in training volume on total RNA levels or rRNA expression within a relatively short and training-intensive intervention, though it should be noted that the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume.
+Indeed, both training protocols utilized in the present study increased muscle strength and induced muscle hypertrophy to a similar degree.
+From a general perspective, albeit volume is an important determinant of increases in muscle strength and mass
+[@RN1767;<!-- schoenfeld 2016 dose response  -->
+@RN2492],<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
+differences in organization of training loads is likely of minor importance when training volumes are equated over time
+[@RN2572].<!-- Grgic 2017 volume equated periodization schemes led to similar muscle hyeprtrophy -->
+It is important to note that RT in the current study was performed with the same volume in the first four sessions, something that could have been more than enough to maximize rRNA transcription in previously untrained individuals.
+This is supported by the observation that pre-rRNA increased rapidly initially in both protocols with minimal changes in response to subsequent sessions, regardless of exercise volume.
+The CONST protocol in the present study corresponded to volumes used in the moderate volume condition in a previous study from our lab (three sets in two exercises activating knee extensor muscles) [@RN2358].
+There, higher levels of total RNA were observed after four sessions in the moderate compared to a low volume protocol [@RN2358].
+Interestingly, using a progressive volume protocol in well-trained participants, increases in total RNA have been reported throughout six weeks of training
+[@RN2225].<!-- Haun total RNA increase througout a 6 week intervention -->
+Altough this observation was done in well-trained participants performing a high volume protocol without a control condition with constant volume, compared to constant volume protocols
+[@RN1809;<!-- Brook 2016 -->
+@RN2358; and the present study],<!-- Hammarström 2020 -->
+progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
+
+In conclusion, RT-induced ribosome accumulation reached peak values in the initial phase of RT (8 sessions) and was interconnected with increases in UBF protein levels.
+The rate of total RNA accumulation predicted RT-induced muscle hypertrophy.
+Fluctuations in training volume did not transfer to fluctuations in ribosomal biogenesis, but training cessation led to decreased ribosomal content.
+
+<!-- NOTES -->
+<!-- UBF and S6 protein levels may be differently regulated in men and rats as more consecutive bouts of electrical stimulation in rats does not prtoduce the same pattern seen in our study 
+ Kotani 2019 10.1152/japplphysiol.00665.2018 increase in t UBF at 6 h post ex but not 48 despite 2 weeks training (3 x week).
+S6 increase at 48 h and 6 bouts but return to sed values at 18 bouts.
+18 and 28 S increased over sed at 48 h post ex with no further increase beyond 3 bouts 45S maximal at three bouts -->
+
+# Materials and Methods
 
@@
 Exclusion criteria were consumption of dietary supplements or medication with known effects on muscle metabolism, injuries causing impaired strength and/or affecting their ability to perform RT, symptoms or history of disease, and known adverse reactions to local anesthetics.
-Participants were allocated to either an training group (TRAIN, *n* = 11) or a non-training control group (CTRL, *n* = 8; see Table 1 for participant characteristics; see Figure 1A for overview of the intervention).
+Participants were allocated to either an training group (TRAIN, *n* = 11) or a non-training control group (CTRL, *n* = 8; see Table 3 for participant characteristics; see Figure 1A for overview of the intervention).
 TRAIN performed a 12 session RT protocol lasting for 3-4 weeks, consisting of 10 repetition maximum (RM) unilateral knee-extension, with the two legs conducting RT with different volume profiles, allowing within-participant comparison of the effects of volume regimes.
@@
 
-```{r table1}
-
-# Read dxa data
-
-dxa_whole <- read_excel("./data/tr010_dxa_standarddefinitions.xlsx") %>%
-  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:`HeleVev%fett`) %>%
-  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:`HeleVev%fett`) %>%
-  mutate(value = gsub("kg", "", value), 
-         value = gsub("[[:space:]]", "", value, fixed = FALSE), 
-         value = as.numeric(value)) %>%
-
-  separate(Etternavn, into = c("Study", "participant")) %>%
-  dplyr::select(participant, Date = Målingsdato, variable, value) %>%
-  filter(!(participant %in% c("P08", "P20", "P17"))) %>%
-  group_by(participant) %>%
-  mutate(time = if_else(Date == min(Date), "S0", 
-                        if_else(Date == max(Date), "post1w", 
-                                "S12")), 
-         leg = if_else(variable == "BeinhøyreMagermasse", "R", "L")) %>%
-  filter(variable %in% c("HeleBeinmasse", 
-                         "HeleFettmasse", 
-                         "HeleVevmasse", 
-                         "HeleMagermasse" ,
-                         "Vekt", 
-                         "HeleTotalmasse")) %>%
-  pivot_wider(names_from = variable, values_from = value) %>%
-  mutate(fat_rel = 100*(HeleFettmasse/HeleTotalmasse)) %>%
-  dplyr::select(participant, time, Vekt,HeleTotalmasse, fat_rel) %>%
-  filter(time == "S0")
-
-# DXA weight stature
-
-read_excel("./data/tr010_age_participant.xlsx") %>%
-  inner_join(dxa_whole) %>%
-  mutate(age = difftime(recruited, dob)/365.2425, 
-         bmi = weight/(height/100)^2) %>%
-  dplyr::select(participant, group, height, weight,age, sex,bmi, fat_rel) %>%
-
-  group_by(group, sex) %>%
-  summarise(n = n(), 
-            Age = mean(age, na.rm = TRUE), 
-            age.sd = sd(age, na.rm = TRUE),
-            mass = mean(weight), 
-            mass.sd = sd(weight),
-            stature = mean(height),
-            stature.sd = sd(height), 
-            BMI = mean(bmi), 
-            bmi.sd = sd(bmi), 
-            relfat = mean(fat_rel), 
-            relfat.sd = mean(fat_rel) ) %>%
-  ungroup() %>%
-  mutate(n = as.character(n), 
-         age = paste0(sprintf("%.1f", round(Age, 1)), " (", sprintf("%.1f", round(age.sd, 1)), ")"),
-         mass = paste0(sprintf("%.1f", round(mass, 1)), " (", sprintf("%.1f", round(mass.sd, 1)), ")"),
-         stature = paste0(sprintf("%.1f", round(stature, 1)), " (", sprintf("%.1f", round(stature.sd, 1)), ")"),
-         BMI = paste0(sprintf("%.1f", round(BMI, 1)), " (", sprintf("%.1f", round(bmi.sd, 1)), ")"),
-         fat = paste0(sprintf("%.1f", round(relfat, 1)), " (", sprintf("%.1f", round(relfat.sd, 1)), ")"))%>%
-  dplyr::select(group, sex, n, age, mass, stature, BMI, fat) %>%
-  pivot_longer(names_to = "variable", values_to = "values", n:fat) %>%
-  pivot_wider(names_from = c(sex,group), values_from = values) %>%
-
-  mutate(variable =  factor(variable,
-                          levels = c("n", "age", "mass", "stature", "BMI", "fat"), 
-                          labels = c("n", "Age (yrs)", "Body mass, (kg)", "Stature (cm)", "Body mass index (kg/m\\textsuperscript{2})", "Body fat (\\%)"))) %>%
-
-  dplyr::select(variable, F_experiment, M_experiment, F_control, M_control) %>%
-
-  kable(caption = "Participant characteristics.", 
-        booktabs = TRUE, 
-        col.names = c("", "Female", "Male", "Female", "Male"),
-        escape = FALSE,
-        format = "latex") %>%
-  add_header_above(c(" ", "Experimental group" = 2, "Control group" = 2))
-
-```
+All participants gave their informed written informed consent prior to data collection.
+The study was conducted according to the Declaration of Helsinki, approved by the local ethics committee (no.
+2017-10-23) and the Norwegian center for research data (ref: 51549/3/AH), and pre-registered (DOI: 10.17605/OSF.IO/WA96Y).
 
@@
 Primers were further evaluated by agarose gel electrophoresis which confirmed primer sizes and non-template control experiments confirming no amplification without template.
-Primer sequences and their respective average performances are shown in Table 2.
+Primer sequences and their respective average performances are shown in Table 4.
 ```{r check-primer-sizes, include = FALSE}
@@
 
-```{r primerstable}
-# Read qpcr data
-qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")
-
-primer_performance <- qdat %>%
-  group_by(target) %>%
-  summarise(mean.cq = mean(cq, na.rm = TRUE), 
-            sd.cq = sd(cq, na.rm = TRUE), 
-            mean.eff = mean(eff))
-
-merge_custom <- function(ft, x, columns){
-  z <- rle(x)
-  rows_at <- cumsum(z$lengths) - z$lengths + 1
-
-  for(i in seq_along(rows_at)){
-    for(j in columns)
-      ft <- merge_at(x = ft, i = seq( rows_at[i], rows_at[i] + z$lengths[i] - 1), j = j)
-  }
-
-  ft
-}
-
-# Read primer data 
-primer_table <- read_excel("./data/tr010_primers.xlsx", sheet = 1) %>%
-  dplyr::select(target = primer_id, forward, reverse, full_name, symbol_ext) %>%
-  inner_join(primer_performance) %>%
-  mutate(performance = paste0(sprintf("%.1f", round(mean.cq,1)), " (", sprintf("%.1f",        round(sd.cq,1)), "), E = ", sprintf("%.2f", round(mean.eff,2))), 
-         forward = paste0("F: 5'-", forward, "-3'"), 
-         reverse = paste0("R: 3'-", reverse, "-5'")) %>%
-
-  dplyr::select(symbol_ext, full_name, forward, reverse, performance) %>%
-  pivot_longer(names_to = "primer", values_to = "seq", forward:reverse) %>%
-dplyr::select(Symbol = symbol_ext, 'Transcript name' = full_name, Sequence = seq, 'Mean Cq (SD) and efficiency' = performance) %>%
-    filter(!(Symbol %in% c("RPL32", "MYH7", "MYH2", "MYH1")))
-
-primer_table %>%
-  kable(format = "latex", 
-        caption = "Primer sequences and average performance.",
-        booktabs = TRUE) %>%
-  collapse_rows(columns = c(1, 2, 4)) %>%
-  landscape()
-
-# 
-# flextable() %>%
-# merge_custom(x = primer_table$Symbol, columns = c(1, 2, 4)) %>%
-# add_header_lines(values = "Table 2. Primer sequences and average performance.") %>%
-#     fontsize(size = 8, part = "body") %>%
-#   fontsize(size = 10, part = "header") %>%
-# autofit()
-#
-
-```
-
 ## Immunoblotting
@@
 All samples from the same participant were run on the same gel and all samples were run in at least duplicates.
+When samples were compared between participants, signals were expressed per a calibrator sample from each participant that in turn was measured on a separate gel with all calibrator samples.
 Separated samples were transferred to &#956m PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 Glycin, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h.
@@
 CTRL) were treated as population (fixed) effects and leg nested within participant included as group level (random) effects.
-These analyses were performed on data with matching time points between TRAIN and CTRL with the exeption that all post-training data from TRAIN were included (post-training and de-training).
+These analyses were performed on data with matching time points between TRAIN and CTRL with the exception that all post-training data from TRAIN were included (post-training and de-training).
 Relative interactions between groups were estimated as $\Delta$ TRAIN - $\Delta$ CTRL.
@@
 Segmented models were fitted with time and volume condition as population effects and legs nested within participants as group level effects.
-
 Muscle strength, muscle thickness, protein and total RNA data was modeled after log transformation.
-Gene abundance data were fitted with number of sessions as a categorical variable in comparisons of volume conditions, and Cq values converted to counts as suggested by Matz et al.
-[-@RN1964].
-A Poisson-lognormal model was used to fit these count, using data from all genes and including group level effects for each technical duplicate, controlling for technical errors during sample preparation.
+
+Gene abundance data were fitted with number of sessions as a categorical variable in comparisons of volume conditions, and Cq values converted to counts as suggested by Matz et al[@RN1964].
+A Poisson-lognormal model was used to fit these count data, using data from all genes and including group level effects for each technical duplicate, controlling for technical errors during sample preparation.
 An offset consisting of a normalization factor proportional to the amount of muscle used to prepare cDNA was used to model gene abundance per tissue weight.
@@
 
-A linear model was used to estimate the increase per session and average total RNA for every leg in the TR group.
+A linear model was used to estimate the increase per session and average total RNA for every leg in the TRAIN group.
 These estimates were then used to estimate the effect of training-induced increase per session and average total RNA abundance on muscle hypertrophy.
@@
 Mean-centering of the independent variable was done to obtain an estimate of the average RNA concentration per leg.
-This also assured that the slope and intercept did not correlate, something that could lead to colinearity issues in subsequent modeling.
+This also assured that the slope and intercept did not correlate, something that could lead to issues with colinearity in subsequent modeling.
 A mixed effects model was subsequently fitted with differences in muscle thickness pre- to post-training as the dependent variable and estimated percentage per session increases in total RNA, the mean total RNA scaled as standard deviations from the mean and sex as independent variables.
-Leg nested within participant was used as group levels effects.
+Participants was used as group levels effect.
 The robustness of this model was assessed by leave-one-out analysis on the level of individual data points in the relationship between total RNA and sessions and on the level of participants (see Results).
@@
 All models were fitted using a Bayesian framework using either the brms [@RN2562] or MCMCglmm [@RN1992] package written for R [@Rcore].
-Inference about effects of interest was done based on point estimates and their 95% credible intervals (CI).
-Credible intervals not containing null effects were interpreted as robust.
-Models were fitted with default priors which also makes CI analogous to confidence intervals but with the interpretation being that the CI contains the true population value with the specified certainty (95%), given the data.
+Inference about effects of interest was drawn based on point estimates and their 95% credible intervals (CI).
+Credible intervals not containing null effects were interpreted as robust effects.
+Models were fitted with default priors.
+CIs were interpreted as containing the true population value with the specified certainty (95%), given the data and priors.
 Fitting performance was assessed by confirming convergence of at least four different chains of MCMC samples (graphically assessed and confirmed with $\hat{R}\approx 1$).
@@
 
-# Results
-```{r training-data}
-
-m2 <- readRDS("./data/derivedData/training-data/m2.RDS")
-# Model containing interaction 
-# readRDS("./data/derivedData/training-data/m1.RDS")
-
-# Create a hypothesis and convert to percentages
-
-h <- hypothesis(m2, c("(weekW2 / Intercept) * 100 = 0",
-                 "(weekW3 / Intercept) * 100 = 0"))$hypothesis
-
-load.inr.w2 <- paste0(sprintf("%.0f",h[1, 2]), 
-                      "%, 95% CI: [",
-                      sprintf("%.0f",h[1, 4]), 
-                      ", ",
-                      sprintf("%.0f",h[1, 5]), 
-                      "]")
-load.inr.w3 <- paste0(sprintf("%.0f",h[2, 2]), 
-                      "% [",
-                      sprintf("%.0f",h[2, 4]), 
-                      ", ",
-                      sprintf("%.0f",h[2, 5]), 
-                      "]")
-
-# Approximate percentages increase in strength and muscle mass
-
-m1.isom <- readRDS("./data/derivedData/strength-bayes/m1.isom.RDS") 
-m1.isok <- readRDS("./data/derivedData/strength-bayes/m1.isok.RDS")
-
-# Change in percentage-points between groups 
-isok.h <- hypothesis(m1.isok, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"
-                                ))$hypothesis
-
-isom.h <- hypothesis(m1.isom, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"
-                                ))$hypothesis
-
-isok_pre_post <-  paste0(sprintf("%.1f",isok.h[1, 2]), 
-                      "% [",
-                      sprintf("%.1f",isok.h[1, 4]), 
-                      ", ",
-                      sprintf("%.1f",isok.h[1, 5]), 
-                      "]")
-
-isok_pre_post_aprox <- paste0(sprintf("%.1f",isok.h[1, 2]))
-
-isok_pre_detrain <-  paste0(sprintf("%.1f",isok.h[2, 2]), 
-                      "% [",
-                      sprintf("%.1f",isok.h[2, 4]), 
-                      ", ",
-                      sprintf("%.1f",isok.h[2, 5]), 
-                      "]")
-
-isok_pre_detrain_approx <- paste0(sprintf("%.1f",isok.h[2, 2]))
-
-isom_pre_post <-  paste0(sprintf("%.1f",isom.h[1, 2]), 
-                      "% [",
-                      sprintf("%.1f",isom.h[1, 4]), 
-                      ", ",
-                      sprintf("%.1f",isom.h[1, 5]), 
-                      "]")
-isom_pre_detrain <-  paste0(sprintf("%.1f",isom.h[2, 2]), 
-                      "% [",
-                      sprintf("%.1f",isom.h[2, 4]), 
-                      ", ",
-                      sprintf("%.1f",isom.h[2, 5]), 
-                      "]")
-
-isom_pre_post_approx <- paste0(sprintf("%.1f",isom.h[1, 2]))
-isom_pre_detrain_approx <- paste0(sprintf("%.1f",isom.h[2, 2]))
-
-## Muscle thickness
-
-m1.us <- readRDS("./data/derivedData/us-freq-bayes/m1us.RDS")
-
-us.h <- hypothesis(m1.us, c("(((txpost_int_train - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0", "(((txpost_int_detrain - txbaseline_int_train)/(txbaseline_int_train + Intercept)) - (txpost_con_train/Intercept)) * 100 = 0"))$hypothesis
-
-us_pre_post_approx <- paste0(sprintf("%.1f",us.h[1, 2]))
-us_pre_detrain_approx <- paste0(sprintf("%.1f",us.h[2, 2]))
+# Acknowledgements
+
+The authors express their gratitude to the participants who made this research possible.
+
+# Conflict of interest
+
+The authors do not have any conflict of interest to declare.
+
+# Data availability statement
+
+Data and code are available at `github.com/dhammarstrom/ribo-accum-paper`
+
+# References
+<div id="refs"></div>
+
+```{r setup-tables, include=FALSE}
+
+source("./R/libs.R")
+library(flextable); library(officer)
 
@@
 
-In TRAIN, all participants successfully completed their prescribed RT on both legs, with the two volume conditions resulting in diverging volume profiles (load &#215; repetitions) over the course of the study (Figure 1B).
-Exercise intensities (resistance at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with each block consisting of four training sessions.
-Concomitantly, in TRAIN, isokinetic strength and thickness of *m.
-vastus lateralis* increased from baseline to after Session 12 compared to CTRL (isokinetic strength $\sim$ `r isok_pre_post_aprox`%-point difference; muscle thickness $\sim$ `r us_pre_post_approx`%-point difference, Figure 1C and D), a difference that was sustained to after eight days of de-training ($\sim$ `r isok_pre_detrain_approx`%-point and $\sim$ `r us_pre_detrain_approx`%-point difference in change in isokinetic strength and muscle thickness, respectively; Figure 1C and D).
-Isometric strength showed the same general pattern to after Session 12 ($\sim$ `r us_pre_detrain_approx`%-point difference), though with considerably larger degrees of uncertainty, as indicated by wider 95% CI normalization compared to CTRL after de-training ($\sim$ `r isom_pre_detrain_approx`%-point; Figure 1C).
-No differences were observed between volume conditions for either strength or muscle thickness.
-
-```{r figure1, fig.height=(23*0.75)/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).", echo=FALSE}
-
-figure1 <- readRDS("./Figures/rds/figure1.RDS")
-
-figure1
-
-```
-
-```{r western-data}
-segmented_protein_results <- readRDS("./data/derivedData/western-analysis/segmented_results.RDS")
-
-s6_slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[6,4]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[6,6]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[6,7]) - 1)),
-                   "]")
-
-s6_slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[7,4]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[7,6]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[7,7]) - 1)),
-                   "]")
-
-s6_slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[8,4]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[8,6]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[8,7]) - 1)),
-                   "]")
-
-ubf_slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[14,4]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[14,6]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[14,7]) - 1)),
-                   "]")
-
-ubf_slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[15,4]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[15,6]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[15,7]) - 1)),
-                   "]")
-
-ubf_slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(segmented_protein_results[16,4]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[16,6]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(segmented_protein_results[16,7]) - 1)),
-                   "]")
-
-s6.m3 <- readRDS("./data/derivedData/western-analysis/s6_detrain.RDS")
-ubf.m3 <- readRDS("./data/derivedData/western-analysis/ubf_detrain.RDS")
-
-ubf.sum <- summary(ubf.m3)$fixed
-
-var_s12 <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[3,1]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(ubf.sum[3,3]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(ubf.sum[3,4]) - 1)),
-                   "]")
-
-const_detrain <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[2,1]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(ubf.sum[2,3]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(ubf.sum[2,4]) - 1)),
-                   "]")
-
-var_const_interaction <- paste0(sprintf("%.1f", 100 * (exp(ubf.sum[4,1]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(ubf.sum[4,3]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(ubf.sum[4,4]) - 1)),
-                   "]")
-
-h <- hypothesis(ubf.m3, "condvar + timepost1w:condvar = 0")$hypothesis %>%
-  data.frame()
-
-var_change <- paste0(sprintf("%.1f", 100 * (exp(h[1,2]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(h[1,4]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(h[1,5]) - 1)),
-                   "]")
-
-```
-
-For both rpS6 and UBF, protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
-This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12 and elevated after eight days of rest (Figure 2A and B), however with no differences being observed after the first training session (48 hrs).
-Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`).
-After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
-For rpS6, de-training affect protein levels, which remained similar between volume conditions period and remained similar between volume conditions.
-At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point.
-No differences were observed between volume conditions for either transcripts (Figure 2D and E).
-
-```{r figure2, fig.height=(23*0.75)/2.54, fig.width=8.9/2.54, fig.align="center", fig.cap="Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF.
-Black points and error bars represents statistically robust results (a 95\\% CI not containing 0).
-C shows western-blots and total protein stains from a representative participant.
-mRNA data is normalized per total RNA.", echo=FALSE}
-
-figure2 <- readRDS("./Figures/rds/figure2.RDS")
-
-figure2
-
-```
-
-```{r total-rna}
-
-# Segmented model analysis in text 
-# Read the segmented model 
-tc.m2 <- readRDS("./data/derivedData/total-rna-analysis/tc.m2.RDS")
-
-time1.hyp <- hypothesis(tc.m2, "time1 = 0")
-time2.hyp <- hypothesis(tc.m2, c("time2 + time1 = 0")) # The slope (on log scale) between session 4 and 8
-time3.hyp <- hypothesis(tc.m2, c("time2 + time1 +time3 = 0")) # The slope on log scale between session 8 and 12
-
-slope0_4 <- paste0(sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,2]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,4]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(time1.hyp$hypothesis[1,5]) - 1)),
-                   "]")
-
-slope4_8 <- paste0(sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,2]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,4]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(time2.hyp$hypothesis[1,5]) - 1)),
-                   "]")
-
-slope8_12 <- paste0(sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,2]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,4]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(time3.hyp$hypothesis[1,5]) - 1)),
-                   "]")
-
-# Estimate changes to end of periods
-
-sessions <- hypothesis(tc.m2, c("time1 * 4 = 0", 
-                                     "(time1 * 8) + (time2 * 4)= 0", 
-                                     "(time1 * 12) + (time2 * 8) + (time3 * 4)= 0"))
-
-## Effect of detraining
-
-detrain.hyp <- hypothesis(tc.m2, "detraindetrain = 0")
-
-detrain_rna <- paste0(sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,2]) - 1)),
-                           "%, [",
-                      sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,4]) - 1)), 
-                   ", ",
-                       sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,5]) - 1)),
-                   "]")
-
-fourth.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,2]) - 1)),
-                           "% [",
-                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,4]) - 1)), 
-                   ", ",
-                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[1,5]) - 1)),
-                   "]")
-
-eight.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,2]) - 1)),
-                           "% [",
-                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,4]) - 1)), 
-                   ", ",
-                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[2,5]) - 1)),
-                   "]")
-
-twelth.session <- paste0(sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,2]) - 1)),
-                           "% [",
-                      sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,4]) - 1)), 
-                   ", ",
-                       sprintf("%.1f", 100 * (exp(sessions$hypothesis[3,5]) - 1)),
-                   "]")
-
-```
-
-A single session of RT (Session 1) led to robust increases in pre-rRNA 47S ETS and 45S ETS abundance per unit tissue weight, measured as changes from baseline to 48 hrs after exercise within TRAIN (Figure 3B), as well as compared to CTRL (Figure 3C).
-After Session 1, pre-rRNA 47S ETS and 45S ETS levels remained at similar levels at all measured time-points in TRAIN (Figure 3D), confirmed in comparison to CTRL after Session 12 (Figure 3C).
-Other rRNA transcripts showed increases in response to training with slightly different temporal patterns with exception of rRNA 5S which did not change and rRNA 5.8 which tended to follow other mature transcript spliced from pre-rRNA 45S ETS, but without statistical robustness (Figure 3C and D).
-After eight days of rest, 18S and 28S remained at elevated levels compared to CTRL (Figure 3B).
-This general pattern of rRNA expression was reflected by total RNA expression per unit tissue weight, which increased robustly and steadily in TRAIN throughout the initial part of the intervention (Figure 3E and G), leading to robust increase compared to CTRL after Session 12 (Figure 3F), followed by decreased levels after de-training (`r detrain_rna`).
-For both rRNA expression and total RNA levels, the training-associated increases in abundances occurred predominately during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from sessions four to eight (`r slope4_8` increase per session) and from sessions eight to twelve `r slope8_12`, corresponding to `r fourth.session`, `r eight.session` and `r twelth.session` increases from baseline to 48 hrs after session four, eight and twelve, respectively.
-In TRAIN, the two volume conditions led to similar changes for most variables (Figure 3D and G), with 45S ETS abundance only showing differential expression, evident as robustly higher levels in VAR compared to CONST after the 12th session (Figure 3D), coinciding with the increased training volume towards the end of the intervention for this condition.
-
-```{r figure3, fig.height=(23*0.75)/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Total RNA and ribosomal RNA subspecies in response to trainin.
-Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (A).
-There were no robust differences between different volume conditions in the training group (A).
-B shows primer locations targeting different ribosomal RNA subspecies.
-Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and other after the training period (C).
-Error bars shows 95\\% CI.
-Asterisk in C indicates robust differences between volume conditions (a 95\\% CI of pairwise differences not containing 0).", echo=FALSE}
-
-figure3 <- readRDS("./Figures/rds/figure3.RDS")
-
-figure3
-
-```
-
-```{r table3-regression-models-muscle-thickness}
+\pagebreak
+
+<!---BLOCK_LANDSCAPE_START--->
+
+# Tables
+
+```{r table1-regression-models-muscle-thickness}
 # Save model
@@
                           "UBF protein levels (SD-units)", 
-                          "Session 1-4\\textsuperscript{b}",
-                          "Session 4-8\\textsuperscript{c}", 
-                          "Session 8-12\\textsuperscript{d}", 
+                          "Session 1-4",
+                          "Session 4-8", 
+                          "Session 8-12", 
                           "De-training",
@@
                           "rpS6 protein levels (SD-units)", 
-                          "Session 1-4\\textsuperscript{b}",
-                          "Session 4-8\\textsuperscript{c}", 
-                          "Session 8-12\\textsuperscript{d}", 
+                          "Session 1-4",
+                          "Session 4-8", 
+                          "Session 8-12", 
                           "De-training",
@@
 
+std_border = fp_border(color="black", width = 2)
+
   rbind(reg_results, reg_results_s6) %>%
@@
 
-    kable(col.names = c("Coefficient", 
-                        "Estimate\\textsuperscript{a}", 
-                        "SD", 
-                        "Lower 95\\% CI", 
-                        "Upper 95\\% CI"), 
-          caption = "Effect of UBF and rpS6 levels, sessions and de-training on RNA-levels.",
-          format = "latex", 
-          escape = FALSE,
-          booktabs = TRUE) %>%
-    row_spec(9, hline_after = TRUE) %>%
-    footnote(alphabet = c("The dependent variable is total RNA levels (log).",
-    "Slope in response to session 1-4.", 
-    "Change in slope in session 4-8.",
-    "Change in slope in session 8-12."), 
-             footnote_as_chunk = TRUE,
-             threeparttable = TRUE, 
-             escape = FALSE,
-             general_title = "")
+  flextable() %>%
+  set_header_labels(coef = "Coefficient", 
+                    Est.Error = "SD", 
+                    lwr = "Lower 95% CI", 
+                    upr = "Upper 95% CI") %>%
+    add_header_lines(values = "Table 1. Effect of UBF and rpS6 levels, sessions and de-training on RNA-levels.") %>%
+
+ # bold(i = c(5, 6, 13, 14, 20, 21), j = c(2:6), bold = TRUE) %>%
+
+ # hline(i = c(9, 17), part="body", border = std_border ) %>%
+      fontsize(size = 8, part = "body") %>%
+    fontsize(size = 10, part = "header") %>%
+      fontsize(size = 7, part = "footer") %>%
+
+  footnote( i = 2, j = 2,
+            value = as_paragraph(
+              c("The dependent variable is total RNA levels (log)")
+            ),
+            ref_symbols = c("a"),
+            part = "header", inline = TRUE) %>%
+    footnote( i = c(3, 4, 5, 12, 13, 14), j = 1,
+            value = as_paragraph(
+              c("Slope in response to session 1-4", 
+                "Change in slope in session 4-8", 
+                "Change in slope in session 8-12")
+            ),
+            ref_symbols = c("b","c","d"),
+            part = "body", inline = TRUE) %>%
+    hline(i = 9, border = std_border) %>%
+
+  autofit()
 
@@
 
-```{r combined-analysis}
-
-# Save model 
-comb.m1 <- summary(readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS"))$fixed
-
-ubf_cor_tot_rna <- paste0(sprintf("%.1f", 100 * (exp(comb.m1[2,1]) - 1)),
-                           "% [",
-                   sprintf("%.1f", 100 * (exp(comb.m1[2,3]) - 1)), 
-                   ", ",
-                   sprintf("%.1f", 100 * (exp(comb.m1[2,4]) - 1)),
-                   "]")
-
-```
-
-```{r table4-regression-models-muscle-thickness}
+\pagebreak
+
+```{r table2-regression-models-muscle-thickness}
 # Save model
@@
       data.frame(coef = c("Intercept", 
+                          "Baseline muscle thickness",
                           "Sex (Male)", 
-                          "Mean Total RNA",
-                          "Mean Total RNA increase per session", 
+                          "Mean RNA at session 6 (SD units)",
+                          "Mean Total RNA increase per session (%)", 
                           "Between Participant variation",
-                          "Between Participant:Leg variation", 
                           "Residual SD"))) %>%
@@
   dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)
+
+std_border = fp_border(color="gray")
 
@@
 
-      kable(col.names = c("Coefficient", 
-                        "Estimate\\textsuperscript{a}", 
-                        "SD", 
-                        "Lower 95\\% CI", 
-                        "Upper 95\\% CI"), 
-          caption = "Total RNA as a predictor of muscle growth.",
-          format = "latex", 
-          escape = FALSE,
-          booktabs = TRUE) %>%
-
-   footnote(alphabet = c("The dependent variable is $\\\\Delta$ Muscle thickness (mm)"),
-
-             threeparttable = TRUE, 
-             escape = FALSE,
-             general_title = "")
+  flextable() %>%
+  set_header_labels(coef = "Coefficient", 
+                    Est.Error = "SD", 
+                    lwr = "Lower 95% CI", 
+                    upr = "Upper 95% CI") %>%
+    add_header_lines(values = "Table 2. Total RNA as a predictor of muscle growth.") %>%
+
+ # bold(i = c(5, 6, 13, 14, 20, 21), j = c(2:6), bold = TRUE) %>%
+
+ # hline(i = c(9, 17), part="body", border = std_border ) %>%
+      fontsize(size = 8, part = "body") %>%
+    fontsize(size = 10, part = "header") %>%
+
+  footnote( i = 2, j = 2,
+            value = as_paragraph(
+              c("The dependent variable is \U0394 Muscle thickness (mm)")
+            ),
+            ref_symbols = c("a"),
+            part = "header") %>%
+
+  autofit()
 
@@
 
-In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coninciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 3).
-In contrast, no evidence was found for a relationship between total RNA and rpS6 protein levels (Table 3).
-
-In TRAIN, there was a robust positive relationship between rates of increase in total RNA in response to training and muscle growth measured as increases in *m.
-vastus lateralis* thickness (Table 4, Figure 4A), with changes in total RNA over the course of the training intervention being estimated in each leg using a regression model containing number of sessions as the independent variable.
-Conversely, there was a tendency towards a negative relationship between average total RNA levels at Session 6 and changes in muscle thickness (Table 4, Figure 4B), with the average total RNA levels estimated as the predicted value at Session 6 (estimated as the intercept-term) from the model used to estimate the rate of total RNA increase per session.
-
-To assess the robustness of the model for predicting muscle growth, individual relationships between sessions and total RNA levels (Figure 4C) were recalculated after removal of single data points from each participant.
-The model predicting muscle growth was refitted using new estimates of changes in total RNA abundances and increases thereof per session.
-Each refitted model resulted in slightly different estimates (displayed as means and 95% CI in Figure 4D).
-No single data point influenced the r4esults in any meaningful way.
-Next we assessed the robustness by iteratively removing one participant from the data set, similarly this showed that estimates of the effect of total RNA increase on muscle growth was robust but the effect of average total RNA estimates were more variable (Participant 8 and 3 in Figure 4D).
-
-```{r figure4, fig.height=17.25/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 4).
+\pagebreak
+
+```{r table3-characteristics}
+
+# Read dxa data
+
+dxa_whole <- read_excel("./data/tr010_dxa_standarddefinitions.xlsx") %>%
+  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:`HeleVev%fett`) %>%
+  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:`HeleVev%fett`) %>%
+  mutate(value = gsub("kg", "", value), 
+         value = gsub("[[:space:]]", "", value, fixed = FALSE), 
+         value = as.numeric(value)) %>%
+
+  separate(Etternavn, into = c("Study", "participant")) %>%
+  dplyr::select(participant, Date = Målingsdato, variable, value) %>%
+  filter(!(participant %in% c("P08", "P20", "P17"))) %>%
+  group_by(participant) %>%
+  mutate(time = if_else(Date == min(Date), "S0", 
+                        if_else(Date == max(Date), "post1w", 
+                                "S12")), 
+         leg = if_else(variable == "BeinhøyreMagermasse", "R", "L")) %>%
+  filter(variable %in% c("HeleBeinmasse", 
+                         "HeleFettmasse", 
+                         "HeleVevmasse", 
+                         "HeleMagermasse" ,
+                         "Vekt", 
+                         "HeleTotalmasse")) %>%
+  pivot_wider(names_from = variable, values_from = value) %>%
+  mutate(fat_rel = 100*(HeleFettmasse/HeleTotalmasse)) %>%
+  dplyr::select(participant, time, Vekt,HeleTotalmasse, fat_rel) %>%
+  filter(time == "S0")
+
+# DXA weight stature
+
+read_excel("./data/tr010_age_participant.xlsx") %>%
+  inner_join(dxa_whole) %>%
+  mutate(age = difftime(recruited, dob)/365.2425, 
+         bmi = weight/(height/100)^2) %>%
+  dplyr::select(participant, group, height, weight,age, sex,bmi, fat_rel) %>%
+
+  group_by(group, sex) %>%
+  summarise(n = n(), 
+            Age = mean(age, na.rm = TRUE), 
+            age.sd = sd(age, na.rm = TRUE),
+            mass = mean(weight), 
+            mass.sd = sd(weight),
+            stature = mean(height),
+            stature.sd = sd(height), 
+            BMI = mean(bmi), 
+            bmi.sd = sd(bmi), 
+            relfat = mean(fat_rel), 
+            relfat.sd = mean(fat_rel) ) %>%
+  ungroup() %>%
+  mutate(n = as.character(n), 
+         age = paste0(sprintf("%.1f", round(Age, 1)), " (", sprintf("%.1f", round(age.sd, 1)), ")"),
+         mass = paste0(sprintf("%.1f", round(mass, 1)), " (", sprintf("%.1f", round(mass.sd, 1)), ")"),
+         stature = paste0(sprintf("%.1f", round(stature, 1)), " (", sprintf("%.1f", round(stature.sd, 1)), ")"),
+         BMI = paste0(sprintf("%.1f", round(BMI, 1)), " (", sprintf("%.1f", round(bmi.sd, 1)), ")"),
+         fat = paste0(sprintf("%.1f", round(relfat, 1)), " (", sprintf("%.1f", round(relfat.sd, 1)), ")"))%>%
+  dplyr::select(group, sex, n, age, mass, stature, BMI, fat) %>%
+  pivot_longer(names_to = "variable", values_to = "values", n:fat) %>%
+  pivot_wider(names_from = c(sex,group), values_from = values) %>%
+
+  mutate(variable =  factor(variable,
+                          levels = c("n", "age", "mass", "stature", "BMI", "fat"), 
+                          labels = c("n", "Age (yrs)", "Body mass, (kg)", "Stature (cm)", "Body mass index (kg/m<sup>2</sup>)", "Body fat (%)"))) %>%
+
+  dplyr::select(variable, F_experiment, M_experiment, F_control, M_control) %>%
+  flextable() %>% 
+  italic(i = ~variable == "n", j = 1, italic = TRUE) %>%
+  compose(part = "body", i = 5, j = 1,
+         value = as_paragraph("Body mass index (kg m",
+                              as_sup("-2"), ")")) %>%
+
+  set_header_labels(variable = "",
+    F_experiment = "Female", M_experiment = "Male",
+    F_control = "Female", M_control = "Male") %>%
+
+   add_header_row(values = c("", "Experimental group", "Experimental group", "Control group", "Control group"), top = TRUE) %>%
+  merge_at(i = 1, j = 2:3, part = "header") %>%
+  merge_at(i = 1, j = 4:5, part = "header") %>%
+  align(align = "center", part = "header") %>%
+    fontsize(size = 8, part = "body") %>%
+    fontsize(size = 10, part = "header") %>%
+  add_header_lines(values = "Table 3. Participant characteristics.") %>%
+  autofit()
+
+```
+
+\pagebreak
+
+```{r table-4-primers}
+
+# Read qpcr data
+qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")
+
+primer_performance <- qdat %>%
+  group_by(target) %>%
+  summarise(mean.cq = mean(cq, na.rm = TRUE), 
+            sd.cq = sd(cq, na.rm = TRUE), 
+            mean.eff = mean(eff))
+
+merge_custom <- function(ft, x, columns){
+  z <- rle(x)
+  rows_at <- cumsum(z$lengths) - z$lengths + 1
+
+  for(i in seq_along(rows_at)){
+    for(j in columns)
+      ft <- merge_at(x = ft, i = seq( rows_at[i], rows_at[i] + z$lengths[i] - 1), j = j)
+  }
+
+  ft
+}
+
+# Read primer data 
+primer_table <- read_excel("./data/tr010_primers.xlsx", sheet = 1) %>%
+  dplyr::select(target = primer_id, forward, reverse, full_name, symbol_ext) %>%
+  inner_join(primer_performance) %>%
+  mutate(performance = paste0(sprintf("%.1f", round(mean.cq,1)), " (", sprintf("%.1f",        round(sd.cq,1)), "), E = ", sprintf("%.2f", round(mean.eff,2))), 
+         forward = paste0("F: 5'-", forward, "-3'"), 
+         reverse = paste0("R: 3'-", reverse, "-5'")) %>%
+
+  dplyr::select(symbol_ext, full_name, forward, reverse, performance) %>%
+  pivot_longer(names_to = "primer", values_to = "seq", forward:reverse) %>%
+dplyr::select(Symbol = symbol_ext, 'Transcript name' = full_name, Sequence = seq, 'Mean Cq (SD) and efficiency' = performance)
+
+primer_table %>%
+  flextable() %>%
+  merge_custom(x = primer_table$Symbol, columns = c(1, 2, 4)) %>%
+  add_header_lines(values = "Table 4. Primer sequences and average performance.") %>%
+      fontsize(size = 8, part = "body") %>%
+    fontsize(size = 10, part = "header") %>%
+  autofit()
+
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->
+
+\pagebreak
+
+# Figure legends
+
+**Figure 1.** Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).
+Intervals in C and D indicate 95% CI.
+
+**Figure 2.** Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF.
+Black points and error bars represent statistically robust results (a 95\% CI not containing 0).
+C shows western blots and total protein stains from a representative participant.
+mRNA data are normalized per total RNA.
+Intervals in A, B, D, and E indicate 95% CI.
+
+**Figure 3.** Total RNA and ribosomal RNA subspecies in response to resistance training.
+A shows primer locations targeting different ribosomal RNA subspecies in qPCR analyses.
+Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (B).
+There were no robust differences between different volume conditions in the training group (B).
+Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and after the training period (C).
+Total RNA increased compared to CTRL after training and normalized after de-training (E, F).
+The time-course analysis revealed the greatest increase during the first four sessions (G).
+Error bars shows 95\% CI.
+Asterisk in D indicates robust differences between volume conditions (a 95\% CI of pairwise differences not containing 0).
+Points in D and G show abundances after de-training for reference.
+
+**Figure 4.** Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 4).
 Values are averaged over values from men and women.
-Individual plots of estimates total RNA increases over time is shown in C together with results from leave-one-out analysis (D).
-Leave-one-out analysis shows the effect of removing a single participant (black point and error-bars) and individual values from the total RNA per time estimates where green points represents bounds of the 95\\% CI and yellow points represents mean estimates.", echo=FALSE}
-
-figure4 <- readRDS("./Figures/rds/figure4_x.RDS")
-
-figure4
-
-```
-
-# Discussion
-Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
-These markers accumulated progressively during the initial part of the intervention before it leveled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
-This increase in total RNA was interconnected with increases in UBF protein abundance, suggesting UBF levels *per se* to play a role in regulation of rRNA transcription regulation in response to RT.
-Total RNA increases was not affected by weekly fluctuations in training volume, however, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
-Finally, individual rates of increases in total RNA abundance predicted the magnitude of muscle growth, confirming the likely link between ribosomal biogenesis and muscle mass accretion 
-[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
-@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
-and muscle hypertrophy
-[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
-@RN1755; <!-- Stec Ribosome bio higher in high responders -->
-@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->
-
-Total RNA seems to be a valid proxy marker of ribosomal density, as most of the RNA is assumed to be ribosomal RNA
-[@RN2223],<!-- Young  -->
-which in turn is a valid marker of translational capacity
-[@RN2145].
-Several studies have shown that total RNA content is altered by RT
-<!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
-[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
-@RN2180;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --> 
-@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
-@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
-@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
-@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
-@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
-@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
-@RN1897<!-- 11% 36 sessions young -->],
-as was also the case in the present data set.
-However, the time course of total RNA/rRNA changes in response to RT has so far remained speculative, with no study investigating responses to prolonged interventions with multiple sampling time points.
-In the present data, RT led to a clear session-to-session increase in total RNA per unit tissue weight in response to the first four session, whereupon the changes gradually leveled out before peaking after the 8^th^ session, with the peak increase from baseline being $\sim$ 50%, defining an accumulation phase.
-This corroborates well with previous suggestions of peak values being reached within four to nine sessions in young males and females
-[@RN1809;
-@RN2358], and may be essential for preparing muscle fibers for subsequent growth
-[@RN1809;<!-- Correlation between increase and muscle growth -->
-@RN1644; <!--Figueiredo correlation between incr and muscle hypertrophy -->
-@RN2358].<!-- Early accumulation of RNA predicts volume depdent growth -->
-After the 8^th^ session, no meaningful increase or decrease were observed for total RNA/rRNA content within the training period, suggesting a plateau phase with attenuated net synthesis of novel ribosomes.
-Within this last part of the intervention, synthesis of novel rRNA still seemed to be elevated per weight unit muscle tissue compared to baseline, as suggested by sustained elevation of pre-rRNA transcripts, coinciding with peak values of UBF protein levels.
-This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth [@RN2142].<!-- Figueiredo review details the concept of rna concentration -->
-
-The observed rates of RNA accumulation over the entirety of the intervention were found to be a determinant of changes in muscle thickness (after controlling for average total RNA levels).
-Individuals with higher rates of accumulation showed larger accretion of muscle mass.
-This supports the notion that ribosomal biogenesis is an important determinant of RT-induced muscle hypertrophy, with previous studies showing that increases in total RNA are positively correlated with increases in muscle mass
-[@RN1644;<!-- Figueriedo  -->
-@RN1897;
-@RN2055],<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
-<!-- Reidy correlation between total RNA and muscle hypertrophy-->
-differs between individuals displaying low vs. high levels of muscle hypertrophy in response to RT
-[@RN1755]<!-- Stec 2016  -->
-an contribute to explain RT volume-dependent changes in muscle mass and strength
-[@RN2358].
-In addition, supression of ribosomal biogenesis in *in vitro* models leads to halted muscle cellular growth in some 
-[@RN1632;<!-- Nader 2005 blockage of mTOR halts ribo bio and growth -->
-@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
-@RN1755]<!-- Stec specific pol I inhibition -->
-but not all studies
-[@RN1929].<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
-Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA was instead associated with a tendency towards lowered muscle growth.
-Overall, the rate of increases in ribosomal density thus seems to be a better predictor of individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
-Interestingly, the interaction between rRNA synthesis rate and muscle mass accretion (but not between ribosomal content and muscle mass accretion) may shed light on observed differences in muscular responses to RT between young and old individuals.
-Whereas aged muscle display higher levels of total RNA at rest 
-[@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
-they show reduced changes in total RNA levels in response to RT
-[@RN1809],<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
-potentially explaining their alleged poorer overall hypertrophic responses [@RN1809].
-Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
-Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.
-
-<!-- rpS6 dissconect from total RNA levels. -->
-In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6.
-Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period.
-Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
-[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
-but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
-[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
-Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for.
-Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT.
-Additionally, ribosomal proteins may have extra-ribosomal functions
-affecting their expression
-[@RN2588].<!-- Warner 2009 General description of   -->
-
-UBF levels robustly explained total RNA levels over the entire course of the intervention.
-These analyses were done with the number of sessions accounted for, allowing unbiased estimates.
-Unrealistically strong relationships could have been otherwise expected as both the dependent variable (total RNA) and the covariate (UBF levels) varies with the number of sessions.
-From a mechanistic perspective, UBF is an important transcription factor for rDNA transcription as it, in its active state recruits a secondary transcription factor (SL1) to the rDNA promoter and enables transcription by RNA Pol I
-[@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
-Activation of UBF is controlled by the mechanosensitive mTOR pathway, and rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
-[@RN1632; <!-- Nader rapamycin block UBF availability  -->
-@RN2564].<!-- Hannan 2003 -->
-<!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1.
-The effect of blockage of mTOR through rapamycin is not rescued by hypophosphorylated UBF -->
-Evidence from human exercise studies confirms training-induced activation of UBF through phosphorylation
-[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
-@RN1644].<!-- Figueriedo ubf increase total and active form (ser388) --> 
-In addition to exercise-induced activation of UBF, mechanical loading also leads to increased levels of total UBF
-[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
-@RN1644]. Increases in UBF was determined to be rapamycin insensitive after synergist ablation in mice
-[@RN1072]<!-- Goodman 2011 UBF increase due to synergist ablation is rapamycin insensitive https://doi.org/10.1113/jphysiol.2011.218255 --> 
-pointing to an effect observed in cell models where c-Myc induces UBF mRNA transcription [@RN2565].<!-- Pootinga 2004 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
-Interestingly the availability of UBF *per se* has been shown to regulate rRNA transcription
-[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->  through control of rDNA gene activity
-[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->
-Together with our observations, this underlines the importance of UBF as a regulator of RT-induced ribosomal biogenesis.
-
-After eight days of de-training, total RNA and rRNA levels per weight unit muscle tissue returned toward baseline levels, though without concomitant reversal of muscle thickness, which remained at elevate levels.
-This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
-This supports the idea that ribosomal biogenesis is a cellular activity on demand, possibly relating to its relative expense [@RN1920] also in muscle tissue.
-Based on this notion, and the fact that RT volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including induction of c-Myc expression, mTOR activation
-[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
-@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
-@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
-subsequent total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
-and subsequent training outcomes [@RN1767; @RN2358], we hypothesized that fluctuations in training volume would be reflected in markers of ribosomal biogenesis.
-When comparing VAR to CONST in the present study we found only one part of the pre-rRNA, 45S ETS, to be differentially expressed and only so after Session 12 in favor of VAR together with a tendency towards rescued UBF levels after de-training in response to increased volume in the VAR but not CONST protocol.
-These observations do not give support to a clear effect of fluctuations in training volume on total RNA levels or rRNA expression within a relatively short and training-intensive intervention, though it should be noted that the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume.
-Indeed, both training protocols utilized in the present study increased muscle strength and induced muscle hypertrophy to a similar degree.
-From a general perspective, albeit volume is an important determinant of increases in muscle strength and mass
-[@RN1767;<!-- schoenfeld 2016 dose response  -->
-@RN2492],<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
-differences in organization of training loads is likely of minor importance when training volumes are equated over time
-[@RN2572].<!-- Grgic 2017 volume equated periodization schemes led to similar muscle hyeprtrophy -->
-It is important to note that RT in the current study was performed with the same volume in the first four sessions, something that could have been more than enough to maximize rRNA transcription in previously untrained individuals.
-This is supported by the observation that pre-rRNA increased rapidly initially in both protocols with minimal changes in response to subsequent sessions, regardless of exercise volume.
-The CONST protocol in the present study corresponded to volumes used in the moderate volume condition in a previous study from our lab (three sets in two exercises activating knee extensor muscles) [@RN2358].
-There, higher levels of total RNA were observed after four sessions in the moderate compared to a low volume protocol [@RN2358].
-Interestingly, using a progressive volume protocol in well-trained participants, increases in total RNA have been reported throughout six weeks of training
-[@RN2225].<!-- Haun total RNA increase througout a 6 week intervention -->
-Altough this observation was done in well-trained participants performing a high volume protocol without a control condition with constant volume, compared to constant volume protocols
-[@RN1809;<!-- Brook 2016 -->
-@RN2358; and the present study],<!-- Hammarström 2020 -->
-progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
-
-In conclusion, RT-induced ribosome accumulation reached peak values in the initial phase of RT (8 sessions) and was interconnected with increases in UBF protein levels.
-The rate of total RNA accumulation predicted RT-induced muscle hypertrophy.
-Fluctuations in training volume did not transfer to fluctuations in ribosomal biogenesis, but training cessation led to decreased ribosomal content.
-
-<!-- NOTES -->
-<!-- UBF and S6 protein levels may be differently regulated in men and rats as more consecutive bouts of electrical stimulation in rats does not prtoduce the same pattern seen in our study 
- Kotani 2019 10.1152/japplphysiol.00665.2018 increase in t UBF at 6 h post ex but not 48 despite 2 weeks training (3 x week).
-S6 increase at 48 h and 6 bouts but return to sed values at 18 bouts.
-18 and 28 S increased over sed at 48 h post ex with no further increase beyond 3 bouts 45S maximal at three bouts -->
-
-# References
-<div id="refs"></div>
+Individual plots of estimates total RNA increases over time are shown in C together with results from leave-one-out analysis (D).
+Leave-one-out analysis shows the effect of removing a single participant (black point and error bars) and individual values from the total RNA per time estimates where green points represent bounds of the 95\% CI and yellow points represent mean estimates.

### 2021-12-10 Figures/Figure1.R

@@ -587,7 +587,7 @@ saveRDS(figure1, "./Figures/rds/figure1.RDS")
 # height of figure = full page = 23 cm
 
 
-ggsave("figures/figure1.pdf", plot = figure1, width = 8.9*2, height = 23 * 0.75, 
+ggsave("figures/figure1.pdf", plot = figure1, width = 7.7*2, height = 23 * 0.75, 
        dpi = 600,
        units = "cm", device=cairo_pdf)
 

### 2021-12-10 Figures/Figure2.R

@@ -111,10 +111,10 @@ prot_fold_change <- tx_results_western %>%
   theme(strip.placement = "top",
         strip.background = element_rect(fill = "white", color = "white"), 
         strip.text = element_text(size = 7, hjust = 0),
-        legend.position = c(0.65, 0.5),
+        legend.position = "bottom",
         legend.text = element_text(size = 7, margin = margin(t = 0.1, b= 0.1,r = 0.1,l = 0.1, unit = "pt")),
         legend.key.size = unit(0, "cm"),
-        legend.margin = margin(-0.5, 0, -0.5, 0, "cm"),
+        legend.margin = margin(-0.5, 0, 0, 0, "cm"),
         legend.spacing.x = unit(0.1, 'cm'),
         legend.direction = "vertical",
         axis.title.x = element_blank()) 
@@ -144,7 +144,8 @@ prot_interaction <- tx_results_western %>%
   facet_wrap( ~ target, ncol = 1, strip.position = "top") +
   
   scale_color_manual(values = c("gray50", "gray10")) +
-  
+  scale_x_continuous(breaks = c(0.5, 1, 1.5, 2, 2.5, 3), 
+                     labels = c("", 1, "", 2, "", 3)) +
   plot_theme() +
   theme(strip.background = element_rect(color = "white", fill = "white"), 
 
@@ -243,7 +244,7 @@ volume_diff_mrna <-   rrna_timecourse$estimated_diff %>%
     scale_color_manual(values = c("gray50", "gray10")) +
     scale_y_continuous(limits = c(0.2, 1.8), 
                        breaks = c(0.5, 1, 1.5)) +
-  scale_x_discrete(labels = c("0", "1", "4", "5", "8", "9", "12", "\nDe-train")) +
+  scale_x_discrete(labels = c("0", "1", "4", "5", "8", "9", "12", "\nDe-\ntrain")) +
     
     plot_theme() +
     theme(strip.background = element_rect(color = "white", fill = "white"), 
@@ -270,12 +271,13 @@ figure2 <- plot_grid(
 
     
   plot_grid(prot_fold_change, prot_interaction, 
-                     ncol = 2, rel_widths = c(0.5, 0.5)), 
+                     ncol = 2, rel_widths = c(0.5, 0.5)),
+  NULL,
   reps_west,
   
[…]

### 2021-12-10 Figures/Figure3.R

@@ -347,7 +347,7 @@ for(i in 1:length(targets)) {
     geom_point(data = filter(dat, time == "post1w"), 
                position = position_nudge(x = 1)) +
     
-    scale_color_manual(values = c(color.scale[1], color.scale[2])) +
+    scale_color_manual(values = c(color.scale[2], color.scale[1])) +
     
     scale_y_continuous(limits = c(floor(min(dat$emmean, na.rm = TRUE)), 
                                   ceiling(max(dat$emmean, na.rm = TRUE))),
@@ -612,17 +612,20 @@ tc.m1.predictions  <- tc.m1 %>%
   summarise(m = median(prediction), 
             lwr =  quantile(prediction, 0.025), 
             upr = quantile(prediction, 0.975)) %>%
+  mutate(cond = factor(cond, levels = c("const", "var"), 
+                       labels = c("CONST", "VAR"))) %>%
   print()
 
 
 
 
 rna_tc_fig <-  tc.m1.predictions %>%
+
   filter(time.c < 12.5) %>% 
   ggplot(aes(time.c, exp(m), group = paste(group, cond))) + 
   
   geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)), alpha = 0.15) +
-  geom_line(aes(color = cond), size = 1.5) + 
+  geom_line(aes(color = cond), size = 1.2) + 
   
   geom_errorbar(data = filter(tc.m1.predictions, time.c > 12.5),
                 aes(time.c, exp(m), ymin = exp(lwr), ymax = exp(upr)), width = 0.2, 
@@ -689,11 +692,11 @@ figure3 <- plot_grid(
 
 saveRDS(figure3, "./Figures/rds/figure3.RDS")
 
-# Width of figure = 1x columns 8.9 cm
+# Width of figure = 1x columns 7.7 cm
 # height of figure = full page = 23 cm
 
 
-ggsave("figures/figure3.pdf", plot = figure3, width = 8.9*2, height = 23 * 0.75, 
+ggsave("figures/figure3.pdf", plot = figure3, width = 7.7 * 2, height = 23 * 0.75, 
        dpi = 600,
        units = "cm", device=cairo_pdf)
 

### 2021-12-10 Figures/Figure4.R

@@ -335,7 +335,7 @@ saveRDS(figure4, "./Figures/rds/figure4_x.RDS")
 # height of figure = full page = 23 cm
 
 
-ggsave("figures/figure4.pdf", plot = figure4, width = 8.9 * 2, height = 23 * 0.75, 
+ggsave("figures/figure4.pdf", plot = figure4, width = 7.7 * 2, height = 23 * 0.75, 
        dpi = 600,
        units = "cm", device=cairo_pdf)
 

### 2021-12-10 R/ubf-tot-rna-model.R

@@ -196,8 +196,6 @@ saveRDS(comb.m1, "./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")
 saveRDS(comb.s6.m1, "./data/derivedData/ubf-tot-rna-model/comb_m1_s6.RDS")
 
 
-
-
  ### Estimation of effects of total RNA on muscle growth ###########################
 
 # A function to test for colinearity

### 2021-12-13 manuscript.Rmd
sorted files and added manual edits

@@
   inner_join(dxa_whole) %>%
-  mutate(age = difftime(recruited, dob)/365.2425, 
+  mutate(age = age, 
          bmi = weight/(height/100)^2) %>%

### 2022-01-17 response-to-reviews-rev1.Rmd
added reviers comments

---
title: "APH-2021-12-0652: Response to the reviewers"
author: "Daniel Hammarström"
date: "2022-01-17"
output: pdf_document
---

## Referee 1

> This manuscript investigates the induction of ribosome biogenesis and accumulation of RNA/ribosomes during the early phase of resistance training. Subjects performed unilateral RE bilaterally but with differing loading protocols. One leg trained with a constant volume whereas the other trained with variable volume. Control subjects were biopsied at relevant time points. The manuscript presents evidence in favor of that skeletal muscle in the naïve state is very responsive with respect to ribosome biogenesis following acute RE. With repeated training sessions this induction is maintained but not further stimulated. Most of the effect on total RNA was seen during the first training block. UBF levels were related to RNA accumulation and rate of total RNA accumulation predicted RT-induced muscle hypertrophy. With eight days of detraining, around 20% of total RNA was lost, indicating that the elevated RNA pool needs to be maintained not to decrease.
The manuscript is highly interesting and ambitious in its design. I think that it adds to the published literature and could potentially be a future important reference piece. However, I have some concerns and comments that I’d like for the authors to address, as I think that they can improve the manuscript further.

> The western blot membranes in the paper does not really translate into the presented data. In my opinion the bands look rather faint and equal in intensity across the time course. Could the authors perhaps provide raw data and membranes for RPS6 and UBF just for the reviewers?

> Is pre-rRNA 47S ETS and 45S ETS per unit tissue weight a previously reported way to present qPCR data? Does the data look the same in relation to 1 or more house keeping genes?

One transcribed pre-rRNA results in a single end product, one ribosome and in the context of protein synthesis the amount of ribosomes per cell or tissue mass is highly correlated with protein synthesis (Milward 1976, West 2016). In contrast to rRNA processing, mRNA transcription and translation could to some degree be seen as competitive between different mRNA.

This translates to different normalization strategies. The use of a houskeeping gene when investigating ribosomal biogenesis is not interesting as it does not

> Could the authors please comment on the apparent disconnect between increase in RNA and RPS6 protein levels?

> The concept that the increase in RNA content predicts subsequent hypertrophy is very interesting. Even more intriguing is the finding that total RNA mid training negatively correlates to muscle hypertrophy is fascinating! I would encourage the authors to deepen the discussion on these key data points, especially in the light of more and more evidence in favor of ribosome specialization. Could a high percentage of new ribosomes be advantageous for growth in more ways than just contributing greater numbers?

The referee raises an interesting point

> Please comment on that UBF levels where maintained at the detraining time point but total RNA decreasing. Would it be relevant to analyze UBF phosphorylation?

> The fact that RNA decreased at the detraining time point is interesting and indicates the importance of actively maintaing translational capacity not to loose it. I think the discussion could benefit from including a short section of the role for ribophagy in this setting?

> The methods in the abstract must be improved. In its current form it lacks key information such as training duration, the number of controls are states as n=7 but n=8 in the M&M section and I’d like for the authors to try and specify biopsy time points.

> Page 3, line 48-49. I encourage the authors to consider including an additional reference (Figueiredo et al 2021 J Phys). This paper shows the specific response of ribosome biogenesis in response to RE and not Endurance type exercise.

> Page 4, line 8: Please clarify “numberical lowering”. Could this be rephrased?

> Page 4, line 10: Double “..”

> Page 4, line 15: Double “..”

> Page 4, line 26: Should “to” be removed?

> Page 11, line 20: n=7 or 8? Please check abstract

> Page 11, line 35: Should “a” be removed? Please double check if it’s 7 or 8 days of detraining. Abstract states 8.

> Page 12, line 19: please clarify “leg being training in the rest period between sets in the first leg. Does this means that they training continuously without break?

> Page 13, line 52: 2-4 passes were made with the Bard Magnum 12-14 gauge needles but only 1-2 aliquots of the samples were frozen. What happened to the other material? If I misunderstood, can the text be rewritten for clarity?

> Page 14, line 8: Was the spike-in used to normalize RNA extraction efficiency? If yes, please state this clearly.

> Page 14, line 49: Does “normalized” refer to denatured?

> Page 15, line 15: Does the authors refer to amplicon size when they write primer sizes?

> Page 31, Figure legends: I think that the figure legends could benefit from a bit more substance. For example Fig 1A, perhaps indicate bar color as black (CONST) and yellow (VAR)?

## Referee: 2

> The authors sought to perform a time course investigation of ribosome density and biogenesis markers after knee extensor training.  Phenotype variables were also assessed.  Training led to increases in muscle growth (VL thickness) as well as in total RNA, rRNA, UBF and rpS6.  Training volume did not play a role in these adaptations.  Interestingly, the rate of total RNA increases predicted hypertrophy.  Likewise, training cessation led to decrements in total RNA abundance.
I love this study.  This continues to contribute to our knowledge regarding the role of ribosome biogenesis in muscle hypertrophy, and is delicate human work with multiple time point biopsies.
I have minor comments in order of appearance, and congratulate the authors on their fine work.

>  ABSTRACT: Reads well and concise.  Nice work, and no comments.

> INTRODUCTION 
- Line 7, minor: “promotes” should read as “promote”
- Line 14, minor: “with subsequent repeated bouts” should read as “and subsequent repeated bouts”
- Line 18, minor: “resting synthetic rate of muscle protein” better reads as “basal muscle protein synthesis rates”
- Lines 48-50, minor: “bouts lead to accumulation of mature rRNA thus also total RNA and presumably functional ribosomes”…this sentence is long, and the word “thus” seems to make it a run-on.  I’d rephrase the sentence to read more coherently.
Page 4
- Lines 10 and 15, minor: each sentence has two stops.
- Line 33, minor: no need to use the phrase “per se”
RESULTS
- Line 51 (and throughout the results), clarification: is “9.2%-point difference” the same as “9.2% difference”?  If so, adjust and remove “point” from all descriptors as this is confusing.
- Other than this, the results, while dense and full of data, are well-written.
DISCUSSION
Page 7
- Line 17, minor: again, “per se” can be removed.
- Line 19, minor: “was not affected” better reads as “were not affected”
- Line 39, minor: “session” should read as “sessions”

> I urge the authors to be creative somewhere in the discussion by perhaps adding a paragraph after the one found in lines 29-56.  We’ve done a lot of work in the ribosome biogenesis area, and we’ve thought a good bit about total RNA/unit muscle with regard to what it represents.  The authors correctly articulated this as a proxy of ribosome density.  However, as muscle tissue and myofibers volumetrically increase, this will affect the metric (the same holds true with mitochondrial biogenesis and using CS activity as a proxy of mitochondrial volume density).  An easy way to interpret this is in the following example:
o Muscle tissue increases 5% in thickness with RT (and myofibers scale accordingly as well)
o Total RNA/unit muscle does not change in value
Most would interpret this as ribosome biogenesis not occurring.  However, biogenesis likely occurred because, as muscle grew, ribosome biogenesis scaled with growth thus leading to no change in RNA concentration.  Our laboratory discusses this important concept with regard to RT and mitochondrial biogenesis (PMID: 32162291, PMID: 34646153).

> Anyhow, this is no critique of the authors’ hard and excellent work.  However, I think the authors re-iterating this point as a stand-alone paragraph would benefit readership.  In addition, I believe taking a moment to define what the total RNA/ribosome density metric conceptually represents, and then (in this newly added paragraph) briefly walking the reader through each phase of the data, and what could be going on with biogenesis in the context of a growing muscle would be interesting to see.  For instance, in Figure 3G, my interpretation (much like the authors’) is that ribosome biogenesis is very rapid at the onset of RT where little functional growth is likely occurring.  And like the authors, I believe this likely was needed for eventual growth to occur.  However, the “no change” in RNA concentrations session 9-12 as muscle is continuing to grow may not indicate a plateau in biogenesis given that this is the time during which muscle growth is likely starting to ramp up.  Hence, biogenesis rates could be equally as ramped here compared to sessions 1-9 of RT.

METHODS

Page 11

- Lines 9-19: Authors stated that 18 participants were recruited, but then go on to say that n=11 TRAIN and n=8 CTRL completed.  Please reconcile.
- Line 35: Authors stated 7 days after detraining, but in the discussion (page 7, line 19) they mention after 8 days of detraining.  Please reconcile.
Page 12
- Line 39, minor: “restitution” should read as “rest”.  Please change accordingly, and if used throughout the paper, please alter.
Page 15
- Line 36, minor: “Glycin” should read as “Glycine”.  In addition, and molar unit is needed.

### 2022-01-17 R/qpcr-compile.R

@@ -20,13 +20,15 @@ targets <- c("UBTF F4R4",
              "rRNA18S F2R2",
              "rRNA45S F1R1",
              "RPL32 F1R1", 
+             "RPL3 F2R2",
+             "RPL3L F4R4",
              "rpS6 F2R2",
              "MyHC1 F1R1",
              "MyHC2A F5R5",
              "MyHC2X F5R5",
              "Lambda KIT")  
 
-batch.filtered<- batch %>%
+batch.filtered <- batch %>%
   filter(target %in% targets) %>%
   print()
 

### 2022-01-18 response-to-reviews-rev1.Rmd
fixed all figures for revision

@@
 Could the authors perhaps provide raw data and membranes for RPS6 and UBF just for the reviewers?
+
+```{r}
+## Raw data of western blot membranes
+
+western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
+# Save data 
+western_data <- readRDS("./data/derivedData/western-compile/western_data_calibrated.RDS")
+
+western_data %>%
+  mutate(detrain = if_else(time == "post1w", "detrain", "train"), 
+         time = if_else(time %in% c("post1w", "postctrl"), "12", 
+                        if_else(time == "S1c", "1", 
+                                gsub("S", "", time))), 
+         group = if_else(cond == "ctrl_leg", "ctrl", "train"), 
+         time = as.numeric(time), 
+         time.fact = as.factor(time)) %>%
+
+  group_by(target, participant, group, time.fact, detrain) %>%
+  summarise(expression = mean(expression)) %>%
+
+  ggplot(aes(time.fact, expression, color = group)) + 
+  geom_violin() + 
+  geom_jitter(height = 0, width = 0.1) +
+  facet_wrap(~ target)
+
+```
 

### 2022-01-18 Figures/Figure1.R

@@ -18,7 +18,7 @@ bayes <- TRUE
 
 source("./R/figure-source.R")
 source("./R/libs.R")
-
+library(grid)
 
 
 
@@ -26,9 +26,8 @@ source("./R/libs.R")
 
 
 
-control_color <- "white"
-
-experiment_color <- "#4abdcb"
+control_color <- "#b42a51"
+experiment_color <- "#4f81bd"
 
 
 
@@ -51,7 +50,10 @@ design_exp <- expand.grid(cond = c("Variable volume", "Constant volume"),
                         if_else(cond == "Variable volume" & session %in% c(9, 10, 11, 12), 9, sets))) %>%
   ggplot(aes(session, sets, fill = cond)) +
   
-  annotate("text", x = -2, y = 11.0, label = c("Experimental group"), size = textsize + 0.5) +  
+  annotate("text", x = -2, y = 11.2, 
+           label = c("Experimental group"), 
+           size = textsize + 0.5, 
+           fontface = 2) +  
   
   geom_bar(stat = "identity", position = position_dodge(width = 0.55),
                  
@@ -64,9 +66,9 @@ design_exp <- expand.grid(cond = c("Variable volume", "Constant volume"),
   
   scale_y_continuous(breaks = c(0, 3, 6, 9, 12), limits = c(0,12), expand = c(0,0), 
                      labels = c("", "3", "6", "9", "")) +
-  scale_fill_manual(values = c(color.scale[2], color.scale[1])) +
+  scale_fill_manual(values = c(color.scale[4], color.scale[2])) +
 
-  annotate("text", x = c(0, 0, -2), y = c(9.1, 8.2, 6), 
+  annotate("text", x = c(0, 0, -1.8), y = c(9.1, 8.2, 6), 
            label = c("Muscle biopsy", "Muscle thickness", "Strength tests"), 
            vjust = 0, hjust = 1, alpha = 0.8, size = textsize) + 
   
@@ -137,6 +139,13 @@ design_exp <- expand.grid(cond = c("Variable volume", "Constant volume"),
 
 ### Design graph: Control group #####
 
+
+# Rounded corners can be achieved by a hack
+
+# devtools::install_github("teunbrand/elementalist")
+library(elementalist)
+
+
 design_ctrl <- expand.grid(cond = c("Variable volume", "Constant volume"),
                                     session = seq(1:12), 
                                     sets = rep(6)) %>%
[…]

### 2022-01-18 Figures/Figure2.R

@@ -9,7 +9,7 @@
 ##
 ##-------------------------------------
 ## Notes: 
-# Figure 2 contains wetsern blot data 
+# Figure 2 contains western blot data 
 # and corresponding mRNA data (rps6 and UBF)
 #
 #
@@ -88,28 +88,29 @@ prot_fold_change <- tx_results_western %>%
                         labels = c("Control", "Training", 
                                    "Training + de-training"))) %>%
 
-  ggplot(aes(time, Estimate, fill = group)) + 
+  ggplot(aes(time, Estimate, fill = group, shape = group)) + 
   
   
   geom_hline(yintercept = 1, lty = 2, color = "gray80") +
   
   geom_errorbar(aes(ymin = CI.Lower, ymax = CI.Upper), 
-                position = position_dodge(width = 0.2), 
-                width = 0) +
-  geom_point(position = position_dodge(width = 0.2), shape = 21) + 
+                position = position_dodge(width = 0.4), 
+                size = error.size,
+                width = 0.2) +
+  geom_point(position = position_dodge(width = 0.4)) + 
   
   facet_wrap(~ target , ncol = 1, strip.position = "top", scales = "free") +
  # scale_y_continuous(limits = c(0.5, 3), breaks = c(1, 2, 3), 
  #                    expand = c(0,0)) +
   
-  scale_fill_manual(values = c(color.scale[5], color.scale[6],color.scale[8])) +
-  
+  scale_fill_manual(values = c(color.scale[1], color.scale[2],color.scale[4])) +
+  scale_shape_manual(values = c(21, 22, 24)) +
   
   labs(y = "Fold change from Baseline") +
   plot_theme()  +
   theme(strip.placement = "top",
         strip.background = element_rect(fill = "white", color = "white"), 
-        strip.text = element_text(size = 7, hjust = 0),
+        strip.text = element_text(size = 7, hjust = 0, face = 2),
         legend.position = "bottom",
         legend.text = element_text(size = 7, margin = margin(t = 0.1, b= 0.1,r = 0.1,l = 0.1, unit = "pt")),
         legend.key.size = unit(0, "cm"),
@@ -133,22 +134,27 @@ prot_interaction <- tx_results_western %>%
                                         "Post-training\n+de-training")), 
          comparison = fct_rev(comparison), 
[…]

### 2022-01-18 Figures/Figure3.R

@@ -33,6 +33,10 @@ primers <- read_excel("./data/tr010_primers.xlsx") %>%
 
 
 
+# Rounded corners can be achieved by a hack
+
+# devtools::install_github("teunbrand/elementalist")
+library(elementalist)
 
 
 
@@ -49,12 +53,12 @@ primers_fig <- data.frame(x = c(1, 13500),
   # 28S segment
   annotate("segment", x = 7925, xend = 12990, y = 1.5, yend = 1.5, size = 2.5) +
   
-  geom_segment(data = primers, aes(y = c(1.54, 1.52, 1.528, 1.54, 1.55, 1.56), 
-                                   yend = c(1.54, 1.52, 1.528, 1.54, 1.55, 1.56), 
+  geom_segment(data = primers, aes(y = c(1.55, 1.52, 1.528, 1.55, 1.55, 1.575), 
+                                   yend = c(1.55, 1.52, 1.528, 1.55, 1.55, 1.575), 
                                    x= on45S_start, xend = on45S_end), 
                size = 4) +
   
-  geom_text(data = primers, aes(y = c(1.54, 1.52, 1.528, 1.544, 1.55, 1.56),
+  geom_text(data = primers, aes(y = c(1.55, 1.52, 1.528, 1.552, 1.55, 1.575),
                                       x= on45S_end, label = symbol_ext), 
                   size = 2.2, 
           hjust = 0,
@@ -76,6 +80,10 @@ primers_fig <- data.frame(x = c(1, 13500),
   scale_y_continuous(limits = c(1.46, 1.58)) +
   plot_theme() + 
   theme(axis.title = element_blank(), 
+        plot.background = element_rect_round(color = "black", 
+                                             size = 0.25, 
+                                             radius = unit(0.25, "cm"),
+                                             fill = "#fefeda"),
         axis.text = element_blank(), 
         axis.line = element_blank(),
         axis.ticks = element_blank())
@@ -115,7 +123,7 @@ anno.df <- data.frame(target = unique(complete_rrna_comp$target),
                      "rRNA 5.8S",
                      "rRNA 5S" ), 
            time = factor("S1", levels = c("S1", "post")), 
-           estimate = 3.5) %>%
+           estimate = 3.85) %>%
   mutate(target = factor(target, levels = c("rRNA18S F2R2",
                                             "rRNA5.8S F2R2",
[…]

### 2022-01-18 Figures/Figure4.R

@@ -111,16 +111,22 @@ loo_panel <- res %>%
     
 #  geom_density_ridges(color = "white") +
   
-  geom_point(position = position_nudge(y = 0.2), shape = 21, alpha = 0.4) +
+  geom_point(position = position_nudge(y = 0.2),
+             size = 1.2,
+             shape = 21, 
+             alpha = 0.4) +
   
   facet_wrap( ~ coef, scales = "free") +
   
   scale_fill_manual(values = c(color.scale[5], color.scale[4], color.scale[4])) +
 
-  geom_point(data = res2, aes(estimate, Participant, fill = NULL)) +
+
   geom_errorbarh(data = res2, aes(fill = NULL, x = estimate, y = Participant, 
                                   xmax =  upr, xmin = lwr), 
                  height = 0) + 
+  
+  geom_point(data = res2, aes(estimate, Participant), 
+             fill = color.scale[3], shape = 22) +
   
   
   # Adds full model estimates
@@ -146,7 +152,8 @@ loo_panel <- res %>%
   
   plot_theme() + 
   theme(strip.background = element_rect(color = "white", fill = "white"),
-        strip.text = element_text(size = 8, hjust = 0),
+        
+        strip.text = element_text(size = 7, hjust = 0),
         axis.title.y = element_text(size = 7), 
         legend.position = "none")
 
@@ -170,8 +177,8 @@ prediction_slope <- predict_df %>%
   
   inner_join(leg) %>%
   mutate(cond = factor(cond, levels = c("const", "var"), 
-                       labels = c("Constant volume", 
-                                  "Variable volume")))  %>%
+                       labels = c("CONST", 
+                                  "VAR")))  %>%
 
   group_by(participant, leg, cond) %>%
   summarise(slope = mean(slope), 
@@ -196,13 +203,15 @@ prediction_slope <- predict_df %>%
                      breaks = c(-3, -2, -1, 0, 1, 2, 3, 4, 5), 
                      labels = c("", -2, "", 0, "", 2, "", 4, "")) +
   
-  scale_fill_manual(values = c(color.scale[1], color.scale[2])) +
+  scale_fill_manual(values = c(color.scale[2], color.scale[4])) +
   labs(x = "Total RNA increase per session (%)", 
        y = "Muscle thickness change (mm)") +
   plot_theme() + 
   theme(axis.title.y = element_text(size = 7), 
[…]

### 2022-01-18 R/figure-source.R

@@ -23,7 +23,7 @@ single.color   <- "#7f7f7f"
 
 # line sizes
 line_size <- 0.3
-
+error.size <- 0.3
 # text sizes
 label.size <- 10
 
@@ -36,7 +36,14 @@ sequential.colors <- c("#1b9e77",
                        "#66a61e",
                        "#e6ab02")
 
+
+# Set Font to Arial 
+
+windowsFonts()
+
+
 # Functions for plotting  #####
+
 
 plot_theme <- function() {
   
@@ -50,8 +57,8 @@ plot_theme <- function() {
           axis.title = element_text(color = "black", size = 7),
           legend.title = element_blank(), 
           legend.background = element_rect(fill = "white"),
-          legend.margin = margin(t = 0, r = 1, b = 1, l = 1, unit = "pt"),
-          legend.key = element_rect(fill = "white"),
+       #   legend.margin = margin(t = 0, r = 1, b = 1, l = 1, unit = "pt"),
+          legend.key = element_rect(fill = "white", color = "white"),
           legend.position = c(0.85, 0.9)) 
   
   

### 2022-01-18 R/figure-source.R
small updates

@@ -23,7 +23,7 @@ single.color   <- "#7f7f7f"
 
 # line sizes
 line_size <- 0.3
-error.size <- 0.3
+error.size <- 0.25
 # text sizes
 label.size <- 10
 

### 2022-01-19 response-to-reviews-rev1.Rmd
misc changes to reviwer responses

--- 2022-01-17 response-to-reviews-rev1.Rmd
@@
 output: pdf_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 Could the authors perhaps provide raw data and membranes for RPS6 and UBF just for the reviewers?
+
+We agree with the referee that the blots presented in Figure 2 were faint, especially for UBF.
+The selected blot was made for illustrative purposes as we had ordered the samples according to sampling time and condition in that particular blot.
+Unfortunately this blot was in the "fainter end" of the spectrum.
+We have replaced the blot with one that we feel is more representative.
+
+The raw aggregated data (averaged over replicates per samples) does show a clear increase in signal over the time course (Figure 1).
+We acknowledge that the western blot technique is associated with considerable technical variation.
+
+```{r}
+## Raw data of western blot membranes
+library(tidyverse)
+
+western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
+
+western_data %>%
+  group_by(sample, target) %>%
+  summarise(n = n(), 
+            m = mean(expression, na.rm = TRUE), 
+            s = sd(expression, na.rm = TRUE), 
+            cv = s/m) %>%
+
+  group_by(target) %>%
+  summarise(cv = mean(cv, na.rm = TRUE)) %>%
+  print()
+
+## ICC
+library(lme4)
+ubf <-   lmer(expression ~ gel + (1|sample), 
+            data = western_data[western_data$target == "t-UBF",])
+
+s6 <-   lmer(expression ~ gel + (1|sample), 
+            data = western_data[western_data$target == "t-s6",])
+
+data.frame(VarCorr(ubf))[1, 4] / sum(data.frame(VarCorr(ubf))[, 4])
+
+data.frame(VarCorr(s6))[1, 4] / sum(data.frame(VarCorr(s6))[, 4])
+
+  ggplot(aes(sample, n)) + geom_point()
+  print()
+
+western_data %>%
+  mutate(detrain = if_else(time == "post1w", "detrain", "train"), 
+         time = if_else(time %in% c("post1w", "postctrl"), "12", 
+                        if_else(time == "S1c", "1", 
+                                gsub("S", "", time))), 
+         time = if_else(detrain == "detrain", "13", time),
+         group = if_else(cond == "ctrl_leg", "ctrl", "train"), 
+         time = as.numeric(time), 
+         time.fact = as.factor(time)) %>%
+
+  group_by(target, participant, group, time.fact, detrain) %>%
+  summarise(expression = mean(expression)) %>%
+
+  ggplot(aes(time.fact, expression, color = group)) + 
+  geom_violin() + 
+  geom_jitter(height = 0, width = 0.1) +
+  facet_wrap(~ target)
+
+```
 

### 2022-01-19 R/western-compile.R

@@ -16,6 +16,8 @@ source("./R/libs.R")
 
 ### Load data  round 1 ####
 
+
+
 western_data_round1 <- read_excel("./data/wetlab/western/tr010_western_round1.xlsx", na = "NA", sheet = "total_protein") %>%
   filter(participant != "LADDER") %>%
   dplyr::select(participant, ExtractionNR, round, gel, well, mean.gray1:total.protein2) %>%

### 2022-02-01 response-to-reviews-rev1.Rmd
updates revision

--- 2022-01-17 response-to-reviews-rev1.Rmd
@@
 output: pdf_document
+editor_options: 
+  chunk_output_type: console
+  markdown: 
+    wrap: 72
+bibliography: references.bib
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+```
 
@@
 
-> This manuscript investigates the induction of ribosome biogenesis and accumulation of RNA/ribosomes during the early phase of resistance training.
-Subjects performed unilateral RE bilaterally but with differing loading protocols.
-One leg trained with a constant volume whereas the other trained with variable volume.
-Control subjects were biopsied at relevant time points.
-The manuscript presents evidence in favor of that skeletal muscle in the naïve state is very responsive with respect to ribosome biogenesis following acute RE.
-With repeated training sessions this induction is maintained but not further stimulated.
-Most of the effect on total RNA was seen during the first training block.
-UBF levels were related to RNA accumulation and rate of total RNA accumulation predicted RT-induced muscle hypertrophy.
-With eight days of detraining, around 20% of total RNA was lost, indicating that the elevated RNA pool needs to be maintained not to decrease.
-The manuscript is highly interesting and ambitious in its design.
-I think that it adds to the published literature and could potentially be a future important reference piece.
-However, I have some concerns and comments that I’d like for the authors to address, as I think that they can improve the manuscript further.
+> This manuscript investigates the induction of ribosome biogenesis and
+> accumulation of RNA/ribosomes during the early phase of resistance
+> training. Subjects performed unilateral RE bilaterally but with
+> differing loading protocols. One leg trained with a constant volume
+> whereas the other trained with variable volume. Control subjects were
+> biopsied at relevant time points. The manuscript presents evidence in
+> favor of that skeletal muscle in the naïve state is very responsive
+> with respect to ribosome biogenesis following acute RE. With repeated
+> training sessions this induction is maintained but not further
+> stimulated. Most of the effect on total RNA was seen during the first
+> training block. UBF levels were related to RNA accumulation and rate
+> of total RNA accumulation predicted RT-induced muscle hypertrophy.
+> With eight days of detraining, around 20% of total RNA was lost,
+> indicating that the elevated RNA pool needs to be maintained not to
+> decrease. The manuscript is highly interesting and ambitious in its
+> design. I think that it adds to the published literature and could
+> potentially be a future important reference piece. However, I have
+> some concerns and comments that I'd like for the authors to address,
+> as I think that they can improve the manuscript further.
 
-> The western blot membranes in the paper does not really translate into the presented data.
-In my opinion the bands look rather faint and equal in intensity across the time course.
-Could the authors perhaps provide raw data and membranes for RPS6 and UBF just for the reviewers?
+> The western blot membranes in the paper does not really translate into
+> the presented data. In my opinion the bands look rather faint and
+> equal in intensity across the time course. Could the authors perhaps
+> provide raw data and membranes for RPS6 and UBF just for the
+> reviewers?
+
+We agree with the referee that the blots presented in Figure 2 were
+faint, especially for UBF. The selected blot was made for illustrative
+purposes as we had ordered the samples according to sampling time and
+condition in that particular blot. Unfortunately this blot was in the
+"fainter end" of the spectrum. We have replaced the blot with one that
+we feel is more representative. We are providing the quantified blots in
+a separate document for the reviewers.
+
+The protein samples were extracted from Trizol preparations, something
+that has been proven difficult but possible (see e.g. [@wen2020;
+@kopec2]). Some of our samples were also difficult to dissolve after
+precipitation and this sometimes gave unsatisfactory results during the
+western blot procedure. Problematic samples were re-extracted and
+re-run. Importantly, assessment of sample quality and any re-extraction
+was performed on samples given a random sample id. De-coding of the
+random sample id was performed only at data analysis, and to create runs
+with ordered samples.
+
+```{r}
+## Raw data of western blot membranes
+library(tidyverse)
+
+western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
+
+### CV of duplicate measurements
+
+western_cv <- western_data %>%
+  group_by(sample, target) %>%
+  summarise(n = n(), 
+            m = mean(expression, na.rm = TRUE), 
+            s = sd(expression, na.rm = TRUE), 
+            cv = s/m) %>%
+
+  group_by(target) %>%
+  summarise(cv = mean(cv, na.rm = TRUE)) %>%
+  pull(cv)
+
+cvs <- paste0(round(min(western_cv) * 100, 1), " and ", round(max(western_cv) * 100, 1), "%")
+
+```
+
+The raw aggregated data (averaged over replicates per samples) does show
+a clear increase in signal over the time course (\@ref(fig:western-fig)). We acknowledge that the western blot technique is associated with
+considerable technical variation. The average coefficient of variation
+(CV) in replicates were `r cvs` for RPS6 and UBF respectively, both within the
+expected range of CV from western blot data [@degasperi2014].
+
+For transparency, we have included assay performance in the the revised version of the manuscript.
+
+```{r western-fig, fig.cap = "Raw data on arbitrary unit scale and normalized to baseline samples.", fig.width = 7, fig.height = 8}
+
+color.scale <- c("#ffffff", "#4f81bd", "#7f7f7f", "#b42a51")
+
+library(lme4) ; library(cowplot)
+
+absolute_data <- western_data %>%
+  mutate(detrain = if_else(time == "post1w", "detrain", "train"), 
+         time = if_else(time %in% c("post1w", "postctrl"), "12",
+                        if_else(time == "S1c", "1", gsub("S", "", time))), 
+         time = if_else(detrain == "detrain", "13", time), 
+         group = if_else(cond == "ctrl_leg", "ctrl", "train"), 
+         time = as.numeric(time), 
+         time.fact = as.factor(time), 
+         group = factor(group, levels = c("ctrl", "train"), 
+                        labels = c("Control group", 
+                                   "Intervention group"))) %>%
+
+  group_by(target, participant, group, time.fact, detrain) %>%
+
+  summarise(expression = mean(expression, na.rm = TRUE)) %>%
+
+  ungroup() %>%
+
+  ggplot(aes(time.fact, expression, fill = group)) +
+  geom_boxplot() + 
+  geom_jitter(height = 0, width = 0.1, shape = 21) +
+
+  scale_fill_manual(values = c(color.scale[1], color.scale[2])) +
+
+  facet_wrap(~ target) +
+
+  labs(x = "Time-point", 
+       y = "Western blot signal (AU)", 
+       title = "Raw western blot data averaged over sample", 
+       fill = "Group")  +
+
+  theme(panel.background = element_rect(fill = "#fefeda"), 
+        panel.grid.major.y = element_line(color = "gray80"), 
+        panel.grid.major.x = element_blank(), 
+        legend.position = "bottom")
+
+relative_data <- western_data %>%
+  mutate(detrain = if_else(time == "post1w", "detrain", "train"), 
+         time = if_else(time %in% c("post1w", "postctrl"), "12",
+                        if_else(time == "S1c", "1", gsub("S", "", time))), 
+         time = if_else(detrain == "detrain", "13", time), 
+         group = if_else(cond == "ctrl_leg", "ctrl", "train"), 
+         time = as.numeric(time), 
+         time.fact = as.factor(time), 
+         group = factor(group, levels = c("ctrl", "train"), 
+                        labels = c("Control group", 
+                                   "Intervention group"))) %>%
+
+  group_by(target, participant, group, time.fact, detrain) %>%
+  mutate(baseline = ifelse(time == 0, expression,NA )) %>%
+
+  group_by(target, participant, leg) %>% 
+  mutate(baseline = mean(baseline, na.rm = TRUE), 
+         expression = expression / baseline) %>%
+
+  group_by(target, participant, group, time.fact, detrain) %>%
+
+  summarise(expression = mean(expression, na.rm = TRUE)) %>%
+
+  ungroup() %>%
+
+  ggplot(aes(time.fact, expression, fill = group)) +
+  geom_boxplot() + 
+  geom_jitter(height = 0, width = 0.1, shape = 21) +
+
+  facet_wrap(~ target) +
+    scale_fill_manual(values = c(color.scale[1], color.scale[2])) +
+  labs(x = "Time-point", 
+       y = "Western blot signal (AU) / Baseline values", 
+       title = "Raw western blot data averaged over sample normalized to baseline", 
+       fill = "Group")  +
+
+  theme(panel.background = element_rect(fill = "#fefeda"), 
+        panel.grid.major.y = element_line(color = "gray80"), 
+        panel.grid.major.x = element_blank(), 
+        legend.position = "none")
+
+plot_grid(absolute_data, relative_data, 
+          ncol = 1, rel_heights = c(0.55, 0.45))
+
+```
 
@@
 
-One transcribed pre-rRNA results in a single end product, one ribosome and in the context of protein synthesis the amount of ribosomes per cell or tissue mass is highly correlated with protein synthesis (Milward 1976, West 2016).
-In contrast to rRNA processing, mRNA transcription and translation could to some degree be seen as competitive between different mRNA.
-
-This translates to different normalization strategies. The use of a houskeeping gene when investigating ribosomal biogenesis is not interesting as it does not
+One transcribed pre-rRNA results in a single end product, one ribosome, and in the context of protein synthesis, the amount of ribosomes per cell or tissue mass is highly correlated with protein synthesis [@millward1973; @west2016].
+In contrast to rRNA processing, mRNA transcription and subsequent translation could, to some degree, be seen as competitive between different mRNA.
+This translates to different normalization strategies where we believe that the most biologically relevant perspective on rRNA is expressing it to the amount of tissue probed in the qPCR.
+The use of a housekeeping or reference gene in this respect could be misleading as the relative stability of the reference gene must be determined to some quantity of interest.
+Normalization of mRNA data using reference genes assumes that such stability has been determined as relative to the mRNA pool [@andersen2004].
+We have however noted that in response to RT, relative stability is hard to determine as the whole mRNA pool is increasing [@khan2020].
+This again brings us to the basic problem of normalization, what is the underlying quantity of interest?
+We have not attempted to determine relative stability of reference genes in the present data set because of our past experiences, therefore we are unwilling to explore the effect of using a reference gene, as the reviewer suggests.
+A simple alternative to determining the relative stability of a reference gene is to use an external (exogenous) reference together with the known quantity of tissue used in sample preparations.
+The use of exogenous reference genes has previously been used to achieve "custom normalization" to increase biological relevance (e.g.
+[@ellefsen2008]).
+We have also previously used the same approach as in the present work to conclude that RT leads to increased abundance of pre-rRNA and rRNA in muscle tissue [@hammarström2020].
 
 > Could the authors please comment on the apparent disconnect between increase in RNA and RPS6 protein levels?
+
+We interpret the results from our analysis as affected by a time-lag between increases in synthesis of total RNA (presumably rRNA) and RPS6.
+The model used "controls" for both quantities being increased over the whole training period, but within a specific time-point there exists no relationship between RPS6 and total RNA.
+This separates our analysis from others who have shown correlations between RPS6 and rRNA without controlling for, in their case, degree of stimuli [@nakada2016] and without examining the time course of their increase.
+As we have already pointed to in the manuscript, our results suggests different temporal regulation or turnover rates in total RNA and RPS6 protein levels.
 
@@
 
-The referee raises an interesting point
+The reviewer raises an interesting point, that we interpret as suggesting that newly synthesized ribosomes would be more beneficial for muscle growth in response to a new stimuli as novel ribosomes are potentially specialized for growth rather than cellular status quo.
+In light of recent studies and overviews highlighting ribosomal heterogoneity, our results are worth following up in relation to a specialized ribosome.
+
+We have added a section in the discussion highlighting this aspect of the results.
 
 > Please comment on that UBF levels where maintained at the detraining time point but total RNA decreasing. Would it be relevant to analyze UBF phosphorylation?
+
+Although not robustly decreased, the overall tendency of total UBF protein was decrease in response to the de-training period.
+However, as the reviewer points to, we did not measure the "active" UBF nor UBF in the context of rDNA chromatin regulation.
+It is possible that in the de-training period a reduced stimuli (possibly through reduced mTORC1 stimulation [@nader2005]) reduced interaction between UBF and rDNA.
+Unfortunately, a limited amount of tissue (together with other resources) prevented us from doing more elaborate analyses, including UBF phosphorylation or UBF/rDNA interactions.
 
@@
 > The methods in the abstract must be improved.
-In its current form it lacks key information such as training duration, the number of controls are states as n=7 but n=8 in the M&M section and I’d like for the authors to try and specify biopsy time points.
+In its current form it lacks key information such as training duration, the number of controls are states as n=7 but n=8 in the M&M section and I'd like for the authors to try and specify biopsy time points.
 
-> Page 3, line 48-49.
-I encourage the authors to consider including an additional reference (Figueiredo et al 2021 J Phys).
-This paper shows the specific response of ribosome biogenesis in response to RE and not Endurance type exercise.
+We thank the reviewer for catching this typo. We have updated the abstract with the correct information and additional information regarding biopsy time points.
 
-> Page 4, line 8: Please clarify “numberical lowering”. Could this be rephrased?
+> Page 3, line 48-49. I encourage the authors to consider including an
+> additional reference (Figueiredo et al 2021 J Phys). This paper shows
+> the specific response of ribosome biogenesis in response to RE and not
+> Endurance type exercise.
 
-> Page 4, line 10: Double “..”
+> Page 4, line 8: Please clarify "numberical lowering". Could this be
+> rephrased?
 
-> Page 4, line 15: Double “..”
+> Page 4, line 10: Double ".."
 
-> Page 4, line 26: Should “to” be removed?
+> Page 4, line 15: Double ".."
+
+> Page 4, line 26: Should "to" be removed?
 
@@
 
-> Page 11, line 35: Should “a” be removed? Please double check if it’s 7 or 8 days of detraining. Abstract states 8.
+> Page 11, line 35: Should "a" be removed? Please double check if it's 7
+> or 8 days of detraining. Abstract states 8.
 
-> Page 12, line 19: please clarify “leg being training in the rest period between sets in the first leg. Does this means that they training continuously without break?
+> Page 12, line 19: please clarify "leg being training in the rest
+> period between sets in the first leg. Does this means that they
+> training continuously without break?
 
-> Page 13, line 52: 2-4 passes were made with the Bard Magnum 12-14 gauge needles but only 1-2 aliquots of the samples were frozen.
-What happened to the other material?
-If I misunderstood, can the text be rewritten for clarity?
+> Page 13, line 52: 2-4 passes were made with the Bard Magnum 12-14
+> gauge needles but only 1-2 aliquots of the samples were frozen. What
+> happened to the other material? If I misunderstood, can the text be
+> rewritten for clarity?
 
-> Page 14, line 8: Was the spike-in used to normalize RNA extraction efficiency? If yes, please state this clearly.
+> Page 14, line 8: Was the spike-in used to normalize RNA extraction
+> efficiency? If yes, please state this clearly.
 
-> Page 14, line 49: Does “normalized” refer to denatured?
+> Page 14, line 49: Does "normalized" refer to denatured?
 
-> Page 15, line 15: Does the authors refer to amplicon size when they write primer sizes?
+> Page 15, line 15: Does the authors refer to amplicon size when they
+> write primer sizes?
 
-> Page 31, Figure legends: I think that the figure legends could benefit from a bit more substance. For example Fig 1A, perhaps indicate bar color as black (CONST) and yellow (VAR)?
+> Page 31, Figure legends: I think that the figure legends could benefit
+> from a bit more substance. For example Fig 1A, perhaps indicate bar
+> color as black (CONST) and yellow (VAR)?
 
@@
 
-> The authors sought to perform a time course investigation of ribosome density and biogenesis markers after knee extensor training.
-Phenotype variables were also assessed.
-Training led to increases in muscle growth (VL thickness) as well as in total RNA, rRNA, UBF and rpS6.
-Training volume did not play a role in these adaptations.
-Interestingly, the rate of total RNA increases predicted hypertrophy.
-Likewise, training cessation led to decrements in total RNA abundance.
-I love this study.  This continues to contribute to our knowledge regarding the role of ribosome biogenesis in muscle hypertrophy, and is delicate human work with multiple time point biopsies.
-I have minor comments in order of appearance, and congratulate the authors on their fine work.
+> The authors sought to perform a time course investigation of ribosome
+> density and biogenesis markers after knee extensor training. Phenotype
+> variables were also assessed. Training led to increases in muscle
+> growth (VL thickness) as well as in total RNA, rRNA, UBF and rpS6.
+> Training volume did not play a role in these adaptations.
+> Interestingly, the rate of total RNA increases predicted hypertrophy.
+> Likewise, training cessation led to decrements in total RNA abundance.
+> I love this study. This continues to contribute to our knowledge
+> regarding the role of ribosome biogenesis in muscle hypertrophy, and
+> is delicate human work with multiple time point biopsies. I have minor
+> comments in order of appearance, and congratulate the authors on their
+> fine work.
 
@@
 
-> INTRODUCTION 
-- Line 7, minor: “promotes” should read as “promote”
-- Line 14, minor: “with subsequent repeated bouts” should read as “and subsequent repeated bouts”
-- Line 18, minor: “resting synthetic rate of muscle protein” better reads as “basal muscle protein synthesis rates”
-- Lines 48-50, minor: “bouts lead to accumulation of mature rRNA thus also total RNA and presumably functional ribosomes”…this sentence is long, and the word “thus” seems to make it a run-on.
-I’d rephrase the sentence to read more coherently.
-Page 4
-- Lines 10 and 15, minor: each sentence has two stops.
-- Line 33, minor: no need to use the phrase “per se”
-RESULTS
-- Line 51 (and throughout the results), clarification: is “9.2%-point difference” the same as “9.2% difference”?  If so, adjust and remove “point” from all descriptors as this is confusing.
-- Other than this, the results, while dense and full of data, are well-written.
-DISCUSSION
-Page 7
-- Line 17, minor: again, “per se” can be removed.
-- Line 19, minor: “was not affected” better reads as “were not affected”
-- Line 39, minor: “session” should read as “sessions”
+> INTRODUCTION - Line 7, minor: "promotes" should read as "promote" -
+> Line 14, minor: "with subsequent repeated bouts" should read as "and
+> subsequent repeated bouts" - Line 18, minor: "resting synthetic rate
+> of muscle protein" better reads as "basal muscle protein synthesis
+> rates" - Lines 48-50, minor: "bouts lead to accumulation of mature
+> rRNA thus also total RNA and presumably functional ribosomes"...this
+> sentence is long, and the word "thus" seems to make it a run-on. I'd
+> rephrase the sentence to read more coherently. Page 4 - Lines 10 and
+> 15, minor: each sentence has two stops. - Line 33, minor: no need to
+> use the phrase "per se" RESULTS - Line 51 (and throughout the
+> results), clarification: is "9.2%-point difference" the same as "9.2%
+> difference"? If so, adjust and remove "point" from all descriptors as
+> this is confusing. - Other than this, the results, while dense and
+> full of data, are well-written. DISCUSSION Page 7 - Line 17, minor:
+> again, "per se" can be removed. - Line 19, minor: "was not affected"
+> better reads as "were not affected" - Line 39, minor: "session" should
+> read as "sessions"
 
-> I urge the authors to be creative somewhere in the discussion by perhaps adding a paragraph after the one found in lines 29-56.
-We’ve done a lot of work in the ribosome biogenesis area, and we’ve thought a good bit about total RNA/unit muscle with regard to what it represents.
-The authors correctly articulated this as a proxy of ribosome density.
-However, as muscle tissue and myofibers volumetrically increase, this will affect the metric (the same holds true with mitochondrial biogenesis and using CS activity as a proxy of mitochondrial volume density).
-An easy way to interpret this is in the following example:
-o Muscle tissue increases 5% in thickness with RT (and myofibers scale accordingly as well)
-o Total RNA/unit muscle does not change in value
-Most would interpret this as ribosome biogenesis not occurring.
-However, biogenesis likely occurred because, as muscle grew, ribosome biogenesis scaled with growth thus leading to no change in RNA concentration.
-Our laboratory discusses this important concept with regard to RT and mitochondrial biogenesis (PMID: 32162291, PMID: 34646153).
+> I urge the authors to be creative somewhere in the discussion by
+> perhaps adding a paragraph after the one found in lines 29-56. We've
+> done a lot of work in the ribosome biogenesis area, and we've thought
+> a good bit about total RNA/unit muscle with regard to what it
+> represents. The authors correctly articulated this as a proxy of
+> ribosome density. However, as muscle tissue and myofibers
+> volumetrically increase, this will affect the metric (the same holds
+> true with mitochondrial biogenesis and using CS activity as a proxy of
+> mitochondrial volume density). An easy way to interpret this is in the
+> following example: o Muscle tissue increases 5% in thickness with RT
+> (and myofibers scale accordingly as well) o Total RNA/unit muscle does
+> not change in value Most would interpret this as ribosome biogenesis
+> not occurring. However, biogenesis likely occurred because, as muscle
+> grew, ribosome biogenesis scaled with growth thus leading to no change
+> in RNA concentration. Our laboratory discusses this important concept
+> with regard to RT and mitochondrial biogenesis (PMID: 32162291, PMID:
+> 34646153).
 
-> Anyhow, this is no critique of the authors’ hard and excellent work.
-However, I think the authors re-iterating this point as a stand-alone paragraph would benefit readership.
-In addition, I believe taking a moment to define what the total RNA/ribosome density metric conceptually represents, and then (in this newly added paragraph) briefly walking the reader through each phase of the data, and what could be going on with biogenesis in the context of a growing muscle would be interesting to see.
-For instance, in Figure 3G, my interpretation (much like the authors’) is that ribosome biogenesis is very rapid at the onset of RT where little functional growth is likely occurring.
-And like the authors, I believe this likely was needed for eventual growth to occur.
-However, the “no change” in RNA concentrations session 9-12 as muscle is continuing to grow may not indicate a plateau in biogenesis given that this is the time during which muscle growth is likely starting to ramp up.
-Hence, biogenesis rates could be equally as ramped here compared to sessions 1-9 of RT.
+> Anyhow, this is no critique of the authors' hard and excellent work.
+> However, I think the authors re-iterating this point as a stand-alone
+> paragraph would benefit readership. In addition, I believe taking a
+> moment to define what the total RNA/ribosome density metric
+> conceptually represents, and then (in this newly added paragraph)
+> briefly walking the reader through each phase of the data, and what
+> could be going on with biogenesis in the context of a growing muscle
+> would be interesting to see. For instance, in Figure 3G, my
+> interpretation (much like the authors') is that ribosome biogenesis is
+> very rapid at the onset of RT where little functional growth is likely
+> occurring. And like the authors, I believe this likely was needed for
+> eventual growth to occur. However, the "no change" in RNA
+> concentrations session 9-12 as muscle is continuing to grow may not
+> indicate a plateau in biogenesis given that this is the time during
+> which muscle growth is likely starting to ramp up. Hence, biogenesis
+> rates could be equally as ramped here compared to sessions 1-9 of RT.
 
@@
 
-- Lines 9-19: Authors stated that 18 participants were recruited, but then go on to say that n=11 TRAIN and n=8 CTRL completed.  Please reconcile.
-- Line 35: Authors stated 7 days after detraining, but in the discussion (page 7, line 19) they mention after 8 days of detraining.  Please reconcile.
-Page 12
-- Line 39, minor: “restitution” should read as “rest”.  Please change accordingly, and if used throughout the paper, please alter.
-Page 15
-- Line 36, minor: “Glycin” should read as “Glycine”.  In addition, and molar unit is needed.
+-   Lines 9-19: Authors stated that 18 participants were recruited, but
+    then go on to say that n=11 TRAIN and n=8 CTRL completed. Please
+    reconcile.
+-   Line 35: Authors stated 7 days after detraining, but in the
+    discussion (page 7, line 19) they mention after 8 days of
+    detraining. Please reconcile. Page 12
+-   Line 39, minor: "restitution" should read as "rest". Please change
+    accordingly, and if used throughout the paper, please alter. Page 15
+-   Line 36, minor: "Glycin" should read as "Glycine". In addition, and
+    molar unit is needed.

### 2022-02-01 manuscript.Rmd
updates revision

@@
 csl: resources/acta-physiologica.csl
-bibliography: resources/bibliography.bib
+bibliography:
+  - resources/bibliography.bib
+  - references.bib
 header-includes: 
@@
 
-**Methods:** A training group (n=11) performed unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg.
-Vastus lateralis biopsies obtained at rest and 48 hrs after selected training sessions and after eight days of de-training and from a non-training control group (n=7) were assayed for ribosome abundance and biogenesis markers.
+**Methods:** A training group (n = 11) performed <span custom-style="CHANGE">12 sessions (3-4 sessions per week) of</span> unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg.
+<span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from</span> vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline and 48 hrs after sessions 1, 4, 5, 8, 9, 12</span> and after eight days of de-training and from a non-training control group (n = 8).
 In addition, muscle thickness was measured before and after training and control periods.
@@
 Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
+<span custom-style="CHANGE">Furthermore, in light of accumulating evidence of tissue specific and heterogeneous ribosomal populations [@chaillou2019; @chaillou2015; @gupta2014], our results suggests a potential role of specialized newly synthesized ribosomes in muscle hypertrophy.
+Our results are however limited in terms of probing any such heterogeneity.
+</span>
 Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.
@@
 [@RN2225].<!-- Haun total RNA increase througout a 6 week intervention -->
-Altough this observation was done in well-trained participants performing a high volume protocol without a control condition with constant volume, compared to constant volume protocols
+Although this observation was done in well-trained participants performing a high volume protocol without a control condition with constant volume, compared to constant volume protocols
 [@RN1809;<!-- Brook 2016 -->
-@RN2358; and the present study],<!-- Hammarström 2020 -->
+@RN2358],<!-- Hammarström 2020 -->
 progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
@@
 After an additional 15-sec restitution period, participants were instructed to push against the lever with maximal effort.
-Within the same assessment session, participants remained seated in the dynamometer for measurement perfomed on both legs.
+Within the same assessment session, participants remained seated in the dynamometer for measurement performed on both legs.
 The first measurement was alternated between legs every other session.
@@
 Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA).
+```{r}
+library(tidyverse)
+
+western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
+
+### CV of duplicate measurements
+
+western_cv <- western_data %>%
+  group_by(sample, target) %>%
+  summarise(n = n(), 
+            m = mean(expression, na.rm = TRUE), 
+            s = sd(expression, na.rm = TRUE), 
+            cv = s/m) %>%
+
+  group_by(target) %>%
+  summarise(cv = mean(cv, na.rm = TRUE)) %>%
+  pull(cv)
+
+cvs <- paste0(round(min(western_cv) * 100, 1), " and ", round(max(western_cv) * 100, 1), "%")
+```
+<span custom-style="CHANGE">The average coefficient of variation across replicates were `r cvs`, for RPS6 and UBF, respectively.</span>
 

### 2022-02-02 response-to-reviews-rev1.Rmd
added revision

@@
 title: "APH-2021-12-0652: Response to the reviewers"
-author: "Daniel Hammarström"
+author: "Daniel Hammarström, Sjur J. Øfsteng, Nicolai B. Jacobsen, Krister B. Flobergseter, Bent R. Rønnestad, Stian Ellefsen"
 date: "2022-01-17"
@@
 
-For transparency, we have included assay performance in the the revised version of the manuscript.
+For transparency, we have included assay performance (CV) in the the revised version of the manuscript.
 
@@
 The use of a housekeeping or reference gene in this respect could be misleading as the relative stability of the reference gene must be determined to some quantity of interest.
-Normalization of mRNA data using reference genes assumes that such stability has been determined as relative to the mRNA pool [@andersen2004].
+Normalization of mRNA data using reference genes assumes that such stability has been determined as relative to the mRNA pool [@andersen2004], or any other quantity.
 We have however noted that in response to RT, relative stability is hard to determine as the whole mRNA pool is increasing [@khan2020].
 This again brings us to the basic problem of normalization, what is the underlying quantity of interest?
-We have not attempted to determine relative stability of reference genes in the present data set because of our past experiences, therefore we are unwilling to explore the effect of using a reference gene, as the reviewer suggests.
+We have not attempted to determine relative stability of reference genes in the present data set because of our past experiences with the RT model, therefore we are unwilling to explore the effect of using a reference gene, as the reviewer suggests.
 A simple alternative to determining the relative stability of a reference gene is to use an external (exogenous) reference together with the known quantity of tissue used in sample preparations.
@@
 
+We thank the reviewer for raising this question as we have attempted to broaden our discussion.
 We interpret the results from our analysis as affected by a time-lag between increases in synthesis of total RNA (presumably rRNA) and RPS6.
-The model used "controls" for both quantities being increased over the whole training period, but within a specific time-point there exists no relationship between RPS6 and total RNA.
+The model used "controls" for both quantities being increased over the whole training period, within a specific time-point there exists no relationship between RPS6 and total RNA.
 This separates our analysis from others who have shown correlations between RPS6 and rRNA without controlling for, in their case, degree of stimuli [@nakada2016] and without examining the time course of their increase.
-As we have already pointed to in the manuscript, our results suggests different temporal regulation or turnover rates in total RNA and RPS6 protein levels.
+As we have already pointed to in the original manuscript, our results suggests different temporal regulation or turnover rates in total RNA and RPS6 protein levels.
+Since ribosomal proteins exists in excess (and are thus not rate limiting for ribosomal biogenesis) [@lam], it is tempting to assume constitutive expression is only indirectly related to the process of ribosomal biogensis.
+We have expanded the discussion by raising the point that unbound ribosomal proteins (including RPS6) may be passively increased by myonuclear accretion.
+Although somewhat speculative, this notion is supported by recent evidence suggesting a specific role of satellite cell fusion in increasing availability of ribosomal proteins together with increased rRNA transcription [@murach2022].
+This could in part help explain our observations since an increased myonuclear pool is presumably maintained during short-term de-training but rRNA transcription is decreased (and/or ribophagy increased).
+
+Our data are however limited to a single ribosomal protein without data supporting accretion of myonuclei. This limitation was also added to the discussion.
 
@@
 Although not robustly decreased, the overall tendency of total UBF protein was decrease in response to the de-training period.
+The variation presumably contributed to relationship with total RNA abundance.
 However, as the reviewer points to, we did not measure the "active" UBF nor UBF in the context of rDNA chromatin regulation.
@@
 Unfortunately, a limited amount of tissue (together with other resources) prevented us from doing more elaborate analyses, including UBF phosphorylation or UBF/rDNA interactions.
+We noted this limitation in the discussion.
 
@@
 I think the discussion could benefit from including a short section of the role for ribophagy in this setting?
+
+We thank the reviewer for this suggestion, we have incorporated a section connecting our discussion to the emerging field of ribophagy.
 
@@
 
-> Page 3, line 48-49. I encourage the authors to consider including an
-> additional reference (Figueiredo et al 2021 J Phys). This paper shows
-> the specific response of ribosome biogenesis in response to RE and not
-> Endurance type exercise.
+> Page 3, line 48-49.
+I encourage the authors to consider including an additional reference (Figueiredo et al 2021 J Phys).
+This paper shows the specific response of ribosome biogenesis in response to RE and not Endurance type exercise.
 
-> Page 4, line 8: Please clarify "numberical lowering". Could this be
-> rephrased?
+We thank the reviewer for this suggestion.
 
+> Page 4, line 8: Please clarify "numberical lowering". Could this be rephrased?
 > Page 4, line 10: Double ".."
+> Page 4, line 15: Double ".."
+> Page 4, line 26: Should "to" be removed?
+> Page 11, line 20: n=7 or 8? Please check abstract
+> Page 11, line 35: Should "a" be removed? Please double check if it's 7 or 8 days of detraining. Abstract states 8.
 
-> Page 4, line 15: Double ".."
+We thank the reviewer for pointing out the above typos, they have been corrected.
 
-> Page 4, line 26: Should "to" be removed?
+> Page 12, line 19: please clarify "leg being training in the rest period between sets in the first leg. Does this means that they training continuously without break?
 
-> Page 11, line 20: n=7 or 8? Please check abstract
+Yes, sets were performed with rest periods for the other leg.
+The protocol still allowed for complete rest between sets ($\sim$ 60 sec) and adjustment of loads if necessary.
+A clarification has ben added to this section.
 
-> Page 11, line 35: Should "a" be removed? Please double check if it's 7
-> or 8 days of detraining. Abstract states 8.
+> Page 13, line 52: 2-4 passes were made with the Bard Magnum 12-14 gauge needles but only 1-2 aliquots of the samples were frozen.
+What happened to the other material?
+If I misunderstood, can the text be rewritten for clarity?
 
-> Page 12, line 19: please clarify "leg being training in the rest
-> period between sets in the first leg. Does this means that they
-> training continuously without break?
+Muscle tissue from all passes were processed together and aliquots were weighed and frozen.
+The section has been updated.
+Unfortunately, a malfunctioning freeze-dryer destroyed a portion of the material.
+The second aliquot was pooled for RNA extraction when available.
 
-> Page 13, line 52: 2-4 passes were made with the Bard Magnum 12-14
-> gauge needles but only 1-2 aliquots of the samples were frozen. What
-> happened to the other material? If I misunderstood, can the text be
-> rewritten for clarity?
+> Page 14, line 8: Was the spike-in used to normalize RNA extraction efficiency? If yes, please state this clearly.
 
-> Page 14, line 8: Was the spike-in used to normalize RNA extraction
-> efficiency? If yes, please state this clearly.
+The spike-in was used to for normalization of rRNA to muscle weight as previously described [hammarström2020].
+The section has been updated to notify the reader of the purpose of adding the external reference.
 
@@
 
-> Page 15, line 15: Does the authors refer to amplicon size when they
-> write primer sizes?
+This refers to diluting all samples to a common protein concentration. The sentence has been updated for clarity.
 
-> Page 31, Figure legends: I think that the figure legends could benefit
-> from a bit more substance. For example Fig 1A, perhaps indicate bar
-> color as black (CONST) and yellow (VAR)?
+> Page 15, line 15: Does the authors refer to amplicon size when they write primer sizes?
+
+Yes, the sentence should read amplicon sizes. This has been updated.
+
+> Page 31, Figure legends: I think that the figure legends could benefit from a bit more substance. For example Fig 1A, perhaps indicate bar color as black (CONST) and yellow (VAR)?
+
+The figure legends has been updated together with an overall update of figure appearances.
 
@@
 
-> The authors sought to perform a time course investigation of ribosome
-> density and biogenesis markers after knee extensor training. Phenotype
-> variables were also assessed. Training led to increases in muscle
-> growth (VL thickness) as well as in total RNA, rRNA, UBF and rpS6.
-> Training volume did not play a role in these adaptations.
-> Interestingly, the rate of total RNA increases predicted hypertrophy.
-> Likewise, training cessation led to decrements in total RNA abundance.
-> I love this study. This continues to contribute to our knowledge
-> regarding the role of ribosome biogenesis in muscle hypertrophy, and
-> is delicate human work with multiple time point biopsies. I have minor
-> comments in order of appearance, and congratulate the authors on their
-> fine work.
+> The authors sought to perform a time course investigation of ribosome density and biogenesis markers after knee extensor training.
+Phenotype variables were also assessed.
+Training led to increases in muscle growth (VL thickness) as well as in total RNA, rRNA, UBF and rpS6.
+Training volume did not play a role in these adaptations.
+Interestingly, the rate of total RNA increases predicted hypertrophy.
+Likewise, training cessation led to decrements in total RNA abundance.
+I love this study.
+This continues to contribute to our knowledge regarding the role of ribosome biogenesis in muscle hypertrophy, and is delicate human work with multiple time point biopsies.
+I have minor comments in order of appearance, and congratulate the authors on their fine work.
+
+We are greatful for the reviwers kind words.
 
@@
 
-> INTRODUCTION - Line 7, minor: "promotes" should read as "promote" -
+> INTRODUCTION
+
+> Line 7, minor: "promotes" should read as "promote" -
 > Line 14, minor: "with subsequent repeated bouts" should read as "and
-> subsequent repeated bouts" - Line 18, minor: "resting synthetic rate
-> of muscle protein" better reads as "basal muscle protein synthesis
-> rates" - Lines 48-50, minor: "bouts lead to accumulation of mature
-> rRNA thus also total RNA and presumably functional ribosomes"...this
-> sentence is long, and the word "thus" seems to make it a run-on. I'd
-> rephrase the sentence to read more coherently. Page 4 - Lines 10 and
-> 15, minor: each sentence has two stops. - Line 33, minor: no need to
-> use the phrase "per se" RESULTS - Line 51 (and throughout the
-> results), clarification: is "9.2%-point difference" the same as "9.2%
-> difference"? If so, adjust and remove "point" from all descriptors as
-> this is confusing. - Other than this, the results, while dense and
-> full of data, are well-written. DISCUSSION Page 7 - Line 17, minor:
-> again, "per se" can be removed. - Line 19, minor: "was not affected"
-> better reads as "were not affected" - Line 39, minor: "session" should
-> read as "sessions"
+> subsequent repeated bouts" 
+Line 18, minor: "resting synthetic rate of muscle protein" better reads as "basal muscle protein synthesis rates"
+Lines 48-50, minor: "bouts lead to accumulation of mature rRNA thus also total RNA and presumably functional ribosomes"...this sentence is long, and the word "thus" seems to make it a run-on.
+I'd rephrase the sentence to read more coherently.
 
-> I urge the authors to be creative somewhere in the discussion by
-> perhaps adding a paragraph after the one found in lines 29-56. We've
-> done a lot of work in the ribosome biogenesis area, and we've thought
-> a good bit about total RNA/unit muscle with regard to what it
-> represents. The authors correctly articulated this as a proxy of
-> ribosome density. However, as muscle tissue and myofibers
-> volumetrically increase, this will affect the metric (the same holds
-> true with mitochondrial biogenesis and using CS activity as a proxy of
-> mitochondrial volume density). An easy way to interpret this is in the
-> following example: o Muscle tissue increases 5% in thickness with RT
-> (and myofibers scale accordingly as well) o Total RNA/unit muscle does
-> not change in value Most would interpret this as ribosome biogenesis
-> not occurring. However, biogenesis likely occurred because, as muscle
-> grew, ribosome biogenesis scaled with growth thus leading to no change
-> in RNA concentration. Our laboratory discusses this important concept
-> with regard to RT and mitochondrial biogenesis (PMID: 32162291, PMID:
-> 34646153).
+> Page 4 - Lines 10 and
+> 15, minor: each sentence has two stops.
+
+> Line 33, minor: no need to use the phrase "per se"
+
+> RESULTS - Line 51 (and throughout the results), clarification: is "9.2%-point difference" the same as "9.2% difference"? If so, adjust and remove "point" from all descriptors as
+this is confusing.
+
+The "%-point difference" referes to a differences in percentage points, i.e.
+one condition increases in 2% the other 11.2%, the percentage-point difference is 9.2%.
+This is different from where percentages are stated without the "-point".
+Although confusing we belive that it is the correct way of reporting this such results.
+
+> Other than this, the results, while dense and full of data, are well-written.
+
+> DISCUSSION Page 7 - Line 17, minor: again, "per se" can be removed. - 
+> Line 19, minor: "was not affected" better reads as "were not affected" 
+> Line 39, minor: "session" should read as "sessions"
+
+We thank the reviewer for spotting mistakes and suggesting improvements to the text.
+
+> I urge the authors to be creative somewhere in the discussion by perhaps adding a paragraph after the one found in lines 29-56.
+We've done a lot of work in the ribosome biogenesis area, and we've thought a good bit about total RNA/unit muscle with regard to what it represents.
+The authors correctly articulated this as a proxy of ribosome density.
+However, as muscle tissue and myofibers volumetrically increase, this will affect the metric (the same holds true with mitochondrial biogenesis and using CS activity as a proxy of mitochondrial volume density).
+An easy way to interpret this is in the following example: o Muscle tissue increases 5% in thickness with RT (and myofibers scale accordingly as well) o Total RNA/unit muscle does not change in value Most would interpret this as ribosome biogenesis not occurring.
+However, biogenesis likely occurred because, as muscle grew, ribosome biogenesis scaled with growth thus leading to no change in RNA concentration.
+Our laboratory discusses this important concept with regard to RT and mitochondrial biogenesis (PMID: 32162291, PMID: 34646153).
 
 > Anyhow, this is no critique of the authors' hard and excellent work.
-> However, I think the authors re-iterating this point as a stand-alone
-> paragraph would benefit readership. In addition, I believe taking a
-> moment to define what the total RNA/ribosome density metric
-> conceptually represents, and then (in this newly added paragraph)
-> briefly walking the reader through each phase of the data, and what
-> could be going on with biogenesis in the context of a growing muscle
-> would be interesting to see. For instance, in Figure 3G, my
-> interpretation (much like the authors') is that ribosome biogenesis is
-> very rapid at the onset of RT where little functional growth is likely
-> occurring. And like the authors, I believe this likely was needed for
-> eventual growth to occur. However, the "no change" in RNA
-> concentrations session 9-12 as muscle is continuing to grow may not
-> indicate a plateau in biogenesis given that this is the time during
-> which muscle growth is likely starting to ramp up. Hence, biogenesis
-> rates could be equally as ramped here compared to sessions 1-9 of RT.
+However, I think the authors re-iterating this point as a stand-alone paragraph would benefit readership.
+In addition, I believe taking a moment to define what the total RNA/ribosome density metric conceptually represents, and then (in this newly added paragraph) briefly walking the reader through each phase of the data, and what could be going on with biogenesis in the context of a growing muscle would be interesting to see.
+For instance, in Figure 3G, my interpretation (much like the authors') is that ribosome biogenesis is very rapid at the onset of RT where little functional growth is likely occurring.
+And like the authors, I believe this likely was needed for eventual growth to occur.
+However, the "no change" in RNA concentrations session 9-12 as muscle is continuing to grow may not indicate a plateau in biogenesis given that this is the time during which muscle growth is likely starting to ramp up.
+Hence, biogenesis rates could be equally as ramped here compared to sessions 1-9 of RT.
 
-METHODS
+We thank the reviewer for this suggestion.
+We agree with the reviwer and feel that the manuscript also communicates this message.
+This is an important point to underline.
+We felt that the section served well as part of the mentioned section and added the following:
 
-Page 11
+This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth.
+This balance, measured as a constant ribosomal density in a growing cell, still requires the biosynthesis of ribosomes to match the volumetric expansion of the cell.
+As such, indirect measures of translational capacity through the density of ribosomes in periods of marked muscle hypertrophy may mask an absolute increase in ribosomes.
 
--   Lines 9-19: Authors stated that 18 participants were recruited, but
+> METHODS
+> Page 11
+> -   Lines 9-19: Authors stated that 18 participants were recruited, but
     then go on to say that n=11 TRAIN and n=8 CTRL completed. Please
     reconcile.
--   Line 35: Authors stated 7 days after detraining, but in the
+> -   Line 35: Authors stated 7 days after detraining, but in the
     discussion (page 7, line 19) they mention after 8 days of
     detraining. Please reconcile. Page 12
--   Line 39, minor: "restitution" should read as "rest". Please change
+>   Line 39, minor: "restitution" should read as "rest". Please change
     accordingly, and if used throughout the paper, please alter. Page 15
--   Line 36, minor: "Glycin" should read as "Glycine". In addition, and
+>   Line 36, minor: "Glycin" should read as "Glycine". In addition, and
     molar unit is needed.
+
+The above points has been addressed in the revised manuscript.

### 2022-02-02 manuscript.Rmd
added revision

@@
 
-Skeletal muscle is a critical target for interventions that promotes health across the lifespan [@RN763]<!-- Wolfe 2006 -->,
+Skeletal muscle is a critical target for interventions that promote health across the lifespan [@RN763]<!-- Wolfe 2006 -->,
 with resistance training (RT) being the advocated remedy.  
 Prolonged RT leads to changes in the balance between muscle protein breakdown and synthesis, with one bout of resistance exercise acutely increasing protein synthesis for up to 48 hrs after exercise [@RN786],<!-- Phillips 1997 -->
-with subsequent repeated bouts leading to accumulation of muscle protein over time
+<span custom-style="CHANGE">and</span> subsequent repeated bouts leading to accumulation of muscle protein over time
 [@RN1272<!-- Phillips review 2014 -->;
 @RN1642]<!-- Brook 2015 experiment -->. 
-In recent years, this view has been supplemented<!-- supported? --> by evidence suggesting that chronic RT leads to increased resting synthetic rate of muscle protein
+In recent years, this view has been supplemented<!-- supported? --> by evidence suggesting that chronic RT leads to increased <span custom-style="CHANGE">basal muscle protein synthesis rates</span>
 [@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> 
@@
 
-Resistance exercise is a potent stimuli for rRNA transcription as a single session leads to increases in pre-rRNA
+Resistance exercise is a potent <span custom-style="CHANGE">and specific [@figueiredo2021]</span> stimuli for rRNA transcription as a single session leads to increases in pre-rRNA
 [@RN1656<!-- A single session upregulates 45 S 24 h post ex -->;
-@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->]
-and repeated bouts lead to accumulation of mature rRNA thus also total RNA and presumably functional ribosomes
+@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->].
+Repeated bouts lead to the accumulation of mature rRNA reflected in total RNA and presumably functional ribosomes
 [@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
@@
 Using voluntary contractions, peak values were reported after nine sessions, followed by a slight decrease to after 18 sessions [@RN1809],
-resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), whereby a numerical lowering occurred to after the last training session of the 12 wk interventions (31 sessions).
+resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), <span custom-style="CHANGE">followed by lower levels measured after</span> the last training session of the 12 wk interventions (31 sessions)
 [@RN2358].
-Interestingly, during the initial phase of RT, total RNA accumulation seems to be volume-dependent, as three sets per exercise in leg exercises led to augmented total RNA and rRNA levels compared to one set per exercise, coinciding with the differences in muscle hypertrophy seen after 12 weeks of RT.
+Interestingly, during the initial phase of RT, total RNA accumulation seems to be volume-dependent, as three sets per exercise in leg exercises led to augmented total RNA and rRNA levels compared to one set per exercise, coinciding with the differences in muscle hypertrophy seen after 12 weeks of RT
 [@RN2358].
@@
 
-Based on these observations we hypothesize that (1) ribosome accumulation occurs during the early phase (3-4 weeks) of RT, within which this accumulation (2) reaches a plateau when RT volume is kept constant, (3) displays fluctuations in response to fluctuating training volume and (4) is partially reversed to one week after cessation of RT.
+Based on these observations we hypothesize that (1) ribosome accumulation occurs during the early phase (3-4 weeks) of RT, within which this accumulation (2) reaches a plateau when RT volume is kept constant, (3) displays fluctuations in response to fluctuating training volume and (4) is partially reversed one week after cessation of RT.
 In addition to addressing these hypotheses we aimed to relate RNA accumulation to total UBF levels and muscle growth.
 We utilized a within-participant unilateral training model where one leg was assigned constant volume (CONST, 6 sets per session) and the contra lateral leg variable volume (VAR, 6, 3, and 9 sets in sessions 1-4, 5-8, and 9-12, respectively).
-Effects of training *per se* were assessed by comparison to a non-training control group (CTRL).
+Effects of training were assessed by comparison to a non-training control group (CTRL).
 
@@
 These markers accumulated progressively during the initial part of the intervention before it leveled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
-This increase in total RNA was interconnected with increases in UBF protein abundance, suggesting UBF levels *per se* to play a role in regulation of rRNA transcription regulation in response to RT.
-Total RNA increases was not affected by weekly fluctuations in training volume, however, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
+This increase in total RNA was interconnected with increases in UBF protein abundance, suggesting UBF levels to play a role in regulation of rRNA transcription regulation in response to RT.
+Total RNA increases <span custom-style="CHANGE">were not affected</span> by weekly fluctuations in training volume, however, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
 Finally, individual rates of increases in total RNA abundance predicted the magnitude of muscle growth, confirming the likely link between ribosomal biogenesis and muscle protein accretion 
@@
 However, the time course of total RNA/rRNA changes in response to RT has so far remained speculative, with no study investigating responses to prolonged interventions with multiple sampling time points.
-In the present data, RT led to a clear session-to-session increase in total RNA per unit tissue weight in response to the first four session, whereupon the changes gradually leveled out before peaking after the 8^th^ session, with the peak increase from baseline being $\sim$ 50%, defining an accumulation phase.
+In the present data, RT led to a clear session-to-session increase in total RNA per unit tissue weight in response to the first four <span custom-style="CHANGE">sessions</span>, whereupon the changes gradually leveled out before peaking after the 8^th^ session, with the peak increase from baseline being $\sim$ 50%, defining an accumulation phase.
 This corroborates well with previous suggestions of peak values being reached within four to nine sessions in young males and females
@@
 Within this last part of the intervention, synthesis of novel rRNA still seemed to be elevated per weight unit muscle tissue compared to baseline, as suggested by sustained elevation of pre-rRNA transcripts, coinciding with peak values of UBF protein levels.
-This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth [@RN2142].<!-- Figueiredo review details the concept of rna concentration -->
+This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth [@RN2142].<!-- Figueiredo review details the concept of rna concentration --> <span custom-style="CHANGE">This balance, measured as a constant ribosomal density in a growing cell, still requires the biosynthesis of ribosomes to match the volumetric expansion of the cell.
+As such, indirect measures of translational capacity through the density of ribosomes in periods of marked muscle hypertrophy may mask an absolute increase in ribosomes.</span>
 
@@
 Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
-<span custom-style="CHANGE">Furthermore, in light of accumulating evidence of tissue specific and heterogeneous ribosomal populations [@chaillou2019; @chaillou2015; @gupta2014], our results suggests a potential role of specialized newly synthesized ribosomes in muscle hypertrophy.
-Our results are however limited in terms of probing any such heterogeneity.
-</span>
+<span custom-style="CHANGE">Furthermore, in light of accumulating evidence of tissue specific and heterogeneous ribosomal populations [@chaillou2019; @chaillou2015; @gupta2014], our results suggests a potential role of specialized, newly synthesized ribosomes in muscle hypertrophy although we acknowledge that the present study is limited in terms of probing any such heterogeneity.</span>
 Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.
@@
 Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT.
-Additionally, ribosomal proteins may have extra-ribosomal functions
-affecting their expression
+<span custom-style="CHANGE">Differences in temporal regulation between total RNA and ribosomal proteins may be explained by observations in cell culture indicating that excess amounts of ribosomal proteins are synthesized and imported into the nucleus [@lam].
+This abundance assures the availability of ribosomal proteins for ribosome assembly upon rRNA transcription while also indicating that the availability of ribosomal proteins is not rate-limiting for ribosomal biogenesis [@lam].
+Whether RT leads to an expanded pool of unbound ribosomal proteins, possibly through satellite cell expansion and myonuclear accretion in humans [@abousawan2021; @murach2022] remains to be determined.
+However, a disconnect between RPS6 and total RNA after de-training could be explained by a greater number of myonuclei supporting a higher constitutive expression of ribosomal proteins.
+Interestingly, in a recent study, Murach and colleagues showed that newly acquired myonuclei (from satellite cell fusion) contribute to the myofibers ribosomal machinery [@murach2022].
+However, the present study's observations are limited to a single ribosomal protein and lack data supporting RT-induced myonuclear accretion.
+A parallel mechanism that could help explain our observations is the possibility of extra-ribosomal functions in selected ribosomal proteins affecting their expression independent of ribosomal biogenesis.
 [@RN2588].<!-- Warner 2009 General description of   -->  
+</span>
 
@@
 Together with our observations, this underlines the importance of UBF as a regulator of RT-induced ribosomal biogenesis.
+<span custom-style="CHANGE"> However, the lack of measurements of UBF in the context of active chromatin interaction may have inhibited us from further explaining the role of UBF in response to the de-training period.</span>
 
@@
 This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
-This supports the idea that ribosomal biogenesis is a cellular activity on demand, possibly relating to its relative expense [@RN1920] also in muscle tissue.
+<span custom-style="CHANGE">The magnitude of the decrease in total RNA ($\sim$ 20\%) was similar as observed in untrained muscle in response to complete unloading during a similar time frame (7-10 days) [@gamrin1998; @figueiredo2020].
+This may indicate that the increase in total RNA induced by RT is more easily lost than total RNA levels measured during homeostatic conditions (removal of 2-3 sessions of RT during eight days vs.
+limb immobilization).
+In addition to reduced synthesis of new ribosomes, degradation of existing ribosomes (ribophagy) likely contributes to the de-training effect seen on total and rRNA levels.
+Such degradation is highly sensitive to inactivity evident from rats subjected to hindlimb suspension [@figueiredo2020].
+The relative contribution to the number of functional ribosomes of pathways regulating synthesis and degradation of ribosomes remains to be determined.
+Future studies will likely show that the balance between ribosome biogenesis and ribophagy is specific to cell types and stimuli.
+Such complexity likely exists as both selective and non-selective pathways for ribophagy exists, activated by different upstream mechanisms [@beese2020] in addition to different effectors controlling ribosome biogenesis.
+For example, in skeletal muscle and in contrast to inactivity-induced RNA degradation [@figueiredo2020], Kim et al.
+reported that impaired translational capacity in a model of cancer cachexia was primarily driven by a reduced rDNA transcription [@kim2021].</span>
+
+The de-training effect on total RNA and rRNA seen in the present study supports the idea that ribosomal biogenesis is a cellular activity on demand, possibly relating to its relative expense [@RN1920] also in muscle tissue.</span>
 Based on this notion, and the fact that RT volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including induction of c-Myc expression, mTOR activation
@@
 
-Eighteen volunteers were recruited to the study.
+Nineteen volunteers were recruited to the study.
 Eligible participants were non-smokers between 18 and 35 years of age with a training history of less than one RT session per week during the six months leading up to the study.
@@
 CTRL did not partake in RT and were instructed to continue their everyday activities.
-Muscle biopsies were sampled bilaterally in TRAIN before and 48 hrs after the first session, as well as 48 hrs after the fourth, fifth, eight, ninth and twelfth session, and after a seven days of de-training.
+Muscle biopsies were sampled bilaterally in TRAIN before and 48 hrs after the first session, as well as 48 hrs after the fourth, fifth, eight, ninth and twelfth session, and after eight days of de-training.
 Muscle biopsies were obtained from CTRL at three occasions; at baseline and 48 hrs and 3-5 weeks (average (SD) `r ctrl`) after the first sampling event.
@@
 Inter-set rest periods were 90 sec.
-Throughout the intervention RT sessions were alternatingly initiated by training the right and left leg, changing every other session, with the contralateral leg being trained in the rest period between sets of the first leg.
+Throughout the intervention RT sessions were alternatingly initiated by training the right and left leg, changing every other session.
+<span custom-style="CHANGE">The contralateral leg was trained in the rest period between sets of the first leg, still allowing for complete rest between efforts ($\sim$ 60 sec)</span>.
 The second session of each four-session block (session 2, 6 and 10) was performed at a sub-maximal resistance (~90% of the previous session) with the same number of repetitions (10).
@@
 Participants were instructed to gradually increase their effort during three warm-up repetitions (50, 60 and, 70% of subjective maximal effort).
-After a 30-sec restitution period participants were instructed to perform three repetitions with maximal effort in the concentric phase.
+After a 30-sec <span custom-style="CHANGE">rest</span> period participants were instructed to perform three repetitions with maximal effort in the concentric phase.
 Sixty seconds after the isokinetic test the lever automatically moved to a 60&deg; angle and participants were instructed to push against the lever enough to see feedback from the visual feedback system.
@@
 Two to four passes were made to get sufficient material.
-One to two aliquots of the samples were quickly dissected free from connective and fat tissue, weighed and frozen in isopentan chilled to -80&deg;C and stored at -80&deg;C until further processing.<!-- At baseline, after twelve weeks of training and, after the de-training period, a further aliquot from each sample was selected for immunohistochemistry analysis and subsequently immersed in 4% formalin for fixation.-->
+<span custom-style="CHANGE"> Material from all passes was quickly dissected free from connective and fat tissue and divided into one to two aliquots (depending on amount of available material).
+Aliquots were weighed and frozen</span> in isopentan chilled to -80&deg;C and stored at -80&deg;C until further processing.<!-- At baseline, after twelve weeks of training and, after the de-training period, a further aliquot from each sample was selected for immunohistochemistry analysis and subsequently immersed in 4% formalin for fixation.-->
 
 ## RNA and protein extraction
-Frozen muscle tissue was homogenized in 1 ml of Trizol (ThermoFisher Scientific, Oslo, Norway) spiked with 0.04 ng of an external, non-mammalian, RNA spike-in (Lambda PolyA External Standard Kit, Takara Bio Europe, Saint-Germain-en-Laye, France).
+Frozen muscle tissue was homogenized in 1 ml of Trizol (ThermoFisher Scientific, Oslo, Norway) spiked with 0.04 ng of an external, non-mammalian, RNA spike-in (Lambda PolyA External Standard Kit, Takara Bio Europe, Saint-Germain-en-Laye, France).<span custom-style="CHANGE"> The addition of the external spike-in allowed for normalization of target RNA to muscle weight, see below</span>.
 Mechanical disruption of the samples was achieved using Zirconium Oxide Beads (0.5 mm, Next Advance, Inc., New York, USA) and a bead mill (Bullet blender, Next Advance).
@@
 Protein concentrations were measured (Pierce Detergent Compatible Bradford Assay, ThermoFisher Scientific).
-Sample were normalized in 4X Laemmli buffer (Bio-Rad Norway AS, Oslo, Norway), boiled (95&deg;C, 5 min) and stored at -20&deg;C before later use.
+<span custom-style="CHANGE">Sample were normalized to a common protein concentration, 4X Laemmli buffer  (Bio-Rad Norway AS, Oslo, Norway) was added and samples were boiled</span> (95&deg;C, 5 min) and stored at -20&deg;C before later use.
 
@@
 Melt curves were collected from all reactions to confirm single product amplification.
-Primers were further evaluated by agarose gel electrophoresis which confirmed primer sizes and non-template control experiments confirming no amplification without template.
+Primers were further evaluated by agarose gel electrophoresis which confirmed <span custom-style="CHANGE">amplicon</span> sizes and non-template control experiments confirming no amplification without template.
 Primer sequences and their respective average performances are shown in Table 4.
@@
 When samples were compared between participants, signals were expressed per a calibrator sample from each participant that in turn was measured on a separate gel with all calibrator samples.
-Separated samples were transferred to &#956m PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 Glycin, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h.
+Separated samples were transferred to PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 mM Glycine, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h.
 Membranes were then stained to confirm transfer and enable total protein quantification using a reversible protein stain (Pierce Reversible Protein Stain, Thermo Fisher Scientiﬁc).
@@
 
-**Figure 1.** Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).
+**Figure 1.**  (A) Study design showing muscle biopsy sampling, thickness and strength assessments time points together with number of sets per session (CONST blue bars, VAR red bars).
+Assessments time points in the negative control group is shown in the lower panel.
+(B) Observed training loads in response to CONST and VAR volume protocols.
+Training outcomes are shown as within condition changes and in comparison to the control group (muscle strength, C; muscle thickness D).
 Intervals in C and D indicate 95% CI.

### 2022-02-03 response-to-reviews-rev1.Rmd
updates

@@
 title: "APH-2021-12-0652: Response to the reviewers"
-author: "Daniel Hammarström, Sjur J. Øfsteng, Nicolai B. Jacobsen, Krister B. Flobergseter, Bent R. Rønnestad, Stian Ellefsen"
-date: "2022-01-17"
+author: 
+  - Daniel Hammarström
+  - Sjur J. Øfsteng
+  - Nicolai B. Jacobsen
+  - Krister B. Flobergseter
+  - Bent R. Rønnestad
+  - Stian Ellefsen
+date: "2022-02-02"
 output: pdf_document
@@
 As we have already pointed to in the original manuscript, our results suggests different temporal regulation or turnover rates in total RNA and RPS6 protein levels.
-Since ribosomal proteins exists in excess (and are thus not rate limiting for ribosomal biogenesis) [@lam], it is tempting to assume constitutive expression is only indirectly related to the process of ribosomal biogensis.
+Since ribosomal proteins exists in excess (and are thus not rate limiting for ribosomal biogenesis) [@lam2007], it is tempting to assume constitutive expression is only indirectly related to the process of ribosomal biogensis.
 We have expanded the discussion by raising the point that unbound ribosomal proteins (including RPS6) may be passively increased by myonuclear accretion.
@@
 The above points has been addressed in the revised manuscript.
+
+# References

### 2022-02-03 manuscript.Rmd
updates

@@
 title: "Ribosome accumulation during early phase resistance training in humans"
-author: Daniel Hammarström^1,2,£^*, Sjur J. Øfsteng^2,£^*, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^, Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
+author: Daniel Hammarström^1,2,£^&ast;, Sjur J. Øfsteng^2,£^&ast;, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^, Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
 date: ''
@@
 Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT.
-<span custom-style="CHANGE">Differences in temporal regulation between total RNA and ribosomal proteins may be explained by observations in cell culture indicating that excess amounts of ribosomal proteins are synthesized and imported into the nucleus [@lam].
-This abundance assures the availability of ribosomal proteins for ribosome assembly upon rRNA transcription while also indicating that the availability of ribosomal proteins is not rate-limiting for ribosomal biogenesis [@lam].
+<span custom-style="CHANGE">Differences in temporal regulation between total RNA and ribosomal proteins may be explained by observations in cell culture indicating that excess amounts of ribosomal proteins are synthesized and imported into the nucleus [@lam2007].
+This abundance assures the availability of ribosomal proteins for ribosome assembly upon rRNA transcription while also indicating that the availability of ribosomal proteins is not rate-limiting for ribosomal biogenesis [@lam2007].
 Whether RT leads to an expanded pool of unbound ribosomal proteins, possibly through satellite cell expansion and myonuclear accretion in humans [@abousawan2021; @murach2022] remains to be determined.

### 2022-02-03 manuscript.Rmd
last rev

@@
 
-<!-- The formatting of the manuscript is based on faseb journal, find csl-files at https://github.com/citation-style-language/styles -->
-
 <!-- Figure texts are included in the end of the manuscript file -->
@@
 <div custom-style="correspondence_style">
-&ast; Correspondance
+&ast; Correspondence
 
@@
 **Methods:** A training group (n = 11) performed <span custom-style="CHANGE">12 sessions (3-4 sessions per week) of</span> unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg.
-<span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from</span> vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline and 48 hrs after sessions 1, 4, 5, 8, 9, 12</span> and after eight days of de-training and from a non-training control group (n = 8).
+<span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline, 48 hrs after sessions 1, 4, 5, 8, 9 and 12</span> and after eight days of de-training, as well as</span> from a non-training control group (n = 8).
 In addition, muscle thickness was measured before and after training and control periods.
@@
 **Results:** Training led to muscle growth (`r hypertrophy` vs.
-control) with concomitant increases in total RNA, rRNA, UBF, and rpS6 with no differences between volume conditions.
+control) with concomitant increases in total RNA, rRNA, UBF, and RPS6 with no differences between volume conditions.
 Total RNA increased most rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline).
@@
 
-For both ribosomal protein S6 (rpS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
-This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12 and elevated after eight days of rest (Figure 2A and B), however with no robust differences being observed after the first training session (48 hrs).
+For both ribosomal protein S6 (RPS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with RPS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
+This general pattern was confirmed when comparing TRAIN to CTRL where UBF and RPS6 protein levels were higher in TRAIN compared to CTRL after Session 12, and remained elevated after eight days of rest (Figure 2A and B), with no robust differences being observed after the first training session (48 hrs).
 Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`).
 After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
-For rpS6, de-training did not affect protein levels, which remained similar between volume conditions.
-At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point.
+For RPS6, de-training did not affect protein levels, which remained similar between volume conditions.
+At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while RPS6 showed no robust differences between TRAIN and CTRL at any time point.
 No differences were observed between volume conditions for either transcripts (Figure 2D and E).
@@
 In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coinciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 1).
-In contrast, no evidence was found for a relationship between total RNA and rpS6 protein levels (Table 1).
+In contrast, no evidence was found for a relationship between total RNA and RPS6 protein levels (Table 1).
 
@@
 # Discussion
-Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
-These markers accumulated progressively during the initial part of the intervention before it leveled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
+Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and RPS6 protein in previously untrained individuals compared to a non-training control group.
+These markers accumulated progressively during the initial part of the intervention before they levelled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
 This increase in total RNA was interconnected with increases in UBF protein abundance, suggesting UBF levels to play a role in regulation of rRNA transcription regulation in response to RT.
-Total RNA increases <span custom-style="CHANGE">were not affected</span> by weekly fluctuations in training volume, however, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
+Total RNA increases <span custom-style="CHANGE">were not affected</span> by weekly fluctuations in training volume.
+However, eight days of de-training led to lowered levels of total RNA and rRNA content, suggesting that training cessation halts ribosome biogenesis.
 Finally, individual rates of increases in total RNA abundance predicted the magnitude of muscle growth, confirming the likely link between ribosomal biogenesis and muscle protein accretion 
@@
 [@RN1809;
-@RN2358], and may be essential for preparing muscle fibers for subsequent growth
+@RN2358], and may be essential for preparing muscle fibres for subsequent growth
 [@RN1809;<!-- Correlation between increase and muscle growth -->
@@
 After the 8^th^ session, no meaningful increase or decrease were observed for total RNA or rRNA content within the training period, suggesting a plateau phase with attenuated net synthesis of novel ribosomes.
-Within this last part of the intervention, synthesis of novel rRNA still seemed to be elevated per weight unit muscle tissue compared to baseline, as suggested by sustained elevation of pre-rRNA transcripts, coinciding with peak values of UBF protein levels.
+Within this last part of the intervention, synthesis of novel rRNA still seemed to be elevated *per* weight unit muscle tissue compared to baseline, as suggested by sustained elevation of pre-rRNA transcripts, coinciding with peak values of UBF protein levels.
 This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth [@RN2142].<!-- Figueiredo review details the concept of rna concentration --> <span custom-style="CHANGE">This balance, measured as a constant ribosomal density in a growing cell, still requires the biosynthesis of ribosomes to match the volumetric expansion of the cell.
-As such, indirect measures of translational capacity through the density of ribosomes in periods of marked muscle hypertrophy may mask an absolute increase in ribosomes.</span>
+As such, indirect measures of translational capacity such as the concentration of total RNA may mask the absolute increase in ribosomes that occurs during periods of muscle hypertrophy.</span>
 
@@
 This supports the notion that ribosomal biogenesis is an important determinant of RT-induced muscle hypertrophy, with previous studies showing that increases in total RNA are positively correlated with increases in muscle mass
-[@RN1644;<!-- Figueriedo  -->
-@RN1897;
-@RN2055],<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
+[@RN1644;@RN1897;@RN2055],<!-- Figueriedo  --><!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
 <!-- Reidy correlation between total RNA and muscle hypertrophy-->
@@
 In addition, supression of ribosomal biogenesis in *in vitro* models leads to halted muscle cellular growth in some 
-[@RN1632;<!-- Nader 2005 blockage of mTOR halts ribo bio and growth -->
-@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
-@RN1755]<!-- Stec specific pol I inhibition -->
+[@RN1632;@RN1754;@RN1755]<!-- Nader 2005 blockage of mTOR halts ribo bio and growth --><!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS--><!-- Stec specific pol I inhibition -->
 but not all studies
 [@RN1929].<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
-Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA was instead associated with a tendency towards lowered muscle growth.
-Overall, the rate of increases in ribosomal density thus seems to be a better predictor of individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
+Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA were instead associated with a tendency towards lowered muscle growth.
+Overall, the rate of increases in ribosomal density hence seems to be a better predictor for individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
 Interestingly, the interaction between rRNA synthesis rate and muscle mass accretion (but not between ribosomal content and muscle mass accretion) may shed light on observed differences in muscular responses to RT between young and old individuals.
@@
 potentially explaining their alleged poorer overall hypertrophic responses [@RN1809].
-Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
-<span custom-style="CHANGE">Furthermore, in light of accumulating evidence of tissue specific and heterogeneous ribosomal populations [@chaillou2019; @chaillou2015; @gupta2014], our results suggests a potential role of specialized, newly synthesized ribosomes in muscle hypertrophy although we acknowledge that the present study is limited in terms of probing any such heterogeneity.</span>
+Whether these cellular characteristics are related to e.g. differences in fibre type distributions [@RN1707] remains to be determined.
+<span custom-style="CHANGE">Furthermore, ribosomal accumulation is unlikely to be the only ribosome-derived trait that is important for training responsiveness.
+Evidence suggests that mechanical loading may lead to changes in ribosome characteristics [@chaillou2015] potentially leading to heterogeneous tissue-specific ribosome populations [@chaillou2019;  @gupta2014].
+Our results support a model where specialized, newly synthesized ribosomes contribute to muscle hypertrophy as the increase in ribosomal content but not absolute levels predicted muscle growth.
+We, however, acknowledge that the present study does not provide substantial insight into this perspective.</span>
 Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.
 
-<!-- rpS6 dissconect from total RNA levels. -->
-In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6.
-Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period.
-Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
-[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
-but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
-[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
-Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for.
-Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT.
-<span custom-style="CHANGE">Differences in temporal regulation between total RNA and ribosomal proteins may be explained by observations in cell culture indicating that excess amounts of ribosomal proteins are synthesized and imported into the nucleus [@lam2007].
-This abundance assures the availability of ribosomal proteins for ribosome assembly upon rRNA transcription while also indicating that the availability of ribosomal proteins is not rate-limiting for ribosomal biogenesis [@lam2007].
-Whether RT leads to an expanded pool of unbound ribosomal proteins, possibly through satellite cell expansion and myonuclear accretion in humans [@abousawan2021; @murach2022] remains to be determined.
-However, a disconnect between RPS6 and total RNA after de-training could be explained by a greater number of myonuclei supporting a higher constitutive expression of ribosomal proteins.
-Interestingly, in a recent study, Murach and colleagues showed that newly acquired myonuclei (from satellite cell fusion) contribute to the myofibers ribosomal machinery [@murach2022].
-However, the present study's observations are limited to a single ribosomal protein and lack data supporting RT-induced myonuclear accretion.
-A parallel mechanism that could help explain our observations is the possibility of extra-ribosomal functions in selected ribosomal proteins affecting their expression independent of ribosomal biogenesis.
+<!-- RPS6 dissconect from total RNA levels. -->
+In the present study, training induced increases in rRNA and total RNA coincided with increases in RPS6.
+Conversely, changes in total RNA levels and RPS6 in response to de-training did not correspond, as RPS6 protein levels remained elevated after the de-training period.
+The training induced increases in RPS6 seen in the present study are in agreement to those previously reported in young men
+[@RN1642],<!-- Brook 2015 RPS6 increase in resposne to RT over control -->
+but not in elderly men and women, where a decrease was observed in response to training despite increases in total RNA and rRNA
+[@RN1755].<!-- Stec 2016 RPS6 decreased 22% after four weeks of training in older  -->
+Although increases were seen for both RPS6 and and total RNA, RPS6 did not explain variation in total RNA (after controlling for the number of sessions).
+Together with a disconnect after the de-training period, this suggest that regulation of RPS6 expression and ribosomal RNA transcription display differential temporal responses to RT.
+<span custom-style="CHANGE">Such difference in temporal regulation of ribosomal RNA and proteins has previously been deduced in cell culture experiments.
+Briefly, inhibition of protein degradation led to accumulation of ribosomal proteins suggesting that excess amounts of ribosomal proteins are synthesized, imported into the nucleus and rapidly degraded if not incorporated into ribosomes [@lam2007].
+This inherent capacity of cells to provide sufficient access to ribosomal proteins also suggests rRNA transcription (and not synthesis of ribosomal proteins) is not rate-limiting during ribosomal biogenesis [@lam2007].
+Currently, it remains unknown if RT leads to expansion of the pool of unbound ribosomal proteins in humans.
+Indeed, in the present study, the disconnect between RPS6 and total RNA after de-training may have been associated with increased numbers of myonuclei, rather than accumulation of RPS6 in existing myonuclei, providing an alternative explanation to the elevated RPS6 levels (which accordingly may have been distributed over a larger number of nuclei).
+Interestingly, in a recent study, Murach and colleagues showed that newly acquired myonuclei (from satellite cell fusion) contribute to the ribosomal pool in myofibers [@murach2022].
+Importantly, the present study was not designed to investigate these perspectives, and observations are limited to one single ribosomal protein, with no investigation of RT-induced myonuclear accretion.
+A parallel mechanism that could futher help explain our observations is the possibility of extra-ribosomal functions in selected ribosomal proteins affecting their expression independent of ribosomal biogenesis
 [@RN2588].<!-- Warner 2009 General description of   -->  
@@
 
-UBF levels robustly explained total RNA levels over the entire course of the intervention.
-These analyses were done with the number of sessions accounted for, allowing unbiased estimates.
+UBF levels robustly explained total RNA levels over the course of the intervention.
+As these analyses were done while accunting for the number of training sessions, estimates are likely to be unbiased.
 Unrealistically strong relationships could have been otherwise expected as both the dependent variable (total RNA) and the covariate (UBF levels) varies with the number of sessions.
@@
 pointing to an effect observed in cell models where c-Myc induces UBF mRNA transcription [@RN2565].<!-- Pootinga 2004 MAD1 and c-Myc controls UBF transcription UBF is required for c-Myc induced rDNA transcription -->
-Interestingly the availability of UBF *per se* has been shown to regulate rRNA transcription
+Interestingly the availability of UBF has been shown to regulate rRNA transcription
 [@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->  through control of rDNA gene activity
@@
 This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
-<span custom-style="CHANGE">The magnitude of the decrease in total RNA ($\sim$ 20\%) was similar as observed in untrained muscle in response to complete unloading during a similar time frame (7-10 days) [@gamrin1998; @figueiredo2020].
-This may indicate that the increase in total RNA induced by RT is more easily lost than total RNA levels measured during homeostatic conditions (removal of 2-3 sessions of RT during eight days vs.
-limb immobilization).
-In addition to reduced synthesis of new ribosomes, degradation of existing ribosomes (ribophagy) likely contributes to the de-training effect seen on total and rRNA levels.
-Such degradation is highly sensitive to inactivity evident from rats subjected to hindlimb suspension [@figueiredo2020].
-The relative contribution to the number of functional ribosomes of pathways regulating synthesis and degradation of ribosomes remains to be determined.
-Future studies will likely show that the balance between ribosome biogenesis and ribophagy is specific to cell types and stimuli.
-Such complexity likely exists as both selective and non-selective pathways for ribophagy exists, activated by different upstream mechanisms [@beese2020] in addition to different effectors controlling ribosome biogenesis.
-For example, in skeletal muscle and in contrast to inactivity-induced RNA degradation [@figueiredo2020], Kim et al.
-reported that impaired translational capacity in a model of cancer cachexia was primarily driven by a reduced rDNA transcription [@kim2021].</span>
+<span custom-style="CHANGE">The magnitude of the detraining-associated decrease in total RNA ($\sim$ 20\%) is similar to that seen in response to unloading of untrained human muscle over a similar time frame (7-10 days) [@gamrin1998; @figueiredo2020].
+This suggests that RT-induced increases in ribosomal content is easily lost, likely driven by a combination of reduced synthesis and enhanced degradation (ribophagy).
+While the relative contribution of these two remains to be determined, inactivity has previously been shown to induce ribophagy in rat muscle (subjected to hindlimb suspension) [@figueiredo2020].
+Future studies should investigate the interrelationship between ribosome biogenesis and ribophagy in response to training and detraining.
+We suspect that their relative importance for ribosomal homeostasis will vary substantially between different physiological perturbations, as they are regulated by a complex interconnection of stimuli and pathways<!--REF?-->.
+For example, Kim et al.
+showed that cancer cachexia is associated with reduced rDNA transcription (and hence reduced translational capacity) [@kim2021], contrasting the inactivity-driven reduction in translational capacity which to a larger degree seems to involve ribophagy [@figueiredo2020].</span>
 
@@
 @RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
-@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
+@RN791]<!-- Burd ex volume protein synthesis and anabolic [RPS6] phospho signalling-->,
 subsequent total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
@@
 Two to four passes were made to get sufficient material.
-<span custom-style="CHANGE"> Material from all passes was quickly dissected free from connective and fat tissue and divided into one to two aliquots (depending on amount of available material).
-Aliquots were weighed and frozen</span> in isopentan chilled to -80&deg;C and stored at -80&deg;C until further processing.<!-- At baseline, after twelve weeks of training and, after the de-training period, a further aliquot from each sample was selected for immunohistochemistry analysis and subsequently immersed in 4% formalin for fixation.-->
+<span custom-style="CHANGE">Material from all passes was quickly dissected free from connective and fat tissue and divided into one to two aliquots (depending on amount of available material).
+Aliquots were weighed and frozen</span> in isopentan chilled to -80&deg;C and stored at -80&deg;C until further processing.
+<span custom-style="CHANGE">Due to difficulties during the sampling procedure, we could not obtain a sample from one participant's leg belonging to TRAIN at baseline.</span><!-- At baseline, after twelve weeks of training and, after the de-training period, a further aliquot from each sample was selected for immunohistochemistry analysis and subsequently immersed in 4% formalin for fixation.-->
 
@@
 When samples were compared between participants, signals were expressed per a calibrator sample from each participant that in turn was measured on a separate gel with all calibrator samples.
-Separated samples were transferred to PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 mM Glycine, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h.
+<span custom-style="CHANGE">Due to technical difficulties, the calibrator sample from one participant was excluded from analysis, reducing the sample size in analyses between participants in TRAIN to n = 10 (Table 1).<span> Separated samples were transferred to PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 mM Glycine, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h.
 Membranes were then stained to confirm transfer and enable total protein quantification using a reversible protein stain (Pierce Reversible Protein Stain, Thermo Fisher Scientiﬁc).
@@
 Descriptive data are presented as mean and standard deviation (SD).
-The effect of training on muscle strength, muscle thickness, UBF/rpS6 protein, total RNA and gene abundances were assessed using mixed effects regression models.
+The effect of training on muscle strength, muscle thickness, UBF/RPS6 protein, total RNA and gene abundances were assessed using mixed effects regression models.
 Time and group (TRAIN vs.
@@
 
-# Model for rpS6 predicting total RNA
+# Model for RPS6 predicting total RNA
 comb.s6.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1_s6.RDS")
@@
       data.frame(coef = c("Intercept", 
-                          "rpS6 protein levels (SD-units)", 
+                          "RPS6 protein levels (SD-units)", 
                           "Session 1-4",
@@
                     upr = "Upper 95% CI") %>%
-    add_header_lines(values = "Table 1. Effect of UBF and rpS6 levels, sessions and de-training on RNA-levels.") %>%
+    add_header_lines(values = "Table 1. Effect of UBF and RPS6 levels, sessions and de-training on RNA-levels.") %>%
 
@@
             value = as_paragraph(
-              c("The dependent variable is total RNA levels (log)")
+              c("The dependent variable is total RNA levels (log), n = 10")
             ),
@@
                 "Change in slope in session 4-8", 
-                "Change in slope in session 8-12")
+                "Change in slope in session 8-12.")
             ),
@@
 dxa_whole <- read_excel("./data/tr010_dxa_standarddefinitions.xlsx") %>%
-  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:`HeleVev%fett`) %>%
-  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:`HeleVev%fett`) %>%
+  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:HeleTotalmasse) %>%
+  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:HeleTotalmasse) %>%
   mutate(value = gsub("kg", "", value), 
-         value = gsub("[[:space:]]", "", value, fixed = FALSE), 
+         value = str_replace_all(value,"\\s", ""), 
          value = as.numeric(value)) %>%
@@
 # Figure legends
-
+<span custom-style="CHANGE">
 **Figure 1.**  (A) Study design showing muscle biopsy sampling, thickness and strength assessments time points together with number of sets per session (CONST blue bars, VAR red bars).
@@
 
-**Figure 2.** Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF.
-Black points and error bars represent statistically robust results (a 95\% CI not containing 0).
+**Figure 2.** Protein (A and B) and mRNA abundances (D and E) of RPS6 and UBF.
+Non-transparent gray points and error bars represent statistically robust results (a 95\% CI not containing 0).
 C shows western blots and total protein stains from a representative participant.
@@
 **Figure 3.** Total RNA and ribosomal RNA subspecies in response to resistance training.
-A shows primer locations targeting different ribosomal RNA subspecies in qPCR analyses.
-Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (B).
-There were no robust differences between different volume conditions in the training group (B).
-Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and after the training period (C).
-Total RNA increased compared to CTRL after training and normalized after de-training (E, F).
-The time-course analysis revealed the greatest increase during the first four sessions (G).
+(A) shows primer locations targeting different ribosomal RNA subspecies in qPCR analyses.
+Ribosomal RNA species measured by qPCR and compared to control was affected by training (B, C), but did not show clear differences between volume conditions (D).
+Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (E, F).
+Time-course analysis revealed the greatest increase during the first four sessions (G).
 Error bars shows 95\% CI.
@@
 
-**Figure 4.** Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 4).
-Values are averaged over values from men and women.
+**Figure 4.** Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 2).
+Model estimates shown as black lines with 95\% CI are averaged over values from men and women.
 Individual plots of estimates total RNA increases over time are shown in C together with results from leave-one-out analysis (D).
-Leave-one-out analysis shows the effect of removing a single participant (black point and error bars) and individual values from the total RNA per time estimates where green points represent bounds of the 95\% CI and yellow points represent mean estimates.
+Leave-one-out analysis shows the effect of removing a single participant (grey point and error bars) and individual values from the total RNA per time estimates where red points represent bounds of the 95\% CI and circles represent mean estimates.
+</span>

### 2022-02-04 response-to-reviews-rev1.Rmd
revision 1 complete

@@
 ---
-title: "APH-2021-12-0652: Response to the reviewers"
+title: "APH-2021-12-0652: Responses to reviewers comments"
 author: 
@@
   - Stian Ellefsen
-date: "2022-02-02"
-output: pdf_document
+date: "2022-02-04"
+output: html_document
 editor_options: 
@@
     wrap: 72
-bibliography: references.bib
+bibliogPlease find our raphy: references.bib
 ---
@@
 
-## Referee 1
+# Response to decision letter
 
-> This manuscript investigates the induction of ribosome biogenesis and
+**Authors answer**: We would first like to thank both reviewers for their interest in our manuscript and their questions and constructive criticism.
+We feel that the revised manuscript has benefited from this first round of reviews.
+We have addressed the reviewer's comments below.
+We have updated the figures to match the journals guidelines, we are sorry not to have noticed the very clear instruction prior to the first submission.
+We have spotted some grammatical errors and changed these.
+We also noted that information was missing regarding one missing biopsy from one participant at baseline and technical difficulties in the calibrator sample of one participant for protein analyses.
+The latter affecting the sample size in a subset of analyses (n = 10 instead of 11; see revised manuscript Methods and Materials and Table 1).
+
+## Reviewer 1
+
+> **Reviewer comment:** This manuscript investigates the induction of ribosome biogenesis and
 > accumulation of RNA/ribosomes during the early phase of resistance
@@
 > equal in intensity across the time course. Could the authors perhaps
-> provide raw data and membranes for RPS6 and UBF just for the
+> provide raw data and membranes for rpS6 and UBF just for the
 > reviewers?
-
-We agree with the referee that the blots presented in Figure 2 were
-faint, especially for UBF. The selected blot was made for illustrative
-purposes as we had ordered the samples according to sampling time and
-condition in that particular blot. Unfortunately this blot was in the
-"fainter end" of the spectrum. We have replaced the blot with one that
-we feel is more representative. We are providing the quantified blots in
-a separate document for the reviewers.
-
-The protein samples were extracted from Trizol preparations, something
-that has been proven difficult but possible (see e.g. [@wen2020;
-@kopec2]). Some of our samples were also difficult to dissolve after
-precipitation and this sometimes gave unsatisfactory results during the
-western blot procedure. Problematic samples were re-extracted and
-re-run. Importantly, assessment of sample quality and any re-extraction
-was performed on samples given a random sample id. De-coding of the
-random sample id was performed only at data analysis, and to create runs
-with ordered samples.
 
@@
 
-The raw aggregated data (averaged over replicates per samples) does show
-a clear increase in signal over the time course (\@ref(fig:western-fig)). We acknowledge that the western blot technique is associated with
-considerable technical variation. The average coefficient of variation
-(CV) in replicates were `r cvs` for RPS6 and UBF respectively, both within the
-expected range of CV from western blot data [@degasperi2014].
+**Authors answer:** We agree with the referee that the blots presented in Figure 2 were faint, especially for UBF.
+The selected blot was made for illustrative purposes as we had ordered the samples according to sampling time and condition in that particular blot.
+Unfortunately this blot was in the "fainter end" of the spectrum.
+We have replaced the blot with one that we feel is more representative.
+We are providing the quantified blots in a separate document for the reviewers.
+
+The protein samples were extracted from Trizol preparations, something that has been proven difficult but possible (see e.g.
+[@wen2020; @kopec2]).
+Some of our samples were also difficult to dissolve after precipitation and this sometimes gave unsatisfactory results during the western blot procedure.
+Problematic samples were re-extracted and re-run.
+Importantly, assessment of sample quality and any re-extraction was performed on samples given a random sample id.
+De-coding of the random sample id was performed only at data analysis, and to create runs
+with ordered samples.
+
+The raw aggregated data (averaged over replicates per samples) does show a clear increase in signal over the time course (Figure 1 in supporting document).
+We acknowledge that the western blot technique is associated with considerable technical variation.
+The average coefficient of variation (CV) in replicates were `r cvs` for rpS6 and UBF respectively, both within the
+expected range of CV from western blot data [@degasperi2014].}
 
@@
 
-> Is pre-rRNA 47S ETS and 45S ETS per unit tissue weight a previously reported way to present qPCR data? Does the data look the same in relation to 1 or more house keeping genes?
+> **Reviewer comment:** Is pre-rRNA 47S ETS and 45S ETS per unit tissue weight a previously reported way to present qPCR data? Does the data look the same in relation to 1 or more house keeping genes?
 
-One transcribed pre-rRNA results in a single end product, one ribosome, and in the context of protein synthesis, the amount of ribosomes per cell or tissue mass is highly correlated with protein synthesis [@millward1973; @west2016].
+**Authors answer:** One transcribed pre-rRNA results in a single end product, one ribosome, and in the context of protein synthesis, the amount of ribosomes per cell or tissue mass is highly correlated with protein synthesis [@millward1973; @west2016].
 In contrast to rRNA processing, mRNA transcription and subsequent translation could, to some degree, be seen as competitive between different mRNA.
@@
 
-> Could the authors please comment on the apparent disconnect between increase in RNA and RPS6 protein levels?
+> **Reviewer comment:** Could the authors please comment on the apparent disconnect between increase in RNA and rpS6 protein levels?
 
-We thank the reviewer for raising this question as we have attempted to broaden our discussion.
-We interpret the results from our analysis as affected by a time-lag between increases in synthesis of total RNA (presumably rRNA) and RPS6.
-The model used "controls" for both quantities being increased over the whole training period, within a specific time-point there exists no relationship between RPS6 and total RNA.
-This separates our analysis from others who have shown correlations between RPS6 and rRNA without controlling for, in their case, degree of stimuli [@nakada2016] and without examining the time course of their increase.
-As we have already pointed to in the original manuscript, our results suggests different temporal regulation or turnover rates in total RNA and RPS6 protein levels.
+**Authors answer:** We thank the reviewer for raising this question as we have attempted to broaden our discussion.
+We interpret the results from our analysis as affected by a time-lag between increases in synthesis of total RNA (presumably rRNA) and rpS6.
+The model used "controls" for both quantities being increased over the whole training period, within a specific time-point there exists no relationship between rpS6 and total RNA.
+This separates our analysis from others who have shown correlations between rpS6 and rRNA without controlling for, in their case, degree of stimuli [@nakada2016] and without examining the time course of their increase.
+As we have already pointed to in the original manuscript, our results suggests different temporal regulation or turnover rates in total RNA and rpS6 protein levels.
 Since ribosomal proteins exists in excess (and are thus not rate limiting for ribosomal biogenesis) [@lam2007], it is tempting to assume constitutive expression is only indirectly related to the process of ribosomal biogensis.
-We have expanded the discussion by raising the point that unbound ribosomal proteins (including RPS6) may be passively increased by myonuclear accretion.
+We have expanded the discussion by raising the point that unbound ribosomal proteins (including rpS6) may be passively increased by myonuclear accretion.
 Although somewhat speculative, this notion is supported by recent evidence suggesting a specific role of satellite cell fusion in increasing availability of ribosomal proteins together with increased rRNA transcription [@murach2022].
@@
 
-> The concept that the increase in RNA content predicts subsequent hypertrophy is very interesting.
+> **Reviewer comment:** The concept that the increase in RNA content predicts subsequent hypertrophy is very interesting.
 Even more intriguing is the finding that total RNA mid training negatively correlates to muscle hypertrophy is fascinating!
@@
 
-The reviewer raises an interesting point, that we interpret as suggesting that newly synthesized ribosomes would be more beneficial for muscle growth in response to a new stimuli as novel ribosomes are potentially specialized for growth rather than cellular status quo.
+**Authors answer:** The reviewer raises an interesting point, that we interpret as suggesting that newly synthesized ribosomes would be more beneficial for muscle growth in response to a new stimuli as novel ribosomes are potentially specialized for growth rather than cellular status quo.
 In light of recent studies and overviews highlighting ribosomal heterogoneity, our results are worth following up in relation to a specialized ribosome.
@@
 
-Although not robustly decreased, the overall tendency of total UBF protein was decrease in response to the de-training period.
+**Authors answer:** Although not robustly decreased, the overall tendency of total UBF protein was decrease in response to the de-training period.
 The variation presumably contributed to relationship with total RNA abundance.
@@
 
-> The fact that RNA decreased at the detraining time point is interesting and indicates the importance of actively maintaing translational capacity not to loose it.
+> **Reviewer comment:** The fact that RNA decreased at the detraining time point is interesting and indicates the importance of actively maintaing translational capacity not to loose it.
 I think the discussion could benefit from including a short section of the role for ribophagy in this setting?
 
-We thank the reviewer for this suggestion, we have incorporated a section connecting our discussion to the emerging field of ribophagy.
+**Authors answer:** We thank the reviewer for this suggestion, we have incorporated a section connecting our discussion to the emerging field of ribophagy.
 
-> The methods in the abstract must be improved.
+> **Reviewer comment:** The methods in the abstract must be improved.
 In its current form it lacks key information such as training duration, the number of controls are states as n=7 but n=8 in the M&M section and I'd like for the authors to try and specify biopsy time points.
 
-We thank the reviewer for catching this typo. We have updated the abstract with the correct information and additional information regarding biopsy time points.
+**Authors answer:** We thank the reviewer for catching this typo. We have updated the abstract with the correct information and additional information regarding biopsy time points.
 
-> Page 3, line 48-49.
+> **Reviewer comment:** Page 3, line 48-49.
 I encourage the authors to consider including an additional reference (Figueiredo et al 2021 J Phys).
@@
 
-We thank the reviewer for this suggestion.
+**Authors answer:** We thank the reviewer for this suggestion.
 
+**Reviewer comment:**
 > Page 4, line 8: Please clarify "numberical lowering". Could this be rephrased?
@@
 
-We thank the reviewer for pointing out the above typos, they have been corrected.
+**Authors answer:** We thank the reviewer for pointing out the above typos, they have been corrected.
 
+**Reviewer comment:**
 > Page 12, line 19: please clarify "leg being training in the rest period between sets in the first leg. Does this means that they training continuously without break?
 
-Yes, sets were performed with rest periods for the other leg.
+**Authors answer:** Yes, sets were performed with rest periods for the other leg.
 The protocol still allowed for complete rest between sets ($\sim$ 60 sec) and adjustment of loads if necessary.
-A clarification has ben added to this section.
+A clarification has been added to this section.
 
+**Reviewer comment:**
 > Page 13, line 52: 2-4 passes were made with the Bard Magnum 12-14 gauge needles but only 1-2 aliquots of the samples were frozen.
@@
 
-Muscle tissue from all passes were processed together and aliquots were weighed and frozen.
+**Authors answer:** Muscle tissue from all passes were processed together and aliquots were weighed and frozen.
 The section has been updated.
@@
 
+**Reviewer comment:**
 > Page 14, line 8: Was the spike-in used to normalize RNA extraction efficiency? If yes, please state this clearly.
 
-The spike-in was used to for normalization of rRNA to muscle weight as previously described [hammarström2020].
+**Authors answer:** The spike-in was used to for normalization of rRNA to muscle weight as previously described [hammarström2020].
 The section has been updated to notify the reader of the purpose of adding the external reference.
 
+**Reviewer comment:**
 > Page 14, line 49: Does "normalized" refer to denatured?
 
-This refers to diluting all samples to a common protein concentration. The sentence has been updated for clarity.
+**Authors answer:** This refers to diluting all samples to a common protein concentration. The sentence has been updated for clarity.
 
+**Reviewer comment:**
 > Page 15, line 15: Does the authors refer to amplicon size when they write primer sizes?
 
-Yes, the sentence should read amplicon sizes. This has been updated.
+**Authors answer:** Yes, the sentence should read amplicon sizes. This has been updated.
 
+**Reviewer comment:**
 > Page 31, Figure legends: I think that the figure legends could benefit from a bit more substance. For example Fig 1A, perhaps indicate bar color as black (CONST) and yellow (VAR)?
 
-The figure legends has been updated together with an overall update of figure appearances.
+**Authors answer:** The figure legends has been updated together with an overall update of figure appearances.
 
-## Referee: 2
+## Reviewer 2
 
+**Reviewer comment:**
 > The authors sought to perform a time course investigation of ribosome density and biogenesis markers after knee extensor training.
@@
 
-We are greatful for the reviwers kind words.
+**Authors answer:** We are grateful for the reviewers kind words.
 
+**Reviewer comment:**
 > ABSTRACT: Reads well and concise. Nice work, and no comments.
@@
 
-The "%-point difference" referes to a differences in percentage points, i.e.
+**Authors answer:** We have corrected in line with the reviewers notes.
+The "%-point difference" refers to a differences in percentage points, i.e.
 one condition increases in 2% the other 11.2%, the percentage-point difference is 9.2%.
 This is different from where percentages are stated without the "-point".
-Although confusing we belive that it is the correct way of reporting this such results.
+Although confusing we believe that it is the correct way of reporting this such results.
 
+**Reviewer comment:**
 > Other than this, the results, while dense and full of data, are well-written.
@@
 
-We thank the reviewer for spotting mistakes and suggesting improvements to the text.
+**Authors answer:** We thank the reviewer for spotting mistakes and suggesting improvements to the text. These have been corrected.
 
+**Reviewer comment:**
 > I urge the authors to be creative somewhere in the discussion by perhaps adding a paragraph after the one found in lines 29-56.
@@
 
-We thank the reviewer for this suggestion.
-We agree with the reviwer and feel that the manuscript also communicates this message.
+**Authors answer:** We thank the reviewer for this suggestion.
+We agree with the reviewer and feel that the manuscript also communicates this message.
 This is an important point to underline.
@@
 
-This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth.
+\textcolor{red}{"This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth.
 This balance, measured as a constant ribosomal density in a growing cell, still requires the biosynthesis of ribosomes to match the volumetric expansion of the cell.
-As such, indirect measures of translational capacity through the density of ribosomes in periods of marked muscle hypertrophy may mask an absolute increase in ribosomes.
+As such, indirect measures of translational capacity such as the concentration of total RNA may mask the absolute increase in ribosomes that occurs during periods of muscle hypertrophy."}
 
+**Reviewer comment:**
 > METHODS
@@
 
-The above points has been addressed in the revised manuscript.
+**Authors answer:** The above points has been addressed in the revised manuscript, thank you for your detailed reading of the manuscript.
 

### 2022-02-04 manuscript.Rmd
revision 1 complete

@@
 
-<div custom-style="runningtitle_style">**Running title:** Ribosome biogenesis in early phase resistance training</div>
+<div custom-style="runningtitle_style">**Short title:** Ribosome biogenesis in early phase resistance training</div>
 
@@
 
-**Aim:** To describe the time course of ribosome accumulation during resistance training, its relation to training volume, UBF protein levels, and muscle growth.
+**Aim:** To describe <span custom-style="CHANGE">ribosome biogenesis during resistance training, its relation to training volume and muscle growth</span>.
 
 **Methods:** A training group (n = 11) performed <span custom-style="CHANGE">12 sessions (3-4 sessions per week) of</span> unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg.
-<span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline, 48 hrs after sessions 1, 4, 5, 8, 9 and 12</span> and after eight days of de-training, as well as</span> from a non-training control group (n = 8).
-In addition, muscle thickness was measured before and after training and control periods.
+<span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline, 48 hrs after sessions 1, 4, 5, 8, 9, and 12,</span> and after eight days of de-training, and from a control group (n = 8).
+Muscle thickness was measured before and after the intervention</span>.
 
 **Results:** Training led to muscle growth (`r hypertrophy` vs.
-control) with concomitant increases in total RNA, rRNA, UBF, and RPS6 with no differences between volume conditions.
-Total RNA increased most rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline).
-After controlling for time, total RNA abundance was associated with UBF protein levels (`r ubf_model` per unit UBF), and the rate of increase in total RNA levels predicted hypertrophy (`r rna_hypertrophy` per %-point increase in total RNA per session).
+control) with concomitant increases in total RNA, rRNA, UBF, and rpS6 with no differences between volume conditions.<span custom-style="CHANGE"> Total RNA increased rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline).
+Total RNA abundance was associated with UBF protein levels (`r ubf_model` per unit UBF), and the rate of increase in total RNA levels predicted hypertrophy</span> (`r rna_hypertrophy` per %-point increase in total RNA per session).
 After de-training, total RNA decreased (`r detrain_rna`) without muscle mass changes indicating halted biosynthesis of ribosomes.
@@
 
-**Keywords:** muscle hypertrophy, resistance training, ribosome biogenesis
+**Keywords:** Muscle hypertrophy, Resistance training, Ribosome biogenesis
 
@@
 <span custom-style="CHANGE">and</span> subsequent repeated bouts leading to accumulation of muscle protein over time
-[@RN1272<!-- Phillips review 2014 -->;
-@RN1642]<!-- Brook 2015 experiment -->. 
+[@RN1272;@RN1642]<!-- Phillips review 2014 --><!-- Brook 2015 experiment -->. 
 In recent years, this view has been supplemented<!-- supported? --> by evidence suggesting that chronic RT leads to increased <span custom-style="CHANGE">basal muscle protein synthesis rates</span>
-[@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> 
-@RN1866;<!-- Wilkingson 2008 -->
-@RN1897],<!-- Reidy 2017 Increased rested state protein synthesis -->
+[@RN1521;@RN1866;@RN1897],<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --><!-- Wilkingson 2008 --><!-- Reidy 2017 Increased rested state protein synthesis -->
 which has been postulated to be associated with increased translational capacity, i.e. accumulation of ribosomes
-[@RN1897;<!-- Reidy 2017 -->
-@RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->.
+[@RN1897;@RN1912]<!-- Reidy 2017 --><!-- Figueiredo 2017 Ribosome biogenesis review -->.
 This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, which is closely connected to protein synthesis
-[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
-@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
+[@RN1754;@RN2145]<!-- West 2016 correlation between total RNA and protein synthesis in C2C12--><!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
 and muscle hypertrophy
-[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
-@RN1755; <!-- Stec Ribosome bio higher in high responders -->
-@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->
+[@RN1644;@RN1755;@RN2358].<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp --><!-- Stec Ribosome bio higher in high responders --><!-- Volume dependent RNA accum relates to volume depend hyper -->
 Conversely, inhibition of ribosomal RNA (rRNA) transcription and inhibition of its up-stream transcription factors act to diminish muscle cell growth
-[@RN1755<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->;
-@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
-@RN1810]<!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->.
+[@RN1755;@RN1754;@RN1810]<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty --><!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS--><!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->.
 
 Biosynthesis of novel ribosomes is a complex, highly coordinated and energy demanding process that involves synthesis of both ribosomal proteins and the four mature rRNA transcripts
-[@RN1920;<!-- Warner review yeast -->
-@RN1820;<!-- Moss review -->
-@RN1810]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
+[@RN1920;@RN1820;@RN1810]<!-- Warner review yeast --><!-- Moss review --><!-- Inhibition of Pol II affects trasncription at Pol I -->.
 Ribosomal accumulation is believed to be determined by the rates of pre-rRNA transcription by RNA polymerase I (Pol I), which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
@@
 Specifically, activation of the of the upstream binding factor (UBF) through phosphorylation is needed to initiate transcription
-[@RN2582; <!-- Tuan UBF phosphorylation important for SL1-UBF interaction -->
-@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
+[@RN2582;@RN2563].<!-- Tuan UBF phosphorylation important for SL1-UBF interaction --><!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
 Such activation is at least partly controlled by the mechanosensitive mTOR pathway, with its inhibition being associated with blocked UBF phosphorylation and subsequent rRNA transcription
-[@RN1632; <!-- Nader rapamycin block UBF availability  -->
-@RN2564].<!-- Hannan mTOR  -->
+[@RN1632;@RN2564].<!-- Nader rapamycin block UBF availability  --><!-- Hannan mTOR  -->
 Interestingly the availability of UBF *per se* has been shown to be a determinant of rRNA transcription
@@
 Resistance exercise is a potent <span custom-style="CHANGE">and specific [@figueiredo2021]</span> stimuli for rRNA transcription as a single session leads to increases in pre-rRNA
-[@RN1656<!-- A single session upregulates 45 S 24 h post ex -->;
-@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->].
+[@RN1656;@RN1037].<!-- A single session upregulates 45 S 24 h post ex --><!-- Nader 2014 45S up in response to 4 h after acute exercise -->
 Repeated bouts lead to the accumulation of mature rRNA reflected in total RNA and presumably functional ribosomes
-[@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
-@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
-@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
-@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
-@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
-@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
-@RN1897<!-- 11% 36 sessions young -->].
+[@RN1656;@RN1520;@RN2358;@RN1755;@RN1644;@RN1809;@RN1897].<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old --><!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr--><!-- Four sessions ~29 and 18 % in mod vs.
+low volume|31 sessions 18/12% young--><!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD--><!-- Figuieredo 2015 16 sessions 49% increase young --><!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young--><!-- 11% 36 sessions young -->
 However, the true time course of ribosomal transcription and accumulation in response to RT remains largely unstudied, with a mere few studies having investigated exercise-induced changes in rRNA over multiple time-points, all of which are either limited to a selected few time-points or a limited time frame.
@@
 
-For both ribosomal protein S6 (RPS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with RPS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
-This general pattern was confirmed when comparing TRAIN to CTRL where UBF and RPS6 protein levels were higher in TRAIN compared to CTRL after Session 12, and remained elevated after eight days of rest (Figure 2A and B), with no robust differences being observed after the first training session (48 hrs).
+For both ribosomal protein S6 (rpS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`.
+This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12, and remained elevated after eight days of rest (Figure 2A and B), with no robust differences being observed after the first training session (48 hrs).
 Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`).
 After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`).
-For RPS6, de-training did not affect protein levels, which remained similar between volume conditions.
-At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while RPS6 showed no robust differences between TRAIN and CTRL at any time point.
+For rpS6, de-training did not affect protein levels, which remained similar between volume conditions.
+At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point.
 No differences were observed between volume conditions for either transcripts (Figure 2D and E).
@@
 In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coinciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 1).
-In contrast, no evidence was found for a relationship between total RNA and RPS6 protein levels (Table 1).
+In contrast, no evidence was found for a relationship between total RNA and rpS6 protein levels (Table 1).
 
@@
 # Discussion
-Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and RPS6 protein in previously untrained individuals compared to a non-training control group.
+Here, we confirm that resistance training leads to increased abundance of markers of ribosome density, measured as total RNA, ribosomal RNA subspecies and rpS6 protein in previously untrained individuals compared to a non-training control group.
 These markers accumulated progressively during the initial part of the intervention before they levelled out, establishing a plausible time course for changes in ribosomal concentration in response to RT that plateaus after $\sim$ 8 sessions.
@@
 Finally, individual rates of increases in total RNA abundance predicted the magnitude of muscle growth, confirming the likely link between ribosomal biogenesis and muscle protein accretion 
-[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
-@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
+[@RN1754;@RN2145]<!-- West 2016 correlation between total RNA and protein synthesis in C2C12--><!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
 and muscle hypertrophy
-[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
-@RN1755; <!-- Stec Ribosome bio higher in high responders -->
-@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->
+[@RN1644;@RN1755;@RN2358].<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp --><!-- Stec Ribosome bio higher in high responders --><!-- Volume dependent RNA accum relates to volume depend hyper -->
 
@@
 <!--@RN2266<  Hortobagyi 2000 NO change in RNA per wet weight after immobilization or 12 weeks retraining low N-->
-[@RN2225;<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 -->
-@RN2180;<!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --> 
-@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
-@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
-@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
-@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
-@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
-@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
-@RN1897<!-- 11% 36 sessions young -->],
+[@RN2225;@RN2180;@RN1656;@RN1520;@RN2358;@RN1755;@RN1644;@RN1809;@RN1897],<!-- Haun 2019 used a progressive volume troughout a six week study study RNA increased during six weeks (12% week 3; 27% week 6 --><!-- Haddad 2005 Unloading through limb suspension leads to lowered total RNA levels, resistance training increased --><!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old --> <!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr--><!-- Four sessions ~29 and 18 % in mod vs.
+low volume|31 sessions 18/12% young--><!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD--><!-- Figuieredo 2015 16 sessions 49% increase young --><!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young--><!-- 11% 36 sessions young -->
 as was also the case in the present data set.
@@
 This corroborates well with previous suggestions of peak values being reached within four to nine sessions in young males and females
-[@RN1809;
-@RN2358], and may be essential for preparing muscle fibres for subsequent growth
-[@RN1809;<!-- Correlation between increase and muscle growth -->
-@RN1644; <!--Figueiredo correlation between incr and muscle hypertrophy -->
-@RN2358].<!-- Early accumulation of RNA predicts volume depdent growth -->
+[@RN1809;@RN2358],
+and may be essential for preparing muscle fibres for subsequent growth
+[@RN1809;@RN1644;@RN2358].<!-- Correlation between increase and muscle growth --><!--Figueiredo correlation between incr and muscle hypertrophy --><!-- Early accumulation of RNA predicts volume depdent growth -->
 After the 8^th^ session, no meaningful increase or decrease were observed for total RNA or rRNA content within the training period, suggesting a plateau phase with attenuated net synthesis of novel ribosomes.
@@
 
-<!-- RPS6 dissconect from total RNA levels. -->
-In the present study, training induced increases in rRNA and total RNA coincided with increases in RPS6.
-Conversely, changes in total RNA levels and RPS6 in response to de-training did not correspond, as RPS6 protein levels remained elevated after the de-training period.
-The training induced increases in RPS6 seen in the present study are in agreement to those previously reported in young men
-[@RN1642],<!-- Brook 2015 RPS6 increase in resposne to RT over control -->
+<!-- rpS6 dissconect from total RNA levels. -->
+In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6.
+Conversely, changes in total RNA levels and rpS6 in response to de-training did not correspond, as rpS6 protein levels remained elevated after the de-training period.
+The training induced increases in rpS6 seen in the present study are in agreement to those previously reported in young men
+[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
 but not in elderly men and women, where a decrease was observed in response to training despite increases in total RNA and rRNA
-[@RN1755].<!-- Stec 2016 RPS6 decreased 22% after four weeks of training in older  -->
-Although increases were seen for both RPS6 and and total RNA, RPS6 did not explain variation in total RNA (after controlling for the number of sessions).
-Together with a disconnect after the de-training period, this suggest that regulation of RPS6 expression and ribosomal RNA transcription display differential temporal responses to RT.
+[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
+Although increases were seen for both rpS6 and and total RNA, rpS6 did not explain variation in total RNA (after controlling for the number of sessions).
+Together with a disconnect after the de-training period, this suggest that regulation of rpS6 expression and ribosomal RNA transcription display differential temporal responses to RT.
 <span custom-style="CHANGE">Such difference in temporal regulation of ribosomal RNA and proteins has previously been deduced in cell culture experiments.
@@
 Currently, it remains unknown if RT leads to expansion of the pool of unbound ribosomal proteins in humans.
-Indeed, in the present study, the disconnect between RPS6 and total RNA after de-training may have been associated with increased numbers of myonuclei, rather than accumulation of RPS6 in existing myonuclei, providing an alternative explanation to the elevated RPS6 levels (which accordingly may have been distributed over a larger number of nuclei).
+Indeed, in the present study, the disconnect between rpS6 and total RNA after de-training may have been associated with increased numbers of myonuclei, rather than accumulation of rpS6 in existing myonuclei, providing an alternative explanation to the elevated rpS6 levels (which accordingly may have been distributed over a larger number of nuclei).
 Interestingly, in a recent study, Murach and colleagues showed that newly acquired myonuclei (from satellite cell fusion) contribute to the ribosomal pool in myofibers [@murach2022].
@@
 Activation of UBF is controlled by the mechanosensitive mTOR pathway, and rapamycin, a specific mTOR inhibitor, blocks UBF from recruiting SL1 and subsequent rRNA transcription
-[@RN1632; <!-- Nader rapamycin block UBF availability  -->
-@RN2564].<!-- Hannan 2003 -->
+[@RN1632; @RN2564].<!-- Nader rapamycin block UBF availability  --><!-- Hannan 2003 -->
 <!-- Hannan 2003 10.1128/MCB.23.23.8862-8877.2003 mTOR is believed to play a role in phosphorylating UBF leading to recruitment of SL-1.
@@
 Evidence from human exercise studies confirms training-induced activation of UBF through phosphorylation
-[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
-@RN1644].<!-- Figueriedo ubf increase total and active form (ser388) --> 
+[@RN1746;@RN1644].<!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase --><!-- Figueriedo ubf increase total and active form (ser388) --> 
 In addition to exercise-induced activation of UBF, mechanical loading also leads to increased levels of total UBF
-[@RN1746; <!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase -->
-@RN1644]. Increases in UBF was determined to be rapamycin insensitive after synergist ablation in mice
+[@RN1746;@RN1644].<!-- Figueiredo 2016 cold water inhibits UBF activation and increase rRNA increase --> Increases in UBF was determined to be rapamycin insensitive after synergist ablation in mice
 [@RN1072]<!-- Goodman 2011 UBF increase due to synergist ablation is rapamycin insensitive https://doi.org/10.1113/jphysiol.2011.218255 --> 
@@
 Future studies should investigate the interrelationship between ribosome biogenesis and ribophagy in response to training and detraining.
-We suspect that their relative importance for ribosomal homeostasis will vary substantially between different physiological perturbations, as they are regulated by a complex interconnection of stimuli and pathways<!--REF?-->.
+We suspect that their relative importance for ribosomal homeostasis will vary substantially between different physiological perturbations, as they are regulated by a complex interconnection of stimuli and pathways [@beese2020;@RN1920;@RN1820].
 For example, Kim et al.
@@
 Based on this notion, and the fact that RT volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including induction of c-Myc expression, mTOR activation
-[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
-@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
-@RN791]<!-- Burd ex volume protein synthesis and anabolic [RPS6] phospho signalling-->,
+[@RN2358;@RN1837;@RN791]<!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc --><!-- Ahtianinen 2015 volume affects mTOR differently --><!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
 subsequent total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
@@
 From a general perspective, albeit volume is an important determinant of increases in muscle strength and mass
-[@RN1767;<!-- schoenfeld 2016 dose response  -->
-@RN2492],<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
+[@RN1767;@RN2492],<!-- schoenfeld 2016 dose response  --><!-- Ralston 2017 Weekly sets volume on muscle strength  -->
 differences in organization of training loads is likely of minor importance when training volumes are equated over time
@@
 Although this observation was done in well-trained participants performing a high volume protocol without a control condition with constant volume, compared to constant volume protocols
-[@RN1809;<!-- Brook 2016 -->
-@RN2358],<!-- Hammarström 2020 -->
+[@RN1809;@RN2358],<!-- Brook 2016 --><!-- Hammarström 2020 -->
 progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.
@@
 ```
-<span custom-style="CHANGE">The average coefficient of variation across replicates were `r cvs`, for RPS6 and UBF, respectively.</span>
+<span custom-style="CHANGE">The average coefficient of variation across replicates were `r cvs`, for rpS6 and UBF, respectively.</span>
 
@@
 Descriptive data are presented as mean and standard deviation (SD).
-The effect of training on muscle strength, muscle thickness, UBF/RPS6 protein, total RNA and gene abundances were assessed using mixed effects regression models.
+The effect of training on muscle strength, muscle thickness, UBF/rpS6 protein, total RNA and gene abundances were assessed using mixed effects regression models.
 Time and group (TRAIN vs.
@@
 
-# Model for RPS6 predicting total RNA
+# Model for rpS6 predicting total RNA
 comb.s6.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1_s6.RDS")
@@
       data.frame(coef = c("Intercept", 
-                          "RPS6 protein levels (SD-units)", 
+                          "rpS6 protein levels (SD-units)", 
                           "Session 1-4",
@@
                     upr = "Upper 95% CI") %>%
-    add_header_lines(values = "Table 1. Effect of UBF and RPS6 levels, sessions and de-training on RNA-levels.") %>%
+    add_header_lines(values = "Table 1. Effect of UBF and rpS6 levels, sessions and de-training on RNA-levels.") %>%
 
@@
 
-**Figure 2.** Protein (A and B) and mRNA abundances (D and E) of RPS6 and UBF.
+**Figure 2.** Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF.
 Non-transparent gray points and error bars represent statistically robust results (a 95\% CI not containing 0).
